### 2016-11-25 results/rbdata.Rmd
First commit: add content in progress

---
title: "RB Data"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE,
                      tidy = TRUE, tidy.opts = list(width.cutoff = 60),
                      cache = TRUE)
#devtools::install_github("statlab/permuter")
require(permuter)
require(dplyr)
require(reshape2)
require(ggplot2)
require(xtable)
options(xtable.comment = FALSE, xtable.include.rownames = FALSE, xtable.align = "p{3.5cm}")
require(grid)
require(gridExtra)

set.seed(737324921) # Generated from random.org Timestamp: 2016-11-07 19:21:48 UTC

ga <- read.csv("GA1102.csv", header = TRUE, stringsAsFactors = FALSE)
```

# Data Cleaning

### Missing data
There's a bunch of missing data. 
First, we notice there are 14 subjects who were only observed during the baseline period.

```{r baseline_only, results = "asis"}
numdays <- table(ga$SUBJID)
baseline_only <- names(numdays)[numdays < 14]
unique_numdays <- as.data.frame(table(numdays))
names(unique_numdays) <- NULL
print(xtable(unique_numdays,
      caption = "Number of subjects observed for 7 or 14 days"),
      include.rownames = FALSE, comment = FALSE)

ga$missing <- ifelse(ga$SUBJID %in% baseline_only, 1, 0)
```

Next, we check how many individuals were assigned to each treatment group.
52 subjects have treatment values of `.`.
They appear to never have been surveyed during week 2: although there are records in the dataset, all values are 0.
These 52 include the fourteen who were only observed during week 1.
We remove all these missing subjects.
We are left with a total of 66 subjects in the placebo group and 70 subjects in the active treatment group.

```{r baseline_only2, results = "asis"}
num_treated <- table(ga$tr)
baseline_only2 <- as.data.frame(num_treated/14)
colnames(baseline_only2) <- c("Treatment", "Number of subjects")
print(xtable(baseline_only2, digits = 0,
             caption = "Number of subjects in each treatment group"), 
      include.rownames = FALSE, comment = FALSE)

ga$missing <- ga$missing + ifelse(ga$tr == ".", 1, 0)
```

## Exploratory Data Analysis

We notice several things patterns from Figure \ref{fig:visualize_missing}, showing missing data patterns.
First, some of the sites have many more subjects than others: these are sites 1, 4, and 12.
Second, some of the sites have a large proportion of missing data, such as sites 3, 4, 5, and 14.
Finally, we notice that sites 3, 6, and 14 have very few subjects, so we will have low power at these sites.
For example, after removing the missing values, site 14 only has one treated subject and two placebos, so the site will only have 3 unique permutation statistics.

```{r visualize_missing, fig.align="center", fig.cap = "Number of subjects in each treatment group, by site ID."}
ga %>%
  mutate(subjid_by_site = SUBJID %% 100) %>%
  ggplot(aes(x = subjid_by_site, y = factor(SITEID))) + 
  geom_tile(aes(fill = tr)) + 
  xlab("Subject ID Number") +
  ylab("Site ID") + 
  scale_fill_discrete(name="Treatment",
                      breaks=c(".", "1", "2"),
                      labels=c("Not assigned", "Placebo", "Drug")) +
  ggtitle("Treatment Assignment Patterns")
```

```{r remove_missing, include = FALSE}
ga <- ga %>% filter(tr != ".") %>% mutate(tr = as.numeric(tr)-1)
```

The data are not in the right shape to analyze a single variable.
Below, we include a function to take the raw data and reshape it to analyze one variable of interest.

```{r data_structure}

data_by_subjid_visitnum <- ga %>% group_by(SUBJID, VISITNUM)

reshape_ga_continuous <- function(variable, data = data_by_subjid_visitnum){
  data <- data %>% mutate_(myvariable = variable)
  summarized <- data %>% summarise("variable" = mean(as.numeric(myvariable)), "tr" = unique(tr), "SITEID" = unique(SITEID))
  cleaned <- dcast(summarized, SUBJID + tr + SITEID ~ VISITNUM, value.var = "variable")
  colnames(cleaned) <- c("SUBJID", "tr", "SITEID", "Baseline", "Outcome")
  cleaned <- ungroup(cleaned) %>% 
    mutate(difference = Outcome - Baseline)
  return(cleaned)
}

```
Figure \ref{fig:hist_continuous_vars} shows the distribution of the final outcome measures between treatment and placebo groups.
Based on these distributions, we would expect to detect a difference in outcomes between the drug and placebo groups for the `daily_heart`, `daily_hrdq`, and `heart_freq` measures.

```{r hist_continuous_vars, fig.height=6, fig.width=12, fig.cap = "Distribution of continuous outcomes"}
continuous_vars <- c("daily_heart", "daily_regurg", "daily_dysp", 
                    "daily_hrdq",
                    "heart_freq", "regurg_freq", "dysp_freq")
plot_distrs <- lapply(continuous_vars, function(variable){
  p <- reshape_ga_continuous(variable) %>% mutate(tr = factor(tr)) %>%
        ggplot(aes(Outcome)) +
        geom_density(alpha = 0.6, aes(fill = tr))+ 
#        geom_histogram(binwidth = 0.5, position = "dodge", aes(fill = tr)) +
        scale_fill_discrete(name="Experimental\nCondition",
                       breaks=c("0", "1"),
                       labels=c("Placebo", "Drug")) + 
        xlab(variable)
  return(p)
})

tmp <- ggplot_gtable(ggplot_build(plot_distrs[[1]]))
leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
legend <- tmp$grobs[[leg]]

plot_distrs <- lapply(plot_distrs, function(x) x + theme(legend.position="none"))
plot_distrs[[length(plot_distrs)+1]] <- legend
do.call(grid.arrange, c(plot_distrs, nrow = 2))
```

### Ordinal variables

Several variables are measured on an ordinal scale but treated as continuous values.
For instance, heartburn frequency is measured on a 7-point scale, with levels 0, 1, 2, 3, 4-5, 6-10, and more than 10.
Values in the 4-5 range are coded as 4.5 and in the 6-10 range are coded as 8.
We do not necessarily believe that these are accurate representations of the data.
We recode the ordinal scales as factors on a scale from 1 to 7.

```{r recode}
recode_ordinal <- function(variable){
  levels <- 
  variable <- as.factor(variable)
  levels(variable) <- seq_along(unique(variable))
  return(variable)
}

ga <- ga %>% mutate(heart_sev_f = recode_ordinal(heart_sev),
                    regurg_sev_f = recode_ordinal(regurg_sev),
                    dysp_sev_f = recode_ordinal(dysp_sev),
                    heart_freq_f = recode_ordinal(heart_freq),
                    regurg_freq_f = recode_ordinal(regurg_freq),
                    dysp_freq_f = recode_ordinal(dysp_freq)
                    )
```
We must put the data in a usable form so there is only one measurement for each individual.
For each person, we take the median response from week 2 and subtract it from their median response from week 1.

```{r data_structure_ordinal}

data_by_subjid_day <- ga %>% 
  mutate(dayofweek = rep(1:7, length.out = nrow(ga))) %>%
  group_by(SUBJID, dayofweek)

# The reshape function below reduces the 14 days into a single measure with the following steps:
# 1) Take the difference in measurements between the first day of treatment week and the first day of baseline week, the second day of treatment week and the second day of baseline week, and so on.
# 2) Take the median of these seven numbers.
# We have decided not to use this method; it assumes that day j in weeks 1 and 2 are paired, which may not be right.

# reshape_ga_ordinal <- function(variable, data = data_by_subjid_day){
#   data <- data %>% mutate_(myvariable = variable)
#   day_diff <- data %>% 
#     summarise("variable" = diff(as.numeric(myvariable)), "tr" = unique(tr), "SITEID" = unique(SITEID))
#   summarized <- day_diff %>%
#     summarise("medoverdays" = median(variable), "tr" = unique(tr), "SITEID" = unique(SITEID))
#   summarized <- ungroup(summarized)
#   return(summarized)
# }
reshape_ga_ordinal <- function(variable, data = data_by_subjid_visitnum){
  data <- data %>% mutate_(myvariable = variable)
  summarized <- data %>% summarise("variable" = median(as.numeric(myvariable)), "tr" = unique(tr), "SITEID" = unique(SITEID))
  cleaned <- dcast(summarized, SUBJID + tr + SITEID ~ VISITNUM, value.var = "variable")
  colnames(cleaned) <- c("SUBJID", "tr", "SITEID", "Baseline", "Outcome")
  cleaned <- ungroup(cleaned) %>% 
    mutate(difference = Outcome - Baseline)
  return(cleaned)
}
```
Figure \ref{fig:hist_ordinal_vars} displays the distribution of these newly-coded ordinal variables in the treatment and control groups.
There are certainly differences in distributions, but it is not clear how different they are.

```{r hist_ordinal_vars, fig.height=6, fig.width=12, fig.cap = "Distribution of ordinal outcomes"}
ordinal_vars <- c("heart_sev_f", "regurg_sev_f", "dysp_sev_f",
                    "heart_freq_f", "regurg_freq_f", "dysp_freq_f") 
plot_distrs <- lapply(ordinal_vars, function(variable){
  p <- reshape_ga_ordinal(variable) %>% 
        mutate(tr = factor(tr)) %>%
        ggplot(aes(difference, ..density..)) +
        geom_bar(aes(y = (..count..)/sum(..count..), fill = tr), position = "dodge") + 
        scale_y_continuous(labels = scales::percent) +
        scale_fill_discrete(name="Experimental\nCondition",
                       breaks=c("0", "1"),
                       labels=c("Placebo", "Drug")) +
        xlab(variable)
  return(p)
})

tmp <- ggplot_gtable(ggplot_build(plot_distrs[[1]]))
leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
legend <- tmp$grobs[[leg]]

plot_distrs <- lapply(plot_distrs, function(x) x + theme(legend.position="none"))
plot_distrs[[length(plot_distrs)+1]] <- legend
do.call(grid.arrange, c(plot_distrs, nrow = 3))
```

# Analysis - Primary endpoint
Let's first restrict attention to the primary endpoint, `daily_hrdq`.
We must transform the data to contain the variables of interest: the baseline score (averaged over 7 days of no treatment), the outcome score (averaged over 7 days of treatment), an indicator for which treatment was received (`1` for the active drug, `0` for placebo), the site ID, and subject ID.

Throughout, we compare two models: 
the first model considers the final outcome measure, adjusting for site ID and perhaps also the baseline score.
The second model considers the difference between outcome and baseline score as its dependent variable, adjusting for site ID.

We compare three testing methods: 
the usual parametric ANOVA,
a permutation test considering the difference in means as test statistic, stratifying by site,
and a permutation test in the style of Freedman and Lane (1983), using a linear regression to adjust for covariates.

```{r hrdq, results = "asis"}
daily_hrdq <- reshape_ga_continuous("daily_hrdq")
summary_stats_hrdq <- cbind(summary(daily_hrdq$difference[daily_hrdq$tr == 1]),
                            summary(daily_hrdq$difference[daily_hrdq$tr == 0]))
colnames(summary_stats_hrdq) <- c("Drug", "Placebo")
print(xtable(summary_stats_hrdq,
             caption = "Summary statistics for change in primary endpoint (daily hrdq) for drug and placebo groups."),
      include.rownames = TRUE)
```

## Parametric ANOVA
First, we do a standard parametric ANOVA using the two models.
Under model 1, using the outcome as response and adjusting for the baseline value, treatment has a significant effect.
Under model 2, the significance goes away.

```{r hrdq_parametric, results = "asis"}
# method 0: parametric ANOVA
lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = daily_hrdq)
print(xtable(summary(aov(lm1))), include.rownames = TRUE)
daily_hrdq$lm1_resid <- residuals(lm1)

lm2 <- lm(difference ~ tr + factor(SITEID), data = daily_hrdq)
print(xtable(summary(aov(lm2))), include.rownames = TRUE)
daily_hrdq$lm2_resid <- residuals(lm2)
```

## Stratified permutation test

Suppose each individual, $i = 1, \dots, N$ has two potential outcomes $(Y_i(1), Y_i(0))$ that indicate their response to the active drug and to the placebo, respectively.
We randomly assign treatment to individuals; treatment determines which of $Y_i(1)$ and $Y_i(0)$ we observe.
We are unable to observe both.
Suppose we wish to test the null hypothesis that individual by individual, treatment has no effect.
This is referred to as the ``sharp null'' hypothesis:

$$H_0: Y_i(1) = Y_i(0), i = 1, \dots, N.$$

Then, whether individual $i$ received the drug or placebo amounts to an arbitrary label.
Once we observe their response under treatment $T_i$, we know what it would have been under treatment $1-T_i$, for $T_i \in \{0,1\}$; namely, it would be the same.

Treatment was completely randomized at each site, independently across sites.
Assuming that drop-out from the study was independent of treatment assignment (which is true if treatment was blinded and assigned at random), then we may condition on the number of individuals who received the drug and who received the placebo at each site.
Any assignment of treatments which preserves the number of treated and controls at each site is valid and has equal probability of occurring.
Therefore, by using this principle of equal probabilities and by imputing the unobserved potential outcomes assuming that the null hypothesis is true, we may obtain the permutation distribution of any statistic under the null hypothesis.

We compare two test statistics: the usual difference in means between treatment and placebo groups, and the difference in means within each site, aggregated by taking the sum of their absolute values over sites.
The first statistic is more comparable to what one would obtain from an ANOVA and it is readily interpretable.
It does not directly account for the stratification by site, so variation between sites may be hidden.
The second statistic is useful for testing the two-sided alternative hypothesis that treatment has some effect. 
It may be more powerful than the simple difference in means if the effect of the drug varies across sites.
For instance, if the drug has a positive effect at one site but a negative effect at another site, this test statistic would be large; 
if we used the simple difference in means without accounting for sites, then the positive effect and negative effect may cancel each other out and appear as no effect.

Figure \ref{fig:hrdq_unadjusted_permutation} shows the permutation distributions of the simple difference in means.
Figure \ref{fig:hrdq_unadj_perm_withinstrata} shows the permutation distribution of the absolute value of the difference of means within strata, summed across strata.
There are several points to notice:
* The second test is far less powerful than the first.  This is likely because some sites had very few patients.  These sites have correspondingly few unique permutations, and so they often contribute the same amount to the summed difference in means statistic.  It is impossible to detect a difference at small significance levels when there are only 3 patients.
* The summed difference of means only makes sense for testing against a two-sided alternative, since taking the absolute value removes the sign of the effect. The permutation distribution is strictly non-negative and skewed to the right.
* This statistic is less interpretable than the difference in means because it doesn't correspond directly to any feature of the distribution of outcomes.

Since the summed difference in means statistic has such lower power, we abandon it for the rest of the analyses.

```{r hrdq_unadjusted_permutation, fig.height = 3, fig.cap = "Permutation distribution of the difference of means for daily_hrdq"}
# method 1 : do permutation of differences
observed_diff_means1 <- mean(daily_hrdq[daily_hrdq$tr == 1, ]$Outcome) - mean(daily_hrdq[daily_hrdq$tr == 0, ]$Outcome)
diff_means_distr1 <- stratified_two_sample(group = daily_hrdq$tr, response = daily_hrdq$Outcome, stratum = daily_hrdq$SITEID, reps = 1e4)
diff_means_pvalue1 <- t2p(observed_diff_means1, diff_means_distr1, alternative = "two-sided")

observed_diff_means2 <- mean(daily_hrdq[daily_hrdq$tr == 1, ]$difference) - mean(daily_hrdq[daily_hrdq$tr == 0, ]$difference)
diff_means_distr2 <- stratified_two_sample(group = daily_hrdq$tr, response = daily_hrdq$difference, stratum = daily_hrdq$SITEID, reps = 1e4)
diff_means_pvalue2 <- t2p(observed_diff_means2, diff_means_distr2, alternative = "two-sided")

data.frame("perm" = c(diff_means_distr1, diff_means_distr2),
           "model" = c(rep("Outcome", length(diff_means_distr1)), 
                       rep("Differenced", length(diff_means_distr2))),
           "xintercept" = c(rep(observed_diff_means1, length(diff_means_distr1)),
                            rep(observed_diff_means2, length(diff_means_distr2)))) %>%
  ggplot(aes(x = perm, fill = model)) +
  geom_histogram() +
  facet_grid(~model) +
  geom_vline(aes(xintercept = xintercept))

```

```{r hrdq_unadj_perm_withinstrata, fig.height = 3, fig.cap = "Permutation distribution of the difference of means within strata, summed across strata, for daily_hrdq"}

obs_diff_means_bystrata1 <- sum(abs(within_group_mean(group = daily_hrdq$tr, 
                                                      response = daily_hrdq$Outcome,
                                                      stratum = daily_hrdq$SITEID, 
                                                      groups = unique(daily_hrdq$tr), 
                                                      strata = unique(daily_hrdq$SITEID)
                                                      )))
diff_means_distr_bystrata1 <- stratified_two_sample(group = daily_hrdq$tr, response = daily_hrdq$Outcome, stratum = daily_hrdq$SITEID, stat = 'mean_within_strata', reps = 1e4)
diff_means_bystrata_pvalue1 <- t2p(obs_diff_means_bystrata1, diff_means_distr_bystrata1, alternative = "two-sided")

obs_diff_means_bystrata2 <- sum(abs(within_group_mean(group = daily_hrdq$tr, 
                                                      response = daily_hrdq$difference,
                                                      stratum = daily_hrdq$SITEID, 
                                                      groups = unique(daily_hrdq$tr), 
                                                      strata = unique(daily_hrdq$SITEID)
                                                      )))
diff_means_distr_bystrata2 <- stratified_two_sample(group = daily_hrdq$tr, response = daily_hrdq$difference, stratum = daily_hrdq$SITEID, stat = 'mean_within_strata', reps = 1e4)
diff_means_bystrata_pvalue2 <- t2p(obs_diff_means_bystrata2, diff_means_distr_bystrata2, alternative = "two-sided")

data.frame("perm" = c(diff_means_distr_bystrata1, diff_means_distr_bystrata2),
           "model" = c(rep("Outcome", length(diff_means_distr_bystrata1)), 
                       rep("Differenced", length(diff_means_distr_bystrata2))),
           "xintercept" = c(rep(obs_diff_means_bystrata1, length(diff_means_distr_bystrata1)),
                            rep(obs_diff_means_bystrata2, length(diff_means_distr_bystrata2)))) %>%
  ggplot(aes(x = perm, fill = model)) +
  geom_histogram() +
  facet_grid(~model) +
  geom_vline(aes(xintercept = xintercept))
```

## Covariate-adjusted permutation test
We would like to test for a difference in outcomes between the active drug and placebo group, but control for other covariates.
In particular, when the outcome is the average response during the second week of follow-up, we would like to control for the average response during the first week of follow-up and the location.
When the outcome is the difference in responses between the second and first weeks, we would just like to control for location.
We will use the approximate permutation test derived by Freedman and Lane (1983) to do so.

Let $Y$ denote the response, $Z$ denote the treatment indicator, and $X$ denote a covariate which may be correlated with $Y$.
We may write the following equation:

$$ Y = \beta_0 + \beta_1 X + \beta_2 Z + \varepsilon$$

Under the sharp null $H_0$, $Z$ has no effect on $Y$.
In other words, the null hypothesis is that $\beta_2 = 0$.
In a randomized experiment such as this, there are two ways we may test this null hypothesis.

### Standard linear model
First, we may permute the treatment assignments $Z$ directly.
Treatment was assigned at random within each site.
This ensures that $Z$ and $\varepsilon$ are statistically independent, conditional on site.
Therefore, we may conduct a test by permuting treatment assignments $Z$ within site, independently across sites, and calculating a test statistic for each such permutation.
We choose to use the $t$ statistic from the linear model as the test statistic.

```{r tr_permutation, fig.height = 3, fig.cap = "Permutation distribution of the stratified permutation test between daily_hrdq and treatment after controlling for covariates in a linear model."}
# method 2 : linear model

observed_t1 <- summary(lm1)[["coefficients"]]["tr", "t value"]
lm1_t_distr <-  replicate(1e4, {
  daily_hrdq$tr_perm <- permute_within_groups(daily_hrdq$tr, daily_hrdq$SITEID)
  lm1_perm <- lm(Outcome ~ Baseline + tr_perm + factor(SITEID), data = daily_hrdq)
  summary(lm1_perm)[["coefficients"]]["tr_perm", "t value"]
})

lm_pvalue_1 <- t2p(observed_t1, lm1_t_distr, alternative = "two-sided")

observed_t2 <- summary(lm2)[["coefficients"]]["tr", "t value"]
lm2_t_distr <-  replicate(1e4, {
  daily_hrdq$tr_perm <- permute_within_groups(daily_hrdq$tr, daily_hrdq$SITEID)
  lm2_perm <- lm(difference ~ tr_perm + factor(SITEID), data = daily_hrdq)
  summary(lm2_perm)[["coefficients"]]["tr_perm", "t value"]
})

lm_pvalue_2 <- t2p(observed_t2, lm2_t_distr, alternative = "two-sided")

data.frame("perm" = c(lm1_t_distr, lm2_t_distr),
           "model" = c(rep("Outcome", length(lm1_t_distr)), 
                       rep("Differenced", length(lm2_t_distr))),
           "xintercept" = c(rep(observed_t1, length(lm1_t_distr)),
                            rep(observed_t2, length(lm2_t_distr)))) %>%
  ggplot(aes(x = perm, fill = model)) +
  geom_histogram() +
  facet_grid(~model) +
  geom_vline(aes(xintercept = xintercept)) +
  xlab("t Statistic") +
  ylab("Frequency") +
  ggtitle("Permutation Distributions")
```

### Linear model residuals

Let's take an alternative view of the problem.
We still write $Y = \beta_0 + \beta_1 X + \beta_2 Z + \varepsilon$.
However, now, we do not treat the $\varepsilon$ as random.
They are simply defined to be the diffence between $Y$ and the data's linear projection onto the plane $\beta_0 + \beta_1 X + \beta_2 Z$.

If the null hypothesis is true, then $\varepsilon = Y - \beta_0 - \beta_1X$.
Therefore, we may estimate the errors $\hat{\varepsilon}$ by $Y - \hat{Y}$, where $\hat{Y}$ is obtained by regressing $Y$ on $X$ but not $Z$.
They approximate the true errors $\varepsilon$ from the equation $Y = \beta_0 + \beta_1 X + \beta_2 Z + \varepsilon$.
The $\varepsilon$ are orthogonal to $X$ and $Z$.
Therefore, within sites, these estimated $\hat{\varepsilon}$ are approximately exchangeable.

We construct a permutation distribution using several steps:

1. Estimate $\hat{\varepsilon}$ by $Y - \hat{\beta}_0 - \hat{\beta}_1 X$, where $\hat{\beta}_0$ and $\hat{\beta}_1$ are obtained by regressing $Y$ on $X$.
2. Construct permuted errors $\hat{\varepsilon}^\pi$ by permuting the $\hat{\varepsilon}$ within sites.
3. Construct permuted responses $Y^\pi = \hat{\beta}_0 + \hat{\beta}_1 X + \hat{\varepsilon}$.
4. Regress $Y^\pi$ on $X$ and $Z$. The test statistic is the $t$-statistic for the coefficient of $Z$.

For this dataset, we are guaranteed that treatment $Z$ is independent of $\varepsilon$, as it is randomized within site.
We should also check that site ID is uncorrelated with $\hat{varepsilon}$.
If not, then the $\hat{\varepsilon}$ are not exchangeable across sites.
However, our permutations are done within sites, so even if this is violated, there is not a big issue.
We check these associations using residual plots in Figure \ref{fig:residual_plots}.
Indeed, the distribution of residuals looks nearly equal between treatment groups in both models.
The distribution of residuals varies a bit across sites, but they all appear roughly centered around 0.
Sites 4 and 5 have larger variance in the differenced model.
As mentioned, this should not be an issue since we permute treatments within sites, independently across sites.
(In general, when treatment is not randomly assigned, this condition is necessary for the permutation test to be valid.)

```{r residual_plots, echo = FALSE, message = FALSE, fig.height = 5, fig.cap = "Residual plots of the linear regression of daily_hrdq on treatment and covariates."}
# Residual plots to check that treatment and covariates are uncorrelated with errors

p11 <- daily_hrdq %>% 
  select(-Baseline, -Outcome, -difference) %>% 
  melt(id.vars = c("SUBJID", "tr", "SITEID")) %>%
  filter(variable == "lm1_resid") %>% 
  ggplot(aes(x = factor(tr), y = value)) +
  geom_boxplot() + 
  ylim(-2.25, 2.25)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Treatment") + ylab("Residual") + 
  ggtitle("Outcome Model")
p21 <- daily_hrdq %>% 
  select(-Baseline, -Outcome, -difference) %>%  
  melt(id.vars = c("SUBJID", "tr", "SITEID")) %>%
  filter(variable == "lm1_resid") %>% 
  ggplot(aes(x = factor(SITEID), y = value)) +
  geom_boxplot() +  
  ylim(-2.25, 2.25)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Site ID") + ylab("Residual") + 
  ggtitle("Outcome Model")
p12 <- daily_hrdq %>% 
  select(-Baseline, -Outcome, -difference) %>% 
  melt(id.vars = c("SUBJID", "tr", "SITEID")) %>%
  filter(variable == "lm2_resid") %>% 
  ggplot(aes(x = factor(tr), y = value)) +
  geom_boxplot() +  
  ylim(-2.25, 2.25)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Treatment") + ylab("Residual") +
  ggtitle("Differences Model")
p22 <- daily_hrdq %>% 
  select(-Baseline, -Outcome, -difference) %>%  
  melt(id.vars = c("SUBJID", "tr", "SITEID")) %>%
  filter(variable == "lm2_resid") %>% 
  ggplot(aes(x = factor(SITEID), y = value)) +
  geom_boxplot() +  
  ylim(-2.25, 2.25)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Site ID") + ylab("Residual") +
  ggtitle("Differences Model")
grid.arrange(p12, p11, p22, p21, nrow = 2, ncol = 2)
```

```{r fl_permutation, fig.height = 3, fig.cap = "Permutation distribution of the Freedman-Lane tests of correlation between daily_hrdq and treatment after controlling for covariates."}
# method 3 : freedman lane perm residuals

lm1_no_tr <- lm(Outcome~Baseline + factor(SITEID), data = daily_hrdq)
lm1_resid <- residuals(lm1_no_tr)
lm1_yhat <- fitted(lm1_no_tr)
observed_t1 <- summary(lm1)[["coefficients"]]["tr", "t value"]
lm1_t_distr <-  replicate(1e4, {
  lm1_resid_perm <- permute_within_groups(lm1_resid, daily_hrdq$SITEID)
  daily_hrdq$response_fl <-  lm1_yhat + lm1_resid_perm
  lm1_perm <- lm(response_fl ~ Baseline + tr + factor(SITEID), data = daily_hrdq)
  summary(lm1_perm)[["coefficients"]]["tr", "t value"]
})

fl_pvalue_1 <- t2p(observed_t1, lm1_t_distr, alternative = "two-sided")

lm2_no_tr <- lm(difference ~ factor(SITEID), data = daily_hrdq)
lm2_resid <- residuals(lm2_no_tr)
lm2_yhat <- fitted(lm2_no_tr)
observed_t2 <- summary(lm2)[["coefficients"]]["tr", "t value"]
lm2_t_distr <-  replicate(1e4, {
  lm2_resid_perm <- permute_within_groups(lm2_resid, daily_hrdq$SITEID)
  daily_hrdq$response_fl <-  lm2_yhat + lm2_resid_perm
  lm2_perm <- lm(response_fl ~ tr + factor(SITEID), data = daily_hrdq)
  summary(lm2_perm)[["coefficients"]]["tr", "t value"]
})

fl_pvalue_2 <- t2p(observed_t2, lm2_t_distr, alternative = "two-sided")

data.frame("perm" = c(lm1_t_distr, lm2_t_distr),
           "model" = c(rep("Outcome", length(lm1_t_distr)), 
                       rep("Differenced", length(lm2_t_distr))),
           "xintercept" = c(rep(observed_t1, length(lm1_t_distr)),
                            rep(observed_t2, length(lm2_t_distr)))) %>%
  ggplot(aes(x = perm, fill = model)) +
  geom_histogram() +
  facet_grid(~model) +
  geom_vline(aes(xintercept = xintercept))

```

Table \ref{primary_endpoint_res} shows the p-values for each of the four methods, using the two models (treatment controlling for baseline, and the difference of treatment minus baseline).
For the models using the treatment measure as the dependent variable, the effect of `daily_hrdq` is significant at the $0.05$ level in two of the five models
and is significant at the $0.1$ level in four of the five models.
It is never significant in the models using the difference from baseline to treatment.
As expected, the stratified permutation test using the sum of differences in means across sites has low power.
We restrict our attention to the model using treatment measurement and controlling for the baseline, as this is what RB's original analysis does.
Now, we run this procedure for all continuous variables.
Table \ref{primary_endpoint_res} shows the results.

```{r primary_endpoint_res, results = "asis", fig.cap = "Comparison of p-values for two measures (average outcome during treatment vs. difference of average outcome and average baseline) of the primary endpoint."}

outcome_pvalues <- c(summary(aov(lm1))[[1]]["tr", "Pr(>F)"],
                     diff_means_pvalue1,
                     diff_means_bystrata_pvalue1,
                     lm_pvalue_1,
                     fl_pvalue_1)
differenced_pvalues <- c(summary(aov(lm2))[[1]]["tr", "Pr(>F)"],
                     diff_means_pvalue2,
                     diff_means_bystrata_pvalue2,
                     lm_pvalue_2,
                     fl_pvalue_2)
pvalues_table <- cbind(differenced_pvalues, outcome_pvalues)

tests <- c("Parametric ANOVA", "Unadjusted permutation", "Unadjusted permutation (summed across strata)", "Linear regression permutation", "Residual permutation")
rownames(pvalues_table) <- tests
colnames(pvalues_table) <- c("Differences", "Outcome")

print(xtable(pvalues_table, digits = 3,
             caption = "Comparison of p-values for two measures (average outcome during treatment vs. difference of average outcome and average baseline) of the primary endpoint."),
      include.rownames = TRUE)
```

```{r results_table_make}
set.seed(919547773) # Generated on Random.org Timestamp: 2016-11-09 15:39:29 UTC

continuous_vars <- c("heart_freq", "regurg_freq", "dysp_freq",
                    "daily_heart", "daily_regurg", "daily_hrdq",
                    "daily_dysp")

tests <- c("Parametric ANOVA", "Unadjusted permutation", "Linear regression permutation", "Residual permutation")
pvalues_table_contin <- as.data.frame(matrix(NA, nrow = length(continuous_vars), ncol = 4))
i <- 0

for(col in continuous_vars){
  i <- i+1
  tmpdata <- reshape_ga_continuous(col)
  lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = tmpdata)
  pvalues_table_contin[i, 1] <- summary(aov(lm1))[[1]]["tr", "Pr(>F)"]

  observed_diff_means <- mean(tmpdata[tmpdata$tr == 1, ]$Outcome) - mean(tmpdata[tmpdata$tr == 0, ]$Outcome)
  diff_means_distr <- stratified_two_sample(group = tmpdata$tr, response = tmpdata$Outcome, stratum = tmpdata$SITEID, reps = 1e3)
  pvalues_table_contin[i, 2] <- t2p(observed_diff_means, diff_means_distr, alternative = "two-sided")

  observed_t <- summary(lm1)[["coefficients"]]["tr", "t value"]
  lm1_t_distr <-  replicate(1e3, {
    tmpdata$tr_perm <- permute_within_groups(tmpdata$tr, tmpdata$SITEID)
  lm1_perm <- lm(Outcome ~ Baseline + tr_perm + factor(SITEID), data = tmpdata)
  summary(lm1_perm)[["coefficients"]]["tr_perm", "t value"]
  })
  pvalues_table_contin[i, 3] <- t2p(observed_t, lm1_t_distr, alternative = "two-sided")

  lm_no_tr <- lm(Outcome ~ Baseline + factor(SITEID), data = tmpdata)
  lm_resid <- residuals(lm_no_tr)
  lm_yhat <- fitted(lm_no_tr)
  observed_t <- summary(lm1)[["coefficients"]]["tr", "t value"]
  lm_t_distr <-  replicate(1e3, {
      lm_resid_perm <- permute_within_groups(lm_resid, tmpdata$SITEID)
      tmpdata$response_fl <-  lm_yhat + lm_resid_perm
      lm_perm <- lm(response_fl ~ Baseline + tr + factor(SITEID), data = tmpdata)
      summary(lm_perm)[["coefficients"]]["tr", "t value"]
})

  pvalues_table_contin[i, 4] <- t2p(observed_t, lm_t_distr, alternative = "two-sided")
}
```
```{r results_table_print, results = "asis", fig.cap = "Comparison of p-values from four tests, for each continuous endpoint."}
rownames(pvalues_table_contin) <- continuous_vars
colnames(pvalues_table_contin) <- tests

print(xtable(pvalues_table_contin, 
             digits = 3,
             align = paste0(c("r|", rep("p{1in}", ncol(pvalues_table_contin))), 
                            collapse = ""),
             caption = "Comparison of p-values from four tests, for each continuous endpoint."),
      include.rownames = TRUE)
```

# Analysis - Ordinal variables

For ordinal variables, it doesn't make sense to do ANOVA.
Instead, we do a multi-aspect analysis to test whether there is any difference in the outcome distributions between the drug and placebo groups.
For this, we use the same stratified permutation test approach as above.
However, we use three different test statistics and combine them into a single, global test using the nonparametric combination (NPC) method.
The test statistics are the difference in mean outcomes between the drug and placebo group, the Anderson-Darling statistic, and the Kolmogorov-Smirnov statistic.
We use the Tippett combining function.

The outcome measures used are the difference in median measures during the treatment week minus the median measures during the baseline week.
Recall from Figure \ref{fig:hist_ordinal_vars} that it is not clear whether the distributions are different.
Table \ref{ordinal_results_print} shows the p-values for the three partial tests plus the global p-value.
None is significant.
It appears that treating the variables as ordinal results in a loss of information and, consequently, power.

```{r ordinal_tests}
# Compute difference in mean value (median outcome) between treatment and placebo
mean_median <- function(outcome, treatment) {
  tr_levels <- sort(unique(treatment))
  mean(outcome[treatment == tr_levels[2]], na.rm = TRUE) - 
    mean(outcome[treatment == tr_levels[1]], na.rm = TRUE)
  }

# Anderson-Darling
ad<-function(outcome, treatment){    
  n <- length(treatment)   
  W <- table(outcome, treatment)
  ecdf <- apply(W, 2, cumsum)
  ecdf_comb <- apply(ecdf, 1, sum)

  ratio <- ecdf[,2]/(ecdf_comb*(n-ecdf_comb))
  j <- length(ratio)
  T_ad <- sum(ratio[seq_len(j-1)])
  return(T_ad)
}

# Kolmogorov-Smirnov
ks <- function(outcome, treatment){
  tr_levels <- sort(unique(treatment))
  # Assume that larger treatment level is active treatment, smaller other is control
  g1 <- outcome[treatment == tr_levels[2]]
  g0 <- outcome[treatment == tr_levels[1]]
  suppressWarnings(ks.test(g1, g0)$statistic)
}

multiaspect_test <- function(outcome, treatment, strata, reps = 1000){
  obs_ts <- c(mean_median(outcome, treatment),
              ad(outcome, treatment),
              ks(outcome, treatment))
  names(obs_ts) <- c("Difference in Means", "Anderson-Darling", "Kolmogorov-Smirnov")
  perm_distr <- t(replicate(reps, {
            perm_treatment <- permute_within_groups(treatment, strata)
            c(mean_median(outcome, perm_treatment),
              ad(outcome, perm_treatment),
              ks(outcome, perm_treatment))
            }))

  partial_pvalues <- sapply(seq_along(obs_ts), function(i){
    t2p(obs_ts[i], perm_distr[, i], "two-sided")
  })
  names(partial_pvalues) <- c("Difference in Means", "Anderson-Darling", "Kolmogorov-Smirnov")
  global_pvalue <- npc(obs_ts, perm_distr, alternatives="two-sided", combine = "tippett")

  return(list(
    "Partial test statistics" = obs_ts,
    "Partial p-values"        = partial_pvalues,
    "Global p-value"          = global_pvalue
  ))
}
```

```{r ordinal_results_make}
set.seed(609726531) # Generated from Random.org Timestamp: 2016-11-09 15:38:50 UTC
ordinal_vars <- c("heart_sev_f", "regurg_sev_f", "dysp_sev_f",
                    "heart_freq_f", "regurg_freq_f", "dysp_freq_f")

pvalues_table_ord <- as.data.frame(matrix(NA, nrow = length(ordinal_vars), ncol = 4))
i <- 0

for(col in ordinal_vars){
  i <- i+1
  tmpdata <- reshape_ga_ordinal(col)
  res <- multiaspect_test(outcome = tmpdata$difference, treatment = tmpdata$tr, strata = tmpdata$SITEID, reps = 1e3)
  pvalues_table_ord[i, 1] <- res$`Partial p-values`[1]
  pvalues_table_ord[i, 2] <- res$`Partial p-values`[2]
  pvalues_table_ord[i, 3] <- res$`Partial p-values`[3]
  pvalues_table_ord[i, 4] <- res$`Global p-value` 
} 
```

```{r ordinal_results_print, results = "asis"}
rownames(pvalues_table_ord) <- ordinal_vars
tests <- c("Difference in Means", "Anderson-Darling", "Kolmogorov-Smirnov", "Global P-value")
colnames(pvalues_table_ord) <- tests

print(xtable(pvalues_table_ord,  
             digits = 3,
             align = paste0(c("r|", rep("p{0.75in}", ncol(pvalues_table_ord))), 
                            collapse = ""),
             caption = "Comparison of p-values from three tests plus a global p-value, for each ordinal endpoint."), 
      include.rownames = TRUE)
```

### 2016-11-25 simulations/simulation.Rmd
First commit: add content in progress

---
title: "ANOVA Comparison Simulations"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE,
                      tidy = TRUE, tidy.opts = list(width.cutoff = 60),
                      cache = TRUE)
#devtools::install_github("statlab/permuter")
require(permuter)
require(dplyr)
require(reshape2)
require(ggplot2)
require(xtable)
options(xtable.comment = FALSE, xtable.include.rownames = FALSE, xtable.align = "p{3.5cm}")
require(grid)
require(gridExtra)
```

# The model

We assume an additive linear model:

$$Y_{ij1} =\beta_0Y_{i0} + \beta_{j} + \gamma_j Z_{ij} + \varepsilon_{ij}$$

for individuals $i = 1, \dots, n_j$, $j = 1, \dots, J$.
$\beta_0$ is the coefficient for the baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being at site $j$, $Z_{ij}$ is the treatment level, $\gamma_j$ is the effect of treatment at site $j$, and $\varepsilon_{ij}$ is an error term.
We will assume that $\beta_0 = 1$.

Suppose there are three sites with $\beta_1 = 1, \beta_2 = 1.5,$ and $\beta_3 = 2$.
Assume that there are 16 individuals per site and treatment assignment is balanced, i.e. 8 people receive each treatment at each site.

set up 1: $\gamma_1 = \gamma_2 = \gamma_3$
set up 2: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$
errors 1: $\varepsilon \sim N(0, \sigma^2)$
errors 2: $\varepsilon \sim t(2)$

```{r generate_sample}
generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
  # Input:
  # gamma = the magnitude of the treatment effect
  # effect = "same effect" or "single site effect" - which sites have a tr effect > 0?
  # errors = "normal" or "heavy"
  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), and SITEID (stratum)

  SITEID <- rep(1:3, times = n)
  N <- sum(n)
  beta <- c(1, 1.5, 2)

  # What is the treatment effect?
  if(effect == "same effect"){
    gamma_vec <- rep(gamma, N)
  } else {
    gamma_vec <- rep(c(gamma, 0, 0), times = n)
  }

  # Generate errors
  if(errors == "normal"){
    epsilon <- rnorm(N)
  } else {
    epsilon <- rt(N, df = 2)
  }

  # Generate covariates
  Y0 <- rnorm(N)
  Z <- rep(0:1, length.out = N)
  site_effect <- rep(beta, times = n)
  Y1 <- Y0 + gamma_vec*Z + site_effect + epsilon
  return(data.frame(Y1, Y0, Z, SITEID))
}

generate_simulated_pvalues <- function(dataset, reps = 1e3){
  # Inputs:
  # dataset = a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), and SITEID (stratum)
  # Returns: a vector of p-values
  # first element is the p-value from the ANOVA
  # second element is the p-value from the stratified two-sample permutation test
  # third element is the p-value from the linear model test, permuting treatment
  # fourth element is the p-value from the Freedman-Lane linear model test, permuting residuals

  # ANOVA
  modelfit <- lm(Y1 ~ Y0 + Z + factor(SITEID), data = dataset)
  resanova <- summary(aov(modelfit))
  anova_pvalue <- resanova[[1]]["Z", "Pr(>F)"]

  # Permutation test
  observed_diff_means <- mean(dataset$Y1[dataset$Z == 1]) - mean(dataset$Y1[dataset$Z == 0])
  diff_means_distr <- stratified_two_sample(group = dataset$Z, response = dataset$Y1, stratum = dataset$SITEID, reps = reps)
  # diff_means_distr2 <-  replicate(reps, {
  #   Z_perm <- permute_within_groups(dataset$Z, dataset$SITEID)
  #   mean(dataset$Y1[Z_perm == 1]) - mean(dataset$Y1[Z_perm == 0])
  # })
  perm_pvalue <- t2p(observed_diff_means, diff_means_distr, alternative = "two-sided")

  # Diffed permutation test
  dataset$diff <- dataset$Y1 - dataset$Y0
  observed_diff_means2 <- mean(dataset$diff[dataset$Z == 1]) - mean(dataset$diff[dataset$Z == 0])
  diff_means_distr2 <- stratified_two_sample(group = dataset$Z, response = dataset$diff, stratum = dataset$SITEID, reps = reps)
  perm_pvalue2 <- t2p(observed_diff_means2, diff_means_distr2, alternative = "two-sided")

  # Permutation of treatment in linear model
  observed_t1 <- summary(modelfit)[["coefficients"]]["Z", "t value"]
  lm1_t_distr <-  replicate(reps, {
    dataset$Z_perm <- permute_within_groups(dataset$Z, dataset$SITEID)
    lm1_perm <- lm(Y1 ~ Y0 + Z_perm + factor(SITEID), data = dataset)
    summary(lm1_perm)[["coefficients"]]["Z_perm", "t value"]
  })
  lm_pvalue <- t2p(observed_t1, lm1_t_distr, alternative = "two-sided")

  # Freedman-Lane linear model residual permutation
  lm2_no_tr <- lm(Y1~Y0 + factor(SITEID), data = dataset)
  lm2_resid <- residuals(lm2_no_tr)
  lm2_yhat <- fitted(lm2_no_tr)
  lm2_t_distr <-  replicate(reps, {
    lm2_resid_perm <- permute_within_groups(lm2_resid, dataset$SITEID)
    dataset$response_fl <-  lm2_yhat + lm2_resid_perm
    lm2_perm <- lm(response_fl ~ Y0 + Z + factor(SITEID), data = dataset)
    summary(lm2_perm)[["coefficients"]]["Z", "t value"]
  })
  fl_pvalue <- t2p(observed_t1, lm2_t_distr, alternative = "two-sided")

  return(c("ANOVA" = anova_pvalue, 
           "Stratified Permutation" = perm_pvalue, 
           "Diffed Stratified Permutation" = perm_pvalue2,
           "LM Permutation" = lm_pvalue, 
           "Freedman-Lane" = fl_pvalue))
}
```

```{r function_power_curves}
compute_power <- function(pvalues){
  sapply((0:99)/100, function(p) mean(pvalues <= p, na.rm = TRUE))
}

plot_power_curves <- function(power_mat, title){
  melt(power_mat) %>% 
  mutate("pvalue" = Var1/100) %>%
  mutate("Method" = Var2) %>%
  ggplot(aes_string(x = "pvalue", y = "value", color = "Method")) +
  geom_line() +
  xlab("P-value") +
  ylab("Power") +
  ggtitle(title)
}

plot_pvalue_hist <- function(pvalue_mat, title){
  melt(pvalue_mat) %>%
  mutate("Method" = Var2) %>%
  ggplot(aes(x = value, fill = Method)) +
  geom_histogram() +
  facet_wrap(~Method) +
  ggtitle(title)
}

plot_pvalue_scatter <- function(pvalue_mat, title){
  pvalue_mat %>%
  as.data.frame() %>%
  select(ANOVA, strat = starts_with("Stratified")) %>%
  ggplot(aes(x = ANOVA, y = strat)) + 
  geom_point() +
  xlim(0, 1) + ylim(0, 1) +
  ylab("Stratified Permutation") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  ggtitle(title)
}
```

```{r design_null, fig.height = 4}
set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC

design0_pvalues <- replicate(100, {
  tmp <- generate_simulated_data(gamma = 0, effect = "same effect", errors = "normal")
  generate_simulated_pvalues(tmp)
})
design0_pvalues <- t(design0_pvalues)
colnames(design0_pvalues) <- c("ANOVA", "Stratified Permutation", "Diffed Permutation", "LM Permutation", "Freedman-Lane")
design0_power <- apply(design0_pvalues, 2, compute_power)

plot_power_curves(design0_power, "No Effect, Normal Errors")

plot_pvalue_hist(design0_pvalues, "No Effect, Normal Errors")

plot_pvalue_scatter(design0_pvalues, "Simulation P-values for Data with \n No Effect, Normal Errors")
```

```{r design1, fig.height = 4}
set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC

design1_pvalues <- replicate(100, {
  tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
  generate_simulated_pvalues(tmp)
})
design1_pvalues <- t(design1_pvalues)
colnames(design1_pvalues) <- c("ANOVA", "Stratified Permutation", "Diffed Permutation", "LM Permutation", "Freedman-Lane")
design1_power <- apply(design1_pvalues, 2, compute_power)

plot_power_curves(design1_power, "Constant Additive Effect of 1, Normal Errors")

plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 1, Normal Errors")

#plot_pvalue_scatter(design1_pvalues, "Simulation P-values for Data with \n Constant Additive Effect of 1, Normal Errors")
```

```{r design2, fig.height = 4}
set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC

design2_pvalues <- replicate(100, {
  tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal")
  generate_simulated_pvalues(tmp)
})
design2_pvalues <- t(design2_pvalues)
colnames(design2_pvalues) <- c("ANOVA", "Stratified Permutation", "Diffed Permutation", "LM Permutation", "Freedman-Lane")
design2_power <- apply(design2_pvalues, 2, compute_power)

plot_power_curves(design2_power, "Effect of 1 at a Single Site, Normal Errors")

plot_pvalue_hist(design2_pvalues, "Effect of 1, Normal Errors")

plot_pvalue_scatter(design2_pvalues, "Simulation P-values for Data with \n Effect of 1 at a Single Site, Normal Errors")
```

```{r design3, fig.height = 4}
set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC

design3_pvalues <- replicate(100, {
  tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "heavy")
  generate_simulated_pvalues(tmp)
})
design3_pvalues <- t(design3_pvalues)
colnames(design3_pvalues) <- c("ANOVA", "Stratified Permutation", "Diffed Permutation", "LM Permutation", "Freedman-Lane")
design3_power <- apply(design3_pvalues, 2, compute_power)

plot_power_curves(design3_power, "Constant Additive Effect of 1, Heavy-tailed Errors")

plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 1, Heavy-tailed Errors")

plot_pvalue_scatter(design3_pvalues, "Simulation P-values for Data with \n Constant Additive Effect of 1, Heavy-tailed Errors")
```

```{r design4, fig.height = 4}
set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC

design4_pvalues <- replicate(100, {
  tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "heavy")
  generate_simulated_pvalues(tmp)
})
design4_pvalues <- t(design4_pvalues)
colnames(design4_pvalues) <- c("ANOVA", "Stratified Permutation", "Diffed Permutation", "LM Permutation", "Freedman-Lane")
design4_power <- apply(design4_pvalues, 2, compute_power)

plot_power_curves(design4_power, "Effect of 1 at a Single Site, Heavy-tailed Errors")

plot_pvalue_hist(design4_pvalues, "Effect of 1, Heavy-tailed Errors")

plot_pvalue_scatter(design4_pvalues, "Simulation P-values for Data with \n Effect of 1 at a Single Site, Heavy-tailed Errors")
```

### 2016-11-25 data/clinical_cleaned.R

@@ -1,0 +1,29 @@
+# Kellie Ottoboni
+# November 14, 2016
+# Script to privatize the RB dataset
+
+require(dplyr)
+require(reshape2)
+
+
+
+ga <- read.csv("GA1102.csv", header = TRUE, stringsAsFactors = FALSE)
+
+# Remove missing
+ga <- ga %>% filter(tr != ".")
+
+# Change treatment values to drug A vs drug B
+ga <- ga %>% mutate(tr = factor(tr, labels = c("A", "B")))
+
+# Rename subject ids
+ga <- ga %>% mutate(SUBJID = as.numeric(factor(SUBJID)))
+
+# Rename site ids
+ga <- ga %>% mutate(SITEID = as.numeric(factor(SITEID)))
+
+# Take means
+ga <- ga %>% group_by(SUBJID, SITEID, VISITNUM, tr, country) %>%
+  summarise_each(funs(mean)) %>% 
+  ungroup()
+
+write.csv(ga, file = "clinical_cleaned.csv", row.names = FALSE)

### 2016-11-29 simulations/simulation.Rmd
Fix simulations: treat baseline as fixed pretreatment

@@
 
-We assume an additive linear model:
+We assume the following linear data-generating process:
 
-$$Y_{ij1} =\beta_0Y_{i0} + \beta_{j} + \gamma_j Z_{ij} + \varepsilon_{ij}$$
+$$Y_{ij1} =\beta_0Y_{ij0} + \beta_{j} + \gamma_j Z_{ij} + \varepsilon_{ij}$$
 
@@
 
-set up 1: $\gamma_1 = \gamma_2 = \gamma_3$
-set up 2: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$
-errors 1: $\varepsilon \sim N(0, \sigma^2)$
-errors 2: $\varepsilon \sim t(2)$
+We use two designs:
+
+* Design 1: $\gamma_1 = \gamma_2 = \gamma_3 = 1$. This is the standard assumption of a constant, additive treatment effect.
+* Design 2: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$.
+This is a constant, additive treatment effect at site 1, but no treatment effect at sites 2 and 3.
+This is a simplistic case of heterogeneous treatment effects.
+
+With these two designs, we vary the distribution of $\varepsilon$.
+In the first case, we use $\varepsilon \sim N(0, 1)$ to mimic the usual ANOVA assumptions.
+In the second case, $\varepsilon \sim t(2)$ so the errors are heavy-tailed.
+Thus, there are four total simulation designs.
+
+We compare five tests:
+
+* ANOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for site.
+* Stratified permutation: we permute treatment assignment within site, then take the difference in means between treated and control outcomes $Y_1$
+* Differenced permutation: we do the same permutation procedure as the stratified permutation test, except we use the difference between outcome and baseline, $Y_1 - Y_0$
+* Linear model (LM) permutation: we use the same stratified permutation procedure as above, except use the $t$-statistic for the coefficient on treatment in the linear regression of $Y_1$ on $Y_0$, $Z$, and site dummies
+* Freedman-Lane test: see the other Rmd document for a full description of this procedure
+
+Throughout our simulations, we first fix $Y_0$ and site ID.
+Treatment $Z$ and the errors $\varepsilon$ are randomly drawn according to their respective distributions.
+Then, $Y_1$ is constructed using the linear data-generating process above.
+We regenerate $Z$, $\varepsilon$, and $Y_1$ 100 times for each design, then compute the empirical power of the five tests.
+
+# Data-generation, tests, and plotting functions
 
@@
   Y1 <- Y0 + gamma_vec*Z + site_effect + epsilon
-  return(data.frame(Y1, Y0, Z, SITEID))
+  return(data.frame(Y1, Y0, Z, gamma_vec, SITEID, site_effect, epsilon))
 }
@@
 
-  # Permutation test
+  # Stratified permutation test of Y1
   observed_diff_means <- mean(dataset$Y1[dataset$Z == 1]) - mean(dataset$Y1[dataset$Z == 0])
   diff_means_distr <- stratified_two_sample(group = dataset$Z, response = dataset$Y1, stratum = dataset$SITEID, reps = reps)
-  # diff_means_distr2 <-  replicate(reps, {
-  #   Z_perm <- permute_within_groups(dataset$Z, dataset$SITEID)
-  #   mean(dataset$Y1[Z_perm == 1]) - mean(dataset$Y1[Z_perm == 0])
-  # })
   perm_pvalue <- t2p(observed_diff_means, diff_means_distr, alternative = "two-sided")
 
-  # Diffed permutation test
+  # Diffed permutation test of Y1-Y0
   dataset$diff <- dataset$Y1 - dataset$Y0
@@
            "Stratified Permutation" = perm_pvalue, 
-           "Diffed Stratified Permutation" = perm_pvalue2,
+           "Differenced Permutation" = perm_pvalue2,
            "LM Permutation" = lm_pvalue, 
@@
   geom_line() +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
   xlab("P-value") +
@@
 
+# Test level: simulation under the null
+
+Before testing for different kinds of effects, we begin checking that the tests have the correct level.
+We follow the procedure described above, using an effect size of $\gamma = 0$ at all sites and using standard normal errors.
+To have the correct level means that the test rate of rejection at level $\alpha$ is $\alpha 100\%$.
+In other words, the p-values are uniformly distributed and the power curve should coincide with the line with slope $1$ through the origin.
+Figure \ref{fig:design_null_plots} demonstrates that this is the case.
+If anything, the differenced stratified permutation test has fewer than $\alpha 100\%$ false positives when using level $\alpha$.
+
 ```{r design_null, fig.height = 4}
@@
 
+tmp <- generate_simulated_data(gamma = 0, effect = "same effect", errors = "normal")
 design0_pvalues <- replicate(100, {
-  tmp <- generate_simulated_data(gamma = 0, effect = "same effect", errors = "normal")
+  tmp$epsilon <- rnorm(nrow(tmp))
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 design0_pvalues <- t(design0_pvalues)
-colnames(design0_pvalues) <- c("ANOVA", "Stratified Permutation", "Diffed Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design0_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design0_power <- apply(design0_pvalues, 2, compute_power)
+```
 
+```{r design_null_plots, fig.height = 4, fig.align="center"}
 plot_power_curves(design0_power, "No Effect, Normal Errors")
@@
 
-plot_pvalue_scatter(design0_pvalues, "Simulation P-values for Data with \n No Effect, Normal Errors")
+#plot_pvalue_scatter(design0_pvalues, "Simulation P-values for Data with \n No Effect, Normal Errors")
 ```
+
+# Design 1: Constant additive effect, normal errors
+
+There is no discernable difference in power between the ANOVA, the differenced stratified permutation test, the LM permutations, or the Freedman-Lane test.
+However, the simple stratified permutation test of $Y_1$ has substantially less power than the other four.
+Without controlling for the baseline values, the variance in $Y_1$ masks the treatment effect.
 
@@
 
+tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
 design1_pvalues <- replicate(100, {
-  tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
+  tmp$epsilon <- rnorm(nrow(tmp))
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
 })
+
 design1_pvalues <- t(design1_pvalues)
-colnames(design1_pvalues) <- c("ANOVA", "Stratified Permutation", "Diffed Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design1_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design1_power <- apply(design1_pvalues, 2, compute_power)
+```
 
+```{r design1_plot, fig.height = 4, fig.align="center"}
 plot_power_curves(design1_power, "Constant Additive Effect of 1, Normal Errors")
@@
 
+# Design 2: Single site effect, normal errors
+
+A similar pattern emerges: the simple stratified permutation test has low power, while the other four power curves roughly coincide.
+The Freedman-Lane test may have the highest power for small p-values, but this could also just be noise.
+All five power curves are closer to the line passing through the origin: since the effect is only present at one site, it is more difficult to detect.
+
 ```{r design2, fig.height = 4}
@@
 
+tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal")
 design2_pvalues <- replicate(100, {
-  tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal")
+  tmp$epsilon <- rnorm(nrow(tmp))
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 design2_pvalues <- t(design2_pvalues)
-colnames(design2_pvalues) <- c("ANOVA", "Stratified Permutation", "Diffed Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design2_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design2_power <- apply(design2_pvalues, 2, compute_power)
+```
 
+```{r design2_plot, fig.height = 4, fig.align="center"}
 plot_power_curves(design2_power, "Effect of 1 at a Single Site, Normal Errors")
@@
 
-plot_pvalue_scatter(design2_pvalues, "Simulation P-values for Data with \n Effect of 1 at a Single Site, Normal Errors")
+#plot_pvalue_scatter(design2_pvalues, "Simulation P-values for Data with \n Effect of 1 at a Single Site, Normal Errors")
 ```
+
+# Design 3: Constant additive effect, heavy-tailed errors
+
+Again, we find that the four power curves coincide while the curve for the stratified permutation test is below the others.
+Here, the difference between the curves is not large.
+Controlling for baseline $Y_0$ does not substantially help reduce variance.
 
@@
 
+tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "heavy")
 design3_pvalues <- replicate(100, {
-  tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "heavy")
+  tmp$epsilon <- rt(nrow(tmp), df = 2)
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 design3_pvalues <- t(design3_pvalues)
-colnames(design3_pvalues) <- c("ANOVA", "Stratified Permutation", "Diffed Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design3_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design3_power <- apply(design3_pvalues, 2, compute_power)
+```
 
+```{r design3_plot, fig.height = 4, fig.align="center"}
 plot_power_curves(design3_power, "Constant Additive Effect of 1, Heavy-tailed Errors")
@@
 
-plot_pvalue_scatter(design3_pvalues, "Simulation P-values for Data with \n Constant Additive Effect of 1, Heavy-tailed Errors")
+#plot_pvalue_scatter(design3_pvalues, "Simulation P-values for Data with \n Constant Additive Effect of 1, Heavy-tailed Errors")
 ```
+
+# Design 4: single site effect, heavy-tailed errors
+
+The noise from the heavy-tailed errors masks the treatment effect so much that controlling for baseline measures makes no difference.
+There is almost no power to detect an effect using any of the five tests.
 
@@
 
+tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "heavy")
 design4_pvalues <- replicate(100, {
-  tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "heavy")
+  tmp$epsilon <- rt(nrow(tmp), df = 2)
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 design4_pvalues <- t(design4_pvalues)
-colnames(design4_pvalues) <- c("ANOVA", "Stratified Permutation", "Diffed Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design4_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design4_power <- apply(design4_pvalues, 2, compute_power)
+```
 
+```{r design4_plot, fig.height = 4, fig.align="center"}
 plot_power_curves(design4_power, "Effect of 1 at a Single Site, Heavy-tailed Errors")
@@
 
-plot_pvalue_scatter(design4_pvalues, "Simulation P-values for Data with \n Effect of 1 at a Single Site, Heavy-tailed Errors")
+#plot_pvalue_scatter(design4_pvalues, "Simulation P-values for Data with \n Effect of 1 at a Single Site, Heavy-tailed Errors")
 ```
+
+# Design 5: Unpredictive baseline
+
+Suppose that the baseline measure is not strongly predictive of the outcome.
+Then, we'd expect that controlling for baseline will not improve the power of the test, and may even make it worse.
+Suppose that in the linear data-generating process above, we have $\beta_0 = 0.25$, meaning that $Y_1$ and $Y_0$ have a correlation of $0.25$.
+
+The power curves for this design look similar to those from Design 1 with one important difference: 
+the stratified permutation test has power very close to ANOVA and the two linear model tests,
+while the differenced permutation test has lower power.
+This suggests that one should be careful when incorporating control variables;
+naively taking the difference $Y_1 - Y_0$ does not capture the correct relationship between baseline and outcome.
+
+```{r design5, fig.height = 4}
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+
+tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
+design5_pvalues <- replicate(100, {
+  tmp$epsilon <- rnorm(nrow(tmp))
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Y1 <- 0.25*tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  generate_simulated_pvalues(tmp)
+})
+
+design5_pvalues <- t(design5_pvalues)
+colnames(design5_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+design5_power <- apply(design5_pvalues, 2, compute_power)
+```
+
+```{r design5_plot, fig.height = 4, fig.align="center"}
+plot_power_curves(design5_power, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
+
+plot_pvalue_hist(design5_pvalues, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
+
+#plot_pvalue_scatter(design5_pvalues, "Simulation P-values for Data with \n Constant Additive Effect of 1, Normal Errors")
+```

### 2016-11-29 simulations/simulation.Rmd
Add a few more simulation scenarios

@@
 
-# Test level: simulation under the null
+## Test level: simulation under the null
 
@@
 plot_pvalue_hist(design0_pvalues, "No Effect, Normal Errors")
-
-#plot_pvalue_scatter(design0_pvalues, "Simulation P-values for Data with \n No Effect, Normal Errors")
 ```
 
-# Design 1: Constant additive effect, normal errors
+## Design 1: Constant additive effect, normal errors
 
@@
 plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 1, Normal Errors")
-
-#plot_pvalue_scatter(design1_pvalues, "Simulation P-values for Data with \n Constant Additive Effect of 1, Normal Errors")
 ```
 
-# Design 2: Single site effect, normal errors
+## Design 2: Single site effect, normal errors
 
@@
 plot_pvalue_hist(design2_pvalues, "Effect of 1, Normal Errors")
-
-#plot_pvalue_scatter(design2_pvalues, "Simulation P-values for Data with \n Effect of 1 at a Single Site, Normal Errors")
 ```
 
-# Design 3: Constant additive effect, heavy-tailed errors
+## Design 3: Constant additive effect, heavy-tailed errors
 
@@
 plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 1, Heavy-tailed Errors")
-
-#plot_pvalue_scatter(design3_pvalues, "Simulation P-values for Data with \n Constant Additive Effect of 1, Heavy-tailed Errors")
 ```
 
-# Design 4: single site effect, heavy-tailed errors
+## Design 4: single site effect, heavy-tailed errors
 
@@
 plot_pvalue_hist(design4_pvalues, "Effect of 1, Heavy-tailed Errors")
-
-#plot_pvalue_scatter(design4_pvalues, "Simulation P-values for Data with \n Effect of 1 at a Single Site, Heavy-tailed Errors")
 ```
 
-# Design 5: Unpredictive baseline
+## Design 5: Unpredictive baseline
 
@@
 plot_pvalue_hist(design5_pvalues, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
+```
 
-#plot_pvalue_scatter(design5_pvalues, "Simulation P-values for Data with \n Constant Additive Effect of 1, Normal Errors")
+# Imbalanced designs
+
+Suppose that instead of having 16 individuals per site, we distribute them unequally across sites:
+site 1 has 8 patients, site 2 has 16 patients, and site 3 has 24 patients.
+This imbalance does not violate any assumptions of the ANOVA.
+However, it may reduce power if the effect is concentrated in sites with fewer patients.
+
+## Design 6: Constant additive effect, normal errors
+
+The power curves here are very similar to those in Design 1.
+This result is expected, since the effect is still present among all patients who received treatment.
+
+```{r design6, fig.height = 4}
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+
+tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal", n = c(8, 16, 24))
+design6_pvalues <- replicate(100, {
+  tmp$epsilon <- rnorm(nrow(tmp))
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  generate_simulated_pvalues(tmp)
+})
+
+design6_pvalues <- t(design6_pvalues)
+colnames(design6_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+design6_power <- apply(design6_pvalues, 2, compute_power)
 ```
+
+```{r design6_plot, fig.height = 4, fig.align="center"}
+plot_power_curves(design6_power, "Constant Additive Effect of 1, Normal Errors")
+
+plot_pvalue_hist(design6_pvalues, "Constant Additive Effect of 1, Normal Errors")
+```
+
+## Design 7: Single site effect, normal errors
+
+There is very low power to detect the effect here, since in the simulated data, the site with only 8 patients was the only site with a nonzero effect of treatment.
+The treatment effect is on the same order of magnitude as the error variance.
+
+```{r design7, fig.height = 4}
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+
+tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal", n = c(8, 16, 24))
+design7_pvalues <- replicate(100, {
+  tmp$epsilon <- rnorm(nrow(tmp))
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  generate_simulated_pvalues(tmp)
+})
+design7_pvalues <- t(design7_pvalues)
+colnames(design7_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+design7_power <- apply(design7_pvalues, 2, compute_power)
+```
+
+```{r design7_plot, fig.height = 4, fig.align="center"}
+plot_power_curves(design7_power, "Effect of 1 at a Single Site, Normal Errors")
+
+plot_pvalue_hist(design7_pvalues, "Effect of 1, Normal Errors")
+```
+
+# Heteroskedastic errors
+
+One assumption of ANOVA is that errors are homoskedastic: they have the same variance across groups.
+Suppose that this does not hold in our data.
+Imagine we have three sites with 16 patients each, but they have different error distributions:
+site 1 has standard normal errors,
+site 2 has normally distributed errors with mean 0 and variance 2,
+and site 3 has errors with a $t$-distribution on 2 degrees of freedom.
+
+## Design 8: heteroskedastic errors, constant treatment effect
+
+```{r design8, fig.height = 4}
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+nn <- 16
+
+tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal", n = rep(nn, 3))
+design8_pvalues <- replicate(100, {
+  tmp$epsilon[tmp$SITEID == 1] <- rnorm(nn)
+  tmp$epsilon[tmp$SITEID == 2] <- rnorm(nn, sd = sqrt(2))
+  tmp$epsilon[tmp$SITEID == 3] <- rt(nn, df = 2)
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  generate_simulated_pvalues(tmp)
+})
+
+design8_pvalues <- t(design8_pvalues)
+colnames(design8_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+design8_power <- apply(design8_pvalues, 2, compute_power)
+```
+
+```{r design8_plot, fig.height = 4, fig.align="center"}
+plot_power_curves(design8_power, "Constant Additive Effect of 1, Heteroskedastic Errors")
+
+plot_pvalue_hist(design8_pvalues, "Constant Additive Effect of 1, Heteroskedastic Errors")
+```
+
+## Design 9: heteroskedastic errors, single site effect
+
+```{r design9, fig.height = 4}
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+nn <- 16
+
+tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal", n = rep(nn, 3))
+design9_pvalues <- replicate(100, {
+  tmp$epsilon[tmp$SITEID == 1] <- rnorm(nn)
+  tmp$epsilon[tmp$SITEID == 2] <- rnorm(nn, sd = sqrt(2))
+  tmp$epsilon[tmp$SITEID == 3] <- rt(nn, df = 2)
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  generate_simulated_pvalues(tmp)
+})
+design9_pvalues <- t(design9_pvalues)
+colnames(design9_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+design9_power <- apply(design9_pvalues, 2, compute_power)
+```
+
+```{r design9_plot, fig.height = 4, fig.align="center"}
+plot_power_curves(design9_power, "Effect of 1 at a Single Site, Heteroskedastic Errors")
+
+plot_pvalue_hist(design9_pvalues, "Effect of 1, Heteroskedastic Errors")
+```

### 2016-11-29 ms/ms.tex
Start rough manuscript outline, add additional file extensions to gitignore

\documentclass[12pt]{article}
\usepackage[breaklinks=true]{hyperref}
\usepackage{color}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{natbib}
\usepackage{array}
\usepackage{booktabs, multicol, multirow}
\usepackage[nohead]{geometry}
\usepackage[singlespacing]{setspace}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{assumption}{Assumption}

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\bit}{\begin{itemize}}
\newcommand{\eit}{\end{itemize}}

\newcommand{\todo}[1]{{\color{red}{TO DO: \sc #1}}}

\newcommand{\reals}{\mathbb{R}}
\newcommand{\integers}{\mathbb{Z}}
\newcommand{\naturals}{\mathbb{N}}
\newcommand{\rationals}{\mathbb{Q}}

\newcommand{\ind}{\mathbb{I}} % Indicator function
\newcommand{\pr}{\mathbb{P}} % Generic probability
\newcommand{\ex}{\mathbb{E}} % Generic expectation
\newcommand{\var}{\textrm{Var}}
\newcommand{\cov}{\textrm{Cov}}

\newcommand{\normal}{N} % for normal distribution (can probably skip this)
\newcommand{\eps}{\varepsilon}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newcommand{\argmax}{\textrm{argmax}}
\newcommand{\argmin}{\textrm{argmin}}
\renewcommand{\baselinestretch}{1.5}

\title{Title}
\author{Kellie Ottoboni \\
Department of Statistics\\
University of California, Berkeley\\ [.2in]
Luigi Salmaso\\
Department of Management Engineering \\
University degli Studi di Padova \\ [.2in]
Fraser Lewis \\
Reckitt Benckiser Group
}\date{Draft \today}
\begin{document}
\maketitle

\newpage

\begin{abstract}
In clinical trials, the standard analysis of the effectiveness of one treatment over another is a parametric analysis of variance (ANOVA), which relies on uncheckable and ostensibly implausible assumptions such as x y and z.
The decision of the effectiveness of the treatment is based upon a single p-value from this hypothesis test, which is essentially meaningless if the assumptions are not met.
One would like to run some sort of robustness check, an additional test or something, whose results can call into question or give evidence in favor of the results from the ANOVA.
Permutation tests present such an opportunity: they require only minimal assumptions which are often guaranteed by the randomization that was conducted.
In contrast to parametric tests, which rely up on stringent assumptions to obtain null distributions with a simple closed form, permutation tests use the assumption of exchangeability or randomization to construct approximate null distributions by simulation, yielding (approximately) exact tests.
In order to mimic the ANOVA that is typically done, we compare several permutation tests based on linear models, which enable one to control for pretreatment covariates and nuisance variables.
We present simulation results showing that the permutation methods maintain power comparable to the ANOVA under a balanced design, but the ANOVA loses power when x y and z.
We illustrate the use of permutation tests alongside the parametric ANOVA using data from a clinical trial across seven (?) sites with different numbers of patients.
These results provide researchers an alternative way to assess the reliability of their statistical analyses and strengthen their confidence in the conclusions.

\end{abstract}

\newpage

\section{Introduction}
\bit
\item what's the problem with doing ANOVA? what are the assumptions and give examples of when they are commonly or clearly violated
\item lit review: cite the vickers paper, what else have people done comparing linear models to permutation/nonparametric tests?
\item why write this paper. what is the goal, where is the opportunity to make an advancement?
\item overview of the paper results
\eit

\section{Methods}

\subsection{Parametric ANOVA}

\subsection{Stratified permutation test}

\subsection{Permutation tests with the linear model}

\section{Simulations}

\bit
\item repeated measures: don't naively control for baseline by taking diffs (cite Frison and Pocock)
\item report relative power or relative efficiency of ANOVA against diffed perm test, LM test, and the Freedman-Lane test
\eit

\section{Clinical data results}

\bit
\item mention that we thought about whether to use the raw outcomes and adjust for baseline, or use outcome minus baseline. The outcome and baseline had a low correlation (include here), so we did not use their difference. As the simulations show, doing this actually reduces power
\eit

\section{Discussion}
\bit
\item This paper adds to the literature comparing parametric and nonparametric tests
\item In randomized control trials, there is no substantial difference between parametric and nonparametric approaches.  They are asymptotically equivalent to begin with (find a citation). But our simulations, using small sample sizes, show that they give comparable results anyway.
\item Results about differenced vs raw outcomes. This isn't new but our simulations and empirical dataset confirm it in the context of permutation tests
\item 
\eit

\end{document}

### 2016-12-07 ms/ms.tex
Progress on manuscript

@@
-\documentclass[12pt]{article}
+\documentclass[11pt]{article}
 \usepackage[breaklinks=true]{hyperref}
@@
 \usepackage{booktabs, multicol, multirow}
-\usepackage[nohead]{geometry}
+\usepackage[nohead, margin=1in]{geometry}
 \usepackage[singlespacing]{setspace}
@@
 
-\title{Title}
+\title{A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Control Trials}
 \author{Kellie Ottoboni \\
@@
 \begin{abstract}
+\noindent\textbf{Background:}
 In clinical trials, the standard analysis of the effectiveness of one treatment over another is a parametric analysis of variance (ANOVA), which relies on uncheckable and ostensibly implausible assumptions such as x y and z.
@@
 One would like to run some sort of robustness check, an additional test or something, whose results can call into question or give evidence in favor of the results from the ANOVA.
+
+\noindent\textbf{Methods:}
 Permutation tests present such an opportunity: they require only minimal assumptions which are often guaranteed by the randomization that was conducted.
@@
 In order to mimic the ANOVA that is typically done, we compare several permutation tests based on linear models, which enable one to control for pretreatment covariates and nuisance variables.
+
+\noindent\textbf{Results:}
 We present simulation results showing that the permutation methods maintain power comparable to the ANOVA under a balanced design, but the ANOVA loses power when x y and z.
 We illustrate the use of permutation tests alongside the parametric ANOVA using data from a clinical trial across seven (?) sites with different numbers of patients.
+
+\noindent\textbf{Conclusions:}
 These results provide researchers an alternative way to assess the reliability of their statistical analyses and strengthen their confidence in the conclusions.
@@
 
+% keywords: permutation test, linear model, ANOVA
 \newpage
 
-\section{Introduction}
+\section{Background}
+% The Background section should explain the background to the study, its aims, a summary of the existing literature and why this study was necessary or its contribution to the field.
+
 \bit
@@
 \section{Methods}
+% The methods section should include:
+
+%the aim, design and setting of the study
+%the characteristics of participants or description of materials
+%a clear description of all processes, interventions and comparisons. Generic drug names should generally be used. When proprietary brands are used in research, include the brand names in parentheses
+%the type of statistical analysis used, including a power calculation if appropriate
+
+\subsection{Problem and notation}
+Suppose we randomly assign two treatments to a group of individuals
+Let $Z$ be treatment assignment, 0 or 1, and we are interested in comparing the relative effectiveness of treatment 1 to treatment 0.
+$Y$ is the outcome of interest observed for all individuals.
+$X$ is a pretreatment covariate that we observe and is related to the outcome (may or may not be causal).  
+Without loss of generality, we suppose that $X$ is univariate, but we could easily extend all of these results to multivariate $X$.
+
+Suppose there is another nuisance variable $\mathcal{J}$. 
+$\mathcal{J}$ is a pretreatment variable used to stratify individuals. 
+There may be some unobservable heterogeneity in outcomes across levels of $\mathcal{J}$.
+
+We observe $\{Y_{ij}(Z_{ij}), Z_{ij}, X_{ij}\}$ for units $i = 1, \dots, n_j$, in strata $j = 1, \dots, J$.
+All units have two potential outcomes, $Y_{ij}(1)$ and $Y_{ij}(0)$, representing their responses to treatments $1$ and $0$, respectively.
+We can never observe both; random assignment of treatment reveals one.
+
+We are interested in the differences in potential outcomes $Y_{ij}(1) - Y_{ij}(0)$.
+This difference represents the effect of treatment for individual $i$ in group $j$: it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.
+Depending on the method of analysis one chooses, different functions of these differences will be considered.
+Any of these quantities may be of clinical interest; it is a matter of philosophy which one prefers.
+A typical problem is to estimate one of these averages.
+Instead, we will focus on hypothesis testing.
+Below, we outline a few approaches to studying the average treatment effect, explain the assumptions that are necessary, and
 
@@
 
+The model is
+
+$$Y_{ij} = \alpha_j + \beta X_{ij} + \gamma Z_{ij} + \eps_{ij}$$
+
+where $\alpha_j$ is a fixed effect for stratum $j$, $\beta$ is the coefficient for the pretreatment covariate,
+$\gamma$ is the coefficient for treatment,
+and $\eps_{ij}$ is an error term.
+The parameter of interest is $\gamma$, which represents the average treatment effect in the sample. 
+We would like to test the null hypothesis $H_0: \gamma = 0$ against
+the two-sided alternative hypothesis $H_1: \gamma \neq 0$.
+
+In the normal theory ANOVA, we make the following assumptions:
+\begin{enumerate}
+\item the $\eps$ are normally distributed with mean $0$ and common variance $\sigma^2$
+\item $\eps \independent X$
+\item \textcolor{red}{ what else? }
+\end{enumerate}
+
+The model is fit using least squares.
+A significance test is done for the estimated coefficient $\hat{\gamma}$, using the $t$-distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
+The test statistic is
+$$ t = \frac{\hat{\gamma}}{\sqrt{ \hat{\sigma}_{\hat{\gamma}}^2}},$$
+where $\hat{\sigma}_{\hat{\gamma}}^2$ is the estimated standard error of $\hat{\gamma}$, obtained from the diagonal element of the inverse covariance matrix of the observed data which corresponds to the treatment.
+\todo{cite freedman linear models textbook}
+
+\todo{Add a paragraph about why this might go wrong, perhaps some citations. All depends on estimating the error variance correctly, because $t$ distribution is the ratio of a normal and a chi-square}
+\bit
+\item This ANOVA model allows for fixed stratum effects, but they are only additive in the outcome.
+The model does not account for differences in the treatment effect across strata.
+In other words, the estimated linear model is not a good estimate of the true data-generating process.
+If effects are not constant across strata, then the coefficient $\gamma$ may be attenuated towards 0.
+A test of the null hypothesis that $\gamma = 0$ will be valid, but will not reflect the true magnitude of treatment effects among individuals.
+\item \textcolor{red}{ what else? }
+\eit
+
 \subsection{Stratified permutation test}
+Suppose we wish to test the null hypothesis that individual by individual, treatment has no effect.
+This is referred to as the ``sharp null'' hypothesis:
+
+$$H_0: Y_{ij}(1) = Y_{ij}(0), i = 1, \dots, n_j, j = 1,\dots, J.$$
+
+Then, which treatment that individual $ij$ received amounts to an arbitrary label.
+Once we observe their response under treatment $Z_{ij}$, we know what it would have been under treatment $1-Z_{ij}$; namely, it would have been the same.
+
+If treatment was completely randomized across strata, independently across strata,
+then we may condition on the number of individuals who received each treatment within each stratum.
+Any assignment of treatments which preserves the number of treated units within each stratum is valid and was just as likely to occur as the randomization that was actually observed.
+By using this principle of equal probabilities and by imputing the unobserved potential outcomes under the sharp null hypothesis, we can construct the permutation distribution of any statistic under the null hypothesis.
+
+The most commonly used statistic is the difference in mean outcomes of people who received drug 1 and the mean outcomes of people who received drug 0.
+This statistic is readily interpretable (``on average, taking the drug changes the outcome by x amount'') and has nice theoretical properties owing to it being the difference of two sums.
+When we want to be sensitive to differences in the magnitude of the effect, the difference in means may not be optimal.
+For an extreme example, imagine that there are $n/2$ males and $n/2$ females in the sample, $n/4$ males and females receive each treatment.
+Those who receive treatment 0 have an outcome of 0, the males who receive treatment 1 have an outcome of 1, and the females who receive treatment 1 have an outcome of -1.
+Then the difference in means between the treatment groups will be $0$, even though the treatment had a positive effect on males and negative effect on females.
+This differential effect will be averaged out.
+Thus, one may want to stratify the sample according to important confounding variables, then take the difference in means within each stratum separately.
+
+There is a great degree of freedom in choosing how exactly to construct such a statistic.
+Two factors must be decided: how to stratify and how to combine the individual means.
+There is a tradeoff in how finely we choose to stratify:
+on the one hand, we need all of the strata to be sufficiently large, otherwise they don't contribute any information to the test statistic,
+ but on the other hand, we want the strata to capture variation in treatment effects.
+\todo{discussion of NPC and cite papers}
 
 \subsection{Permutation tests with the linear model}
+We would like to test for a difference in outcomes between two treatments, but control for other covariates that may be predictive of the outcome.
+For example, if the outcome is the outcome after receiving treatment, we would like to control for the same variable measured before treatment was administered.
+When the outcome is the difference in responses between the second and first weeks, we would just like to control for location.
+We will use the approximate permutation test derived by Freedman and Lane (1983) to do so.
 
-\section{Simulations}
+Under the sharp null $H_0$, $Z$ has no effect on $Y$.
+In other words, the null hypothesis is that $\beta_2 = 0$.
+In a randomized experiment such as this, there are two ways we may test this null hypothesis.
 
+\subsubsection{Standard linear model}
+We may do a variation on the stratified two-sample test described above.
+Instead of using the difference in means as the statistic, we can use the t-statistic for the coefficient on $Z$ in a linear regression.
+If treatment was assigned at random within each stratum, then it is guaranteed that $Z$ and $\varepsilon$ are statistically independent, conditional on stratum.
+Therefore, we may conduct a test by permuting treatment assignments $Z$ within site, independently across sites, and calculating a test statistic for each such permutation.
+
+\subsubsection{Linear model residuals}
+
+Let's take an alternative view of the problem.
+We still write $Y = \beta_0 + \beta_1 X + \beta_2 Z + \varepsilon$.
+However, now, we do not treat the $\varepsilon$ as random.
+They are simply defined to be the diffence between $Y$ and the data's linear projection onto the plane $\beta_0 + \beta_1 X + \beta_2 Z$.
+
+If the null hypothesis is true, then $\varepsilon = Y - \beta_0 - \beta_1X$.
+We may regress $Y$ on $X$ to obtain coefficient estimates $\hat{\beta}_0$ and $\hat{\beta}_1$.
+Then we would predict the outcome to be $\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1X$
+and we may estimate the errors $\hat{\varepsilon}$ by $Y - \hat{Y}$.
+The $\hat{\varepsilon}$ approximate the true errors $\varepsilon$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit.
+Furthermore, under the null hypothesis, the $\varepsilon$ are independent of $X$ and $Z$. \todo{ this is false. fix}
+Therefore, within sites, these estimated $\hat{\varepsilon}$ are approximately exchangeable.
+Note that the exchangeability is only approximate, because $\hat{\varepsilon}$ will have some correlation with $X$ due to the finite sample size. \todo{ fix wording}
+
+We construct a permutation distribution using several steps:
+
+\begin{enumerate}
+\item Estimate $\hat{\varepsilon}$ by $Y - \hat{\beta}_0 - \hat{\beta}_1 X$, where $\hat{\beta}_0$ and $\hat{\beta}_1$ are obtained by regressing $Y$ on $X$.
+\item Construct permuted errors $\hat{\varepsilon}^\pi$ by permuting the $\hat{\varepsilon}$ within sites.
+\item Construct permuted responses $Y^\pi = \hat{\beta}_0 + \hat{\beta}_1 X + \hat{\varepsilon}$.
+\item Regress $Y^\pi$ on $X$ and $Z$. The test statistic is the $t$-statistic for the coefficient of $Z$.
+\end{enumerate}
+
+\subsubsection{Other linear model tests}
+It is important to note that the permutation tests described here are only exact in the context of randomized experiments.
+In observational studies, they are only approximately exact under the best circumstances.
+What's the difference between these two situations?
+In randomized experiments, treatment is assigned at random, and is therefore statistically independent of the covariates $X$ and the errors $\varepsilon$.
+Conversely, in observational studies, treatment may be associated with $X$ and/or $\varepsilon$, often in a way that is difficult or impossible to model.
+Independence, or the weaker condition of exchangeability, is what enables us to construct permutation distributions while holding $X$ fixed.
+When exchangeability doesn't hold, it does not make sense to permute $Z$ while holding $X$ fixed, as we cannot disentangle the effect of one over the other.
+In order to justify the use of one of these linear model tests for a treatment effect in observational data, one must argue that the object being permuted is uncorrelated with the other variables being held fixed.
+This can be done visually using scatterplots and residual plots.
+\todo{citations!}
+
+Others have developed approximate permutation tests which use linear regression to adjust for covariates. 
 \bit
-\item repeated measures: don't naively control for baseline by taking diffs (cite Frison and Pocock)
+\item Cite a few here, 
+\item explain why Freedman-Lane is best. 
+\item Cite some people who have shown by simulation that it works better.
+\eit
+
+\section{Results}
+\subsection{Simulations}
+
+To compare the performance of these tests, we simulated data coming from a randomized experiment, applied the tests to the data, and compared their empirical power over repeated random treatment assignments and random errors.
+We compared the following five tests:
+the $t$-statistic from the parametric ANOVA,
+a stratified permutation test using the difference in mean outcomes\footnote{ We considered using the difference in means within each stratum, aggregated by taking the sum of their absolute values over stratum.
+However this would not be ideal for imbalanced designs, which are common in real-world applications, so we did not consider it in the simulations.
+See the data analysis for details.}
+ (called ``stratified permutation'' in what follows),
+a stratified permutation test difference mean differences, taken by subtracting the baseline measure from the outcome (called ``differenced permutation'' in what follows),
+a stratified permutation test based on the $t$-statistic from the linear regression of outcome on control variables (called ``LM permutation'' in what follows),
+and the Freedman-Lane permutation test.
+
+We assume the following linear data-generating process:
+
+$$Y_{ij1} =\beta_0Y_{ij0} + \beta_{j} + \gamma_j Z_{ij} + \varepsilon_{ij}$$
+
+for individuals $i = 1, \dots, n_j$, $j = 1, 2, 3$.
+$\beta_0$ is the coefficient for the baseline measurement $Y_{ij0}$, 
+$\beta_j$ is the mean effect of being in stratum $j$, 
+$Z_{ij}$ is the treatment level, 
+$\gamma_j$ is the effect of treatment in stratum $j$, 
+and $\varepsilon_{ij}$ is an error term.
+Suppose there are three strata with $\beta_1 = 1, \beta_2 = 1.5,$ and $\beta_3 = 2$.
+Assume that there are $n_j = 16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment within each stratum.
+
+We use two designs:
+\begin{itemize}
+\item Constant additive treatment effect: $\gamma_1 = \gamma_2 = \gamma_3 = \gamma$. This is the implicit assumption when doing inference on estimated treatment effects.
+\item Treatment effect in a single stratum: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$.
+This is a constant, additive treatment effect in stratum 1, but no treatment effect in strata 2 and 3.
+This is a simplistic case of heterogeneous treatment effects.
+\end{itemize}
+
+We draw the baseline measurements once and condition on observing $\{ Y_{ij0}\}_{i = 1,\dots,16, j = 1,\dots, 3}$.
+Then, for a large number of trials, we randomly assign treatment within strata, generate errors, and construct new outcomes $Y_{ij1}$.
+We conduct all five tests on this new data, obtaining a two-sided p-value for each.
+The power curves are estimated by computing the proportion of observed p-values less than or equal to $\alpha$, for $\alpha = 0.01, 0.02, \dots, 1$.
+
+In our first set of simulations, we assume that the baseline measurements $Y_{ij0}$ are standard normally distributed and the pairs $(Y_{ij0}, \varepsilon_{ij})$ are independent across $i$ and $j$.
+We will assume that $\beta_0 = 1$.
+For each of the two simulation designs that vary the treatment effect, we use two distributions of $\varepsilon$.
+In the first case, we use $\varepsilon \sim N(0, 1)$ to fit with the parametric ANOVA assumption of normal errors.
+In the second case, $\varepsilon$ comes from a $t$ distribution with 2 degrees of freedom so the errors are heavy-tailed.
+Thus, there are four total simulation designs.
+
+\todo{Summary of results for normal sims}
+all 3 methods with linear model seem to be equally good. when everything is satisfied, all methods do well;
+
+In our second set of simulations, we assume that the baseline measurements $Y_{ij0}$ are discrete and skewed.
+This is representative of some survey data, where individuals are asked to rate their experience on a scale from 1 to 10, for instance.
+Here, we generate the baseline measures using independent draws from a Poisson distribution with mean 4, and censoring values larger than 10, instead setting them to 10.
+We condition on these observed baseline measures once we have generated them once.
+Then, for each of the large number of trials, we randomly assign treatment within strata and generate errors taking on values $0.5$ and $-0.5$ with equal probability.
+We construct $Y_{ij1}$ using the linear equation above.
+However we don't observe $Y_{ij1}$ but instead $\tilde{Y}_{ij1}$, which is defined as
+
+\begin{displaymath}
+   \tilde{Y}_{ij1} = \left\{
+     \begin{array}{ll}
+       1 & \text{if } \tilde{Y}_{ij1} < 1\\
+       10 & \text{if } \tilde{Y}_{ij1} > 10 \\
+       \lfloor \tilde{Y}_{ij1} \rfloor & \text{otherwise}
+     \end{array}
+   \right.
+\end{displaymath}
+
+The observed outcomes $\tilde{Y}_{ij1}$ that we test are also discrete.
+The assumption of normality is clearly violated here.
+
+\todo{Summary of results for skewed sims}
+
+\textbf{discussion of results}
+ \bit
+\item repeated measures: don't naively control for baseline by taking diffs (\cite{frison_repeated_1992})
 \item report relative power or relative efficiency of ANOVA against diffed perm test, LM test, and the Freedman-Lane test
@@
 
-\section{Clinical data results}
+\subsection{Clinical data results}
 
@@
 As the simulations show, doing this actually reduces power
+\item Freedman-Lane stuff: for the appro
+For this dataset, we are guaranteed that treatment $Z$ is independent of $\varepsilon$, as it is randomized within site.
+Indeed, the distribution of residuals looks nearly equal between treatment groups in both models.
+The distribution of residuals varies a bit across sites, but they all appear roughly centered around 0.
+Sites 4 and 5 have larger variance in the differenced model.
+As mentioned, this should not be an issue since we permute treatments within sites, independently across sites.
+(In general, when treatment is not randomly assigned, this condition is necessary for the permutation test to be valid.)
 \eit
@@
 \section{Discussion}
+% This section should discuss the implications of the findings in context of existing research and highlight limitations of the study.
+
 \bit
@@
 
+% list of abbreviations
+
+\section{Declarations}
+% see http://bmcmedresmethodol.biomedcentral.com/submission-guidelines/preparing-your-manuscript/technical-advance-article
+\textbf{Ethics approval and consent to participate}
+Not applicable
+\textbf{Consent for publication}
+\textbf{Availability of data and material}
+\textbf{Competing interests}
+\textbf{Funding}
+\textbf{Authors' contributions}
+\textbf{Acknowledgements}
+\textbf{Authors' information}
+
 \end{document}

### 2016-12-07 simulations/simulation.Rmd
Changes to text around errors in normal simulations

@@
 ---
-title: "ANOVA Comparison Simulations"
+title: "ANOVA Comparison Simulations: Gaussian Data"
 author: "Kellie Ottoboni"
@@
 
-# The model
+# Normal data
 
@@
 for individuals $i = 1, \dots, n_j$, $j = 1, \dots, J$.
-$\beta_0$ is the coefficient for the baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being at site $j$, $Z_{ij}$ is the treatment level, $\gamma_j$ is the effect of treatment at site $j$, and $\varepsilon_{ij}$ is an error term.
+$\beta_0$ is the coefficient for the standard normally-distributed baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being at site $j$, $Z_{ij}$ is the treatment level, $\gamma_j$ is the effect of treatment at site $j$, and $\varepsilon_{ij}$ is an error term.
 We will assume that $\beta_0 = 1$.
+The observed $(Y_{ij0}, \varepsilon_{ij})$ are independent across $i$ and $j$.
 
@@
   # errors = "normal" or "heavy"
-  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), and SITEID (stratum)
+  # n = number of individuals at each site
+  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), SITEID (stratum), site_effect (beta coefficient per individual), and epsilon (errors)
 

### 2016-12-07 simulations/simulation_skewed.Rmd
Add simulations for skewed discrete measures

--- 2016-11-29 simulations/simulation.Rmd
@@
 ---
-title: "ANOVA Comparison Simulations"
+title: "ANOVA Comparison Simulations: Skewed, Poisson Data"
 author: "Kellie Ottoboni"
@@
 
-# The model
+# Poisson data
 
-We assume the following linear data-generating process:
+As in the other simulations that use Gaussian data, we assume the following linear data-generating process:
 
-$$Y_{ij1} =\beta_0Y_{ij0} + \beta_{j} + \gamma_j Z_{ij} + \varepsilon_{ij}$$
+$$\tilde{Y}_{ij1} =\beta_0Y_{ij0} + \beta_{j} + \gamma_j Z_{ij} + \varepsilon_{ij}$$
 
 for individuals $i = 1, \dots, n_j$, $j = 1, \dots, J$.
-$\beta_0$ is the coefficient for the baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being at site $j$, $Z_{ij}$ is the treatment level, $\gamma_j$ is the effect of treatment at site $j$, and $\varepsilon_{ij}$ is an error term.
+$\beta_0$ is the coefficient for the baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being at site $j$, $Z_{ij}$ is the treatment level indicator,, $\gamma_j$ is the effect of treatment at site $j$, and $\varepsilon_{ij}$ takes the values $0$ or $0.5$ with equal probability.
 We will assume that $\beta_0 = 1$.
+
+Now, instead of $Y_{ij0}$ being normally distributed, we let these baseline measurements come from a truncated Poisson distribution with mean 4, with values greater than 10 censored and set to 10.
+This is reflective of some survey data where individuals are asked to rate their experience on a scale from 1 to 10.
+The data are not only discrete but also skewed:
+more values are in the lower half of the range.
+
+We don't observe $\tilde{Y}_{ij1}$ but $Y_{ij1}$, which is defined as
+
+\begin{displaymath}
+   Y_{ij1} = \left\{
+     \begin{array}{ll}
+       1 & \text{if } \tilde{Y}_{ij1} < 1\\
+       10 & \text{if } \tilde{Y}_{ij1} > 10 \\
+       \lfloor \tilde{Y}_{ij1} \rfloor & \text{otherwise}
+     \end{array}
+   \right.
+\end{displaymath}
 
@@
 ```{r generate_sample}
-generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
+generate_outcome <- function(baseline, tr_effect, site_effect, error){
+  Y1 <- baseline + tr_effect + site_effect + error
+  Y1 <- floor(Y1)
+  Y1 <- pmax(pmin(Y1, 10), 1) 
+  return(Y1)
+}
+
+generate_simulated_data <- function(gamma, effect, n = c(16, 16, 16)){
   # Input:
@@
   # effect = "same effect" or "single site effect" - which sites have a tr effect > 0?
-  # errors = "normal" or "heavy"
-  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), and SITEID (stratum)
+  # n = number of individuals at each site
+  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), SITEID (stratum), site_effect (beta coefficient per individual), and epsilon (errors)
 
@@
 
-  # Generate errors
-  if(errors == "normal"){
-    epsilon <- rnorm(N)
-  } else {
-    epsilon <- rt(N, df = 2)
-  }
-
   # Generate covariates
-  Y0 <- rnorm(N)
+  epsilon <- rbinom(N, 1, 0.5) - 0.5
+  Y0 <- rpois(N, lambda = 4)
+  Y0[Y0 > 10] <- 10
   Z <- rep(0:1, length.out = N)
   site_effect <- rep(beta, times = n)
-  Y1 <- Y0 + gamma_vec*Z + site_effect + epsilon
+  Y1 <- generate_outcome(Y0, gamma_vec*Z, site_effect, epsilon)
   return(data.frame(Y1, Y0, Z, gamma_vec, SITEID, site_effect, epsilon))
@@
 
-tmp <- generate_simulated_data(gamma = 0, effect = "same effect", errors = "normal")
+tmp <- generate_simulated_data(gamma = 0, effect = "same effect")
 design0_pvalues <- replicate(100, {
@@
   tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$epsilon
+  tmp$Y1 <- generate_outcome(tmp$Y0, rep(0, nrow(tmp)), tmp$site_effect)
   generate_simulated_pvalues(tmp)
@@
 plot_pvalue_hist(design0_pvalues, "No Effect, Normal Errors")
-
-#plot_pvalue_scatter(design0_pvalues, "Simulation P-values for Data with \n No Effect, Normal Errors")
 ```
 
-# Design 1: Constant additive effect, normal errors
+## Design 1: Constant additive effect
 
@@
 
-tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
+tmp <- generate_simulated_data(gamma = 0.5, effect = "same effect")
 design1_pvalues <- replicate(100, {
-  tmp$epsilon <- rnorm(nrow(tmp))
   tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
+  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
   generate_simulated_pvalues(tmp)
@@
 ```{r design1_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design1_power, "Constant Additive Effect of 1, Normal Errors")
-
-plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 1, Normal Errors")
-
-#plot_pvalue_scatter(design1_pvalues, "Simulation P-values for Data with \n Constant Additive Effect of 1, Normal Errors")
+p1 <- plot_power_curves(design1_power, "Constant Additive Effect of 0.5")
+p1
+plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 0.5")
 ```
 
-# Design 2: Single site effect, normal errors
+## Design 2: Single site effect
 
@@
 
-tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal")
+tmp <- generate_simulated_data(gamma = 0.5, effect = "single site effect")
 design2_pvalues <- replicate(100, {
-  tmp$epsilon <- rnorm(nrow(tmp))
   tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
+  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
   generate_simulated_pvalues(tmp)
@@
 ```{r design2_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design2_power, "Effect of 1 at a Single Site, Normal Errors")
-
-plot_pvalue_hist(design2_pvalues, "Effect of 1, Normal Errors")
-
-#plot_pvalue_scatter(design2_pvalues, "Simulation P-values for Data with \n Effect of 1 at a Single Site, Normal Errors")
+p2 <- plot_power_curves(design2_power, "Effect of 0.5 in a Single Stratum")
+p2
+plot_pvalue_hist(design2_pvalues, "Effect of 0.5 in a Single Stratum")
 ```
 
-# Design 3: Constant additive effect, heavy-tailed errors
-
-Again, we find that the four power curves coincide while the curve for the stratified permutation test is below the others.
-Here, the difference between the curves is not large.
-Controlling for baseline $Y_0$ does not substantially help reduce variance.
-
-```{r design3, fig.height = 4}
-set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
-
-tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "heavy")
-design3_pvalues <- replicate(100, {
-  tmp$epsilon <- rt(nrow(tmp), df = 2)
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
-  generate_simulated_pvalues(tmp)
-})
-design3_pvalues <- t(design3_pvalues)
-colnames(design3_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
-design3_power <- apply(design3_pvalues, 2, compute_power)
-```
-
-```{r design3_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design3_power, "Constant Additive Effect of 1, Heavy-tailed Errors")
-
-plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 1, Heavy-tailed Errors")
-
-#plot_pvalue_scatter(design3_pvalues, "Simulation P-values for Data with \n Constant Additive Effect of 1, Heavy-tailed Errors")
-```
-
-# Design 4: single site effect, heavy-tailed errors
-
-The noise from the heavy-tailed errors masks the treatment effect so much that controlling for baseline measures makes no difference.
-There is almost no power to detect an effect using any of the five tests.
-
-```{r design4, fig.height = 4}
-set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
-
-tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "heavy")
-design4_pvalues <- replicate(100, {
-  tmp$epsilon <- rt(nrow(tmp), df = 2)
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
-  generate_simulated_pvalues(tmp)
-})
-design4_pvalues <- t(design4_pvalues)
-colnames(design4_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
-design4_power <- apply(design4_pvalues, 2, compute_power)
-```
-
-```{r design4_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design4_power, "Effect of 1 at a Single Site, Heavy-tailed Errors")
-
-plot_pvalue_hist(design4_pvalues, "Effect of 1, Heavy-tailed Errors")
-
-#plot_pvalue_scatter(design4_pvalues, "Simulation P-values for Data with \n Effect of 1 at a Single Site, Heavy-tailed Errors")
-```
-
-# Design 5: Unpredictive baseline
+## Design 3: Unpredictive baseline
 
@@
 
+```{r design3, fig.height = 4}
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+
+tmp <- generate_simulated_data(gamma = 0.5, effect = "same effect")
+design3_pvalues <- replicate(100, {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
+  tmp$Y1 <- generate_outcome(0.1*tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
+  generate_simulated_pvalues(tmp)
+})
+
+design3_pvalues <- t(design3_pvalues)
+colnames(design3_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+design3_power <- apply(design3_pvalues, 2, compute_power)
+```
+
+```{r design3_plot, fig.height = 4, fig.align="center"}
+p3 <- plot_power_curves(design3_power, "Constant Additive Effect of 0.5 \n Unpredictive baseline")
+p3
+plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 0.5 \n Unpredictive baseline")
+```
+
+# Design 4: Constant additive effect, imbalanced design
+
+Suppose that instead of having 16 individuals per site, we distribute them unequally across sites:
+site 1 has 8 patients, site 2 has 16 patients, and site 3 has 24 patients.
+This imbalance does not violate any assumptions of the ANOVA.
+However, it may reduce power if the effect is concentrated in sites with fewer patients.
+
+The power curves here are very similar to those in Design 1.
+This result is expected, since the effect is still present among all patients who received treatment.
+
+```{r design4, fig.height = 4}
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+
+tmp <- generate_simulated_data(gamma = 0.5, effect = "same effect", n = c(8, 16, 24))
+design4_pvalues <- replicate(100, {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
+  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
+  generate_simulated_pvalues(tmp)
+})
+
+design4_pvalues <- t(design4_pvalues)
+colnames(design4_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+design4_power <- apply(design4_pvalues, 2, compute_power)
+```
+
+```{r design4_plot, fig.height = 4, fig.align="center"}
+p4 <- plot_power_curves(design4_power, "Constant Additive Effect of 0.5 \n Imbalanced Design")
+p4
+plot_pvalue_hist(design4_pvalues, "Constant Additive Effect of 0.5 \n Imbalanced Design")
+```
+
+# Design 5: Single site effect, imbalanced design
+
+There is very low power to detect the effect here, since in the simulated data, the site with only 8 patients was the only site with a nonzero effect of treatment.
+The treatment effect is on the same order of magnitude as the error variance.
+
 ```{r design5, fig.height = 4}
@@
 
-tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
+tmp <- generate_simulated_data(gamma = 0.5, effect = "single site effect", n = c(8, 16, 24))
 design5_pvalues <- replicate(100, {
-  tmp$epsilon <- rnorm(nrow(tmp))
   tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- 0.25*tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
+  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
   generate_simulated_pvalues(tmp)
 })
-
 design5_pvalues <- t(design5_pvalues)
@@
 ```{r design5_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design5_power, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
+p5 <- plot_power_curves(design5_power, "Effect of 0.5 at a Single Site, Imbalanced Design")
+p5 
+plot_pvalue_hist(design5_pvalues, "Effect of 0.5 at a Single Site, Imbalanced Design")
+```
 
-plot_pvalue_hist(design5_pvalues, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
-
-#plot_pvalue_scatter(design5_pvalues, "Simulation P-values for Data with \n Constant Additive Effect of 1, Normal Errors")
+```{r final_summary, fig.align = "center"}
+grid.arrange(p1, p2, p4, p5)
 ```

### 2016-12-07 simulations/simulation_skewed.Rmd
Create 2 by 2 plot of power curves for different simulations

@@
 require(gridExtra)
+require(gtable)
 ```
@@
 ```{r final_summary, fig.align = "center"}
-grid.arrange(p1, p2, p4, p5)
+#plots4 <- lapply(list(p1, p2, p4, p5), function(pl) pl + theme(legend.position = "none") + ggtitle(""))
+#do.call(grid.arrange, plots4)
+
+powers <- list(design1_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design1_power)), "Design" = rep("Balanced", nrow(design1_power))), 
+               design2_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Site Effect", nrow(design2_power)), "Design" = rep("Balanced", nrow(design2_power))), 
+               design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design4_power)), "Design" = rep("Imbalanced", nrow(design4_power))), 
+               design5_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Site Effect", nrow(design5_power)), "Design" = rep("Imbalanced", nrow(design5_power))))
+
+all_power_curves <- do.call(rbind, powers)
+twobytwo <- melt(all_power_curves, id.vars = c("Treatment", "Design")) %>% 
+  mutate("pvalue" = rep((1: 100)/100, 5*4)) %>%
+  mutate("Method" = variable) %>%
+  ggplot(aes_string(x = "pvalue", y = "value", color = "Method")) +
+  geom_line() +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
+  xlab("P-value") +
+  ylab("Power") +
+  facet_wrap(Treatment~Design) +
+  ggtitle("Discrete, Skewed Data")
+
+twobytwo
 ```

### 2016-12-07 simulations/simulation.Rmd
Add 2 by 2 power plot to normal simulations

@@
 ```
+
+```{r final_summary, fig.align = "center"}
+#plots4 <- lapply(list(p1, p2, p4, p5), function(pl) pl + theme(legend.position = "none") + ggtitle(""))
+#do.call(grid.arrange, plots4)
+
+powers <- list(design1_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design1_power)), "Errors" = rep("Gaussian", nrow(design1_power))), 
+               design2_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Site Effect", nrow(design2_power)), "Errors" = rep("Gaussian", nrow(design2_power))), 
+               design3_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design3_power)), "Errors" = rep("Heavy-tailed", nrow(design3_power))), 
+               design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Site Effect", nrow(design4_power)), "Errors" = rep("Heavy-tailed", nrow(design4_power))))
+
+all_power_curves <- do.call(rbind, powers)
+twobytwo <- melt(all_power_curves, id.vars = c("Treatment", "Errors")) %>% 
+  mutate("pvalue" = rep((1: 100)/100, 5*4)) %>%
+  mutate("Method" = variable) %>%
+  ggplot(aes_string(x = "pvalue", y = "value", color = "Method")) +
+  geom_line() +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
+  xlab("P-value") +
+  ylab("Power") +
+  facet_wrap(Treatment~Errors) +
+  ggtitle("Gaussian Data")
+
+twobytwo
+```

### 2016-12-07 simulations/simulation.Rmd
Create fig directory and create power plots for manuscript

@@
 
-```{r final_summary, fig.align = "center"}
+```{r final_summary, fig.align = "center", fig.width = 12}
 #plots4 <- lapply(list(p1, p2, p4, p5), function(pl) pl + theme(legend.position = "none") + ggtitle(""))
@@
   facet_wrap(Treatment~Errors) +
-  ggtitle("Gaussian Data")
+  ggtitle("Gaussian Data") +
+  theme(
+    axis.text.x = element_text(size = 12),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 16),
+    title = element_text(size = 16),
+    legend.title = element_text(size = 12),
+    legend.text = element_text(size = 12),
+    strip.text.x = element_text(size = 12),
+    legend.position = "bottom"
+  ) + 
+  guides(color=guide_legend(nrow=2,byrow=TRUE))
 
 twobytwo
+
+pdf(file = "../ms/fig/normal_simulation_power.pdf", width = 8)
+twobytwo
+dev.off()
 ```

### 2016-12-07 simulations/simulation_skewed.Rmd
Create fig directory and create power plots for manuscript

@@
   facet_wrap(Treatment~Design) +
-  ggtitle("Discrete, Skewed Data")
+  ggtitle("Discrete, Skewed Data") +
+    theme(
+    axis.text.x = element_text(size = 12),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 16),
+    title = element_text(size = 16),
+    legend.title = element_text(size = 12),
+    legend.text = element_text(size = 12),
+    strip.text.x = element_text(size = 12),
+    legend.position = "bottom"
+  ) + 
+  guides(color=guide_legend(nrow=2,byrow=TRUE))
 
 twobytwo
+
+pdf(file = "../ms/fig/skewed_simulation_power.pdf", width = 8)
+twobytwo
+dev.off()
 ```

### 2016-12-07 ms/fig/normal_simulation_power_summary.tex
Create power summary tables

\begin{table}[ht]
\centering
\begin{tabular}{p{1.25in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}}
  \hline
Treatment & Errors & ANOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
  \hline
Constant Additive Effect & Gaussian & 0.88 & 0.55 & 0.85 & 0.87 & 0.90 \\ 
  Single Site Effect & Gaussian & 0.15 & 0.18 & 0.15 & 0.18 & 0.17 \\ 
  Constant Additive Effect & Heavy-tailed & 0.35 & 0.26 & 0.35 & 0.38 & 0.38 \\ 
  Single Site Effect & Heavy-tailed & 0.05 & 0.07 & 0.08 & 0.07 & 0.06 \\ 
   \hline
\end{tabular}
\caption{Empirical power at level $0.05$ for Gaussian simulated data} 
\end{table}

### 2016-12-07 ms/fig/skewed_simulation_power_summary.tex
Create power summary tables

\begin{table}[ht]
\centering
\begin{tabular}{p{1.25in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}}
  \hline
Treatment & Design & ANOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
  \hline
Constant Additive Effect & Balanced & 0.97 & 0.12 & 0.85 & 0.96 & 0.97 \\ 
  Single Site Effect & Balanced & 0.46 & 0.04 & 0.31 & 0.44 & 0.46 \\ 
  Constant Additive Effect & Imbalanced & 0.98 & 0.21 & 0.86 & 0.98 & 0.99 \\ 
  Single Site Effect & Imbalanced & 0.19 & 0.07 & 0.10 & 0.19 & 0.19 \\ 
   \hline
\end{tabular}
\caption{Empirical power at level $0.05$ for discrete, skewed simulated data} 
\end{table}

### 2016-12-07 simulations/simulation.Rmd
Create power summary tables

@@
 ```
+
+```{r power_summary_tab, results = "asis"}
+summary05 <- t(sapply(powers, function(x) x[5, c(6, 7, 1:5)]))
+summarytab <- xtable(summary05, align = "rp{1.25in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}",
+       caption = "Empirical power at level $0.05$ for Gaussian simulated data")
+print(summarytab, type="latex", file="../ms/fig/normal_simulation_power_summary.tex")
+summarytab
+```

### 2016-12-07 simulations/simulation_skewed.Rmd
Create power summary tables

@@
 ```
+
+```{r power_summary_tab, results = "asis"}
+summary05 <- t(sapply(powers, function(x) x[5, c(6, 7, 1:5)]))
+summarytab <- xtable(summary05, align = "rp{1.25in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}",
+       caption = "Empirical power at level $0.05$ for discrete, skewed simulated data")
+print(summarytab, type="latex", file="../ms/fig/skewed_simulation_power_summary.tex")
+summarytab
+```

### 2016-12-08 ms/fig/normal_simulation_power_summary.tex
Rerun sims with 1000 reps to get cleaner plots and power estimates

@@
 \centering
-\begin{tabular}{p{1.25in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}}
+\begin{tabular}{p{1.15in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}}
   \hline
@@
   \hline
-Constant Additive Effect & Gaussian & 0.88 & 0.55 & 0.85 & 0.87 & 0.90 \\ 
-  Single Site Effect & Gaussian & 0.15 & 0.18 & 0.15 & 0.18 & 0.17 \\ 
-  Constant Additive Effect & Heavy-tailed & 0.35 & 0.26 & 0.35 & 0.38 & 0.38 \\ 
-  Single Site Effect & Heavy-tailed & 0.05 & 0.07 & 0.08 & 0.07 & 0.06 \\ 
+Constant Additive Effect & Gaussian & 0.89 & 0.55 & 0.86 & 0.89 & 0.89 \\ 
+  Single Stratum Effect & Gaussian & 0.16 & 0.09 & 0.16 & 0.16 & 0.17 \\ 
+  Constant Additive Effect & Heavy-tailed & 0.35 & 0.24 & 0.38 & 0.40 & 0.39 \\ 
+  Single Stratum Effect & Heavy-tailed & 0.07 & 0.08 & 0.08 & 0.08 & 0.08 \\ 
    \hline
@@
 \caption{Empirical power at level $0.05$ for Gaussian simulated data} 
+\label{tab:normal_power}
 \end{table}

### 2016-12-08 ms/fig/skewed_simulation_power_summary.tex
Rerun sims with 1000 reps to get cleaner plots and power estimates

@@
   \hline
-Constant Additive Effect & Balanced & 0.97 & 0.12 & 0.85 & 0.96 & 0.97 \\ 
-  Single Site Effect & Balanced & 0.46 & 0.04 & 0.31 & 0.44 & 0.46 \\ 
-  Constant Additive Effect & Imbalanced & 0.98 & 0.21 & 0.86 & 0.98 & 0.99 \\ 
-  Single Site Effect & Imbalanced & 0.19 & 0.07 & 0.10 & 0.19 & 0.19 \\ 
+Constant Additive Effect & Balanced & 0.97 & 0.14 & 0.85 & 0.97 & 0.98 \\ 
+  Single Stratum Effect & Balanced & 0.42 & 0.05 & 0.28 & 0.43 & 0.42 \\ 
+  Constant Additive Effect & Imbalanced & 0.97 & 0.15 & 0.86 & 0.97 & 0.97 \\ 
+  Single Stratum Effect & Imbalanced & 0.19 & 0.07 & 0.10 & 0.19 & 0.19 \\ 
    \hline
@@
 \caption{Empirical power at level $0.05$ for discrete, skewed simulated data} 
+\label{tab:skewed_power}
 \end{table}

### 2016-12-08 simulations/simulation.Rmd
Rerun sims with 1000 reps to get cleaner plots and power estimates

--- 2016-12-07 simulations/simulation.Rmd
@@
 for individuals $i = 1, \dots, n_j$, $j = 1, \dots, J$.
-$\beta_0$ is the coefficient for the standard normally-distributed baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being at site $j$, $Z_{ij}$ is the treatment level, $\gamma_j$ is the effect of treatment at site $j$, and $\varepsilon_{ij}$ is an error term.
+$\beta_0$ is the coefficient for the standard normally-distributed baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being in stratum $j$, $Z_{ij}$ is the treatment level, $\gamma_j$ is the effect of treatment in stratum $j$, and $\varepsilon_{ij}$ is an error term.
 We will assume that $\beta_0 = 1$.
@@
 
-Suppose there are three sites with $\beta_1 = 1, \beta_2 = 1.5,$ and $\beta_3 = 2$.
-Assume that there are 16 individuals per site and treatment assignment is balanced, i.e. 8 people receive each treatment at each site.
+Suppose there are three strata with $\beta_1 = 1, \beta_2 = 1.5,$ and $\beta_3 = 2$.
+Assume that there are 16 individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
 
@@
 * Design 2: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$.
-This is a constant, additive treatment effect at site 1, but no treatment effect at sites 2 and 3.
+This is a constant, additive treatment effect in stratum 1, but no treatment effect in stratums 2 and 3.
 This is a simplistic case of heterogeneous treatment effects.
@@
 
-* ANOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for site.
-* Stratified permutation: we permute treatment assignment within site, then take the difference in means between treated and control outcomes $Y_1$
+* ANOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for stratum.
+* Stratified permutation: we permute treatment assignment within stratum, then take the difference in means between treated and control outcomes $Y_1$
 * Differenced permutation: we do the same permutation procedure as the stratified permutation test, except we use the difference between outcome and baseline, $Y_1 - Y_0$
-* Linear model (LM) permutation: we use the same stratified permutation procedure as above, except use the $t$-statistic for the coefficient on treatment in the linear regression of $Y_1$ on $Y_0$, $Z$, and site dummies
+* Linear model (LM) permutation: we use the same stratified permutation procedure as above, except use the $t$-statistic for the coefficient on treatment in the linear regression of $Y_1$ on $Y_0$, $Z$, and stratum dummies
 * Freedman-Lane test: see the other Rmd document for a full description of this procedure
 
-Throughout our simulations, we first fix $Y_0$ and site ID.
+Throughout our simulations, we first fix $Y_0$ and stratum ID.
 Treatment $Z$ and the errors $\varepsilon$ are randomly drawn according to their respective distributions.
@@
   # gamma = the magnitude of the treatment effect
-  # effect = "same effect" or "single site effect" - which sites have a tr effect > 0?
+  # effect = "same effect" or "single stratum effect" - which strata have a tr effect > 0?
   # errors = "normal" or "heavy"
-  # n = number of individuals at each site
-  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), SITEID (stratum), site_effect (beta coefficient per individual), and epsilon (errors)
+  # n = number of individuals at each stratum
+  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), stratumID (stratum), stratum_effect (beta coefficient per individual), and epsilon (errors)
 
-  SITEID <- rep(1:3, times = n)
+  stratumID <- rep(1:3, times = n)
   N <- sum(n)
@@
   Z <- rep(0:1, length.out = N)
-  site_effect <- rep(beta, times = n)
-  Y1 <- Y0 + gamma_vec*Z + site_effect + epsilon
-  return(data.frame(Y1, Y0, Z, gamma_vec, SITEID, site_effect, epsilon))
+  stratum_effect <- rep(beta, times = n)
+  Y1 <- Y0 + gamma_vec*Z + stratum_effect + epsilon
+  return(data.frame(Y1, Y0, Z, gamma_vec, stratumID, stratum_effect, epsilon))
 }
@@
   # Inputs:
-  # dataset = a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), and SITEID (stratum)
+  # dataset = a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), and stratumID (stratum)
   # Returns: a vector of p-values
@@
   # ANOVA
-  modelfit <- lm(Y1 ~ Y0 + Z + factor(SITEID), data = dataset)
+  modelfit <- lm(Y1 ~ Y0 + Z + factor(stratumID), data = dataset)
   resanova <- summary(aov(modelfit))
@@
   observed_diff_means <- mean(dataset$Y1[dataset$Z == 1]) - mean(dataset$Y1[dataset$Z == 0])
-  diff_means_distr <- stratified_two_sample(group = dataset$Z, response = dataset$Y1, stratum = dataset$SITEID, reps = reps)
+  diff_means_distr <- stratified_two_sample(group = dataset$Z, response = dataset$Y1, stratum = dataset$stratumID, reps = reps)
   perm_pvalue <- t2p(observed_diff_means, diff_means_distr, alternative = "two-sided")
@@
   observed_diff_means2 <- mean(dataset$diff[dataset$Z == 1]) - mean(dataset$diff[dataset$Z == 0])
-  diff_means_distr2 <- stratified_two_sample(group = dataset$Z, response = dataset$diff, stratum = dataset$SITEID, reps = reps)
+  diff_means_distr2 <- stratified_two_sample(group = dataset$Z, response = dataset$diff, stratum = dataset$stratumID, reps = reps)
   perm_pvalue2 <- t2p(observed_diff_means2, diff_means_distr2, alternative = "two-sided")
@@
   lm1_t_distr <-  replicate(reps, {
-    dataset$Z_perm <- permute_within_groups(dataset$Z, dataset$SITEID)
-    lm1_perm <- lm(Y1 ~ Y0 + Z_perm + factor(SITEID), data = dataset)
+    dataset$Z_perm <- permute_within_groups(dataset$Z, dataset$stratumID)
+    lm1_perm <- lm(Y1 ~ Y0 + Z_perm + factor(stratumID), data = dataset)
     summary(lm1_perm)[["coefficients"]]["Z_perm", "t value"]
@@
   # Freedman-Lane linear model residual permutation
-  lm2_no_tr <- lm(Y1~Y0 + factor(SITEID), data = dataset)
+  lm2_no_tr <- lm(Y1~Y0 + factor(stratumID), data = dataset)
   lm2_resid <- residuals(lm2_no_tr)
@@
   lm2_t_distr <-  replicate(reps, {
-    lm2_resid_perm <- permute_within_groups(lm2_resid, dataset$SITEID)
+    lm2_resid_perm <- permute_within_groups(lm2_resid, dataset$stratumID)
     dataset$response_fl <-  lm2_yhat + lm2_resid_perm
-    lm2_perm <- lm(response_fl ~ Y0 + Z + factor(SITEID), data = dataset)
+    lm2_perm <- lm(response_fl ~ Y0 + Z + factor(stratumID), data = dataset)
     summary(lm2_perm)[["coefficients"]]["Z", "t value"]
@@
 Before testing for different kinds of effects, we begin checking that the tests have the correct level.
-We follow the procedure described above, using an effect size of $\gamma = 0$ at all sites and using standard normal errors.
+We follow the procedure described above, using an effect size of $\gamma = 0$ at all strata and using standard normal errors.
 To have the correct level means that the test rate of rejection at level $\alpha$ is $\alpha 100\%$.
@@
   tmp$epsilon <- rnorm(nrow(tmp))
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$epsilon
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
-design1_pvalues <- replicate(100, {
+design1_pvalues <- replicate(1000, {
   tmp$epsilon <- rnorm(nrow(tmp))
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 
-## Design 2: Single site effect, normal errors
+## Design 2: Single stratum effect, normal errors
 
@@
 The Freedman-Lane test may have the highest power for small p-values, but this could also just be noise.
-All five power curves are closer to the line passing through the origin: since the effect is only present at one site, it is more difficult to detect.
+All five power curves are closer to the line passing through the origin: since the effect is only present at one stratum, it is more difficult to detect.
 
@@
 
-tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal")
-design2_pvalues <- replicate(100, {
+tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "normal")
+design2_pvalues <- replicate(1000, {
   tmp$epsilon <- rnorm(nrow(tmp))
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 ```{r design2_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design2_power, "Effect of 1 at a Single Site, Normal Errors")
+plot_power_curves(design2_power, "Effect of 1 at a Single stratum, Normal Errors")
 
@@
 tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "heavy")
-design3_pvalues <- replicate(100, {
+design3_pvalues <- replicate(1000, {
   tmp$epsilon <- rt(nrow(tmp), df = 2)
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 
-## Design 4: single site effect, heavy-tailed errors
+## Design 4: single stratum effect, heavy-tailed errors
 
@@
 
-tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "heavy")
-design4_pvalues <- replicate(100, {
+tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "heavy")
+design4_pvalues <- replicate(1000, {
   tmp$epsilon <- rt(nrow(tmp), df = 2)
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 ```{r design4_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design4_power, "Effect of 1 at a Single Site, Heavy-tailed Errors")
+plot_power_curves(design4_power, "Effect of 1 at a Single stratum, Heavy-tailed Errors")
 
@@
   tmp$epsilon <- rnorm(nrow(tmp))
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- 0.25*tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- 0.25*tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 
-Suppose that instead of having 16 individuals per site, we distribute them unequally across sites:
-site 1 has 8 patients, site 2 has 16 patients, and site 3 has 24 patients.
+Suppose that instead of having 16 individuals per stratum, we distribute them unequally across strata:
+stratum 1 has 8 patients, stratum 2 has 16 patients, and stratum 3 has 24 patients.
 This imbalance does not violate any assumptions of the ANOVA.
-However, it may reduce power if the effect is concentrated in sites with fewer patients.
+However, it may reduce power if the effect is concentrated in strata with fewer patients.
 
@@
   tmp$epsilon <- rnorm(nrow(tmp))
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 
-## Design 7: Single site effect, normal errors
+## Design 7: Single stratum effect, normal errors
 
-There is very low power to detect the effect here, since in the simulated data, the site with only 8 patients was the only site with a nonzero effect of treatment.
+There is very low power to detect the effect here, since in the simulated data, the stratum with only 8 patients was the only stratum with a nonzero effect of treatment.
 The treatment effect is on the same order of magnitude as the error variance.
@@
 
-tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal", n = c(8, 16, 24))
+tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "normal", n = c(8, 16, 24))
 design7_pvalues <- replicate(100, {
   tmp$epsilon <- rnorm(nrow(tmp))
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 ```{r design7_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design7_power, "Effect of 1 at a Single Site, Normal Errors")
+plot_power_curves(design7_power, "Effect of 1 at a Single stratum, Normal Errors")
 
@@
 Suppose that this does not hold in our data.
-Imagine we have three sites with 16 patients each, but they have different error distributions:
-site 1 has standard normal errors,
-site 2 has normally distributed errors with mean 0 and variance 2,
-and site 3 has errors with a $t$-distribution on 2 degrees of freedom.
+Imagine we have three strata with 16 patients each, but they have different error distributions:
+stratum 1 has standard normal errors,
+stratum 2 has normally distributed errors with mean 0 and variance 2,
+and stratum 3 has errors with a $t$-distribution on 2 degrees of freedom.
 
@@
 design8_pvalues <- replicate(100, {
-  tmp$epsilon[tmp$SITEID == 1] <- rnorm(nn)
-  tmp$epsilon[tmp$SITEID == 2] <- rnorm(nn, sd = sqrt(2))
-  tmp$epsilon[tmp$SITEID == 3] <- rt(nn, df = 2)
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$epsilon[tmp$stratumID == 1] <- rnorm(nn)
+  tmp$epsilon[tmp$stratumID == 2] <- rnorm(nn, sd = sqrt(2))
+  tmp$epsilon[tmp$stratumID == 3] <- rt(nn, df = 2)
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 
-## Design 9: heteroskedastic errors, single site effect
+## Design 9: heteroskedastic errors, single stratum effect
 
@@
 
-tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal", n = rep(nn, 3))
+tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "normal", n = rep(nn, 3))
 design9_pvalues <- replicate(100, {
-  tmp$epsilon[tmp$SITEID == 1] <- rnorm(nn)
-  tmp$epsilon[tmp$SITEID == 2] <- rnorm(nn, sd = sqrt(2))
-  tmp$epsilon[tmp$SITEID == 3] <- rt(nn, df = 2)
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
+  tmp$epsilon[tmp$stratumID == 1] <- rnorm(nn)
+  tmp$epsilon[tmp$stratumID == 2] <- rnorm(nn, sd = sqrt(2))
+  tmp$epsilon[tmp$stratumID == 3] <- rt(nn, df = 2)
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
   generate_simulated_pvalues(tmp)
@@
 ```{r design9_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design9_power, "Effect of 1 at a Single Site, Heteroskedastic Errors")
+plot_power_curves(design9_power, "Effect of 1 at a Single stratum, Heteroskedastic Errors")
 
@@
 powers <- list(design1_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design1_power)), "Errors" = rep("Gaussian", nrow(design1_power))), 
-               design2_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Site Effect", nrow(design2_power)), "Errors" = rep("Gaussian", nrow(design2_power))), 
+               design2_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design2_power)), "Errors" = rep("Gaussian", nrow(design2_power))), 
                design3_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design3_power)), "Errors" = rep("Heavy-tailed", nrow(design3_power))), 
-               design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Site Effect", nrow(design4_power)), "Errors" = rep("Heavy-tailed", nrow(design4_power))))
+               design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design4_power)), "Errors" = rep("Heavy-tailed", nrow(design4_power))))
 
@@
 ```
+
+```{r power_summary_tab, results = "asis"}
+summary05 <- t(sapply(powers, function(x) x[5, c(6, 7, 1:5)]))
+summarytab <- xtable(summary05, align = "rp{1.15in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}",
+       caption = "Empirical power at level $0.05$ for Gaussian simulated data",
+       label = "tab:normal_power") 
+print(summarytab, type="latex", file="../ms/fig/normal_simulation_power_summary.tex")
+summarytab
+```

### 2016-12-08 simulations/simulation_skewed.Rmd
Rerun sims with 1000 reps to get cleaner plots and power estimates

--- 2016-12-07 simulations/simulation_skewed.Rmd
@@
 for individuals $i = 1, \dots, n_j$, $j = 1, \dots, J$.
-$\beta_0$ is the coefficient for the baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being at site $j$, $Z_{ij}$ is the treatment level indicator,, $\gamma_j$ is the effect of treatment at site $j$, and $\varepsilon_{ij}$ takes the values $0$ or $0.5$ with equal probability.
+$\beta_0$ is the coefficient for the baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being at stratum $j$, $Z_{ij}$ is the treatment level indicator,, $\gamma_j$ is the effect of treatment at stratum $j$, and $\varepsilon_{ij}$ takes the values $0$ or $0.5$ with equal probability.
 We will assume that $\beta_0 = 1$.
@@
 * Design 2: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$.
-This is a constant, additive treatment effect at site 1, but no treatment effect at sites 2 and 3.
+This is a constant, additive treatment effect at stratum 1, but no treatment effect at strata 2 and 3.
 This is a simplistic case of heterogeneous treatment effects.
@@
 ```{r generate_sample}
-generate_outcome <- function(baseline, tr_effect, site_effect, error){
-  Y1 <- baseline + tr_effect + site_effect + error
+generate_outcome <- function(baseline, tr_effect, stratum_effect, error){
+  Y1 <- baseline + tr_effect + stratum_effect + error
   Y1 <- floor(Y1)
@@
   # gamma = the magnitude of the treatment effect
-  # effect = "same effect" or "single site effect" - which sites have a tr effect > 0?
-  # n = number of individuals at each site
-  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), SITEID (stratum), site_effect (beta coefficient per individual), and epsilon (errors)
+  # effect = "same effect" or "single stratum effect" - which strata have a tr effect > 0?
+  # n = number of individuals at each stratum
+  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), stratumID (stratum), stratum_effect (beta coefficient per individual), and epsilon (errors)
 
-  SITEID <- rep(1:3, times = n)
+  stratumID <- rep(1:3, times = n)
   N <- sum(n)
@@
   Z <- rep(0:1, length.out = N)
-  site_effect <- rep(beta, times = n)
-  Y1 <- generate_outcome(Y0, gamma_vec*Z, site_effect, epsilon)
-  return(data.frame(Y1, Y0, Z, gamma_vec, SITEID, site_effect, epsilon))
+  stratum_effect <- rep(beta, times = n)
+  Y1 <- generate_outcome(Y0, gamma_vec*Z, stratum_effect, epsilon)
+  return(data.frame(Y1, Y0, Z, gamma_vec, stratumID, stratum_effect, epsilon))
 }
@@
   tmp$epsilon <- rnorm(nrow(tmp))
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
-  tmp$Y1 <- generate_outcome(tmp$Y0, rep(0, nrow(tmp)), tmp$site_effect)
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- generate_outcome(tmp$Y0, rep(0, nrow(tmp)), tmp$stratum_effect, tmp$epsilon)
   generate_simulated_pvalues(tmp)
@@
 tmp <- generate_simulated_data(gamma = 0.5, effect = "same effect")
-design1_pvalues <- replicate(100, {
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+design1_pvalues <- replicate(1000, {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
   tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
-  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
+  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$stratum_effect, tmp$epsilon)
   generate_simulated_pvalues(tmp)
@@
 
-## Design 2: Single site effect
+## Design 2: Single stratum effect
 
@@
 
-tmp <- generate_simulated_data(gamma = 0.5, effect = "single site effect")
-design2_pvalues <- replicate(100, {
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+tmp <- generate_simulated_data(gamma = 0.5, effect = "single stratum effect")
+design2_pvalues <- replicate(1000, {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
   tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
-  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
+  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$stratum_effect, tmp$epsilon)
   generate_simulated_pvalues(tmp)
@@
 design3_pvalues <- replicate(100, {
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
   tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
-  tmp$Y1 <- generate_outcome(0.1*tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
+  tmp$Y1 <- generate_outcome(0.1*tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$stratum_effect, tmp$epsilon)
   generate_simulated_pvalues(tmp)
@@
 tmp <- generate_simulated_data(gamma = 0.5, effect = "same effect", n = c(8, 16, 24))
-design4_pvalues <- replicate(100, {
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+design4_pvalues <- replicate(1000, {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
   tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
-  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
+  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$stratum_effect, tmp$epsilon)
   generate_simulated_pvalues(tmp)
@@
 
-# Design 5: Single site effect, imbalanced design
+# Design 5: Single stratum effect, imbalanced design
 
-There is very low power to detect the effect here, since in the simulated data, the site with only 8 patients was the only site with a nonzero effect of treatment.
+There is very low power to detect the effect here, since in the simulated data, the stratum with only 8 patients was the only stratum with a nonzero effect of treatment.
 The treatment effect is on the same order of magnitude as the error variance.
@@
 
-tmp <- generate_simulated_data(gamma = 0.5, effect = "single site effect", n = c(8, 16, 24))
+tmp <- generate_simulated_data(gamma = 0.5, effect = "single stratum effect", n = c(8, 16, 24))
 design5_pvalues <- replicate(100, {
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
   tmp$epsilon <- rbinom(nrow(tmp), 1, 0.5) - 0.5
-  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$site_effect, tmp$epsilon)
+  tmp$Y1 <- generate_outcome(tmp$Y0, tmp$Z*tmp$gamma_vec, tmp$stratum_effect, tmp$epsilon)
   generate_simulated_pvalues(tmp)
@@
 ```{r design5_plot, fig.height = 4, fig.align="center"}
-p5 <- plot_power_curves(design5_power, "Effect of 0.5 at a Single Site, Imbalanced Design")
+p5 <- plot_power_curves(design5_power, "Effect of 0.5 at a Single stratum, Imbalanced Design")
 p5 
-plot_pvalue_hist(design5_pvalues, "Effect of 0.5 at a Single Site, Imbalanced Design")
+plot_pvalue_hist(design5_pvalues, "Effect of 0.5 at a Single stratum, Imbalanced Design")
 ```
@@
 powers <- list(design1_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design1_power)), "Design" = rep("Balanced", nrow(design1_power))), 
-               design2_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Site Effect", nrow(design2_power)), "Design" = rep("Balanced", nrow(design2_power))), 
+               design2_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design2_power)), "Design" = rep("Balanced", nrow(design2_power))), 
                design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design4_power)), "Design" = rep("Imbalanced", nrow(design4_power))), 
-               design5_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Site Effect", nrow(design5_power)), "Design" = rep("Imbalanced", nrow(design5_power))))
+               design5_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design5_power)), "Design" = rep("Imbalanced", nrow(design5_power))))
 
@@
 ```
+
+```{r power_summary_tab, results = "asis"}
+summary05 <- t(sapply(powers, function(x) x[5, c(6, 7, 1:5)]))
+summarytab <- xtable(summary05, align = "rp{1.25in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}",
+       caption = "Empirical power at level $0.05$ for discrete, skewed simulated data",
+       label = "tab:skewed_power")
+print(summarytab, type="latex", file="../ms/fig/skewed_simulation_power_summary.tex") 
+summarytab 
+```

### 2016-12-08 ms/ms.tex
Include power plots and tables
Add updated sims to ms with some additional discussion

@@
 \usepackage[singlespacing]{setspace}
+\usepackage{graphicx}
 
@@
 
-\todo{Summary of results for normal sims}
-all 3 methods with linear model seem to be equally good. when everything is satisfied, all methods do well;
+Figure~\ref{fig:normal_sim_power} shows the estimated power curves in these four designs using a treatment effect of $\gamma = 1$.
+The best case is when the errors are Gaussian and the treatment effect is constant across strata, while the worst case is when the errors are heavy-tailed and the treatment effect only appears in one stratum.
+\todo{it makes sense that this happens as assumptions are broken in other cases}
+A consistent pattern appears in each case: the stratified permutation test has the lowest power, while the other four tests tend to have very similar power.
+Table~\ref{tab:normal_power} displays the value of these power curves at level $0.05$.
+These numbers show that actually, for $\alpha=0.05$, the discrepancy between the stratified permutation test and the others is not so large when the effect is isolated in a single stratum.
+In this case, the permutation test may have more power than the parametric ANOVA at significance levels that are used in practice.
+
+\begin{figure}
+\centering
+\includegraphics[width = \textwidth]{fig/normal_simulation_power}
+\caption{Empirical power curves for the Gaussian simulated data}
+\label{fig:normal_sim_power}
+\end{figure}
+\begin{center}
+\input{fig/normal_simulation_power_summary.tex}
+\end{center}
 
@@
 
-\todo{Summary of results for skewed sims}
+In these simulations, we examine the effect of sample size.
+First, we consider balanced samples, where each stratum contains 16 individuals, 8 of whom receive one treatment and the other 8 of whom receive the other.
+Next, we consider imbalanced samples, where the smallest stratum has only 8 individuals, the next has 16, and the largest has 24.
+We again consider the four combinations of treatment effects and sample sizes.
+In the case where there is only a non-null treatment effect in one stratum and sample sizes are different, this effect occurs in the smallest stratum.
 
-\textbf{discussion of results}
- \bit
-\item repeated measures: don't naively control for baseline by taking diffs (\cite{frison_repeated_1992})
-\item report relative power or relative efficiency of ANOVA against diffed perm test, LM test, and the Freedman-Lane test
-\eit
+Figure~\ref{fig:skewed_sim_power} shows the estimated power curves in these four designs using a treatment effect of $\gamma = 0.5$.
+When the treatment effect is constant, sample sizes within each stratum don't matter; the power curves look the same up to chance variation.
+When the treatment effect only occurs in one stratum, there is a substantial loss in power due to halving the number of individuals in the stratum where the effect is present.
+In practice, one does not usually know a priori which individuals will be affected by treatment (otherwise, those unaffected by treatment would be excluded from the study altogether),
+so this result suggests using caution when stratifying.
+As in the first set of simulations, the stratified permutation test using the raw outcomes has the lowest power of the five tests.
+Now, however, the stratified permutation test using the difference between outcome and baseline also has lower power than the three using linear models.
+Presumably, this is because the differences are also discrete and lie in a small range of values, so the test statistic does not vary greatly across permutations.
+Table~\ref{tab:skewed_power} confirms this: power for the three linear model methods is high when the treatment is constant across strata, and power for all of the methods suffers when the effect is isolated in a single stratum.
+
+\begin{figure}
+\centering
+\includegraphics[width = \textwidth]{fig/skewed_simulation_power}
+\caption{Empirical power curves for the skewed, discrete simulated data}
+\label{fig:skewed_sim_power}
+\end{figure}
+
+\begin{center}
+\input{fig/skewed_simulation_power_summary.tex}
+\end{center}
+
+It makes sense that in all of these simulations, the permutation test using the difference in outcome and baseline as the response measure is more powerful than the permutation test using the raw outcomes.
+In additional simulations\footnote{see supplementary files on GitHub}, we changed the Gaussian model so that the outcome and baseline had correlation of $0.25$.
+In this case, the result was reversed: the differenced test had low power while the test using the raw outcomes had a power curve closer to the tests that adjust for covariates using a linear model.
+This phenomenon has been studied before \todo{cite}, and it is recommended to use the difference in outcome and baseline only when their correlation is larger than $0.5$ (\cite{frison_repeated_1992}).
 
@@
 
+\bibliographystyle{plainnat}
+\bibliography{references}
+
 \end{document}

### 2016-12-09 results/results.Rmd
Untrack old data analysis files; update data analysis to use cleaned dataset; add cache files to gitignore

--- 2016-11-25 results/rbdata.Rmd
@@
 ---
-title: "RB Data"
+title: "Clinical Trial Data Analysis"
 author: "Kellie Ottoboni"
@@
 
-ga <- read.csv("GA1102.csv", header = TRUE, stringsAsFactors = FALSE)
+clinical <- read.csv("../data/clinical_cleaned.csv", header = TRUE, stringsAsFactors = FALSE)
 ```
@@
 
-### Missing data
-There's a bunch of missing data. 
-First, we notice there are 14 subjects who were only observed during the baseline period.
+We preprocessed the data from its original format in the file "data/clinical_cleaned.R" so that it would be sufficiently anonymized.
+We took the following steps:
 
-```{r baseline_only, results = "asis"}
-numdays <- table(ga$SUBJID)
-baseline_only <- names(numdays)[numdays < 14]
-unique_numdays <- as.data.frame(table(numdays))
-names(unique_numdays) <- NULL
-print(xtable(unique_numdays,
-      caption = "Number of subjects observed for 7 or 14 days"),
-      include.rownames = FALSE, comment = FALSE)
+* Initially, there were 188 subjects.
+52 of these were never assigned to a treatment group and were only observed at baseline.
+This ensures that censoring was independent of treatment, so removing them from the analysis does not introduce any selection bias.
+We are left with 66 subjects in group A and 70 subjects in group B.
+* We renamed treatments, subject IDs, and site IDs to remove identifying information.
+* Each individual was observed for a total of 14 days, 7 during the baseline week and 7 after receiving treatment.
+We averaged the 7 measurements taken during each of these time periods, leaving us with 2 measurements per individual.
 
-ga$missing <- ifelse(ga$SUBJID %in% baseline_only, 1, 0)
-```
+# Exploratory Data Analysis
 
-Next, we check how many individuals were assigned to each treatment group.
-52 subjects have treatment values of `.`.
-They appear to never have been surveyed during week 2: although there are records in the dataset, all values are 0.
-These 52 include the fourteen who were only observed during week 1.
-We remove all these missing subjects.
-We are left with a total of 66 subjects in the placebo group and 70 subjects in the active treatment group.
+We notice several things patterns from Figure \ref{fig:count_by_site}.
+Some of the sites have many more subjects than others: these are sites 1, 3, and 6, and 7.
+Sites 2, 5, and 8 have very few subjects, so we will have low power at these sites.
+For example, after removing the missing values, site 8 only has one person in group B and two in group A, so the site will only have 3 unique permutation statistics.
 
-```{r baseline_only2, results = "asis"}
-num_treated <- table(ga$tr)
-baseline_only2 <- as.data.frame(num_treated/14)
-colnames(baseline_only2) <- c("Treatment", "Number of subjects")
-print(xtable(baseline_only2, digits = 0,
-             caption = "Number of subjects in each treatment group"), 
-      include.rownames = FALSE, comment = FALSE)
-
-ga$missing <- ga$missing + ifelse(ga$tr == ".", 1, 0)
-```
-
-## Exploratory Data Analysis
-
-We notice several things patterns from Figure \ref{fig:visualize_missing}, showing missing data patterns.
-First, some of the sites have many more subjects than others: these are sites 1, 4, and 12.
-Second, some of the sites have a large proportion of missing data, such as sites 3, 4, 5, and 14.
-Finally, we notice that sites 3, 6, and 14 have very few subjects, so we will have low power at these sites.
-For example, after removing the missing values, site 14 only has one treated subject and two placebos, so the site will only have 3 unique permutation statistics.
-
-```{r visualize_missing, fig.align="center", fig.cap = "Number of subjects in each treatment group, by site ID."}
-ga %>%
-  mutate(subjid_by_site = SUBJID %% 100) %>%
-  ggplot(aes(x = subjid_by_site, y = factor(SITEID))) + 
-  geom_tile(aes(fill = tr)) + 
-  xlab("Subject ID Number") +
-  ylab("Site ID") + 
-  scale_fill_discrete(name="Treatment",
-                      breaks=c(".", "1", "2"),
-                      labels=c("Not assigned", "Placebo", "Drug")) +
-  ggtitle("Treatment Assignment Patterns")
-```
-
-```{r remove_missing, include = FALSE}
-ga <- ga %>% filter(tr != ".") %>% mutate(tr = as.numeric(tr)-1)
+```{r count_by_site, fig.align="center", fig.cap = "Number of subjects in each treatment group, by site ID."}
+clinical %>% 
+  filter(VISITNUM == 1) %>% 
+  ggplot(aes(x = factor(SITEID))) + 
+  geom_bar(aes(fill = tr), position = "dodge") +
+  xlab("Site ID") +
+  ylab("Count") +
+  ggtitle("Number of Individuals per Treatment") +
+  labs(fill = "Treatment") 
 ```
@@
 
-data_by_subjid_visitnum <- ga %>% group_by(SUBJID, VISITNUM)
+data_by_subjid_visitnum <- clinical %>% group_by(SUBJID, VISITNUM)
 
-reshape_ga_continuous <- function(variable, data = data_by_subjid_visitnum){
+reshape_data <- function(variable, data = data_by_subjid_visitnum){
   data <- data %>% mutate_(myvariable = variable)
-  summarized <- data %>% summarise("variable" = mean(as.numeric(myvariable)), "tr" = unique(tr), "SITEID" = unique(SITEID))
-  cleaned <- dcast(summarized, SUBJID + tr + SITEID ~ VISITNUM, value.var = "variable")
+  cleaned <- dcast(data, SUBJID + tr + SITEID ~ VISITNUM, value.var = "myvariable")
   colnames(cleaned) <- c("SUBJID", "tr", "SITEID", "Baseline", "Outcome")
@@
 ```
-Figure \ref{fig:hist_continuous_vars} shows the distribution of the final outcome measures between treatment and placebo groups.
-Based on these distributions, we would expect to detect a difference in outcomes between the drug and placebo groups for the `daily_heart`, `daily_hrdq`, and `heart_freq` measures.
+Figure \ref{fig:hist_continuous_vars} shows the distribution of the final outcome measures between the two treatment groups.
+Based on these distributions, we would expect to detect a difference in outcomes between the treatments for the `daily_heart`, `daily_hrdq`, and `heart_freq` measures.
 
@@
 plot_distrs <- lapply(continuous_vars, function(variable){
-  p <- reshape_ga_continuous(variable) %>% mutate(tr = factor(tr)) %>%
+  p <- reshape_data(variable) %>% mutate(tr = factor(tr)) %>%
         ggplot(aes(Outcome)) +
@@
 #        geom_histogram(binwidth = 0.5, position = "dodge", aes(fill = tr)) +
-        scale_fill_discrete(name="Experimental\nCondition",
-                       breaks=c("0", "1"),
-                       labels=c("Placebo", "Drug")) + 
-        xlab(variable)
+#        scale_fill_discrete(name="Experimental\nCondition",
+#                       breaks=c("0", "1"),
+#                       labels=c("A", "B")) + 
+        labs(x = variable, fill = "Treatment")
   return(p)
@@
 
-### Ordinal variables
-
-Several variables are measured on an ordinal scale but treated as continuous values.
-For instance, heartburn frequency is measured on a 7-point scale, with levels 0, 1, 2, 3, 4-5, 6-10, and more than 10.
-Values in the 4-5 range are coded as 4.5 and in the 6-10 range are coded as 8.
-We do not necessarily believe that these are accurate representations of the data.
-We recode the ordinal scales as factors on a scale from 1 to 7.
-
-```{r recode}
-recode_ordinal <- function(variable){
-  levels <- 
-  variable <- as.factor(variable)
-  levels(variable) <- seq_along(unique(variable))
-  return(variable)
-}
-
-ga <- ga %>% mutate(heart_sev_f = recode_ordinal(heart_sev),
-                    regurg_sev_f = recode_ordinal(regurg_sev),
-                    dysp_sev_f = recode_ordinal(dysp_sev),
-                    heart_freq_f = recode_ordinal(heart_freq),
-                    regurg_freq_f = recode_ordinal(regurg_freq),
-                    dysp_freq_f = recode_ordinal(dysp_freq)
-                    )
-```
-We must put the data in a usable form so there is only one measurement for each individual.
-For each person, we take the median response from week 2 and subtract it from their median response from week 1.
-
-```{r data_structure_ordinal}
-
-data_by_subjid_day <- ga %>% 
-  mutate(dayofweek = rep(1:7, length.out = nrow(ga))) %>%
-  group_by(SUBJID, dayofweek)
-
-# The reshape function below reduces the 14 days into a single measure with the following steps:
-# 1) Take the difference in measurements between the first day of treatment week and the first day of baseline week, the second day of treatment week and the second day of baseline week, and so on.
-# 2) Take the median of these seven numbers.
-# We have decided not to use this method; it assumes that day j in weeks 1 and 2 are paired, which may not be right.
-
-# reshape_ga_ordinal <- function(variable, data = data_by_subjid_day){
-#   data <- data %>% mutate_(myvariable = variable)
-#   day_diff <- data %>% 
-#     summarise("variable" = diff(as.numeric(myvariable)), "tr" = unique(tr), "SITEID" = unique(SITEID))
-#   summarized <- day_diff %>%
-#     summarise("medoverdays" = median(variable), "tr" = unique(tr), "SITEID" = unique(SITEID))
-#   summarized <- ungroup(summarized)
-#   return(summarized)
-# }
-reshape_ga_ordinal <- function(variable, data = data_by_subjid_visitnum){
-  data <- data %>% mutate_(myvariable = variable)
-  summarized <- data %>% summarise("variable" = median(as.numeric(myvariable)), "tr" = unique(tr), "SITEID" = unique(SITEID))
-  cleaned <- dcast(summarized, SUBJID + tr + SITEID ~ VISITNUM, value.var = "variable")
-  colnames(cleaned) <- c("SUBJID", "tr", "SITEID", "Baseline", "Outcome")
-  cleaned <- ungroup(cleaned) %>% 
-    mutate(difference = Outcome - Baseline)
-  return(cleaned)
-}
-```
-Figure \ref{fig:hist_ordinal_vars} displays the distribution of these newly-coded ordinal variables in the treatment and control groups.
-There are certainly differences in distributions, but it is not clear how different they are.
-
-```{r hist_ordinal_vars, fig.height=6, fig.width=12, fig.cap = "Distribution of ordinal outcomes"}
-ordinal_vars <- c("heart_sev_f", "regurg_sev_f", "dysp_sev_f",
-                    "heart_freq_f", "regurg_freq_f", "dysp_freq_f") 
-plot_distrs <- lapply(ordinal_vars, function(variable){
-  p <- reshape_ga_ordinal(variable) %>% 
-        mutate(tr = factor(tr)) %>%
-        ggplot(aes(difference, ..density..)) +
-        geom_bar(aes(y = (..count..)/sum(..count..), fill = tr), position = "dodge") + 
-        scale_y_continuous(labels = scales::percent) +
-        scale_fill_discrete(name="Experimental\nCondition",
-                       breaks=c("0", "1"),
-                       labels=c("Placebo", "Drug")) +
-        xlab(variable)
-  return(p)
-})
-
-tmp <- ggplot_gtable(ggplot_build(plot_distrs[[1]]))
-leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
-legend <- tmp$grobs[[leg]]
-
-plot_distrs <- lapply(plot_distrs, function(x) x + theme(legend.position="none"))
-plot_distrs[[length(plot_distrs)+1]] <- legend
-do.call(grid.arrange, c(plot_distrs, nrow = 3))
-```
-
 # Analysis - Primary endpoint
-Let's first restrict attention to the primary endpoint, `daily_hrdq`.
-We must transform the data to contain the variables of interest: the baseline score (averaged over 7 days of no treatment), the outcome score (averaged over 7 days of treatment), an indicator for which treatment was received (`1` for the active drug, `0` for placebo), the site ID, and subject ID.
-
+To illustrate the different methods, let's first restrict attention to the primary endpoint, `daily_hrdq`.
 Throughout, we compare two models: 
@@
 
-We compare three testing methods: 
+We compare four testing methods: 
 the usual parametric ANOVA,
 a permutation test considering the difference in means as test statistic, stratifying by site,
+the stratified permutation test, using the $t$-statistic from the linear regression instead,
 and a permutation test in the style of Freedman and Lane (1983), using a linear regression to adjust for covariates.
@@
 ```{r hrdq, results = "asis"}
-daily_hrdq <- reshape_ga_continuous("daily_hrdq")
-summary_stats_hrdq <- cbind(summary(daily_hrdq$difference[daily_hrdq$tr == 1]),
-                            summary(daily_hrdq$difference[daily_hrdq$tr == 0]))
-colnames(summary_stats_hrdq) <- c("Drug", "Placebo")
+daily_hrdq <- reshape_data("daily_hrdq")
+summary_stats_hrdq <- cbind(summary(daily_hrdq$difference[daily_hrdq$tr == "B"]),
+                            summary(daily_hrdq$difference[daily_hrdq$tr == "A"]))
+colnames(summary_stats_hrdq) <- c("B", "A")
 print(xtable(summary_stats_hrdq,
-             caption = "Summary statistics for change in primary endpoint (daily hrdq) for drug and placebo groups."),
+             caption = "Summary statistics for change in primary endpoint (daily hrdq) for the two treatment groups."),
       include.rownames = TRUE)
@@
 Under model 2, the significance goes away.
-
+(Note that treatment is still coded as A and B, which are characters.
+The `lm` function is smart enough to know that it should be coded as a factor, where we compare the effect of B relative to A.
+In general, be very careful when you're doing this!)
 ```{r hrdq_parametric, results = "asis"}
@@
 
-Suppose each individual, $i = 1, \dots, N$ has two potential outcomes $(Y_i(1), Y_i(0))$ that indicate their response to the active drug and to the placebo, respectively.
-We randomly assign treatment to individuals; treatment determines which of $Y_i(1)$ and $Y_i(0)$ we observe.
+Suppose each individual, $i = 1, \dots, N$ has two potential outcomes $(Y_i(A), Y_i(B))$ that indicate their response to treatments A and B, respectively.
+We randomly assign treatment to individuals; treatment determines which of $Y_i(A)$ and $Y_i(B)$ we observe.
 We are unable to observe both.
@@
 
-$$H_0: Y_i(1) = Y_i(0), i = 1, \dots, N.$$
+$$H_0: Y_i(A) = Y_i(B), i = 1, \dots, N.$$
 
-Then, whether individual $i$ received the drug or placebo amounts to an arbitrary label.
-Once we observe their response under treatment $T_i$, we know what it would have been under treatment $1-T_i$, for $T_i \in \{0,1\}$; namely, it would be the same.
+Then, whether individual $i$ received A or B amounts to an arbitrary label.
+Once we observe their response under one treatment, we know what it would have been under the other; namely, it would be the same.
 
 Treatment was completely randomized at each site, independently across sites.
-Assuming that drop-out from the study was independent of treatment assignment (which is true if treatment was blinded and assigned at random), then we may condition on the number of individuals who received the drug and who received the placebo at each site.
+Assuming that drop-out from the study was independent of treatment assignment (which is true if treatment was blinded and assigned at random), then we may condition on the number of individuals who received A and B at each site.
 Any assignment of treatments which preserves the number of treated and controls at each site is valid and has equal probability of occurring.
@@
 
-We compare two test statistics: the usual difference in means between treatment and placebo groups, and the difference in means within each site, aggregated by taking the sum of their absolute values over sites.
+We compare two test statistics: the usual difference in means between group B and group A, and the difference in means within each site, aggregated by taking the sum of their absolute values over sites.
 The first statistic is more comparable to what one would obtain from an ANOVA and it is readily interpretable.
@@
 # method 1 : do permutation of differences
-observed_diff_means1 <- mean(daily_hrdq[daily_hrdq$tr == 1, ]$Outcome) - mean(daily_hrdq[daily_hrdq$tr == 0, ]$Outcome)
+observed_diff_means1 <- mean(daily_hrdq[daily_hrdq$tr == "B", ]$Outcome) - mean(daily_hrdq[daily_hrdq$tr == "A", ]$Outcome)
 diff_means_distr1 <- stratified_two_sample(group = daily_hrdq$tr, response = daily_hrdq$Outcome, stratum = daily_hrdq$SITEID, reps = 1e4)
@@
 
-observed_diff_means2 <- mean(daily_hrdq[daily_hrdq$tr == 1, ]$difference) - mean(daily_hrdq[daily_hrdq$tr == 0, ]$difference)
+observed_diff_means2 <- mean(daily_hrdq[daily_hrdq$tr == "B", ]$difference) - mean(daily_hrdq[daily_hrdq$tr == "A", ]$difference)
 diff_means_distr2 <- stratified_two_sample(group = daily_hrdq$tr, response = daily_hrdq$difference, stratum = daily_hrdq$SITEID, reps = 1e4)
@@
 ## Covariate-adjusted permutation test
-We would like to test for a difference in outcomes between the active drug and placebo group, but control for other covariates.
+We would like to test for a difference in outcomes between the treatment groups group, but control for other covariates.
 In particular, when the outcome is the average response during the second week of follow-up, we would like to control for the average response during the first week of follow-up and the location.
@@
 
-Let $Y$ denote the response, $Z$ denote the treatment indicator, and $X$ denote a covariate which may be correlated with $Y$.
+Let $Y$ denote the response, $Z$ denote the treatment indicator ($1$ if the person gets treatment B, $0$ if they get A), and $X$ denote a covariate which may be correlated with $Y$.
 We may write the following equation:
@@
 
-observed_t1 <- summary(lm1)[["coefficients"]]["tr", "t value"]
+observed_t1 <- summary(lm1)[["coefficients"]]["trB", "t value"]
 lm1_t_distr <-  replicate(1e4, {
@@
   lm1_perm <- lm(Outcome ~ Baseline + tr_perm + factor(SITEID), data = daily_hrdq)
-  summary(lm1_perm)[["coefficients"]]["tr_perm", "t value"]
+  summary(lm1_perm)[["coefficients"]]["tr_permB", "t value"]
 })
@@
 
-observed_t2 <- summary(lm2)[["coefficients"]]["tr", "t value"]
+observed_t2 <- summary(lm2)[["coefficients"]]["trB", "t value"]
 lm2_t_distr <-  replicate(1e4, {
@@
   lm2_perm <- lm(difference ~ tr_perm + factor(SITEID), data = daily_hrdq)
-  summary(lm2_perm)[["coefficients"]]["tr_perm", "t value"]
+  summary(lm2_perm)[["coefficients"]]["tr_permB", "t value"]
 })
@@
 lm1_yhat <- fitted(lm1_no_tr)
-observed_t1 <- summary(lm1)[["coefficients"]]["tr", "t value"]
+observed_t1 <- summary(lm1)[["coefficients"]]["trB", "t value"]
 lm1_t_distr <-  replicate(1e4, {
@@
   lm1_perm <- lm(response_fl ~ Baseline + tr + factor(SITEID), data = daily_hrdq)
-  summary(lm1_perm)[["coefficients"]]["tr", "t value"]
+  summary(lm1_perm)[["coefficients"]]["trB", "t value"]
 })
@@
 lm2_yhat <- fitted(lm2_no_tr)
-observed_t2 <- summary(lm2)[["coefficients"]]["tr", "t value"]
+observed_t2 <- summary(lm2)[["coefficients"]]["trB", "t value"]
 lm2_t_distr <-  replicate(1e4, {
@@
   lm2_perm <- lm(response_fl ~ tr + factor(SITEID), data = daily_hrdq)
-  summary(lm2_perm)[["coefficients"]]["tr", "t value"]
+  summary(lm2_perm)[["coefficients"]]["trB", "t value"]
 })
@@
 
-outcome_pvalues <- c(summary(aov(lm1))[[1]]["tr", "Pr(>F)"],
+outcome_pvalues <- c(summary(aov(lm1))[[1]]["trB", "Pr(>F)"],
                      diff_means_pvalue1,
@@
                      fl_pvalue_1)
-differenced_pvalues <- c(summary(aov(lm2))[[1]]["tr", "Pr(>F)"],
+differenced_pvalues <- c(summary(aov(lm2))[[1]]["trB", "Pr(>F)"],
                      diff_means_pvalue2,
@@
   i <- i+1
-  tmpdata <- reshape_ga_continuous(col)
+  tmpdata <- reshape_data(col)
   lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = tmpdata)
@@
 
-  observed_diff_means <- mean(tmpdata[tmpdata$tr == 1, ]$Outcome) - mean(tmpdata[tmpdata$tr == 0, ]$Outcome)
+  observed_diff_means <- mean(tmpdata[tmpdata$tr == "B", ]$Outcome) - mean(tmpdata[tmpdata$tr == "A", ]$Outcome)
   diff_means_distr <- stratified_two_sample(group = tmpdata$tr, response = tmpdata$Outcome, stratum = tmpdata$SITEID, reps = 1e3)
@@
 
-  observed_t <- summary(lm1)[["coefficients"]]["tr", "t value"]
+  observed_t <- summary(lm1)[["coefficients"]]["trB", "t value"]
   lm1_t_distr <-  replicate(1e3, {
@@
   lm1_perm <- lm(Outcome ~ Baseline + tr_perm + factor(SITEID), data = tmpdata)
-  summary(lm1_perm)[["coefficients"]]["tr_perm", "t value"]
+  summary(lm1_perm)[["coefficients"]]["tr_permB", "t value"]
   })
@@
   lm_yhat <- fitted(lm_no_tr)
-  observed_t <- summary(lm1)[["coefficients"]]["tr", "t value"]
+  observed_t <- summary(lm1)[["coefficients"]]["trB", "t value"]
   lm_t_distr <-  replicate(1e3, {
@@
       lm_perm <- lm(response_fl ~ Baseline + tr + factor(SITEID), data = tmpdata)
-      summary(lm_perm)[["coefficients"]]["tr", "t value"]
+      summary(lm_perm)[["coefficients"]]["trB", "t value"]
 })
@@
 ```
-
-# Analysis - Ordinal variables
-
-For ordinal variables, it doesn't make sense to do ANOVA.
-Instead, we do a multi-aspect analysis to test whether there is any difference in the outcome distributions between the drug and placebo groups.
-For this, we use the same stratified permutation test approach as above.
-However, we use three different test statistics and combine them into a single, global test using the nonparametric combination (NPC) method.
-The test statistics are the difference in mean outcomes between the drug and placebo group, the Anderson-Darling statistic, and the Kolmogorov-Smirnov statistic.
-We use the Tippett combining function.
-
-The outcome measures used are the difference in median measures during the treatment week minus the median measures during the baseline week.
-Recall from Figure \ref{fig:hist_ordinal_vars} that it is not clear whether the distributions are different.
-Table \ref{ordinal_results_print} shows the p-values for the three partial tests plus the global p-value.
-None is significant.
-It appears that treating the variables as ordinal results in a loss of information and, consequently, power.
-
-```{r ordinal_tests}
-# Compute difference in mean value (median outcome) between treatment and placebo
-mean_median <- function(outcome, treatment) {
-  tr_levels <- sort(unique(treatment))
-  mean(outcome[treatment == tr_levels[2]], na.rm = TRUE) - 
-    mean(outcome[treatment == tr_levels[1]], na.rm = TRUE)
-  }
-
-# Anderson-Darling
-ad<-function(outcome, treatment){    
-  n <- length(treatment)   
-  W <- table(outcome, treatment)
-  ecdf <- apply(W, 2, cumsum)
-  ecdf_comb <- apply(ecdf, 1, sum)
-
-  ratio <- ecdf[,2]/(ecdf_comb*(n-ecdf_comb))
-  j <- length(ratio)
-  T_ad <- sum(ratio[seq_len(j-1)])
-  return(T_ad)
-}
-
-# Kolmogorov-Smirnov
-ks <- function(outcome, treatment){
-  tr_levels <- sort(unique(treatment))
-  # Assume that larger treatment level is active treatment, smaller other is control
-  g1 <- outcome[treatment == tr_levels[2]]
-  g0 <- outcome[treatment == tr_levels[1]]
-  suppressWarnings(ks.test(g1, g0)$statistic)
-}
-
-multiaspect_test <- function(outcome, treatment, strata, reps = 1000){
-  obs_ts <- c(mean_median(outcome, treatment),
-              ad(outcome, treatment),
-              ks(outcome, treatment))
-  names(obs_ts) <- c("Difference in Means", "Anderson-Darling", "Kolmogorov-Smirnov")
-  perm_distr <- t(replicate(reps, {
-            perm_treatment <- permute_within_groups(treatment, strata)
-            c(mean_median(outcome, perm_treatment),
-              ad(outcome, perm_treatment),
-              ks(outcome, perm_treatment))
-            }))
-
-  partial_pvalues <- sapply(seq_along(obs_ts), function(i){
-    t2p(obs_ts[i], perm_distr[, i], "two-sided")
-  })
-  names(partial_pvalues) <- c("Difference in Means", "Anderson-Darling", "Kolmogorov-Smirnov")
-  global_pvalue <- npc(obs_ts, perm_distr, alternatives="two-sided", combine = "tippett")
-
-  return(list(
-    "Partial test statistics" = obs_ts,
-    "Partial p-values"        = partial_pvalues,
-    "Global p-value"          = global_pvalue
-  ))
-}
-```
-
-```{r ordinal_results_make}
-set.seed(609726531) # Generated from Random.org Timestamp: 2016-11-09 15:38:50 UTC
-ordinal_vars <- c("heart_sev_f", "regurg_sev_f", "dysp_sev_f",
-                    "heart_freq_f", "regurg_freq_f", "dysp_freq_f")
-
-pvalues_table_ord <- as.data.frame(matrix(NA, nrow = length(ordinal_vars), ncol = 4))
-i <- 0
-
-for(col in ordinal_vars){
-  i <- i+1
-  tmpdata <- reshape_ga_ordinal(col)
-  res <- multiaspect_test(outcome = tmpdata$difference, treatment = tmpdata$tr, strata = tmpdata$SITEID, reps = 1e3)
-  pvalues_table_ord[i, 1] <- res$`Partial p-values`[1]
-  pvalues_table_ord[i, 2] <- res$`Partial p-values`[2]
-  pvalues_table_ord[i, 3] <- res$`Partial p-values`[3]
-  pvalues_table_ord[i, 4] <- res$`Global p-value` 
-} 
-```
-
-```{r ordinal_results_print, results = "asis"}
-rownames(pvalues_table_ord) <- ordinal_vars
-tests <- c("Difference in Means", "Anderson-Darling", "Kolmogorov-Smirnov", "Global P-value")
-colnames(pvalues_table_ord) <- tests
-
-print(xtable(pvalues_table_ord,  
-             digits = 3,
-             align = paste0(c("r|", rep("p{0.75in}", ncol(pvalues_table_ord))), 
-                            collapse = ""),
-             caption = "Comparison of p-values from three tests plus a global p-value, for each ordinal endpoint."), 
-      include.rownames = TRUE)
-```

### 2016-12-09 results/results.Rmd
Table of p-values from data results

@@
 
-print(xtable(pvalues_table_contin, 
+summarytab <- xtable(pvalues_table_contin, 
              digits = 3,
-             align = paste0(c("r|", rep("p{1in}", ncol(pvalues_table_contin))), 
+             align = paste0(c("r|", rep("p{1.2in}", ncol(pvalues_table_contin))), 
                             collapse = ""),
-             caption = "Comparison of p-values from four tests, for each continuous endpoint."),
-      include.rownames = TRUE)
+             caption = "Comparison of p-values from four tests, for each continuous endpoint.")
+print(summarytab, include.rownames = TRUE)
+
+print(summarytab, include.rownames = TRUE, type="latex", file="../ms/fig/results_pvalue_summary.tex") 
 ```

### 2016-12-15 ms/ms.tex
Table of p-values from data results
Major work on the introduction

@@
 \author{Kellie Ottoboni \\
-Department of Statistics\\
+Department of Statistics; Berkeley Institute for Data Science\\
 University of California, Berkeley\\ [.2in]
 Luigi Salmaso\\
-Department of Management Engineering \\
-University degli Studi di Padova \\ [.2in]
+Department of Management and Engineering \\
+University of Padova \\ [.2in]
 Fraser Lewis \\
-Reckitt Benckiser Group
+RB
 }\date{Draft \today}
@@
 \noindent\textbf{Background:}
-In clinical trials, the standard analysis of the effectiveness of one treatment over another is a parametric analysis of variance (ANOVA), which relies on uncheckable and ostensibly implausible assumptions such as x y and z.
-The decision of the effectiveness of the treatment is based upon a single p-value from this hypothesis test, which is essentially meaningless if the assumptions are not met.
-One would like to run some sort of robustness check, an additional test or something, whose results can call into question or give evidence in favor of the results from the ANOVA.
+In clinical trials, the standard analysis of the effectiveness of one treatment over another is a parametric analysis of covariance (ANCOVA), a method based on linear regression which relies on uncheckable and ostensibly implausible assumptions such as constant treatment effects across groups, normality of errors, and homoskedasticity.
+The decision of the effectiveness of the treatment is based upon a single p-value from this hypothesis test, which may not be valid if the assumptions are not met.
+One would like to run some sort of robustness check, an additional test or something, whose results can call into question or give evidence in favor of the results from the ANCOVA.
 
@@
 In contrast to parametric tests, which rely up on stringent assumptions to obtain null distributions with a simple closed form, permutation tests use the assumption of exchangeability or randomization to construct approximate null distributions by simulation, yielding (approximately) exact tests.
-In order to mimic the ANOVA that is typically done, we compare several permutation tests based on linear models, which enable one to control for pretreatment covariates and nuisance variables.
+In order to mimic the ANCOVA that is typically done, we compare several permutation tests based on linear models, which enable one to control for pretreatment covariates and nuisance variables.
 
 \noindent\textbf{Results:}
-We present simulation results showing that the permutation methods maintain power comparable to the ANOVA under a balanced design, but the ANOVA loses power when x y and z.
-We illustrate the use of permutation tests alongside the parametric ANOVA using data from a clinical trial across seven (?) sites with different numbers of patients.
+We present simulations showing that the permutation methods maintain power comparable to the ANCOVA in randomized trials.
+We illustrate the use of permutation tests alongside the parametric ANCOVA using data from a clinical trial across eight sites with different numbers of patients.
 
 \noindent\textbf{Conclusions:}
-These results provide researchers an alternative way to assess the reliability of their statistical analyses and strengthen their confidence in the conclusions.
+These results suggest that the parametric ANCOVA and linear model permutation tests perform similarly, even when the more stringent assumptions for the parametric test are not met.
 
@@
 % The Background section should explain the background to the study, its aims, a summary of the existing literature and why this study was necessary or its contribution to the field.
+A hypothesis test is a statistical method for determining whether observed data is consistent with a hypothesis about the process that generated the data.
+Medical trials use hypothesis testing to assess the evidence that a treatment has an effect on one or more clinically relevant outcomes.
+The simplest version of this experiment, which involves randomly assigning treatment to a fixed number of individuals in a group and measuring a single outcome, has been studied for nearly a century (see \citet{fisher_design_1935} and \citet{neyman_application_1923} for early references).
+One can conduct hypothesis tests and construct confidence intervals for the estimated treatment effect
+by exploiting the fact that the difference in average outcomes between the treatment and control groups is asymptotically normal with a variance that can be estimated from the data.
 
+Randomized experiments in the real world are rarely this simple:
+individuals are heterogeneous with respect to pretreatment covariates,
+they may not take the treatment they are assigned or may drop out of the study,
+and an individual's assignment to treatment may affect others' outcomes.
+Techniques have been developed to deal with each of these issues.
+In this paper, we focus on the first.
+
+Random assignment of treatment ensures that on average, across all possible randomizations, pretreatment covariates are balanced between treatment groups.
+However, in any particular randomization that is realized, there may be imbalances.
+If these covariates are predictive of the outcome, then even when there is no effect of treatment, differences in outcomes between treatment groups may appear.
+Controlling for such baseline covariates can reduce the variability of estimated treatment effects and yield more powerful hypothesis tests.
+
+Another method for controlling for baseline covariates is stratification:
+individuals are divided into groups based on levels of some covariate.
+A common stratification variable in clinical experiments is location:
+individuals often come from many sites because it is difficult to recruit a sufficient number of participants at one doctor's office or hospital.
+Treatments are randomly assigned  \textit{within strata} to guarantee that the stratum variable is balanced between treatment groups.
+As with other pretreatment covariates, outcomes may vary across strata due to factors unrelated to individuals.
+Controlling for stratum can have the same beneficial effects as controlling for other pretreatment covariates.
+
+\todo{not happy with this paragraph}
+The most common way to control for covariates is linear regression.
+The premise is to project the outcomes onto a linear plane which describes the contribution of each covariate to determining the outcome.
+The coefficient of any particular variable answers the question, if we were to hold fixed all other covariates and increase this variable by one unit, how much would the outcome change?
+This model posits a linear relationship between covariates, treatment, and outcome; if the true relationship is not linear, then it is unclear what such a regression is estimating.
+However, if one knows all covariates related to the outcome and fits a fully saturated linear model (including all covariates and their interactions), the coefficient for the treatment is an asymptotically consistent estimate of the average treatment effect (\cite{lin_agnostic_2013}).
+When controlling for strata, such as location, it is standard to use analysis of covariance (ANCOVA), a particular case of linear regression which allows each stratum to have its own mean outcome level.
+This amounts to choosing the plane of best fit for the data, allowing each stratum to have a different intercept but enforcing that the slope is the same for each.
+
+Hypothesis testing of estimated coefficients requires even stronger assumptions than linearity, such as normally distributed and homoskedastic errors.
+When this holds, the distribution of a coefficient scaled by its estimated variance, has a $t$-distribution.
+Thus, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the linear model, and can be evaluated analytically and quickly.
+However, these assumptions are clearly violated in many cases:
+when the data are discrete or ordinal, when the treatment has a differential effect across strata, and when measurement error sizes differ across strata.
+
+An alternative approach is permutation testing.
+Deliberate randomization induces a distribution for any test statistic under the null hypothesis that treatment has no effect on the outcome,
+because the randomization scheme provides information about all possible ways that treatment may have been assigned 
+and the null hypothesis tells us what each individual's response would be regardless of the assignment (namely, it would be the same).
+Instead of deciding how ``extreme'' the observed test statistic is relative to a parametric reference distribution, such as Student's $t$ or the Gaussian, one judges it against this randomization distribution.
+Such a test is exact, meaning that it controls the type I error rate at the pre-specified level, even in finite samples, whereas parametric hypothesis tests based on asymptotic approximations do not always guarantee good finite sample properties.
+In the past, parametric tests were necessary because asymptotic approximations were the only computationally feasible way to estimate distributions. 
+Now, computational power is no longer a barrier to finding exact (or exact to pre-specified precision) randomization distributions.
+In most cases, a randomization test is preferable to a parametric one:
+``[a] corresponding parametric test is valid only to the extent that it results in the same statistical decision [as the randomization test]'' (\cite{bradley_distribution_1968}).
+
+\todo{ lit review: 
 \bit 
-\item what's the problem with doing ANOVA? what are the assumptions and give examples of when they are commonly or clearly violated
-\item lit review: cite the vickers paper, what else have people done comparing linear models to permutation/nonparametric tests?
-\item why write this paper. what is the goal, where is the opportunity to make an advancement?
-\item overview of the paper results
+\item cite the vickers paper
+\item what else have people done comparing linear models to permutation/nonparametric tests?
 \eit
+}
+
+In this paper, we review ANCOVA and its permutation counterparts, comparing their performance in different scenarios, and illustrating their application with a clinical dataset.
+We find in simulations that even when assumptions are not satisfied, the parametric and permutation tests have comparable power to detect a treatment effect.
+We apply each of the methods to data from a clinical trial seeking to compare the performance of two treatments for gastroesophageal reflux disease (GERD).
+We conclude by discussing implications of these results for practitioners.
 
@@
 
-\subsection{Parametric ANOVA}
+\subsection{Parametric ANCOVA}
 
@@
 
-In the normal theory ANOVA, we make the following assumptions:
+To carry out hypothesis testing using parametric distributions, we make the following assumptions (\cite{freedman_statistical_2005}):
+
 \begin{enumerate}
-\item the $\eps$ are normally distributed with mean $0$ and common variance $\sigma^2$
-\item $\eps \independent X$
-\item \textcolor{red}{ what else? }
+\item Linearity: the data $Y$ are related to $X$ and $Z$ linearly
+\item Constant slopes: stratum only affects the intercept $\alpha_j$, not the slopes $\beta$
+\item IID Errors: the $\eps_{ij}$ are independent and identically distributed with mean $0$ and variance $\sigma^2$
+\item Independence: if $X$ is random, $\eps \independent X$
+\item Normality: the errors are normally distributed
 \end{enumerate}
 
-The model is fit using least squares.
+The coefficients are estimated using least squares.\footnote{
+Note, only Assumptions (1) and (2) are needed to guarantee the existence of a solution.
+If there is a linear relationship between $X$, $Z$, and stratum, the solution will not be unique.
+}
 A significance test is done for the estimated coefficient $\hat{\gamma}$, using the $t$-distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
@@
 $$ t = \frac{\hat{\gamma}}{\sqrt{ \hat{\sigma}_{\hat{\gamma}}^2}},$$
-where $\hat{\sigma}_{\hat{\gamma}}^2$ is the estimated standard error of $\hat{\gamma}$, obtained from the diagonal element of the inverse covariance matrix of the observed data which corresponds to the treatment.
-\todo{cite freedman linear models textbook}
+where $\hat{\sigma}_{\hat{\gamma}}^2$ is the estimated standard error of $\hat{\gamma}$.\footnote{
+$\hat{\sigma}_{\hat{\gamma}}^2$ is estimated by first computing the residual standard error $\hat{\sigma}^2$,
+then taking the diagonal element corresponding to the treatment variable in the inverse covariance matrix of covariates.
+}
 
 \todo{Add a paragraph about why this might go wrong, perhaps some citations. All depends on estimating the error variance correctly, because $t$ distribution is the ratio of a normal and a chi-square}
-\bit
-\item This ANOVA model allows for fixed stratum effects, but they are only additive in the outcome.
+
+The ANCOVA model allows for fixed stratum effects, but they are only additive in the outcome.  
 The model does not account for differences in the treatment effect across strata.  
-In other words, the estimated linear model is not a good estimate of the true data-generating process.
+The estimated linear model may not accurately capture details of the true data-generating process.  
 If effects are not constant across strata, then the coefficient $\gamma$ may be attenuated towards 0.  
 A test of the null hypothesis that $\gamma = 0$ will be valid, but will not reflect the true magnitude of treatment effects among individuals.
-\item \textcolor{red}{ what else? }
-\eit
 
@@
 
+\begin{center}
+\input{fig/results_pvalue_summary.tex}
+\end{center}
+
 \section{Discussion}

### 2016-12-16 ms/ms.tex
Edits and additions to background and methods

--- 2016-12-08 ms/ms.tex
@@
 \noindent\textbf{Background:}
-In clinical trials, the standard analysis of the effectiveness of one treatment over another is a parametric analysis of variance (ANOVA), which relies on uncheckable and ostensibly implausible assumptions such as x y and z.
-The decision of the effectiveness of the treatment is based upon a single p-value from this hypothesis test, which is essentially meaningless if the assumptions are not met.
-One would like to run some sort of robustness check, an additional test or something, whose results can call into question or give evidence in favor of the results from the ANOVA.
+In clinical trials, the standard method of measuring the effectiveness of one treatment over another is a parametric analysis of covariance (ANCOVA), a special case of linear regression which incorporates stratification.
+The method relies on strong, untestable assumptions about the data-generating process such as constant treatment effects across groups, normality of errors, and homoskedasticity.
+The effectiveness of the treatment is decided by the p-value from this hypothesis test, which may not be valid if the assumptions for ANCOVA are not met.
+We would like to understand which assumptions may be relaxed or violated for the ANCOVA to remain valid (but potentially less powerful).
 
@@
 Permutation tests present such an opportunity: they require only minimal assumptions which are often guaranteed by the randomization that was conducted.
-In contrast to parametric tests, which rely up on stringent assumptions to obtain null distributions with a simple closed form, permutation tests use the assumption of exchangeability or randomization to construct approximate null distributions by simulation, yielding (approximately) exact tests.
-In order to mimic the ANOVA that is typically done, we compare several permutation tests based on linear models, which enable one to control for pretreatment covariates and nuisance variables.
+In contrast to parametric tests, which rely up on strong assumptions to obtain a null distribution with a simple closed form, with permutation tests one uses the assumption of exchangeability or randomization to construct approximate null distributions by simulation.
+These tests are exact, even in finite samples.
+We compare ANCOVA to several permutation tests based on linear models, all of which control for pretreatment covariates to increase power and precision.
 
 \noindent\textbf{Results:}
-We present simulation results showing that the permutation methods maintain power comparable to the ANOVA under a balanced design, but the ANOVA loses power when x y and z.
-We illustrate the use of permutation tests alongside the parametric ANOVA using data from a clinical trial across seven (?) sites with different numbers of patients.
+Our simulations show that the permutation methods maintain power comparable to the ANCOVA in randomized trials.
+We illustrate the use of permutation tests alongside the parametric ANCOVA using data from a clinical trial conducted across eight sites,
+each having a different numbers of patients.
 
 \noindent\textbf{Conclusions:}
-These results provide researchers an alternative way to assess the reliability of their statistical analyses and strengthen their confidence in the conclusions.
+These results suggest that the parametric ANCOVA and linear model permutation tests perform similarly, even when not all assumptions for the parametric test are met.
 
@@
 \section{Background}
-% The Background section should explain the background to the study, its aims, a summary of the existing literature and why this study was necessary or its contribution to the field.
+A hypothesis test is a statistical method for determining whether observed data is consistent with a belief about the process that generated the data.
+Medical experiments use hypothesis testing to assess the evidence that a treatment has an effect on one or more clinically relevant outcomes.
+The simplest version of this experiment, which involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome, has been studied for nearly a century (see \citet{fisher_design_1935} and \citet{neyman_application_1923} for early references).
+One can conduct hypothesis tests and construct confidence intervals for the estimated treatment effect
+by exploiting the fact that the difference in average outcomes between the two treatment groups is asymptotically normal with a variance that can be estimated from the data.
 
-\bit
-\item what's the problem with doing ANOVA? what are the assumptions and give examples of when they are commonly or clearly violated
-\item lit review: cite the vickers paper, what else have people done comparing linear models to permutation/nonparametric tests?
-\item why write this paper. what is the goal, where is the opportunity to make an advancement?
-\item overview of the paper results
-\eit
+Randomized experiments in the real world are rarely this simple:
+individuals are heterogeneous with respect to pretreatment covariates,
+there may be more than two treatments under study,
+they may not take the treatment they are assigned or may drop out of the study,
+and an individual's assignment to treatment may affect others' outcomes.
+Techniques have been developed to deal with each of these issues.
+In this paper, we focus on the first.
+
+Random assignment of treatment ensures that pretreatment covariates are balanced between treatment groups on average, across all possible randomizations.
+However, in any particular randomization, there may be imbalances.
+If the imbalanced covariates are associated with the outcome, then even when there is no effect of treatment, we may observe differences in outcomes between treatment groups.
+Controlling for such baseline covariates can reduce the variability of estimated treatment effects and yield more powerful hypothesis tests. \todo{cite}
+
+Stratification is one method for controlling for covariates that are known a priori to be associated with the outcome.
+Individuals are divided into groups based on levels of this covariate, and random assignment of treatments is conducted within each group, independently across groups.
+This guarantees that the stratification variable is balanced between treatment groups.
+The stratification is done \textit{before any outcome data is collected}.
+A common stratification variable in clinical experiments is location:
+individuals often come from many sites because it is difficult to recruit a sufficient number of participants at one doctor's office or hospital.
+As with other pretreatment covariates, outcomes may vary across strata due to factors unrelated to individuals.
+Controlling for stratum can have the same beneficial effects as controlling for other pretreatment covariates.
+
+Linear regression is another common way to control for baseline covariates.
+It is done \textit{after} data are collected and outcomes are observed.
+Whereas stratification guarantees balance on the stratifying variable, other covariates may be imbalanced, so stronger modeling assumptions are needed to balance them after the fact.
+The idea of linear regression is to project the outcomes onto a plane which describes each variable's relationship to the outcome.
+The coefficient of any particular variable answers the question, if we were to hold fixed all other covariates and increase this variable by one unit, how much would we expect the outcome to change?
+This model posits a linear relationship between covariates, treatment, and outcome; 
+if the true relationship is not linear, then regression only estimates the best linear approximation to the conditional expectation of the outcome.
+When controlling for strata, such as location, it is standard to use analysis of covariance (ANCOVA), a particular case of linear regression which allows each stratum to have its own mean outcome level.
+This amounts to choosing the plane of best fit for the data, allowing each stratum to have a different intercept but enforcing that the slope is the same for each.
+
+Hypothesis testing of estimated coefficients requires even stronger assumptions than linearity, such as Gaussian, homoskedastic errors.
+When this holds, a regression coefficient scaled by its estimated variance follows Student's $t$ distribution.
+Thus, if we believe the linear model is correctly specified, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the ANCOVA, and can be evaluated analytically and quickly.
+However, the assumptions necessary for the coefficient to be $t$ distributed are clearly violated in many cases:
+when the data are discrete or ordinal, when the treatment has a differential effect across strata, and when measurement error sizes differ across strata.
+Furthermore, the hypothesis test implicitly assumes that the data were sampled at random from some underlying population, when in fact,
+recruitment of patients to medical experiments is rarely done this way \todo{cite}.
+
+An alternative approach is permutation testing. \todo{citations: fisher 1935, pitman 1937 and 1938}
+Deliberate randomization induces a distribution for any test statistic under the null hypothesis that treatment has no effect on the outcome:
+the randomization scheme provides information about all possible ways that treatment may have been assigned 
+and the null hypothesis tells us what each individual's response would be regardless of the assignment (namely, it would be the same).
+One determines how ``extreme'' the observed test statistic is relative to this randomization distribution, rather than a parametric reference distribution, such as Student's $t$ or the Gaussian.
+Such a test is exact, meaning that it controls the type I error rate at the pre-specified level, even in finite samples, whereas parametric hypothesis tests based on asymptotic approximations do not always guarantee good finite sample properties.
+Permutation tests achieve this by operating conditional on the observed sample; the resulting inference applies to the sample at hand, but does not necessarily generalize to other groups of individuals.
+In the past, parametric tests were necessary because asymptotic approximations were the only computationally feasible way to estimate distributions. 
+Now, computational power is no longer a barrier to finding exact (or exact to pre-specified precision) randomization distributions.
+In most cases, a randomization test is preferable to a parametric one:
+``[a] corresponding parametric test is valid only to the extent that it results in the same statistical decision [as the randomization test]'' (\cite{bradley_distribution_1968}).
+
+There is no hard and fast rule describing the rate at which parametric tests approach the exact permutation solution, as they are both highly dependent on the particular data observed.
+On the one hand, some argue that violations of parametric test assumptions necessitate the use of permutation methods.
+\citet{ludbrook_why_1998} point out that medical trials rarely follow the population sampling model implicit in parametric methods.
+However, the parametric and nonparametric tests seem to perform similarly in most simulations, even when data violate the assumptions of the parametric method.
+Medical trials studying pain or perceived severity scores often use Likert scale data, which is discrete and does not match the normality assumptions of parametric tests.
+However, \citet{winter_five-point_2010} found that the two sample $t$ test and Mann-Whitney test had comparable Type I and II error rates for five-point Likert scale data, suggesting that the violation of normality does not entirely invalidate the parametric test.
+\todo{what other linear model comparisons have people done?}
+Most similar to our question of study, \citet{vickers_parametric_2005} compared the parametric ANCOVA to the Mann-Whitney rank test in the context of randomized experiments, finding that except in extreme situations, ANCOVA outperformed the nonparametric test.
+
+In this paper, we review hypothesis tests for a treatment effect which incorporate covariate adjustment to increase power.  
+We focus on ANCOVA and its permutation counterparts, comparing their performance in different scenarios, and illustrating their application with a clinical dataset.
+We find in simulations that even when assumptions are not satisfied, the parametric and permutation tests have comparable power to detect a treatment effect.
+We apply each of the methods to data from a clinical trial seeking to compare the performance of two treatments for gastroesophageal reflux disease (GERD).
+We conclude by discussing implications of these results for practitioners.
 
@@
 
-\subsection{Problem and notation}
-Suppose we randomly assign two treatments to a group of individuals
-Let $Z$ be treatment assignment, 0 or 1, and we are interested in comparing the relative effectiveness of treatment 1 to treatment 0.
-$Y$ is the outcome of interest observed for all individuals.
-$X$ is a pretreatment covariate that we observe and is related to the outcome (may or may not be causal).  
-Without loss of generality, we suppose that $X$ is univariate, but we could easily extend all of these results to multivariate $X$.
+\subsection{Problem and notation \todo{what do others call this?}}
+Suppose we randomly assign two treatments labelled $0$ and $1$ to a group of individuals.
+We are interested in comparing the relative effectiveness of treatment 1 to treatment 0.
+Let $Z$ indicate treatment assignment, 0 or 1, 
+$Y$ denote the observed outcome of interest,
+and $X$ be a pretreatment covariate that is observed and associated with the outcome.
+For expository clarity, we suppose that $X$ is univariate, but all results are easily extended to the case when $X$ is multivariate.
+Suppose there is a categorical pretreatment variable with $J$ levels used to stratify individuals.
 
-Suppose there is another nuisance variable $\mathcal{J}$. 
-$\mathcal{J}$ is a pretreatment variable used to stratify individuals. 
-There may be some unobservable heterogeneity in outcomes across levels of $\mathcal{J}$.
-
-We observe $\{Y_{ij}(Z_{ij}), Z_{ij}, X_{ij}\}$ for units $i = 1, \dots, n_j$, in strata $j = 1, \dots, J$.
-All units have two potential outcomes, $Y_{ij}(1)$ and $Y_{ij}(0)$, representing their responses to treatments $1$ and $0$, respectively.
-We can never observe both; random assignment of treatment reveals one.
+We observe $\{Y_{ij}, Z_{ij}, X_{ij}\}$ for individuals $i = 1, \dots, n_j$, in strata $j = 1, \dots, J$.
+All individuals have two potential outcomes, $Y_{ij}(1)$ and $Y_{ij}(0)$, representing their responses to treatments $1$ and $0$, respectively.
+We can never observe both; random assignment of treatment reveals $Y_{ij} = Z_{ij}Y_{ij}(1) + (1-Z_{ij})Y_{ij}(0)$.
 
 We are interested in the differences in potential outcomes $Y_{ij}(1) - Y_{ij}(0)$.
-This difference represents the effect of treatment for individual $i$ in group $j$: it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.
-Depending on the method of analysis one chooses, different functions of these differences will be considered.
-Any of these quantities may be of clinical interest; it is a matter of philosophy which one prefers.
-A typical problem is to estimate one of these averages.
-Instead, we will focus on hypothesis testing.
-Below, we outline a few approaches to studying the average treatment effect, explain the assumptions that are necessary, and
+This difference represents the effect of treatment for individual $i$ in group $j$: it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.\footnote{
+Depending on the method of analysis one chooses, various functions of these differences will be considered.
+Any of these quantities may be of clinical interest; it is a matter of philosophy which one prefers.}
+A typical problem is to estimate the average of differences $Y_{ij}(1) - Y_{ij}(0)$.
+Instead of estimation, we will focus on hypothesis testing for whether these differences are nonzero.
 
-\subsection{Parametric ANOVA}
+\subsection{Parametric ANCOVA}
 
@@
 
-where $\alpha_j$ is a fixed effect for stratum $j$, $\beta$ is the coefficient for the pretreatment covariate,
+\noindent where $\alpha_j$ is a fixed effect for stratum $j$, $\beta$ is the coefficient for the pretreatment covariate,
 $\gamma$ is the coefficient for treatment,
@@
 The parameter of interest is $\gamma$, which represents the average treatment effect in the sample. 
+If we believe the linear model is correctly specified, it asserts that for each individual, $Y(1) = Y(0) + \gamma$.
 We would like to test the null hypothesis $H_0: \gamma = 0$ against
@@
 
-In the normal theory ANOVA, we make the following assumptions:
+To carry out the standard parametric hypothesis test, we make the following assumptions (\cite{freedman_statistical_2005}):
+
 \begin{enumerate}
-\item the $\eps$ are normally distributed with mean $0$ and common variance $\sigma^2$
-\item $\eps \independent X$
-\item \textcolor{red}{ what else? }
+\item \textbf{Linearity:} The data $Y$ are related to $X$ and $Z$ linearly.
+\item \textbf{Constant slopes:} Stratum only affects the intercept $\alpha_j$, not the slopes $\beta$ and $\gamma$.
+\item \textbf{IID Errors:} The $\eps_{ij}$ are independent and identically distributed with mean $0$ and variance $\sigma^2$.
+\item \textbf{Independence:} If $X$ is random, $\eps \independent X$.\footnote{$\independent$ means statistical independence; $\perp$ means orthogonal.}
+\item \textbf{Normality:} The errors are normally distributed.
 \end{enumerate}
 
-The model is fit using least squares.
-A significance test is done for the estimated coefficient $\hat{\gamma}$, using the $t$-distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
-The test statistic is
-$$ t = \frac{\hat{\gamma}}{\sqrt{ \hat{\sigma}_{\hat{\gamma}}^2}},$$
-where $\hat{\sigma}_{\hat{\gamma}}^2$ is the estimated standard error of $\hat{\gamma}$, obtained from the diagonal element of the inverse covariance matrix of the observed data which corresponds to the treatment.
-\todo{cite freedman linear models textbook}
+The coefficients are estimated using least squares.\footnote{
+Note, only Assumptions (1) and (2) are needed to guarantee the existence of a solution.
+The solution will be unique as long as there is no linear relationship between $X$, $Z$, and stratum membership.
+}
+Under the null hypothesis, the test statistic 
+$$ T = \frac{\hat{\gamma}}{\sqrt{ \hat{\sigma}_{\hat{\gamma}}^2}},$$
+where $\hat{\sigma}_{\hat{\gamma}}^2$ is the estimated standard error of $\hat{\gamma}$,\footnote{
+$\hat{\sigma}_{\hat{\gamma}}^2$ is estimated by first computing the residual standard error $\hat{\sigma}^2$,
+then taking the diagonal element corresponding to the treatment variable in the inverse covariance matrix of covariates.
+}
+follows the Student $t$ distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
+The p-value for this hypothesis test is the probability that a value drawn from this $t$ distribution is larger in magnitude than $T$.
 
 \todo{Add a paragraph about why this might go wrong, perhaps some citations. All depends on estimating the error variance correctly, because $t$ distribution is the ratio of a normal and a chi-square}
-\bit
-\item This ANOVA model allows for fixed stratum effects, but they are only additive in the outcome.
+
+The ANCOVA model allows for fixed stratum effects, but they are only additive in the outcome.  
 The model does not account for differences in the treatment effect across strata.  
-In other words, the estimated linear model is not a good estimate of the true data-generating process.
+The estimated linear model may not accurately capture details of the true data-generating process.  
 If effects are not constant across strata, then the coefficient $\gamma$ may be attenuated towards 0.  
 A test of the null hypothesis that $\gamma = 0$ will be valid, but will not reflect the true magnitude of treatment effects among individuals.
-\item \textcolor{red}{ what else? }
-\eit
 
@@
 \bit
-\item Cite a few here, 
+\item Cite a few here, \todo{all the Anderson papers}
 \item explain why Freedman-Lane is best. 
@@
 \bit
-\item This paper adds to the literature comparing parametric and nonparametric tests
-\item In randomized control trials, there is no substantial difference between parametric and nonparametric approaches.
+\item This paper adds to the literature comparing parametric and nonparametric tests.
+Our results match those of \citet{vickers_parametric_2005}, where he found that ANCOVA performed the same or better than nonparametric tests.
+%In randomized control trials, there is no substantial difference between parametric and nonparametric approaches.
 They are asymptotically equivalent to begin with (find a citation).
@@
 \item Results about differenced vs raw outcomes. This isn't new but our simulations and empirical dataset confirm it in the context of permutation tests
-\item 
+\item In an ideal case, one knows all covariates related to the outcome and fits a fully saturated linear model (including all covariates and their interactions), the coefficient for the treatment is an asymptotically consistent estimate of the average treatment effect (\cite{lin_agnostic_2013}).
+\item Permutation tests don't come free of assumptions, though.
+\citet{romano_behavior_1990} warns against using permutation tests naively, if items are not truly exchangeable.
+For instance, he points to the case where observations have unequal variances.
+\citet{boik_fisherpitman_1987} illustrates this phenomenon, comparing the normal theory $F$ test to its permutation counterpart and demonstrating by simulation that the latter has larger than nominal level.
 \eit

### 2016-12-19 ms/fig/imbalanced_simulation_power_summary.tex
Add simulations for imbalanced design; update tables and figs; edits to results

\begin{table}[ht]
\centering
\begin{tabular}{|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}}
  \hline
Treatment & Errors & ANCOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
  \hline
No Effect & Gaussian & 0.046 & 0.031 & 0.028 & 0.042 & 0.045 \\ 
  Constant Additive Effect & Gaussian & 0.840 & 0.490 & 0.810 & 0.870 & 0.860 \\ 
  Single Stratum Effect & Gaussian & 0.120 & 0.050 & 0.090 & 0.100 & 0.120 \\ 
   \hline
No Effect & Heavy-tailed & 0.034 & 0.054 & 0.055 & 0.047 & 0.042 \\ 
  Constant Additive Effect & Heavy-tailed & 0.230 & 0.200 & 0.270 & 0.260 & 0.270 \\ 
  Single Stratum Effect & Heavy-tailed & 0.070 & 0.080 & 0.090 & 0.090 & 0.080 \\ 
   \hline
\end{tabular}
\caption{Empirical power at level $0.05$ for Gaussian simulated data with imbalanced treatment groups} 
\label{tab:imbalanced_power}
\end{table}

### 2016-12-19 ms/fig/normal_simulation_power_summary.tex
Add simulations for imbalanced design; update tables and figs; edits to results

@@
   \hline
-Treatment & Errors & ANOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
+Treatment & Errors & ANCOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
   \hline
-Constant Additive Effect & Gaussian & 0.89 & 0.55 & 0.86 & 0.89 & 0.89 \\ 
-  Single Stratum Effect & Gaussian & 0.16 & 0.09 & 0.16 & 0.16 & 0.17 \\ 
-  Constant Additive Effect & Heavy-tailed & 0.35 & 0.24 & 0.38 & 0.40 & 0.39 \\ 
-  Single Stratum Effect & Heavy-tailed & 0.07 & 0.08 & 0.08 & 0.08 & 0.08 \\ 
+No Effect & Gaussian & 0.058 & 0.042 & 0.038 & 0.062 & 0.057 \\ 
+  Constant Additive Effect & Gaussian & 0.886 & 0.545 & 0.858 & 0.891 & 0.892 \\ 
+  Single Stratum Effect & Gaussian & 0.159 & 0.089 & 0.160 & 0.162 & 0.170 \\ 
+   \hline
+No Effect & Heavy-tailed & 0.032 & 0.040 & 0.045 & 0.042 & 0.040 \\ 
+  Constant Additive Effect & Heavy-tailed & 0.355 & 0.240 & 0.378 & 0.400 & 0.393 \\ 
+  Single Stratum Effect & Heavy-tailed & 0.066 & 0.080 & 0.082 & 0.076 & 0.080 \\ 
    \hline

### 2016-12-19 ms/fig/skewed_simulation_power_summary.tex
Add simulations for imbalanced design; update tables and figs; edits to results

@@
   \hline
-Treatment & Design & ANOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
+Treatment & Design & ANCOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
   \hline
-Constant Additive Effect & Balanced & 0.97 & 0.14 & 0.85 & 0.97 & 0.98 \\ 
-  Single Stratum Effect & Balanced & 0.42 & 0.05 & 0.28 & 0.43 & 0.42 \\ 
-  Constant Additive Effect & Imbalanced & 0.97 & 0.15 & 0.86 & 0.97 & 0.97 \\ 
-  Single Stratum Effect & Imbalanced & 0.19 & 0.07 & 0.10 & 0.19 & 0.19 \\ 
+No Effect & Balanced & 0.036 & 0.025 & 0.021 & 0.035 & 0.037 \\ 
+  Constant Additive Effect & Balanced & 0.974 & 0.143 & 0.853 & 0.972 & 0.977 \\ 
+  Single Stratum Effect & Balanced & 0.422 & 0.050 & 0.275 & 0.429 & 0.424 \\ 
+   \hline
+No Effect & Imbalanced & 0.038 & 0.024 & 0.019 & 0.037 & 0.039 \\ 
+  Constant Additive Effect & Imbalanced & 0.971 & 0.147 & 0.860 & 0.970 & 0.972 \\ 
+  Single Stratum Effect & Imbalanced & 0.190 & 0.070 & 0.100 & 0.190 & 0.190 \\ 
    \hline

### 2016-12-19 ms/ms.tex
Add simulations for imbalanced design; update tables and figs; edits to results

@@
 Furthermore, the hypothesis test implicitly assumes that the data were sampled at random from some underlying population, when in fact,
-recruitment of patients to medical experiments is rarely done this way \todo{cite}.
+medical experimenters rarely recruit patients this way.
 
-An alternative approach is permutation testing. \todo{citations: fisher 1935, pitman 1937 and 1938}
+Permutation testing is an alternate approach (\cite{fisher_design_1935, pitman_significance_1937,pitman_significance_1938}).
 Deliberate randomization induces a distribution for any test statistic under the null hypothesis that treatment has no effect on the outcome:
@@
 
-$$Y_{ij} = \alpha_j + \beta X_{ij} + \gamma Z_{ij} + \eps_{ij}$$
+\begin{equation}
+Y_{ij} = \alpha_j + \beta X_{ij} + \gamma Z_{ij} + \eps_{ij}
+\end{equation}
 
@@
 and $\eps_{ij}$ is an error term.
-The parameter of interest is $\gamma$, which represents the average treatment effect in the sample. 
-If we believe the linear model is correctly specified, it asserts that for each individual, $Y(1) = Y(0) + \gamma$.
+The parameter of interest is $\gamma$. 
+If we believe the linear model is the true data-generating process, it asserts that for each individual, $Y(1) = Y(0) + \gamma$.
+However, we needn't take this perspective for $\gamma$ to be a useful quantity; it represents the average treatment effect, holding the other variables fixed.
 We would like to test the null hypothesis $H_0: \gamma = 0$ against
@@
 
-To compare the performance of these tests, we simulated data coming from a randomized experiment, applied the tests to the data, and compared their empirical power over repeated random treatment assignments and random errors.
+To compare the performance of these tests, we simulated data coming from a randomized experiment, using several different data-generating processes.
+We applied the tests to the data and compared their empirical power over repeated random treatment assignments and random errors.
 We compared the following five tests:
-the $t$-statistic from the parametric ANOVA,
+the $t$-statistic from the parametric ANCOVA,
 a stratified permutation test using the difference in mean outcomes\footnote{ We considered using the difference in means within each stratum, aggregated by taking the sum of their absolute values over stratum.
@@
 
-$$Y_{ij1} =\beta_0Y_{ij0} + \beta_{j} + \gamma_j Z_{ij} + \varepsilon_{ij}$$
+\begin{equation}\label{eqn:dgp}
+Y_{ij1} =\alpha_j + \beta_0Y_{ij0} + \gamma_j Z_{ij} + \varepsilon_{ij}
+\end{equation}
 
-for individuals $i = 1, \dots, n_j$, $j = 1, 2, 3$.
+\noindent for individuals $i = 1, \dots, n_j$, $j = 1, 2, 3$.
+$\alpha_j$ is the mean effect of being in stratum $j$, 
 $\beta_0$ is the coefficient for the baseline measurement $Y_{ij0}$, 
-$\beta_j$ is the mean effect of being in stratum $j$, 
 $Z_{ij}$ is the treatment level, 
@@
 This is a simplistic case of heterogeneous treatment effects.
+The standard ANCOVA model does not account for this scenario.
 \end{itemize}
 
-We draw the baseline measurements once and condition on observing $\{ Y_{ij0}\}_{i = 1,\dots,16, j = 1,\dots, 3}$.
-Then, for a large number of trials, we randomly assign treatment within strata, generate errors, and construct new outcomes $Y_{ij1}$.
+\noindent We draw the baseline measurements once, treating them as fixed, and condition on observing $\{ Y_{ij0} : {i = 1,\dots,16, j = 1,\dots, 3}\}$.
+Then, for 1000 trials, we randomly assign treatment to half of the individuals within each stratum, generate random errors, and construct new outcomes $Y_{ij1}$ using Equation~\ref{eqn:dgp}.
 We conduct all five tests on this new data, obtaining a two-sided p-value for each.
@@
 In our first set of simulations, we assume that the baseline measurements $Y_{ij0}$ are standard normally distributed and the pairs $(Y_{ij0}, \varepsilon_{ij})$ are independent across $i$ and $j$.
-We will assume that $\beta_0 = 1$.
+We will assume that $\beta_0 = 1$ and let the treatment effect be $\gamma = 1$
 For each of the two simulation designs that vary the treatment effect, we use two distributions of $\varepsilon$.
@@
 
-Figure~\ref{fig:normal_sim_power} shows the estimated power curves in these four designs using a treatment effect of $\gamma = 1$.
+Figure~\ref{fig:normal_sim_power} shows the estimated power curves in these four designs.
 The best case is when the errors are Gaussian and the treatment effect is constant across strata, while the worst case is when the errors are heavy-tailed and the treatment effect only appears in one stratum.
-\todo{it makes sense that this happens as assumptions are broken in other cases}
+Intuitively, it makes sense that power decreases relative to the Gaussian, constant treatment effect case, as we increasingly obscure the treatment effect by modifying the design.
 A consistent pattern appears in each case: the stratified permutation test has the lowest power, while the other four tests tend to have very similar power.
-Table~\ref{tab:normal_power} displays the value of these power curves at level $0.05$.
+
+Table~\ref{tab:normal_power} displays the size of the test and the power at level $0.05$.
+The three tests based on linear models have slightly higher than nominal level, though the margin is within two standard errors of $0.05$.\footnote{
+The number of tests rejected under the null in 1000 trials has a binomial distribution.
+If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.}
 These numbers show that actually, for $\alpha=0.05$, the discrepancy between the stratified permutation test and the others is not so large when the effect is isolated in a single stratum.
-In this case, the permutation test may have more power than the parametric ANOVA at significance levels that are used in practice.
-
-\begin{figure}
+In this case, the permutation test may have slightly higher power than the parametric ANCOVA at small significance levels like those used in practice.
+\begin{figure}[h]
 \centering
@@
 
-In our second set of simulations, we assume that the baseline measurements $Y_{ij0}$ are discrete and skewed.
+In our second set of simulations, we used the same design as the first simulations, but modified the sizes of the treatment groups.
+Within each stratum, 4 individuals received one treatment and 12 received the second treatment.
+This mimics real-world experiments, where the more expensive treatment is administered less frequently than the other.
+It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exaccerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
+Here, the data have the same variance, so the test size should not be a problem, but power may suffer.\footnote{
+See the supplementary simulation files for a treatment of heterogeneous variances.}
+We omit a plot of the power curves from these simulations, as the patterns appear similar to those in Figure~\ref{fig:normal_sim_power}.
+Table~\ref{tab:imbalanced_power} displays the value of these power curves and the size of the test at level $0.05$.
+In this case, the permutation tests (except for the simple stratified permutation) have slightly higher power than the ANCOVA, but the difference is not substantial.
+%\begin{figure}
+%\centering
+%\includegraphics[width = \textwidth]{fig/imbalanced_simulation_power}
+%\caption{Empirical power curves for the Gaussian simulated data with imbalanced group sizes}
+%\label{fig:imbalanced_sim_power}
+%\end{figure}
+\begin{center}
+\input{fig/imbalanced_simulation_power_summary.tex}
+\end{center}
+
+In our third set of simulations, we assume that the baseline measurements $Y_{ij0}$ are discrete and skewed.
 This is representative of some survey data, where individuals are asked to rate their experience on a scale from 1 to 10, for instance.
-Here, we generate the baseline measures using independent draws from a Poisson distribution with mean 4, and censoring values larger than 10, instead setting them to 10.
-We condition on these observed baseline measures once we have generated them once.
-Then, for each of the large number of trials, we randomly assign treatment within strata and generate errors taking on values $0.5$ and $-0.5$ with equal probability.
-We construct $Y_{ij1}$ using the linear equation above.
+We generate the baseline measures using independent draws from a Poisson distribution with mean 4, and censoring values larger than 10, instead setting them to 10.
+We condition on these observed baseline measures after we have generated them once.
+Then, for each 1000 trials, we randomly assign treatment to half of the individuals in each stratum and generate errors taking on values $0.5$ and $-0.5$ with equal probability.
+We construct $Y_{ij1}$ using Equation~\ref{eqn:dgp} with a treatment effect of $\gamma = 0.5$.
 However we don't observe $Y_{ij1}$ but instead $\tilde{Y}_{ij1}$, which is defined as
@@
 
-The observed outcomes $\tilde{Y}_{ij1}$ that we test are also discrete.
+The observed outcomes $\tilde{Y}_{ij1}$, the baseline measures $Y_{ij0}$, and the errors are discrete.
 The assumption of normality is clearly violated here.
 
-In these simulations, we examine the effect of sample size.
-First, we consider balanced samples, where each stratum contains 16 individuals, 8 of whom receive one treatment and the other 8 of whom receive the other.
-Next, we consider imbalanced samples, where the smallest stratum has only 8 individuals, the next has 16, and the largest has 24.
+We examine the effect of stratum size, rather than error distribution, in this set of simulations.
+First, we consider balanced samples as before, where each stratum contains 16 individuals.
+Next, we consider imbalanced strata, where the smallest stratum has only 8 individuals, the next has 16, and the largest has 24.
 We again consider the four combinations of treatment effects and sample sizes.
@@
 
-Figure~\ref{fig:skewed_sim_power} shows the estimated power curves in these four designs using a treatment effect of $\gamma = 0.5$.
+Figure~\ref{fig:skewed_sim_power} shows the estimated power curves in these four designs.
 When the treatment effect is constant, sample sizes within each stratum don't matter; the power curves look the same up to chance variation.
-When the treatment effect only occurs in one stratum, there is a substantial loss in power due to halving the number of individuals in the stratum where the effect is present.
+When the treatment effect only occurs in one stratum, there is a substantial loss in power because the effect only appears in the smallest stratum.
 In practice, one does not usually know a priori which individuals will be affected by treatment (otherwise, those unaffected by treatment would be excluded from the study altogether),
-so this result suggests using caution when stratifying.
+This result suggests that when stratifying, one must be cautious not to stratify too finely, otherwise treatment effects may be indistinguishable from random variability.
+
 As in the first set of simulations, the stratified permutation test using the raw outcomes has the lowest power of the five tests.
 Now, however, the stratified permutation test using the difference between outcome and baseline also has lower power than the three using linear models.
-Presumably, this is because the differences are also discrete and lie in a small range of values, so the test statistic does not vary greatly across permutations.
+Presumably, this is because the differences $\tilde{Y}_{ij1} - Y_{ij0}$ are also discrete and lie in a small range of values, so the test statistic does not vary greatly across permutations.
 Table~\ref{tab:skewed_power} confirms this: power for the three linear model methods is high when the treatment is constant across strata, and power for all of the methods suffers when the effect is isolated in a single stratum.
+All of the tests have lower than nominal level.
 
@@
 It makes sense that in all of these simulations, the permutation test using the difference in outcome and baseline as the response measure is more powerful than the permutation test using the raw outcomes.
-In additional simulations\footnote{see supplementary files on GitHub}, we changed the Gaussian model so that the outcome and baseline had correlation of $0.25$.
+In additional simulations\footnote{see supplementary files on GitHub}, we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had correlation of $0.25$.
 In this case, the result was reversed: the differenced test had low power while the test using the raw outcomes had a power curve closer to the tests that adjust for covariates using a linear model.
-This phenomenon has been studied before \todo{cite}, and it is recommended to use the difference in outcome and baseline only when their correlation is larger than $0.5$ (\cite{frison_repeated_1992}).
+This phenomenon has been studied before by \citet{frison_repeated_1992}, who recommended to use the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
 
@@
 \citet{boik_fisherpitman_1987} illustrates this phenomenon, comparing the normal theory $F$ test to its permutation counterpart and demonstrating by simulation that the latter has larger than nominal level.
+\item \citet{gail_tests_1988} propose a randomization test based on residuals, using not a linear model but an exponential family model to construct residuals, and find that omitting relevant covariates leads to tests with higher than nominal level.
+\todo{perhaps this is something that one should be mindful of when using these linear model based tests.}
 \eit

### 2016-12-19 simulations/simulation.Rmd
Add simulations for imbalanced design; update tables and figs; edits to results

@@
 ---
-title: "ANOVA Comparison Simulations: Gaussian Data"
+title: "ANCOVA Comparison Simulations: Gaussian Data"
 author: "Kellie Ottoboni"
@@
 With these two designs, we vary the distribution of $\varepsilon$.
-In the first case, we use $\varepsilon \sim N(0, 1)$ to mimic the usual ANOVA assumptions.
+In the first case, we use $\varepsilon \sim N(0, 1)$ to mimic the usual ANCOVA assumptions.
 In the second case, $\varepsilon \sim t(2)$ so the errors are heavy-tailed.
@@
 
-* ANOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for stratum.
+* ANCOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for stratum.
 * Stratified permutation: we permute treatment assignment within stratum, then take the difference in means between treated and control outcomes $Y_1$
@@
   # Returns: a vector of p-values
-  # first element is the p-value from the ANOVA
+  # first element is the p-value from the ANCOVA
   # second element is the p-value from the stratified two-sample permutation test
@@
 
-  # ANOVA
+  # ANCOVA
   modelfit <- lm(Y1 ~ Y0 + Z + factor(stratumID), data = dataset)
@@
 
-  return(c("ANOVA" = anova_pvalue, 
+  return(c("ANCOVA" = anova_pvalue, 
            "Stratified Permutation" = perm_pvalue, 
@@
   as.data.frame() %>%
-  select(ANOVA, strat = starts_with("Stratified")) %>%
-  ggplot(aes(x = ANOVA, y = strat)) + 
+  select(ANCOVA, strat = starts_with("Stratified")) %>%
+  ggplot(aes(x = ANCOVA, y = strat)) + 
   geom_point() +
@@
 
-```{r design_null, fig.height = 4}
+```{r design_null_gaussian, fig.height = 4}
 set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
@@
 tmp <- generate_simulated_data(gamma = 0, effect = "same effect", errors = "normal")
-design0_pvalues <- replicate(100, {
+design0_pvalues <- replicate(1000, {
   tmp$epsilon <- rnorm(nrow(tmp))
@@
 design0_pvalues <- t(design0_pvalues)
-colnames(design0_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design0_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design0_power <- apply(design0_pvalues, 2, compute_power)
@@
 
+```{r design_null_heavy, fig.height = 4}
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+
+tmp <- generate_simulated_data(gamma = 0, effect = "same effect", errors = "heavy")
+design00_pvalues <- replicate(1000, {
+  tmp$epsilon <- rt(nrow(tmp), df = 2)
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$epsilon
+  generate_simulated_pvalues(tmp)
+})
+design00_pvalues <- t(design00_pvalues)
+colnames(design00_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+design00_power <- apply(design00_pvalues, 2, compute_power)
+```
+
+```{r design_null_heavy_plots, fig.height = 4, fig.align="center"}
+plot_power_curves(design00_power, "No Effect, Heavy-tailed Errors")
+
+plot_pvalue_hist(design00_pvalues, "No Effect, Heavy-tailed Errors") 
+```
+
 ## Design 1: Constant additive effect, normal errors
 
-There is no discernable difference in power between the ANOVA, the differenced stratified permutation test, the LM permutations, or the Freedman-Lane test.
+There is no discernable difference in power between the ANCOVA, the differenced stratified permutation test, the LM permutations, or the Freedman-Lane test.
 However, the simple stratified permutation test of $Y_1$ has substantially less power than the other four.
@@
 design1_pvalues <- t(design1_pvalues)
-colnames(design1_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design1_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design1_power <- apply(design1_pvalues, 2, compute_power)
@@
 design2_pvalues <- t(design2_pvalues)
-colnames(design2_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design2_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design2_power <- apply(design2_pvalues, 2, compute_power)
@@
 design3_pvalues <- t(design3_pvalues)
-colnames(design3_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design3_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design3_power <- apply(design3_pvalues, 2, compute_power)
@@
 design4_pvalues <- t(design4_pvalues)
-colnames(design4_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design4_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design4_power <- apply(design4_pvalues, 2, compute_power)
@@
 The power curves for this design look similar to those from Design 1 with one important difference: 
-the stratified permutation test has power very close to ANOVA and the two linear model tests,
+the stratified permutation test has power very close to ANCOVA and the two linear model tests,
 while the differenced permutation test has lower power.
@@
 design5_pvalues <- t(design5_pvalues)
-colnames(design5_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design5_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design5_power <- apply(design5_pvalues, 2, compute_power)
@@
 stratum 1 has 8 patients, stratum 2 has 16 patients, and stratum 3 has 24 patients.
-This imbalance does not violate any assumptions of the ANOVA.
+This imbalance does not violate any assumptions of the ANCOVA.
 However, it may reduce power if the effect is concentrated in strata with fewer patients.
@@
 design6_pvalues <- t(design6_pvalues)
-colnames(design6_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design6_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design6_power <- apply(design6_pvalues, 2, compute_power)
@@
 design7_pvalues <- t(design7_pvalues)
-colnames(design7_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design7_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design7_power <- apply(design7_pvalues, 2, compute_power)
@@
 
-One assumption of ANOVA is that errors are homoskedastic: they have the same variance across groups.
+One assumption of ANCOVA is that errors are homoskedastic: they have the same variance across groups.
 Suppose that this does not hold in our data.
@@
 design8_pvalues <- t(design8_pvalues)
-colnames(design8_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design8_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design8_power <- apply(design8_pvalues, 2, compute_power)
@@
 design9_pvalues <- t(design9_pvalues)
-colnames(design9_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+colnames(design9_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
 design9_power <- apply(design9_pvalues, 2, compute_power)
@@
 ```{r final_summary, fig.align = "center", fig.width = 12}
-#plots4 <- lapply(list(p1, p2, p4, p5), function(pl) pl + theme(legend.position = "none") + ggtitle(""))
-#do.call(grid.arrange, plots4)
-
 powers <- list(design1_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design1_power)), "Errors" = rep("Gaussian", nrow(design1_power))), 
@@
                design3_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design3_power)), "Errors" = rep("Heavy-tailed", nrow(design3_power))), 
-               design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design4_power)), "Errors" = rep("Heavy-tailed", nrow(design4_power))))
+               design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design4_power)), "Errors" = rep("Heavy-tailed", nrow(design4_power))),
+               design0_power %>% as.data.frame() %>% mutate("Treatment" = rep("No Effect", nrow(design0_power)), "Errors" = rep("Gaussian", nrow(design0_power))),
+              design00_power %>% as.data.frame() %>% mutate("Treatment" = rep("No Effect", nrow(design00_power)), "Errors" = rep("Heavy-tailed", nrow(design00_power))))
 
 all_power_curves <- do.call(rbind, powers)
-twobytwo <- melt(all_power_curves, id.vars = c("Treatment", "Errors")) %>% 
+twobytwo <- all_power_curves %>%
+  filter(Treatment != "No Effect") %>%
+  melt(id.vars = c("Treatment", "Errors")) %>% 
   mutate("pvalue" = rep((1: 100)/100, 5*4)) %>%
@@
 summary05 <- t(sapply(powers, function(x) x[5, c(6, 7, 1:5)]))
-summarytab <- xtable(summary05, align = "rp{1.15in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}",
+summary05 <- summary05[c(5, 1, 2, 6, 3, 4), ]
+summarytab <- xtable(summary05, digits = 3, align = "rp{1.15in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}",
        caption = "Empirical power at level $0.05$ for Gaussian simulated data",
        label = "tab:normal_power") 
-print(summarytab, type="latex", file="../ms/fig/normal_simulation_power_summary.tex")
+print(summarytab, hline.after = c(-1, 0, 3, nrow(summarytab)), type="latex", file="../ms/fig/normal_simulation_power_summary.tex")
 summarytab

### 2016-12-19 simulations/simulation_imbalanced.Rmd
Add simulations for imbalanced design; update tables and figs; edits to results

--- 2016-12-19 simulations/simulation.Rmd
@@
 ---
-title: "ANCOVA Comparison Simulations: Gaussian Data"
+title: "ANCOVA Comparison Simulations: Imbalanced Design"
 author: "Kellie Ottoboni"
@@
 Suppose there are three strata with $\beta_1 = 1, \beta_2 = 1.5,$ and $\beta_3 = 2$.
-Assume that there are 16 individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
+Assume that there are 16 individuals per stratum and treatment assignment is \textbf{imbalanced}: 12 people receive each treatment 0 and 4 people receive treatment 1.
 
@@
   Y0 <- rnorm(N)
-  Z <- rep(0:1, length.out = N)
+  Z <- do.call(c, lapply(n, function(x) rep(0:1, times = c(x*(3/4), x*(1/4)))))
   stratum_effect <- rep(beta, times = n)
@@
 ```
-
 ## Design 1: Constant additive effect, normal errors
@@
 tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
-design1_pvalues <- replicate(1000, {
+design1_pvalues <- replicate(100, {
   tmp$epsilon <- rnorm(nrow(tmp))
@@
 tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "normal")
-design2_pvalues <- replicate(1000, {
+design2_pvalues <- replicate(100, {
   tmp$epsilon <- rnorm(nrow(tmp))
@@
 tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "heavy")
-design3_pvalues <- replicate(1000, {
+design3_pvalues <- replicate(100, {
   tmp$epsilon <- rt(nrow(tmp), df = 2)
@@
 tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "heavy")
-design4_pvalues <- replicate(1000, {
+design4_pvalues <- replicate(100, {
   tmp$epsilon <- rt(nrow(tmp), df = 2)
@@
 plot_pvalue_hist(design4_pvalues, "Effect of 1, Heavy-tailed Errors")
-```
-
-## Design 5: Unpredictive baseline
-
-Suppose that the baseline measure is not strongly predictive of the outcome.
-Then, we'd expect that controlling for baseline will not improve the power of the test, and may even make it worse.
-Suppose that in the linear data-generating process above, we have $\beta_0 = 0.25$, meaning that $Y_1$ and $Y_0$ have a correlation of $0.25$.
-
-The power curves for this design look similar to those from Design 1 with one important difference: 
-the stratified permutation test has power very close to ANCOVA and the two linear model tests,
-while the differenced permutation test has lower power.
-This suggests that one should be careful when incorporating control variables;
-naively taking the difference $Y_1 - Y_0$ does not capture the correct relationship between baseline and outcome.
-
-```{r design5, fig.height = 4}
-set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
-
-tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
-design5_pvalues <- replicate(100, {
-  tmp$epsilon <- rnorm(nrow(tmp))
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <- 0.25*tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
-  generate_simulated_pvalues(tmp)
-})
-
-design5_pvalues <- t(design5_pvalues)
-colnames(design5_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
-design5_power <- apply(design5_pvalues, 2, compute_power)
-```
-
-```{r design5_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design5_power, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
-
-plot_pvalue_hist(design5_pvalues, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
-```
-
-# Imbalanced designs
-
-Suppose that instead of having 16 individuals per stratum, we distribute them unequally across strata:
-stratum 1 has 8 patients, stratum 2 has 16 patients, and stratum 3 has 24 patients.
-This imbalance does not violate any assumptions of the ANCOVA.
-However, it may reduce power if the effect is concentrated in strata with fewer patients.
-
-## Design 6: Constant additive effect, normal errors
-
-The power curves here are very similar to those in Design 1.
-This result is expected, since the effect is still present among all patients who received treatment.
-
-```{r design6, fig.height = 4}
-set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
-
-tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal", n = c(8, 16, 24))
-design6_pvalues <- replicate(100, {
-  tmp$epsilon <- rnorm(nrow(tmp))
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
-  generate_simulated_pvalues(tmp)
-})
-
-design6_pvalues <- t(design6_pvalues)
-colnames(design6_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
-design6_power <- apply(design6_pvalues, 2, compute_power)
-```
-
-```{r design6_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design6_power, "Constant Additive Effect of 1, Normal Errors")
-
-plot_pvalue_hist(design6_pvalues, "Constant Additive Effect of 1, Normal Errors")
-```
-
-## Design 7: Single stratum effect, normal errors
-
-There is very low power to detect the effect here, since in the simulated data, the stratum with only 8 patients was the only stratum with a nonzero effect of treatment.
-The treatment effect is on the same order of magnitude as the error variance.
-
-```{r design7, fig.height = 4}
-set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
-
-tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "normal", n = c(8, 16, 24))
-design7_pvalues <- replicate(100, {
-  tmp$epsilon <- rnorm(nrow(tmp))
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
-  generate_simulated_pvalues(tmp)
-})
-design7_pvalues <- t(design7_pvalues)
-colnames(design7_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
-design7_power <- apply(design7_pvalues, 2, compute_power)
-```
-
-```{r design7_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design7_power, "Effect of 1 at a Single stratum, Normal Errors")
-
-plot_pvalue_hist(design7_pvalues, "Effect of 1, Normal Errors")
-```
-
-# Heteroskedastic errors
-
-One assumption of ANCOVA is that errors are homoskedastic: they have the same variance across groups.
-Suppose that this does not hold in our data.
-Imagine we have three strata with 16 patients each, but they have different error distributions:
-stratum 1 has standard normal errors,
-stratum 2 has normally distributed errors with mean 0 and variance 2,
-and stratum 3 has errors with a $t$-distribution on 2 degrees of freedom.
-
-## Design 8: heteroskedastic errors, constant treatment effect
-
-```{r design8, fig.height = 4}
-set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
-nn <- 16
-
-tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal", n = rep(nn, 3))
-design8_pvalues <- replicate(100, {
-  tmp$epsilon[tmp$stratumID == 1] <- rnorm(nn)
-  tmp$epsilon[tmp$stratumID == 2] <- rnorm(nn, sd = sqrt(2))
-  tmp$epsilon[tmp$stratumID == 3] <- rt(nn, df = 2)
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
-  generate_simulated_pvalues(tmp)
-})
-
-design8_pvalues <- t(design8_pvalues)
-colnames(design8_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
-design8_power <- apply(design8_pvalues, 2, compute_power)
-```
-
-```{r design8_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design8_power, "Constant Additive Effect of 1, Heteroskedastic Errors")
-
-plot_pvalue_hist(design8_pvalues, "Constant Additive Effect of 1, Heteroskedastic Errors")
-```
-
-## Design 9: heteroskedastic errors, single stratum effect
-
-```{r design9, fig.height = 4}
-set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
-nn <- 16
-
-tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "normal", n = rep(nn, 3))
-design9_pvalues <- replicate(100, {
-  tmp$epsilon[tmp$stratumID == 1] <- rnorm(nn)
-  tmp$epsilon[tmp$stratumID == 2] <- rnorm(nn, sd = sqrt(2))
-  tmp$epsilon[tmp$stratumID == 3] <- rt(nn, df = 2)
-  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
-  generate_simulated_pvalues(tmp)
-})
-design9_pvalues <- t(design9_pvalues)
-colnames(design9_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
-design9_power <- apply(design9_pvalues, 2, compute_power)
-```
-
-```{r design9_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design9_power, "Effect of 1 at a Single stratum, Heteroskedastic Errors")
-
-plot_pvalue_hist(design9_pvalues, "Effect of 1, Heteroskedastic Errors")
 ```
@@
   facet_wrap(Treatment~Errors) +
-  ggtitle("Gaussian Data") +
+  ggtitle("Gaussian Data, Imbalanced Treatment Groups") +
   theme(
@@
   guides(color=guide_legend(nrow=2,byrow=TRUE))
-
 twobytwo
 
-pdf(file = "../ms/fig/normal_simulation_power.pdf", width = 8)
+pdf(file = "../ms/fig/imbalanced_simulation_power.pdf", width = 8)
 twobytwo
@@
 summary05 <- summary05[c(5, 1, 2, 6, 3, 4), ]
-summarytab <- xtable(summary05, digits = 3, align = "rp{1.15in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}",
-       caption = "Empirical power at level $0.05$ for Gaussian simulated data",
-       label = "tab:normal_power") 
-print(summarytab, hline.after = c(-1, 0, 3, nrow(summarytab)), type="latex", file="../ms/fig/normal_simulation_power_summary.tex")
+summarytab <- xtable(summary05, digits = 3, align = "p{1.25in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}",
+       caption = "Empirical power at level $0.05$ for Gaussian simulated data with imbalanced treatment groups",
+       label = "tab:imbalanced_power",
+       hline.after = c(-1, 0, 3)) 
+print(summarytab, hline.after = c(-1, 0, 3, nrow(summarytab)), type="latex", file="../ms/fig/imbalanced_simulation_power_summary.tex")
 summarytab 

### 2016-12-19 ms/ms.tex
Write results section; create figure with outcome distributions

@@
 
-\begin{equation}
+\begin{equation}\label{eqn:ancova}
 Y_{ij} = \alpha_j + \beta X_{ij} + \gamma Z_{ij} + \eps_{ij}
@@
 \subsection{Clinical data results}
+We compare the methods we discussed on a dataset from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease (GERD).
+Patients were treated at eight sites in two different countries.
+At each site, patients were randomly assigned one of two treatments.
+Patients were observed for a week before receiving treatment and for a week after receiving treatment.
+On each of the fourteen days of observation, patients responded to a survey asking about their heartburn, regurgitation, and dyspepsia frequency and severity.
+These endpoints are measured on a discrete scale.
+There were several additional derived endpoints, such as daily regurgitation, daily ``hrdq'', and daily dyspepsia, which are functions of the survey measures.
+Daily ``hrdq'' was the primary endpoint.
+To reduce day-to-day variation, we averaged the measures in the pretreatment week and post-treatment week for each endpoint.
 
-\bit
-\item mention that we thought about whether to use the raw outcomes and adjust for baseline, or use outcome minus baseline.
-The outcome and baseline had a low correlation (include here), so we did not use their difference.
-As the simulations show, doing this actually reduces power
-\item Freedman-Lane stuff: for the appro
-For this dataset, we are guaranteed that treatment $Z$ is independent of $\varepsilon$, as it is randomized within site.
-Indeed, the distribution of residuals looks nearly equal between treatment groups in both models.
-The distribution of residuals varies a bit across sites, but they all appear roughly centered around 0.
-Sites 4 and 5 have larger variance in the differenced model.
-As mentioned, this should not be an issue since we permute treatments within sites, independently across sites.
-(In general, when treatment is not randomly assigned, this condition is necessary for the permutation test to be valid.)
-\eit
+In our model, we treat site as the stratification variable, as this is the level at which treatment was randomized.
+We do not include country in the model, as a country-level effect should be captured in the site-level effects.
+The linear model for ANCOVA analysis is as in Equation~\ref{eqn:ancova}.
+We chose to use this form of the model, with the actual outcomes as the $Y$ and the baseline measures as control variables, rather than testing the difference in outcome and baseline. 
+The outcome and baseline had a low correlation (for instance, $0.56$ for daily ``hrdq''), so we did not use their difference as the dependent variable. 
+As our simulations and previous work (\cite{frison_repeated_1992}) show, when the correlation between baseline and outcome is low, it is more powerful to control for baseline outcomes using a model.
+
+Figure~\ref{fig:clinical_distr} shows the distribution of each clinical endpoint, separated by treatment group.
+There is a clear visual difference in distributions for daily heartburn (``daily\_heart''), daily ``hrdq'', and heartburn frequency (``heart\_freq'').
+The difference is less clear for daily regurgitation (``daily\_regurg'') and regurgitation frequency (``regurg\_freq'').
+
+Table~\ref{tab:clinical_pvalues} shows the $p$-values for the four tests.
+Overall, the results confirm our expectations based on visual comparison in Figure~\ref{fig:clinical_distr}:
+one or more of the tests reject the null hypothesis that outcomes are the same between treatments for heartburn frequency, daily heartburn, and daily ``hrdq,''
+but not for any of the other endpoints.
+The three tests based on the linear model all give similar results here.
+The $p$-values for the ANCOVA tend to be smaller than the linear model and the residual permutation tests, while the $p$-values stratified, unadjusted permutation test have no consistent pattern.
+These results match our simulations: the parametric test has comparable or slightly higher power than the other permutation tests.
+
+\begin{figure}
+\centering
+\includegraphics[width = \textwidth]{fig/clinical_distr}
+\caption{Distribution of each endpoint for the two treatments A and B}
+\label{fig:clinical_distr}
+\end{figure}
 
@@
 Our results match those of \citet{vickers_parametric_2005}, where he found that ANCOVA performed the same or better than nonparametric tests.
-%In randomized control trials, there is no substantial difference between parametric and nonparametric approaches.
-They are asymptotically equivalent to begin with (find a citation).
-But our simulations, using small sample sizes, show that they give comparable results anyway.
 \item Results about differenced vs raw outcomes. This isn't new but our simulations and empirical dataset confirm it in the context of permutation tests
@@
 \citet{boik_fisherpitman_1987} illustrates this phenomenon, comparing the normal theory $F$ test to its permutation counterpart and demonstrating by simulation that the latter has larger than nominal level.
+\item Errors are very important!
+They are a source of non-exchangeability.
+We can't observe them, so we need to believe that they are actually homogeneous and do some checks on residuals to give evidence.
+\bit
 \item \citet{gail_tests_1988} propose a randomization test based on residuals, using not a linear model but an exponential family model to construct residuals, and find that omitting relevant covariates leads to tests with higher than nominal level.
 \todo{perhaps this is something that one should be mindful of when using these linear model based tests.}
+\item Freedman-Lane stuff: for the appro
+In randomized experiements, we are guaranteed that treatment $Z$ is independent of $\varepsilon$, conditional on stratum.
+Indeed, the distribution of residuals looks nearly equal between treatment groups in both models.
+As mentioned, this should not be an issue since we permute treatments within sites, independently across sites.
+(In general, when treatment is not randomly assigned, this condition is necessary for the permutation test to be valid.)
+\eit
 \eit

### 2016-12-19 results/results.Rmd
Write results section; create figure with outcome distributions

@@
 #                       labels=c("A", "B")) + 
-        labs(x = variable, fill = "Treatment")
+        labs(x = variable, fill = "Treatment") +
+        theme(
+        axis.text.x = element_text(size = 12),
+        axis.text.y = element_text(size = 12),
+        axis.title = element_text(size = 16),
+        title = element_text(size = 16),
+        legend.title = element_text(size = 12),
+        legend.text = element_text(size = 12),
+        strip.text.x = element_text(size = 12)
+        ) 
   return(p)
@@
 plot_distrs[[length(plot_distrs)+1]] <- legend
+
+pdf("../ms/fig/clinical_distr.pdf", width = 8, height = 4)
 do.call(grid.arrange, c(plot_distrs, nrow = 2))
+dev.off()
 ```
@@
 We compare four testing methods: 
-the usual parametric ANOVA,
+the usual parametric ANCOVA,
 a permutation test considering the difference in means as test statistic, stratifying by site,
@@
 
-## Parametric ANOVA
-First, we do a standard parametric ANOVA using the two models.
+## Parametric ANCOVA
+First, we do a standard parametric ANCOVA using the two models.
 Under model 1, using the outcome as response and adjusting for the baseline value, treatment has a significant effect.
@@
 ```{r hrdq_parametric, results = "asis"}
-# method 0: parametric ANOVA
+# method 0: parametric ANCOVA
 lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = daily_hrdq)
@@
 We compare two test statistics: the usual difference in means between group B and group A, and the difference in means within each site, aggregated by taking the sum of their absolute values over sites.
-The first statistic is more comparable to what one would obtain from an ANOVA and it is readily interpretable.
+The first statistic is more comparable to what one would obtain from an ANCOVA and it is readily interpretable.
 It does not directly account for the stratification by site, so variation between sites may be hidden.
@@
 
-tests <- c("Parametric ANOVA", "Unadjusted permutation", "Unadjusted permutation (summed across strata)", "Linear regression permutation", "Residual permutation")
+tests <- c("Parametric ANCOVA", "Unadjusted permutation", "Unadjusted permutation (summed across strata)", "Linear regression permutation", "Residual permutation")
 rownames(pvalues_table) <- tests
@@
 
-tests <- c("Parametric ANOVA", "Unadjusted permutation", "Linear regression permutation", "Residual permutation")
+tests <- c("Parametric ANCOVA", "Stratified permutation", "LM permutation", "Residual permutation")
 pvalues_table_contin <- as.data.frame(matrix(NA, nrow = length(continuous_vars), ncol = 4))
@@
                             collapse = ""),
-             caption = "Comparison of p-values from four tests, for each continuous endpoint.")
+             caption = "Comparison of p-values from four tests, for each continuous endpoint.",
+             label = "tab:clinical_pvalues")
 print(summarytab, include.rownames = TRUE)

### 2016-12-19 ms/ms.tex
Edits to theory section

@@
 We are interested in comparing the relative effectiveness of treatment 1 to treatment 0.
-Let $Z$ indicate treatment assignment, 0 or 1, 
-$Y$ denote the observed outcome of interest,
-and $X$ be a pretreatment covariate that is observed and associated with the outcome.
-For expository clarity, we suppose that $X$ is univariate, but all results are easily extended to the case when $X$ is multivariate.
+Let $\mathbf{Z}$ indicate treatment assignment, 0 or 1, 
+$\mathbf{Y}$ denote the observed outcome of interest,
+and $\mathbf{X}$ be a pretreatment covariate that is observed and associated with the outcome.
+For expository clarity, we suppose that $\mathbf{X}$ is univariate, but all results are easily extended to the case when $\mathbf{X}$ is multivariate.
 Suppose there is a categorical pretreatment variable with $J$ levels used to stratify individuals.
@@
 \begin{enumerate}
-\item \textbf{Linearity:} The data $Y$ are related to $X$ and $Z$ linearly.
+\item \textbf{Linearity:} The data $\mathbf{Y}$ are related to $\mathbf{X}$ and $\mathbf{Z}$ linearly.
 \item \textbf{Constant slopes:} Stratum only affects the intercept $\alpha_j$, not the slopes $\beta$ and $\gamma$.
 \item \textbf{IID Errors:} The $\eps_{ij}$ are independent and identically distributed with mean $0$ and variance $\sigma^2$.
-\item \textbf{Independence:} If $X$ is random, $\eps \independent X$.\footnote{$\independent$ means statistical independence; $\perp$ means orthogonal.}
+\item \textbf{Independence:} If $\mathbf{X}$ is random, $\eps \independent X$.\footnote{$\independent$ means statistical independence; $\perp$ means orthogonal.}
 \item \textbf{Normality:} The errors are normally distributed.
@@
 Note, only Assumptions (1) and (2) are needed to guarantee the existence of a solution.
-The solution will be unique as long as there is no linear relationship between $X$, $Z$, and stratum membership.
+The solution will be unique as long as there is no linear relationship between $\mathbf{X}$, $\mathbf{Z}$, and stratum membership.
 }
@@
 Then, which treatment that individual $ij$ received amounts to an arbitrary label.
-Once we observe their response under treatment $Z_{ij}$, we know what it would have been under treatment $1-Z_{ij}$; namely, it would have been the same.
+Once we observe their response under one treatment, we can impute the other potential outcome; namely, it would have been the same.
+This is a stronger assertion than the null hypothesis for the parametric ANCOVA, which only considers the average treatment effect.
 
-If treatment was completely randomized across strata, independently across strata,
-then we may condition on the number of individuals who received each treatment within each stratum.
+For the permutation test, we condition on the number of individuals who received each treatment within each stratum.
 Any assignment of treatments which preserves the number of treated units within each stratum is valid and was just as likely to occur as the randomization that was actually observed.
-By using this principle of equal probabilities and by imputing the unobserved potential outcomes under the sharp null hypothesis, we can construct the permutation distribution of any statistic under the null hypothesis.
+We can construct the permutation distribution of any statistic under the null hypothesis by using this principle of equal probabilities and imputing the unobserved potential outcomes under the sharp null hypothesis.
 
-The most commonly used statistic is the difference in mean outcomes of people who received drug 1 and the mean outcomes of people who received drug 0.
-This statistic is readily interpretable (``on average, taking the drug changes the outcome by x amount'') and has nice theoretical properties owing to it being the difference of two sums.
-When we want to be sensitive to differences in the magnitude of the effect, the difference in means may not be optimal.
-For an extreme example, imagine that there are $n/2$ males and $n/2$ females in the sample, $n/4$ males and females receive each treatment.
-Those who receive treatment 0 have an outcome of 0, the males who receive treatment 1 have an outcome of 1, and the females who receive treatment 1 have an outcome of -1.
-Then the difference in means between the treatment groups will be $0$, even though the treatment had a positive effect on males and negative effect on females.
-This differential effect will be averaged out.
-Thus, one may want to stratify the sample according to important confounding variables, then take the difference in means within each stratum separately.
+The most commonly used statistic is the difference in mean outcomes of people who received treatment 1 and the mean outcomes of people who received treatment 0.  
+This statistic is readily interpretable (``on average, taking treatment changes the outcome by x amount'') and has nice theoretical properties owing to it being the difference of two sums.
+However, the difference in means may not be optimal if we want to be sensitive to heterogeneous effects.  
+For an extreme example, imagine that there an equal number of males and females in the sample, and each treatment is assigned to half of males and half of females.  
+Everybody who receives treatment 0 has an outcome of 0, but males who receive treatment 1 have an outcome of $1$ and females who receive treatment 1 have an outcome of $-1$.  
+Then the difference in means between the treatment groups is $0$, even though the treatment had nonzero effects on both males and females.  
+This differential effect gets averaged out.
 
+One may want to stratify the sample according to important confounding variables, then compute a statistic for each stratum separately. 
 There is a great degree of freedom in choosing how exactly to construct such a statistic.
-Two factors must be decided: how to stratify and how to combine the individual means.
+Two factors must be decided: how to stratify and how to combine the individual strata statistics.
 There is a tradeoff in how finely we choose to stratify:
-on the one hand, we need all of the strata to be sufficiently large, otherwise they don't contribute any information to the test statistic,
- but on the other hand, we want the strata to capture variation in treatment effects.
-\todo{discussion of NPC and cite papers}
+on the one hand, the strata must be sufficiently large to be informative,
+ but on the other hand, must be fine enough to capture variation in treatment effects.
+Our simulations in Section~\ref{simulations} show a loss of power when effects are concentrated in small strata.
+
+One way to construct a test statistic is to combine the statistics directly, for instance taking the sum of absolute values, into a single statistic on the data.
+Taking the absolute value before summing ensures that effects with different signs do not cancel each other out.
+The permutation test is conducted identically, using this statistic in place of the difference in means.
+Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by \citet{pesarin_permutation_2010}.
+In this framework, we would break down the ``global'' null hypothesis of no effect whatsoever into the intersection of null hypotheses of no treatment effect within each stratum.
+For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies.
+In this case, strata are randomized independently so there are no dependencies to preserve, making this method is equivalent to the first.
 
 \subsection{Permutation tests with the linear model}
-We would like to test for a difference in outcomes between two treatments, but control for other covariates that may be predictive of the outcome.
-For example, if the outcome is the outcome after receiving treatment, we would like to control for the same variable measured before treatment was administered.
-When the outcome is the difference in responses between the second and first weeks, we would just like to control for location.
-We will use the approximate permutation test derived by Freedman and Lane (1983) to do so.
+The tests in the previous section only use information on the treatment and the outcome.
+However, experimenters record additional covariates, some of which may be predictive of the outcome.
+One would like to construct a more powerful test by incorporating these covariates in a way to reduce the variance of the statistic.
+In particular, we can turn the parametric ANCOVA into a nonparametric permutation test.
+We will model the outcome as in Equation~\ref{eqn:ancova}.
+In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes.
+This makes several things ``permutable'' and allows us to construct several different tests.
 
-Under the sharp null $H_0$, $Z$ has no effect on $Y$.
-In other words, the null hypothesis is that $\beta_2 = 0$.
-In a randomized experiment such as this, there are two ways we may test this null hypothesis.
+First, we may do a simple variation on the stratified permutation test described above.
+Instead of using the difference in means as the statistic, we use the $t$ statistic for the coefficient on $\mathbf{Z}$ in a linear regression.
+If treatment was assigned at random within each stratum, then it is guaranteed that $\mathbf{Z}$ and $\mathbf{\varepsilon}$ are statistically independent, conditional on stratum.
+Therefore, we may conduct a test by permuting treatment assignments $\mathbf{Z}$ within site, independently across sites, and calculating a test statistic for each such permutation.
 
-\subsubsection{Standard linear model}
-We may do a variation on the stratified two-sample test described above.
-Instead of using the difference in means as the statistic, we can use the t-statistic for the coefficient on $Z$ in a linear regression.
-If treatment was assigned at random within each stratum, then it is guaranteed that $Z$ and $\varepsilon$ are statistically independent, conditional on stratum.
-Therefore, we may conduct a test by permuting treatment assignments $Z$ within site, independently across sites, and calculating a test statistic for each such permutation.
+\citet{freedman_nonstochastic_1983} propose an alternative test based on the residuals of the linear regression.
+Let's take an alternative view of the problem.
+We still write the outcome in the form of Equation~\ref{eqn:ancova}, but now, we do not treat the $\varepsilon_{ij}$ as random.
+They are simply defined to be the difference between the observed outcome $Y_{ij}$ and the data's linear projection onto the plane $\alpha_j + \beta X_{ij}+ \gamma Z_{ij}$.
 
-\subsubsection{Linear model residuals}
-
-Let's take an alternative view of the problem.
-We still write $Y = \beta_0 + \beta_1 X + \beta_2 Z + \varepsilon$.
-However, now, we do not treat the $\varepsilon$ as random.
-They are simply defined to be the diffence between $Y$ and the data's linear projection onto the plane $\beta_0 + \beta_1 X + \beta_2 Z$.
-
-If the null hypothesis is true, then $\varepsilon = Y - \beta_0 - \beta_1X$.
-We may regress $Y$ on $X$ to obtain coefficient estimates $\hat{\beta}_0$ and $\hat{\beta}_1$.
-Then we would predict the outcome to be $\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1X$
-and we may estimate the errors $\hat{\varepsilon}$ by $Y - \hat{Y}$.
-The $\hat{\varepsilon}$ approximate the true errors $\varepsilon$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit.
-Furthermore, under the null hypothesis, the $\varepsilon$ are independent of $X$ and $Z$. \todo{ this is false. fix}
-Therefore, within sites, these estimated $\hat{\varepsilon}$ are approximately exchangeable.
-Note that the exchangeability is only approximate, because $\hat{\varepsilon}$ will have some correlation with $X$ due to the finite sample size. \todo{ fix wording}
+If the null hypothesis is true, then $\gamma = 0$ and $\varepsilon_{ij} = Y_{ij} - \alpha_{ij} - \beta X_{ij}$ for all $i = 1, \dots, n_j, j = 1, \dots, J$.
+We may regress $\mathbf{Y}$ on $\mathbf{X}$ to obtain coefficient estimates $\hat{\alpha}_j$ and $\hat{\beta}$.
+Then we would predict the outcome to be $\hat{Y}_{ij} = \hat{\alpha}_j + \hat{\beta}X_{ij}$
+and we may estimate the errors $\mathbf{\hat{\varepsilon}}$ by $\mathbf{Y} - \mathbf{\hat{Y}}$.
+The $\mathbf{\hat{\varepsilon}}$ approximate the true errors $\mathbf{\varepsilon}$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit.
+Furthermore, under the null hypothesis, the $\mathbf{\varepsilon}$ are independent $\mathbf{Z}$ within strata. 
+If we make the additional assumption that $\mathbf{\varepsilon}$ are exchangeable against $\mathbf{X}$ within strata, then 
+we may permute the estimated $\hat{\varepsilon}$ within strata, independently across strata.
 
@@
 \begin{enumerate}
-\item Estimate $\hat{\varepsilon}$ by $Y - \hat{\beta}_0 - \hat{\beta}_1 X$, where $\hat{\beta}_0$ and $\hat{\beta}_1$ are obtained by regressing $Y$ on $X$.
-\item Construct permuted errors $\hat{\varepsilon}^\pi$ by permuting the $\hat{\varepsilon}$ within sites.
-\item Construct permuted responses $Y^\pi = \hat{\beta}_0 + \hat{\beta}_1 X + \hat{\varepsilon}$.
-\item Regress $Y^\pi$ on $X$ and $Z$. The test statistic is the $t$-statistic for the coefficient of $Z$.
+\item Estimate $\mathbf{\hat{\varepsilon}}$ by $Y_{ij} - \hat{\alpha}_j - \hat{\beta} X_{ij}$ for all $i$ and $j$, where $\hat{\alpha}_j$ and $\hat{\beta}$ are obtained by regressing $\mathbf{Y}$ on $\mathbf{X}$ and stratum assignment.
+\item Construct permuted errors $\mathbf{\hat{\varepsilon}}^\pi$ by permuting the $\mathbf{\hat{\varepsilon}}$ within strata.
+\item Construct permuted responses $Y_{ij}^\pi = \hat{\alpha}_j + \hat{\beta} X_{ij}+ \hat{\varepsilon}_{ij}^\pi$.
+\item Regress $\mathbf{Y}^\pi$ on $\mathbf{X}$, $\mathbf{Z}$, and stratum. The test statistic is the $t$-statistic for the coefficient of $\mathbf{Z}$.
 \end{enumerate}
 
-\subsubsection{Other linear model tests}
 It is important to note that the permutation tests described here are only exact in the context of randomized experiments.
-In observational studies, they are only approximately exact under the best circumstances.
-What's the difference between these two situations?
-In randomized experiments, treatment is assigned at random, and is therefore statistically independent of the covariates $X$ and the errors $\varepsilon$.
-Conversely, in observational studies, treatment may be associated with $X$ and/or $\varepsilon$, often in a way that is difficult or impossible to model.
-Independence, or the weaker condition of exchangeability, is what enables us to construct permutation distributions while holding $X$ fixed.
-When exchangeability doesn't hold, it does not make sense to permute $Z$ while holding $X$ fixed, as we cannot disentangle the effect of one over the other.
+In observational studies, even under the best circumstances they are only approximately exact.
+In randomized experiments, treatment is assigned at random, and is therefore statistically independent of the covariates $\mathbf{X}$ and the errors $\mathbf{\varepsilon}$.
+Conversely, in observational studies, treatment may be associated with $\mathbf{X}$ and/or $\mathbf{\varepsilon}$, often in a way that is difficult or impossible to model.
+Independence, or the weaker condition of exchangeability, is what enables us to construct permutation distributions while holding $\mathbf{X}$ fixed.
+When exchangeability doesn't hold, it does not make sense to permute $\mathbf{Z}$ while holding $\mathbf{X}$ fixed, as we cannot disentangle the effect of one over the other.
 In order to justify the use of one of these linear model tests for a treatment effect in observational data, one must argue that the object being permuted is uncorrelated with the other variables being held fixed.
-This can be done visually using scatterplots and residual plots.
-\todo{citations!}
+This can be done visually using scatterplots and residual plots (\cite{freedman_nonstochastic_1983}).
 
@@
 \section{Results}
-\subsection{Simulations}
+\subsection{Simulations}\label{simulations}
 
@@
 We will assume that $\beta_0 = 1$ and let the treatment effect be $\gamma = 1$
-For each of the two simulation designs that vary the treatment effect, we use two distributions of $\varepsilon$.
+For each of the two simulation designs that vary the treatment effect, we use two distributions of $\mathbf{\varepsilon}$.
 In the first case, we use $\varepsilon \sim N(0, 1)$ to fit with the parametric ANOVA assumption of normal errors.
-In the second case, $\varepsilon$ comes from a $t$ distribution with 2 degrees of freedom so the errors are heavy-tailed.
+In the second case, $\mathbf{\varepsilon}$ comes from a $t$ distribution with 2 degrees of freedom so the errors are heavy-tailed.
 Thus, there are four total simulation designs.
@@
 The linear model for ANCOVA analysis is as in Equation~\ref{eqn:ancova}.
-We chose to use this form of the model, with the actual outcomes as the $Y$ and the baseline measures as control variables, rather than testing the difference in outcome and baseline. 
+We chose to use this form of the model, with the actual outcomes as the $\mathbf{Y}$ and the baseline measures as control variables, rather than testing the difference in outcome and baseline. 
 The outcome and baseline had a low correlation (for instance, $0.56$ for daily ``hrdq''), so we did not use their difference as the dependent variable. 
@@
 \item Freedman-Lane stuff: for the appro
-In randomized experiements, we are guaranteed that treatment $Z$ is independent of $\varepsilon$, conditional on stratum.
+In randomized experiments, we are guaranteed that treatment $\mathbf{Z}$ is independent of $\mathbf{\varepsilon}$, conditional on stratum.
 Indeed, the distribution of residuals looks nearly equal between treatment groups in both models.

### 2016-12-20 ms/ms.tex
Start conforming to BMC style; update some citations; flesh out methods and discussion

@@
 \documentclass[11pt]{article}
+
+%\documentclass{bmcart}
+
+%%% Load packages
+%\usepackage{amsthm,amsmath}
+%\RequirePackage{natbib}
+%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
+%\RequirePackage{hyperref}
+\usepackage[utf8]{inputenc} %unicode support
+
 \usepackage[breaklinks=true]{hyperref}
@@
 \usepackage{amsmath,amssymb,amsthm}
-\usepackage{natbib}
+\usepackage[numbers]{natbib}
 \usepackage{array}
-\usepackage{booktabs, multicol, multirow}
+%\usepackage{booktabs, multicol, multirow}
 \usepackage[nohead, margin=1in]{geometry}
-\usepackage[singlespacing]{setspace}
+%\usepackage[singlespacing]{setspace}
 \usepackage{graphicx}
+
+%%% Put your definitions there:
+%\startlocaldefs
 
@@
 
+%\endlocaldefs
+
+\newcommand{\parttitle}[1]{\textbf{#1}}
+
+%%% Begin ...
+\begin{document}
+
+%%% Start of article front matter
+%\begin{frontmatter}
+
+%\begin{fmbox}
+%\dochead{Research}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the title of your article here     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \title{A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Control Trials}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors here                   %%
+%%                                          %%
+%% Specify information, if available,       %%
+%% in the form:                             %%
+%%   <key>={<id1>,<id2>}                    %%
+%%   <key>=                                 %%
+%% Comment or delete the keys which are     %%
+%% not used. Repeat \author command as much %%
+%% as required.                             %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%\author[
+%   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
+%   corref={aff1},                       % id of corresponding address, if any
+%   noteref={n1},                        % id's of article notes, if any
+%   email={jane.e.doe@cambridge.co.uk}   % email address
+%]{\inits{JE}\fnm{Jane E} \snm{Doe}}
+%\author[
+%   addressref={aff1,aff2},
+%   email={john.RS.Smith@cambridge.co.uk}
+%]{\inits{JRS}\fnm{John RS} \snm{Smith}}
+
 \author{Kellie Ottoboni \\
@@
 }\date{Draft \today}
-\begin{document}
 \maketitle
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors' addresses here        %%
+%%                                          %%
+%% Repeat \address commands as much as      %%
+%% required.                                %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\newpage
+%\address[id=aff1]{%                           % unique id
+%  \orgname{Department of Zoology, Cambridge}, % university, etc
+%  \street{Waterloo Road},                     %
+%  %\postcode{}                                % post or zip code
+%  \city{London},                              % city
+%  \cny{UK}                                    % country
+%}
+%\address[id=aff2]{%
+%  \orgname{Marine Ecology Department, Institute of Marine Sciences Kiel},
+%  \street{D\"{u}sternbrooker Weg 20},
+%  \postcode{24105}
+%  \city{Kiel},
+%  \cny{Germany}
+%}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter short notes here                   %%
+%%                                          %%
+%% Short notes will be after addresses      %%
+%% on first page.                           %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%\begin{artnotes}
+%%\note{Sample of title note}     % note to the article
+%\note[id=n1]{Equal contributor} % note, connected to author
+%\end{artnotes}
+%
+%\end{fmbox}% comment this for two column layout
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Abstract begins here                 %%
+%%                                          %%
+%% Please refer to the Instructions for     %%
+%% authors on http://www.biomedcentral.com  %%
+%% and include the section headings         %%
+%% accordingly for your article type.       %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\begin{abstractbox}
 
 \begin{abstract}
-\noindent\textbf{Background:}
+\parttitle{Background}
 In clinical trials, the standard method of measuring the effectiveness of one treatment over another is a parametric analysis of covariance (ANCOVA), a special case of linear regression which incorporates stratification.
 The method relies on strong, untestable assumptions about the data-generating process such as constant treatment effects across groups, normality of errors, and homoskedasticity.
-The effectiveness of the treatment is decided by the p-value from this hypothesis test, which may not be valid if the assumptions for ANCOVA are not met.
+The effectiveness of the treatment is decided by the $p$-value from this hypothesis test, which may not be valid if the assumptions for ANCOVA are not met.
 We would like to understand which assumptions may be relaxed or violated for the ANCOVA to remain valid (but potentially less powerful).
 
-\noindent\textbf{Methods:}
+\parttitle{Methods}
 Permutation tests present such an opportunity: they require only minimal assumptions which are often guaranteed by the randomization that was conducted.
@@
 
-\noindent\textbf{Results:}
+\parttitle{Results}
 Our simulations show that the permutation methods maintain power comparable to the ANCOVA in randomized trials.
@@
 
-\noindent\textbf{Conclusions:}
+\parttitle{Conclusions}
 These results suggest that the parametric ANCOVA and linear model permutation tests perform similarly, even when not all assumptions for the parametric test are met.
@@
 
-% keywords: permutation test, linear model, ANOVA
-\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The keywords begin here                  %%
+%%                                          %%
+%% Put each keyword in separate \kwd{}.     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Background}
+%\begin{keyword}
+%\kwd{permutation test}
+%\kwd{linear model}
+%\kwd{ANCOVA}
+%\end{keyword}
+
+% MSC classifications codes, if any
+%\begin{keyword}[class=AMS]
+%\kwd[Primary ]{}
+%\kwd{}
+%\kwd[; secondary ]{}
+%\end{keyword}
+%
+%\end{abstractbox}
+%%
+%%\end{fmbox}% uncomment this for twcolumn layout
+%
+%\end{frontmatter}
+
+\section*{Background}
 A hypothesis test is a statistical method for determining whether observed data is consistent with a belief about the process that generated the data.
 Medical experiments use hypothesis testing to assess the evidence that a treatment has an effect on one or more clinically relevant outcomes.
-The simplest version of this experiment, which involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome, has been studied for nearly a century (see \citet{fisher_design_1935} and \citet{neyman_application_1923} for early references).
+The simplest version of this experiment, which involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome, has been studied for nearly a century (see \cite{fisher_design_1935} and \cite{neyman_application_1923} for early references).
 One can conduct hypothesis tests and construct confidence intervals for the estimated treatment effect
@@
 On the one hand, some argue that violations of parametric test assumptions necessitate the use of permutation methods.
-\citet{ludbrook_why_1998} point out that medical trials rarely follow the population sampling model implicit in parametric methods.
-However, the parametric and nonparametric tests seem to perform similarly in most simulations, even when data violate the assumptions of the parametric method.
+\cite{ludbrook_why_1998} point out that medical trials rarely follow the population sampling model implicit in parametric methods.
+Many people recommend using permutation tests in place of the usual parametric tests for typical analyses, such as ANOVA and generalized linear models (\cite{still_approximate_1981, winkler_permutation_2014}).
+They argue that there are a myriad of ways that the data may violate the necessary assumptions for the test, and so it is more robust to use a permutation test.
+
+However, simulations show that parametric and nonparametric tests seem to perform similarly in most cases, even when data violate the assumptions of the parametric method.
 Medical trials studying pain or perceived severity scores often use Likert scale data, which is discrete and does not match the normality assumptions of parametric tests.
-However, \citet{winter_five-point_2010} found that the two sample $t$ test and Mann-Whitney test had comparable Type I and II error rates for five-point Likert scale data, suggesting that the violation of normality does not entirely invalidate the parametric test.
-\todo{what other linear model comparisons have people done?}
-Most similar to our question of study, \citet{vickers_parametric_2005} compared the parametric ANCOVA to the Mann-Whitney rank test in the context of randomized experiments, finding that except in extreme situations, ANCOVA outperformed the nonparametric test.
+However, \cite{winter_five-point_2010} found that the two sample $t$ test and Mann-Whitney test had comparable Type I and II error rates for five-point Likert scale data, suggesting that the violation of normality does not entirely invalidate the parametric test.
+\cite{vickers_parametric_2005} compared the parametric ANCOVA to the Mann-Whitney rank test in the context of randomized experiments, finding that except in extreme situations, ANCOVA was more powerful than the nonparametric test.
+Most similar to our question of study, \cite{anderson_empirical_1999} found little difference between several permutation tests for coefficients in a linear model alongside the parametric $t$ test.
 
@@
 
-\section{Methods}
+\section*{Methods}
 % The methods section should include:
@@
 
-\subsection{Problem and notation \todo{what do others call this?}}
+\subsection*{Problem and notation \todo{what do others call this?}}
 Suppose we randomly assign two treatments labelled $0$ and $1$ to a group of individuals.
@@
 
-\subsection{Parametric ANCOVA}
+\subsection*{Parametric ANCOVA}
 
@@
 \item \textbf{IID Errors:} The $\eps_{ij}$ are independent and identically distributed with mean $0$ and variance $\sigma^2$.
-\item \textbf{Independence:} If $\mathbf{X}$ is random, $\eps \independent X$.\footnote{$\independent$ means statistical independence; $\perp$ means orthogonal.}
+\item \textbf{Independence:} If $\mathbf{X}$ is random, $\mathbf{\varepsilon}$ is statistically independent of $\mathbf{X}$.
 \item \textbf{Normality:} The errors are normally distributed.
@@
 follows the Student $t$ distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
-The p-value for this hypothesis test is the probability that a value drawn from this $t$ distribution is larger in magnitude than $T$.
-
-\todo{Add a paragraph about why this might go wrong, perhaps some citations. All depends on estimating the error variance correctly, because $t$ distribution is the ratio of a normal and a chi-square}
+The $p$-value for this hypothesis test is the probability that a value drawn from this $t$ distribution is larger in magnitude than $T$.
 
@@
 
-\subsection{Stratified permutation test}
+\subsection*{Stratified permutation test}
 Suppose we wish to test the null hypothesis that individual by individual, treatment has no effect.
@@
 The permutation test is conducted identically, using this statistic in place of the difference in means.
-Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by \citet{pesarin_permutation_2010}.
+Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by \cite{pesarin_permutation_2010}.
 In this framework, we would break down the ``global'' null hypothesis of no effect whatsoever into the intersection of null hypotheses of no treatment effect within each stratum.
@@
 
-\subsection{Permutation tests with the linear model}
+\subsection*{Permutation tests with the linear model}
 The tests in the previous section only use information on the treatment and the outcome.
@@
 
-\citet{freedman_nonstochastic_1983} propose an alternative test based on the residuals of the linear regression.
+Freedman and Lane \cite{freedman_nonstochastic_1983} propose an alternative test based on the residuals of the linear regression.
 Let's take an alternative view of the problem.
@@
 
-It is important to note that the permutation tests described here are only exact in the context of randomized experiments.
-In observational studies, even under the best circumstances they are only approximately exact.
-In randomized experiments, treatment is assigned at random, and is therefore statistically independent of the covariates $\mathbf{X}$ and the errors $\mathbf{\varepsilon}$.
-Conversely, in observational studies, treatment may be associated with $\mathbf{X}$ and/or $\mathbf{\varepsilon}$, often in a way that is difficult or impossible to model.
-Independence, or the weaker condition of exchangeability, is what enables us to construct permutation distributions while holding $\mathbf{X}$ fixed.
-When exchangeability doesn't hold, it does not make sense to permute $\mathbf{Z}$ while holding $\mathbf{X}$ fixed, as we cannot disentangle the effect of one over the other.
-In order to justify the use of one of these linear model tests for a treatment effect in observational data, one must argue that the object being permuted is uncorrelated with the other variables being held fixed.
-This can be done visually using scatterplots and residual plots (\cite{freedman_nonstochastic_1983}).
+Others have developed different variants on approximate permutation tests based on linear regression. 
+There is some disagreement on what constitutes an appropriate permutation scheme;
+everything we have described so far is approximate insofar as we do not know the ``true'' data-generating process described by the $\alpha_j$ and $\beta$,
+only the estimates $\hat{\alpha}_j$ and $\hat{\beta}$.
+\cite{manly_randomization_2006} proposes randomizing the outcomes $\mathbf{Y}$, treating them as though they were randomly assigned to pairs $(X, Z)$ under the null hypothesis.
+Permutations under the null treat all coefficients as 0, which may not reflect the true data-generating process.
+\cite{kennedy_randomization_1995} proposed a permutation method similar in spirit to \cite{freedman_nonstochastic_1983}, but which differs procedurally.
+Both methods attempt to measure the correlation between treatment and unexplained variation in outcomes, but
+instead of regressing pseudo-outcomes $\mathbf{Y}^\pi$ on covariates to obtain a permutation distribution, he recommended regressing $\mathbf{Z}$ on the permuted residuals $\mathbf{\hat{\varepsilon}}^\pi$ and using the $t$ statistic from the resulting coefficient as the test statistic.
 
-Others have developed approximate permutation tests which use linear regression to adjust for covariates. 
-\bit
-\item Cite a few here, \todo{all the Anderson papers}
-\item explain why Freedman-Lane is best. 
-\item Cite some people who have shown by simulation that it works better.
-\eit
+Several authors have compared these tests, theoretically and through simulations \cite{anderson_empirical_1999, anderson_permutation_2001, kennedy_randomization_1996}.
+They find that the Freedman-Lane test of residuals is asymptotically equivalent to the ``oracle'' exact hypothesis test, which we could conduct if we knew which permutations of $\mathbf{Y}$ given $\mathbf{X}$ were equiprobable under the null hypothesis \cite{anderson_permutation_2001}.
+This confirms empirical results, which show that the Freedman-Lane test performs better than other linear model based tests in simulations simulations \cite{anderson_empirical_1999}, though its advantage is small.
+Thus, throughout the rest of the paper, we focus on the two linear regression permutations we described in detail: linear regression after permuting treatment assignments and the Freedman-Lane method of permuting residuals.
 
-\section{Results}
-\subsection{Simulations}\label{simulations}
+\section*{Results}
+\subsection*{Simulations}\label{simulations}
 
@@
 Then, for 1000 trials, we randomly assign treatment to half of the individuals within each stratum, generate random errors, and construct new outcomes $Y_{ij1}$ using Equation~\ref{eqn:dgp}.
-We conduct all five tests on this new data, obtaining a two-sided p-value for each.
-The power curves are estimated by computing the proportion of observed p-values less than or equal to $\alpha$, for $\alpha = 0.01, 0.02, \dots, 1$.
+We conduct all five tests on this new data, obtaining a two-sided $p$-value for each.
+The power curves are estimated by computing the proportion of observed $p$-values less than or equal to $\alpha$, for $\alpha = 0.01, 0.02, \dots, 1$.
 
@@
 In this case, the result was reversed: the differenced test had low power while the test using the raw outcomes had a power curve closer to the tests that adjust for covariates using a linear model.
-This phenomenon has been studied before by \citet{frison_repeated_1992}, who recommended to use the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
+This phenomenon has been studied before by \cite{frison_repeated_1992}, who recommended to use the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
 
-\subsection{Clinical data results}
+\subsection*{Clinical data results}
 We compare the methods we discussed on a dataset from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease (GERD).
@@
 The outcome and baseline had a low correlation (for instance, $0.56$ for daily ``hrdq''), so we did not use their difference as the dependent variable. 
-As our simulations and previous work (\cite{frison_repeated_1992}) show, when the correlation between baseline and outcome is low, it is more powerful to control for baseline outcomes using a model.
+As our simulations and previous work \cite{frison_repeated_1992} show, when the correlation between baseline and outcome is low, it is more powerful to control for baseline outcomes using a model.
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 % This section should discuss the implications of the findings in context of existing research and highlight limitations of the study.
 
-\bit
-\item This paper adds to the literature comparing parametric and nonparametric tests.
-Our results match those of \citet{vickers_parametric_2005}, where he found that ANCOVA performed the same or better than nonparametric tests.
-\item Results about differenced vs raw outcomes. This isn't new but our simulations and empirical dataset confirm it in the context of permutation tests
-\item In an ideal case, one knows all covariates related to the outcome and fits a fully saturated linear model (including all covariates and their interactions), the coefficient for the treatment is an asymptotically consistent estimate of the average treatment effect (\cite{lin_agnostic_2013}).
-\item Permutation tests don't come free of assumptions, though.
-\citet{romano_behavior_1990} warns against using permutation tests naively, if items are not truly exchangeable.
+This paper adds to the literature comparing parametric and nonparametric tests.
+Our results match those of \cite{vickers_parametric_2005, anderson_empirical_1999}, where parametric ANCOVA and regression-based $t$ tests performed the same or better than the comparable nonparametric tests.
+We simulated x y and z violations
+When the ANCOVA assumptions were violated, the linear regression-based tests lost power relative to the best case data-generating process where all assumptions were satisfied.
+However, power did not change \textit{relative to each other}.
+It is a matter of philosophy which one chooses: while the parametric test may be robust to violations of its assumptions, it is somewhat reassuring that the permutation test can exactly match the randomization that was done and makes fewer distributional assumptions.
+
+Permutation tests don't come free of assumptions, though.  \cite{romano_behavior_1990} warns against using permutation tests naively if items are not truly exchangeable. 
 For instance, he points to the case where observations have unequal variances.  
-\citet{boik_fisherpitman_1987} illustrates this phenomenon, comparing the normal theory $F$ test to its permutation counterpart and demonstrating by simulation that the latter has larger than nominal level.
-\item Errors are very important!
-They are a source of non-exchangeability.
-We can't observe them, so we need to believe that they are actually homogeneous and do some checks on residuals to give evidence.
-\bit
-\item \citet{gail_tests_1988} propose a randomization test based on residuals, using not a linear model but an exponential family model to construct residuals, and find that omitting relevant covariates leads to tests with higher than nominal level.
-\todo{perhaps this is something that one should be mindful of when using these linear model based tests.}
-\item Freedman-Lane stuff: for the appro
-In randomized experiments, we are guaranteed that treatment $\mathbf{Z}$ is independent of $\mathbf{\varepsilon}$, conditional on stratum.
-Indeed, the distribution of residuals looks nearly equal between treatment groups in both models.
-As mentioned, this should not be an issue since we permute treatments within sites, independently across sites.
-(In general, when treatment is not randomly assigned, this condition is necessary for the permutation test to be valid.)
-\eit
-\eit
+This is a problem as we cannot observe the errors; it is a leap of faith to suppose that they are homogeneous.
+\cite{boik_fisherpitman_1987} illustrates this phenomenon, comparing the normal theory $F$ test to its permutation counterpart and demonstrating by simulation that the latter has larger than nominal level when the error variances are heterogeneous.
 
-% list of abbreviations
+It is important to note that the permutation tests described here are only approximately exact in the context of randomized experiments.
+In this case, treatment is assigned at random and is therefore statistically independent of the covariates $\mathbf{X}$ and the errors $\mathbf{\varepsilon}$.
+Conversely, in observational studies, treatment may be associated with $\mathbf{X}$ and/or $\mathbf{\varepsilon}$, often in a way that is difficult or impossible to model.
+Independence, or the weaker condition of exchangeability, is what enables us to construct permutation distributions while holding $\mathbf{X}$ fixed.
+When exchangeability doesn't hold, we cannot disentangle the effect of $\mathbf{Z}$ from the effect of $\mathbf{X}$.
+In observational studies, one must provide evidence that items are exchangeable in order to achieve approximate exchangeability.
+One must argue that the object being permuted is uncorrelated with the other variables being held fixed.
+This can be done visually using scatterplots and residual plots \cite{freedman_nonstochastic_1983} and from subject-matter knowledge.
 
-\section{Declarations}
-% see http://bmcmedresmethodol.biomedcentral.com/submission-guidelines/preparing-your-manuscript/technical-advance-article
-\textbf{Ethics approval and consent to participate}
-Not applicable
-\textbf{Consent for publication}
-\textbf{Availability of data and material}
-\textbf{Competing interests}
-\textbf{Funding}
-\textbf{Authors' contributions}
-\textbf{Acknowledgements}
-\textbf{Authors' information}
+One should be mindful of which covariates are important to control for.
+Ideally, one knows all covariates related to the outcome.
+In that case, the coefficient for treatment in a fully saturated linear model (including all covariates and their interactions) is an asymptotically consistent estimate of the average treatment effect \cite{lin_agnostic_2013}.
+It is rare that one knows \textit{all} relevant covariates in practice; failure to control for all relevant variables can introduce bias in certain permutation tests.
+ \cite{gail_tests_1988} propose a randomization test based on residuals, using not a linear model but an exponential family model to construct residuals, and find that omitting relevant covariates leads to tests with higher than nominal level.
 
-\bibliographystyle{plainnat}
-\bibliography{references}
+The method of controlling for baseline covariates matters, too.
+The naive thing to do is to consider change scores, subtracting baseline measures from the outcome and doing hypothesis testing using the change as the dependent variable.
+Our simulations confirm the suggestion of \cite{frison_repeated_1992} to use change scores only when there is a strong correlation between baseline and outcome.
+If the correlation is weak, then the resulting test using the difference may be less powerful than ignoring the baseline altogether.
+Instead, we suggest incorporating the baseline in a regression model.
+This is more general than using differences;
+treating the change scores as the dependent variable is a special case of the linear regression, constraining the coefficient on the baseline measure to be $1$.
+Weak correlations between baseline and outcome occur often in practice, as was the case with our GERD dataset.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Backmatter begins here                   %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\begin{backmatter}
+%
+%\section*{Competing interests}
+%  The authors declare that they have no competing interests.
+%
+%\section*{Author's contributions}
+%    Text for this section \ldots
+%
+%\section*{Acknowledgements}
+%  Text for this section \ldots
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                  The Bibliography                       %%
+%%                                                         %%
+%%  Bmc_mathpys.bst  will be used to                       %%
+%%  create a .BBL file for submission.                     %%
+%%  After submission of the .TEX file,                     %%
+%%  you will be prompted to submit your .BBL file.         %%
+%%                                                         %%
+%%                                                         %%
+%%  Note that the displayed Bibliography will not          %%
+%%  necessarily be rendered by Latex exactly as specified  %%
+%%  in the online Instructions for Authors.                %%
+%%                                                         %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% if your bibliography is in bibtex format, use those commands:
+\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
+\bibliography{references}      % Bibliography file (usually '*.bib' )
+%% for author-year bibliography (bmc-mathphys or spbasic)
+%% a) write to bib file (bmc-mathphys only)
+%% @settings{label, options="nameyear"}
+%% b) uncomment next line
+%%\nocite{label}
+%
+%% or include bibliography directly:
+%% \begin{thebibliography}
+%% \bibitem{b1}
+%% \end{thebibliography}
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%                               %%
+%%% Figures                       %%
+%%%                               %%
+%%% NB: this is for captions and  %%
+%%% Titles. All graphics must be  %%
+%%% submitted separately and NOT  %%
+%%% included in the Tex document  %%
+%%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%%%
+%%% Do not use \listoffigures as most will included as separate files
+%
+%\section*{Figures}
+%  \begin{figure}[h!]
+%  \caption{\csentence{Sample figure title.}
+%      A short description of the figure content
+%      should go here.}
+%      \end{figure}
+%
+%\begin{figure}[h!]
+%  \caption{\csentence{Sample figure title.}
+%      Figure legend text.}
+%      \end{figure}
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%                               %%
+%%% Tables                        %%
+%%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%%% Use of \listoftables is discouraged.
+%%%
+%\section*{Tables}
+%\begin{table}[h!]
+%\caption{Sample table title. This is where the description of the table should go.}
+%      \begin{tabular}{cccc}
+%        \hline
+%           & B1  &B2   & B3\\ \hline
+%        A1 & 0.1 & 0.2 & 0.3\\
+%        A2 & ... & ..  & .\\
+%        A3 & ..  & .   & .\\ \hline
+%      \end{tabular}
+%\end{table}
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%                               %%
+%%% Additional Files              %%
+%%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%\section*{Additional Files}
+%  \subsection*{Additional file 1 --- Sample additional file title}
+%    Additional file descriptions text (including details of how to
+%    view the file, if it is in a non-standard format or the file extension).  This might
+%    refer to a multi-page table or a figure.
+%
+%  \subsection*{Additional file 2 --- Sample additional file title}
+%    Additional file descriptions text.
+%
+%
+%\end{backmatter}
 

### 2016-12-23 ms/ms.tex
Fix bibliography entries
Major edits to manuscript

@@
 Department of Statistics; Berkeley Institute for Data Science\\
-University of California, Berkeley\\ [.2in]
+University of California, Berkeley \\ [.2in]
+Fraser Lewis \\
+RB\\ [.2in]
 Luigi Salmaso\\
 Department of Management and Engineering \\
-University of Padova \\ [.2in]
-Fraser Lewis \\
-RB
+University of Padova
 }\date{Draft \today}
@@
 \parttitle{Background}
-In clinical trials, the standard method of measuring the effectiveness of one treatment over another is a parametric analysis of covariance (ANCOVA), a special case of linear regression which incorporates stratification.
-The method relies on strong, untestable assumptions about the data-generating process such as constant treatment effects across groups, normality of errors, and homoskedasticity.
+In clinical trials, the standard method of measuring the effectiveness of one treatment over another is parametric analysis of covariance (ANCOVA), a special case of linear regression which incorporates stratification.
+The hypothesis test is derived based on strong, untestable assumptions about the data-generating process such as a constant treatment effect across groups and Gaussian, homoskedastic random errors.
 The effectiveness of the treatment is decided by the $p$-value from this hypothesis test, which may not be valid if the assumptions for ANCOVA are not met.
@@
 \parttitle{Methods}
-Permutation tests present such an opportunity: they require only minimal assumptions which are often guaranteed by the randomization that was conducted.
-In contrast to parametric tests, which rely up on strong assumptions to obtain a null distribution with a simple closed form, with permutation tests one uses the assumption of exchangeability or randomization to construct approximate null distributions by simulation.
+Permutation tests are a nonparametric hypothesis test and require only minimal assumptions which are often guaranteed by the randomization that was conducted.
+In contrast to parametric tests, which rely up on strong assumptions to obtain a null distribution with a simple closed form, one uses the assumption of exchangeability or randomization to simulate approximate null distributions for permutation tests.
 These tests are exact, even in finite samples.
-We compare ANCOVA to several permutation tests based on linear models, all of which control for pretreatment covariates to increase power and precision.
+We compare ANCOVA to several permutation tests based on linear models, which control for pretreatment covariates to increase power and precision.
 
 \parttitle{Results}
-Our simulations show that the permutation methods maintain power comparable to the ANCOVA in randomized trials.
-We illustrate the use of permutation tests alongside the parametric ANCOVA using data from a clinical trial conducted across eight sites,
+We simulate randomized trials using a variety of data-generating processes, some of which violate the parametric ANCOVA assumptions.
+The permutation tests based on linear models maintain power comparable to the ANCOVA in every case.
+We illustrate the use of these permutation tests alongside the parametric ANCOVA with data from a clinical trial conducted across eight sites,
 each having a different numbers of patients.
@@
 A hypothesis test is a statistical method for determining whether observed data is consistent with a belief about the process that generated the data.
-Medical experiments use hypothesis testing to assess the evidence that a treatment has an effect on one or more clinically relevant outcomes.
+Medical experiments use hypothesis testing to assess the evidence that a treatment affects one or more clinically relevant outcomes.
 The simplest version of this experiment, which involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome, has been studied for nearly a century (see \cite{fisher_design_1935} and \cite{neyman_application_1923} for early references).
@@
 Randomized experiments in the real world are rarely this simple:
-individuals are heterogeneous with respect to pretreatment covariates,
-there may be more than two treatments under study,
-they may not take the treatment they are assigned or may drop out of the study,
+if individuals are heterogeneous before the study, then their outcomes may differ for reasons besides the treatment;
+there may be more than two treatments under study;
+patients may not take the treatment they are assigned or may drop out of the study;
 and an individual's assignment to treatment may affect others' outcomes.
-Techniques have been developed to deal with each of these issues.
-In this paper, we focus on the first.
+Techniques have been developed to deal with each of these issues, but solutions are rarely as simple as an asymptotic normality argument.
+In this paper, we focus on the first question of heterogeneity and assume that the others do not affect the study.
 
@@
 However, in any particular randomization, there may be imbalances.
-If the imbalanced covariates are associated with the outcome, then even when there is no effect of treatment, we may observe differences in outcomes between treatment groups.
-Controlling for such baseline covariates can reduce the variability of estimated treatment effects and yield more powerful hypothesis tests. \todo{cite}
+If the imbalanced variables are associated with the outcome, then even when treatment has no effect, there may be differences in outcomes between treatment groups.
+Adjusting for such covariates can reduce the variability of treatment effect estimates and yield more powerful hypothesis tests.
 
-Stratification is one method for controlling for covariates that are known a priori to be associated with the outcome.
-Individuals are divided into groups based on levels of this covariate, and random assignment of treatments is conducted within each group, independently across groups.
+Stratification is one method to control for covariates that are known a priori to be associated with the outcome.
+Strata are groups of individuals with similar levels of a covariate, and random assignment of treatments is conducted within each group, independently across groups.
+The stratification is done \textit{before any outcome data is collected}.
 This guarantees that the stratification variable is balanced between treatment groups.
-The stratification is done \textit{before any outcome data is collected}.
 A common stratification variable in clinical experiments is location:
-individuals often come from many sites because it is difficult to recruit a sufficient number of participants at one doctor's office or hospital.
-As with other pretreatment covariates, outcomes may vary across strata due to factors unrelated to individuals.
-Controlling for stratum can have the same beneficial effects as controlling for other pretreatment covariates.
+individuals often come from many sites because it is difficult to recruit a sufficient number of participants at one doctor's office or hospital,
+especially when the object of study is a rare disease or a rare outcome.
+Controlling for stratum can have the same beneficial effects as controlling for other pretreatment covariates, 
+as outcomes may vary across strata due to factors unrelated to individuals.
 
-Linear regression is another common way to control for baseline covariates.
-It is done \textit{after} data are collected and outcomes are observed.
-Whereas stratification guarantees balance on the stratifying variable, other covariates may be imbalanced, so stronger modeling assumptions are needed to balance them after the fact.
-The idea of linear regression is to project the outcomes onto a plane which describes each variable's relationship to the outcome.
-The coefficient of any particular variable answers the question, if we were to hold fixed all other covariates and increase this variable by one unit, how much would we expect the outcome to change?
+Linear regression is another way to control for baseline covariates.
+It is done \textit{after} data are collected and outcomes have been observed.
+While stratification guarantees balance on the stratifying variable, other covariates may still be imbalanced.
+Stronger modeling assumptions may be necessary to balance them after the fact.
+Linear regression is a method which projects the outcomes onto the plane that best summarizes each variable's relationship to the outcome.
+The coefficient of any particular covariate answers the question, if we were to hold fixed all other variables and increase this variable by one unit, how much would we expect the outcome to change?
 This model posits a linear relationship between covariates, treatment, and outcome; 
-if the true relationship is not linear, then regression only estimates the best linear approximation to the conditional expectation of the outcome.
-When controlling for strata, such as location, it is standard to use analysis of covariance (ANCOVA), a particular case of linear regression which allows each stratum to have its own mean outcome level.
-This amounts to choosing the plane of best fit for the data, allowing each stratum to have a different intercept but enforcing that the slope is the same for each.
+if the true relationship is not linear, then linear regression gives the best linear approximation to the conditional expectation of the outcome.
+When controlling for strata, such as location, it is standard to use analysis of covariance (ANCOVA).
+ANCOVA is a particular version of linear regression that allows each stratum to have its own mean outcome level.
+This amounts to fitting a plane to each stratum, allowing each plane to have a different intercept but requiring them to have a common slope.
 
-Hypothesis testing of estimated coefficients requires even stronger assumptions than linearity, such as Gaussian, homoskedastic errors.
+Hypothesis testing of estimated coefficients requires even stronger assumptions than linearity, namely Gaussian, homoskedastic errors.
 When this holds, a regression coefficient scaled by its estimated variance follows Student's $t$ distribution.
-Thus, if we believe the linear model is correctly specified, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the ANCOVA, and can be evaluated analytically and quickly.
-However, the assumptions necessary for the coefficient to be $t$ distributed are clearly violated in many cases:
+Thus, if we believe the linear model is correctly specified, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the ANCOVA model can be evaluated analytically and quickly.
+However, the modeling assumptions are clearly violated in many cases:
 when the data are discrete or ordinal, when the treatment has a differential effect across strata, and when measurement error sizes differ across strata.
 Furthermore, the hypothesis test implicitly assumes that the data were sampled at random from some underlying population, when in fact,
-medical experimenters rarely recruit patients this way.
+medical experimenters rarely recruit patients this way (\cite{ludbrook_why_1998}).
 
@@
 and the null hypothesis tells us what each individual's response would be regardless of the assignment (namely, it would be the same).
-One determines how ``extreme'' the observed test statistic is relative to this randomization distribution, rather than a parametric reference distribution, such as Student's $t$ or the Gaussian.
-Such a test is exact, meaning that it controls the type I error rate at the pre-specified level, even in finite samples, whereas parametric hypothesis tests based on asymptotic approximations do not always guarantee good finite sample properties.
+One determines how ``extreme'' the observed test statistic is relative to this randomization distribution, rather than a parametric reference distribution like Student's $t$ or the standard Gaussian.
+Such a test is exact, meaning that it controls the type I error rate at the pre-specified level even in finite samples, whereas parametric hypothesis tests based on asymptotic approximations do not always guarantee good finite sample properties.
 Permutation tests achieve this by operating conditional on the observed sample; the resulting inference applies to the sample at hand, but does not necessarily generalize to other groups of individuals.
+This may be acceptable when the individuals who make up the sample are not representative of any well-defined population.
+
 In the past, parametric tests were necessary because asymptotic approximations were the only computationally feasible way to estimate distributions. 
@@
 
-However, simulations show that parametric and nonparametric tests seem to perform similarly in most cases, even when data violate the assumptions of the parametric method.
+However, parametric and nonparametric tests seem to perform similarly when compared side-by-side in simulations, even when data violate the assumptions of the parametric method.
 Medical trials studying pain or perceived severity scores often use Likert scale data, which is discrete and does not match the normality assumptions of parametric tests.
@@
 
-In this paper, we review hypothesis tests for a treatment effect which incorporate covariate adjustment to increase power.  
-We focus on ANCOVA and its permutation counterparts, comparing their performance in different scenarios, and illustrating their application with a clinical dataset.
+In this paper, we review several hypothesis tests for a treatment effect which adjust for pretreatment covariates to increase power.  
+We focus on ANCOVA and its permutation counterparts, comparing their performance in different scenarios and illustrating their application with a clinical dataset.
 We find in simulations that even when assumptions are not satisfied, the parametric and permutation tests have comparable power to detect a treatment effect.
-We apply each of the methods to data from a clinical trial seeking to compare the performance of two treatments for gastroesophageal reflux disease (GERD).
+We apply each of the methods to data from a clinical trial which assessed the performance of two treatments for gastroesophageal reflux disease (GERD).
 We conclude by discussing implications of these results for practitioners.
@@
 
-\subsection*{Problem and notation \todo{what do others call this?}}
+\subsection*{Notation}
 Suppose we randomly assign two treatments labelled $0$ and $1$ to a group of individuals.
@@
 For expository clarity, we suppose that $\mathbf{X}$ is univariate, but all results are easily extended to the case when $\mathbf{X}$ is multivariate.
-Suppose there is a categorical pretreatment variable with $J$ levels used to stratify individuals.
+Suppose further that there is a categorical pretreatment variable with $J$ levels used to stratify individuals.
 
@@
 We can never observe both; random assignment of treatment reveals $Y_{ij} = Z_{ij}Y_{ij}(1) + (1-Z_{ij})Y_{ij}(0)$.
+Throughout, we assume that there is no interference between individuals 
+(in other words, $Y_{ij}$ is a function of $(Z_{ij}, Y_{ij}(1), Y_{ij}(0))$ and not any other $Z_{i', j'}$)
+and that there is no non-compliance or study dropout 
+(we actually observe $Y_{ij} = Y_{ij}(Z_{ij})$ for all $i, j$).
 
 We are interested in the differences in potential outcomes $Y_{ij}(1) - Y_{ij}(0)$.
-This difference represents the effect of treatment for individual $i$ in group $j$: it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.\footnote{
-Depending on the method of analysis one chooses, various functions of these differences will be considered.
-Any of these quantities may be of clinical interest; it is a matter of philosophy which one prefers.}
-A typical problem is to estimate the average of differences $Y_{ij}(1) - Y_{ij}(0)$.
+This quantity represents the effect of treatment for individual $i$ in group $j$: 
+it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.\footnote{
+Which function of these differences will be considered depends on the chosen method of analysis.
+Any of these possible quantities may be of clinical interest; it is a matter of philosophy which one prefers.}
+A typical problem is to estimate the average of differences $Y_{ij}(1) - Y_{ij}(0)$ in some group, such as the study sample or in a target population.
 Instead of estimation, we will focus on hypothesis testing for whether these differences are nonzero.
@@
 }
-Under the null hypothesis, the test statistic 
-$$ T = \frac{\hat{\gamma}}{\sqrt{ \hat{\sigma}_{\hat{\gamma}}^2}},$$
-where $\hat{\sigma}_{\hat{\gamma}}^2$ is the estimated standard error of $\hat{\gamma}$,\footnote{
+From this procedure, we may estimate $\hat{\sigma}_{\hat{\gamma}}^2$, the  standard error of $\hat{\gamma}$\footnote{
 $\hat{\sigma}_{\hat{\gamma}}^2$ is estimated by first computing the residual standard error $\hat{\sigma}^2$,
 then taking the diagonal element corresponding to the treatment variable in the inverse covariance matrix of covariates.
-}
+}.
+Under the null hypothesis, the test statistic 
+$$ T = \frac{\hat{\gamma}}{\sqrt{ \hat{\sigma}_{\hat{\gamma}}^2}}$$
 follows the Student $t$ distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
-The $p$-value for this hypothesis test is the probability that a value drawn from this $t$ distribution is larger in magnitude than $T$.
+The $p$-value for this hypothesis test is the probability that a value drawn from the $t$ distribution is larger in magnitude than $T$.
 
-The ANCOVA model allows for fixed stratum effects, but they are only additive in the outcome.  
-The model does not account for differences in the treatment effect across strata.  
-The estimated linear model may not accurately capture details of the true data-generating process.  
-If effects are not constant across strata, then the coefficient $\gamma$ may be attenuated towards 0.  
+The ANCOVA model allows for fixed stratum effects that are additive in the outcome.  
+The model does not account for variation in the treatment effect across strata.  
+The estimated linear model may not accurately capture details of the true data-generating process:
+if effects are not constant across strata, then the coefficient $\gamma$ may be attenuated towards 0.  
 A test of the null hypothesis that $\gamma = 0$ will be valid, but will not reflect the true magnitude of treatment effects among individuals.
@@
 Suppose we wish to test the null hypothesis that individual by individual, treatment has no effect.
-This is referred to as the ``sharp null'' hypothesis:
+This is referred to as the ``sharp'' null hypothesis:
 
-$$H_0: Y_{ij}(1) = Y_{ij}(0), i = 1, \dots, n_j, j = 1,\dots, J.$$
+$$H_0: Y_{ij}(1) = Y_{ij}(0), \forall i = 1, \dots, n_j, j = 1,\dots, J.$$
 
-Then, which treatment that individual $ij$ received amounts to an arbitrary label.
-Once we observe their response under one treatment, we can impute the other potential outcome; namely, it would have been the same.
-This is a stronger assertion than the null hypothesis for the parametric ANCOVA, which only considers the average treatment effect.
+Then, which treatment an individual received amounts to an arbitrary label.
+Once we observe one response under a particular treatment, we can impute the other potential outcome; namely, it would have been the same.
+This null hypothesis is stronger than the null hypothesis for the parametric ANCOVA, which only asserts that the treatment has no effect \textit{on average}.
 
@@
 The most commonly used statistic is the difference in mean outcomes of people who received treatment 1 and the mean outcomes of people who received treatment 0.  
-This statistic is readily interpretable (``on average, taking treatment changes the outcome by x amount'') and has nice theoretical properties owing to it being the difference of two sums.
+This statistic is unbiased and readily interpretable in randomized experiments (``on average, taking treatment changes the outcome by x amount''), and has nice theoretical properties owing to it being the difference of two sums.
 However, the difference in means may not be optimal if we want to be sensitive to heterogeneous effects.  
@@
 
-One may want to stratify the sample according to important confounding variables, then compute a statistic for each stratum separately. 
+To avoid this, one may want to stratify the sample according to important confounding variables, then compute a statistic for each stratum separately. 
 There is a great degree of freedom in choosing how exactly to construct such a statistic.
-Two factors must be decided: how to stratify and how to combine the individual strata statistics.
-There is a tradeoff in how finely we choose to stratify:
-on the one hand, the strata must be sufficiently large to be informative,
+Two factors must be decided: how to stratify and how to combine the statistics across strata.
+There is a tradeoff in how finely we choose to stratify.
+On the one hand, the strata must be sufficiently large to be informative,
  but on the other hand, must be fine enough to capture variation in treatment effects.
@@
 
-One way to construct a test statistic is to combine the statistics directly, for instance taking the sum of absolute values, into a single statistic on the data.
+One way to construct a test statistic is to directly combine the statistics into a single statistic on the data, for instance taking the sum of their absolute values.
 Taking the absolute value before summing ensures that effects with different signs do not cancel each other out.
 The permutation test is conducted identically, using this statistic in place of the difference in means.
-Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by \cite{pesarin_permutation_2010}.
-In this framework, we would break down the ``global'' null hypothesis of no effect whatsoever into the intersection of null hypotheses of no treatment effect within each stratum.
-For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies.
+Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by Pesarin and Salmaso \cite{pesarin_permutation_2010}.
+In this framework, one would break down the ``global'' null hypothesis of no effect whatsoever into the intersection of ``partial'' null hypotheses of no treatment effect within each stratum.
+For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies\footnote{ 
+We considered using the difference in means within each stratum, aggregated by taking the sum of their absolute values over stratum.
+However this would not be ideal for imbalanced designs, which are common in real-world applications, so we did not consider it in the simulations.
+See the clinical data analysis file for a more detailed treatment.}.
 In this case, strata are randomized independently so there are no dependencies to preserve, making this method is equivalent to the first.
+In general, these tests based will not be equivalent.
 
@@
 One would like to construct a more powerful test by incorporating these covariates in a way to reduce the variance of the statistic.
-In particular, we can turn the parametric ANCOVA into a nonparametric permutation test.
-We will model the outcome as in Equation~\ref{eqn:ancova}.
-In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes.
-This makes several things ``permutable'' and allows us to construct several different tests.
+In particular, we will continue to model the outcome as in Equation~\ref{eqn:ancova}, but use fewer assumptions about the data to conduct a nonparametric permutation test.
+In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes,
+making several things ``permutable'' and allowing us to construct several different tests.
 
 First, we may do a simple variation on the stratified permutation test described above.
-Instead of using the difference in means as the statistic, we use the $t$ statistic for the coefficient on $\mathbf{Z}$ in a linear regression.
+Instead of using the difference in means as the statistic, we use the $t$ statistic for the coefficient on $\mathbf{Z}$ in the linear regression.
 If treatment was assigned at random within each stratum, then it is guaranteed that $\mathbf{Z}$ and $\mathbf{\varepsilon}$ are statistically independent, conditional on stratum.
-Therefore, we may conduct a test by permuting treatment assignments $\mathbf{Z}$ within site, independently across sites, and calculating a test statistic for each such permutation.
+Therefore, we may permute treatment assignments $\mathbf{Z}$ within strata, independently across strata, and calculate a test statistic for each such permutation to obtain a null distribution.
 
 Freedman and Lane \cite{freedman_nonstochastic_1983} propose an alternative test based on the residuals of the linear regression.
-Let's take an alternative view of the problem.
-We still write the outcome in the form of Equation~\ref{eqn:ancova}, but now, we do not treat the $\varepsilon_{ij}$ as random.
-They are simply defined to be the difference between the observed outcome $Y_{ij}$ and the data's linear projection onto the plane $\alpha_j + \beta X_{ij}+ \gamma Z_{ij}$.
+They take an alternative view of the problem, still writing the outcome in the form of Equation~\ref{eqn:ancova}, but instead of treating the $\varepsilon_{ij}$ as random they define the errors to be the difference between the observed outcome $Y_{ij}$ and the data's linear projection onto the plane $\alpha_j + \beta X_{ij}+ \gamma Z_{ij}$.
+The $\mathbf{\varepsilon}$ are fixed approximation errors in this framework.
 
 If the null hypothesis is true, then $\gamma = 0$ and $\varepsilon_{ij} = Y_{ij} - \alpha_{ij} - \beta X_{ij}$ for all $i = 1, \dots, n_j, j = 1, \dots, J$.
-We may regress $\mathbf{Y}$ on $\mathbf{X}$ to obtain coefficient estimates $\hat{\alpha}_j$ and $\hat{\beta}$.
-Then we would predict the outcome to be $\hat{Y}_{ij} = \hat{\alpha}_j + \hat{\beta}X_{ij}$
-and we may estimate the errors $\mathbf{\hat{\varepsilon}}$ by $\mathbf{Y} - \mathbf{\hat{Y}}$.
-The $\mathbf{\hat{\varepsilon}}$ approximate the true errors $\mathbf{\varepsilon}$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit.
+We may regress $\mathbf{Y}$ on $\mathbf{X}$ \textit{and not $\mathbf{Z}$} to obtain coefficient estimates $\hat{\alpha}_j$ and $\hat{\beta}$.
+Then we estimate the errors $\mathbf{\hat{\varepsilon}}$ by $\mathbf{Y} - \mathbf{\hat{Y}}$, where $\mathbf{\hat{Y}}$ is the vector of fitted values $\hat{Y}_{ij} = \hat{\alpha}_j + \hat{\beta}X_{ij}$.
+The $\mathbf{\hat{\varepsilon}}$ approximate the true errors $\mathbf{\varepsilon}$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit\footnote{
+\cite{freedman_nonstochastic_1983} advise against using this method if there are large outlier values in the covariate $\mathbf{X}$ or when $\mathbf{X}$ and $\mathbf{Z}$ are highly colinear.
+}.
 Furthermore, under the null hypothesis, the $\mathbf{\varepsilon}$ are independent $\mathbf{Z}$ within strata. 
@@
 
-We construct a permutation distribution using several steps:
+To summarize the steps for constructing a permutation distribution:
 
 \begin{enumerate}
-\item Estimate $\mathbf{\hat{\varepsilon}}$ by $Y_{ij} - \hat{\alpha}_j - \hat{\beta} X_{ij}$ for all $i$ and $j$, where $\hat{\alpha}_j$ and $\hat{\beta}$ are obtained by regressing $\mathbf{Y}$ on $\mathbf{X}$ and stratum assignment.
-\item Construct permuted errors $\mathbf{\hat{\varepsilon}}^\pi$ by permuting the $\mathbf{\hat{\varepsilon}}$ within strata.
-\item Construct permuted responses $Y_{ij}^\pi = \hat{\alpha}_j + \hat{\beta} X_{ij}+ \hat{\varepsilon}_{ij}^\pi$.
-\item Regress $\mathbf{Y}^\pi$ on $\mathbf{X}$, $\mathbf{Z}$, and stratum. The test statistic is the $t$-statistic for the coefficient of $\mathbf{Z}$.
+\item Estimate $\mathbf{\hat{\varepsilon}}$ by $Y_{ij} - \hat{\alpha}_j - \hat{\beta} X_{ij}$ for all $i$ and $j$, where $\hat{\alpha}_j$ and $\hat{\beta}$ are obtained by regressing $\mathbf{Y}$ on $\mathbf{X}$ and stratum assignment, \textit{but not $\mathbf{Z}$}.
+\item Obtain permuted errors $\mathbf{\hat{\varepsilon}}^\pi$ by permuting the $\mathbf{\hat{\varepsilon}}$ within strata.
+\item Compute permuted responses $Y_{ij}^\pi = \hat{\alpha}_j + \hat{\beta} X_{ij}+ \hat{\varepsilon}_{ij}^\pi$.
+\item Regress $\mathbf{Y}^\pi$ on $\mathbf{X}$, $\mathbf{Z}$, and stratum. The test statistic is the $t$ statistic for the coefficient of $\mathbf{Z}$.
 \end{enumerate}
 
-Others have developed different variants on approximate permutation tests based on linear regression. 
+Others have developed variants on these approximate regression-based permutation tests. 
 There is some disagreement on what constitutes an appropriate permutation scheme;
@@
 only the estimates $\hat{\alpha}_j$ and $\hat{\beta}$.
-\cite{manly_randomization_2006} proposes randomizing the outcomes $\mathbf{Y}$, treating them as though they were randomly assigned to pairs $(X, Z)$ under the null hypothesis.
+\cite{manly_randomization_2006} proposes randomizing the outcomes $\mathbf{Y}$, treating them as though they were randomly assigned to pairs $(X_{ij}, Z_{ij})$ under the null hypothesis.
 Permutations under the null treat all coefficients as 0, which may not reflect the true data-generating process.
@@
 Both methods attempt to measure the correlation between treatment and unexplained variation in outcomes, but
-instead of regressing pseudo-outcomes $\mathbf{Y}^\pi$ on covariates to obtain a permutation distribution, he recommended regressing $\mathbf{Z}$ on the permuted residuals $\mathbf{\hat{\varepsilon}}^\pi$ and using the $t$ statistic from the resulting coefficient as the test statistic.
+instead of regressing pseudo-outcomes $\mathbf{Y}^\pi$ on covariates to obtain a permutation distribution, \cite{kennedy_randomization_1995} recommends using the $t$ statistic from regressing $\mathbf{Z}$ on the permuted residuals $\mathbf{\hat{\varepsilon}}^\pi$.
 
-Several authors have compared these tests, theoretically and through simulations \cite{anderson_empirical_1999, anderson_permutation_2001, kennedy_randomization_1996}.
-They find that the Freedman-Lane test of residuals is asymptotically equivalent to the ``oracle'' exact hypothesis test, which we could conduct if we knew which permutations of $\mathbf{Y}$ given $\mathbf{X}$ were equiprobable under the null hypothesis \cite{anderson_permutation_2001}.
-This confirms empirical results, which show that the Freedman-Lane test performs better than other linear model based tests in simulations simulations \cite{anderson_empirical_1999}, though its advantage is small.
-Thus, throughout the rest of the paper, we focus on the two linear regression permutations we described in detail: linear regression after permuting treatment assignments and the Freedman-Lane method of permuting residuals.
+Several authors have compared these tests theoretically and through simulations \cite{anderson_empirical_1999, anderson_permutation_2001, kennedy_randomization_1996}.
+They find that the Freedman-Lane test of residuals is asymptotically equivalent to the ``oracle'' exact hypothesis test which we could conduct if we knew which permutations of $\mathbf{Y}$ given $\mathbf{X}$ were equiprobable under the null hypothesis \cite{anderson_permutation_2001}.
+This confirms empirical results, which show that the Freedman-Lane test performs better than other linear regression based tests in simulations \cite{anderson_empirical_1999}, though its advantage is small.
+Therefore, throughout the rest of the paper we focus on the two linear regression permutations we described in detail: linear regression after permuting treatment assignments and the Freedman-Lane method of permuting residuals.
 
@@
 
-To compare the performance of these tests, we simulated data coming from a randomized experiment, using several different data-generating processes.
+We simulated data coming from a randomized experiment using several different data-generating processes.
 We applied the tests to the data and compared their empirical power over repeated random treatment assignments and random errors.
 We compared the following five tests:
-the $t$-statistic from the parametric ANCOVA,
-a stratified permutation test using the difference in mean outcomes\footnote{ We considered using the difference in means within each stratum, aggregated by taking the sum of their absolute values over stratum.
-However this would not be ideal for imbalanced designs, which are common in real-world applications, so we did not consider it in the simulations.
-See the data analysis for details.}
+the $t$ test from the parametric ANCOVA,
+a stratified permutation test using the difference in mean outcomes
  (called ``stratified permutation'' in what follows),
-a stratified permutation test difference mean differences, taken by subtracting the baseline measure from the outcome (called ``differenced permutation'' in what follows),
-a stratified permutation test based on the $t$-statistic from the linear regression of outcome on control variables (called ``LM permutation'' in what follows),
+a stratified permutation test using the mean change scores, defined as the difference between the baseline measure and the outcome (called ``differenced permutation'' in what follows),
+a stratified permutation test based on the $t$ statistic from the linear regression of outcome on control variables (called ``LM permutation'' in what follows),
 and the Freedman-Lane permutation test.
 
-We assume the following linear data-generating process:
-
+Outcomes were generated according to the linear equation 
 \begin{equation}\label{eqn:dgp}
@@
 and $\varepsilon_{ij}$ is an error term.
-Suppose there are three strata with $\beta_1 = 1, \beta_2 = 1.5,$ and $\beta_3 = 2$.
-Assume that there are $n_j = 16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment within each stratum.
+We used three strata with $\alpha_1 = 1, \alpha_2 = 1.5,$ and $\alpha_3 = 2$.
 
-We use two designs:
+We used two designs:
 \begin{itemize}
-\item Constant additive treatment effect: $\gamma_1 = \gamma_2 = \gamma_3 = \gamma$. This is the implicit assumption when doing inference on estimated treatment effects.
+\item Constant additive treatment effect: $\gamma_1 = \gamma_2 = \gamma_3 = \gamma$. This is the implicit assumption when using a linear model.
 \item Treatment effect in a single stratum: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$.
 This is a constant, additive treatment effect in stratum 1, but no treatment effect in strata 2 and 3.
-This is a simplistic case of heterogeneous treatment effects.
+This is a simplistic example of a heterogeneous treatment effect.
 The standard ANCOVA model does not account for this scenario.
@@
 
-\noindent We draw the baseline measurements once, treating them as fixed, and condition on observing $\{ Y_{ij0} : {i = 1,\dots,16, j = 1,\dots, 3}\}$.
-Then, for 1000 trials, we randomly assign treatment to half of the individuals within each stratum, generate random errors, and construct new outcomes $Y_{ij1}$ using Equation~\ref{eqn:dgp}.
-We conduct all five tests on this new data, obtaining a two-sided $p$-value for each.
-The power curves are estimated by computing the proportion of observed $p$-values less than or equal to $\alpha$, for $\alpha = 0.01, 0.02, \dots, 1$.
+\noindent We drew the baseline measurements once, treating them as fixed, and conditioned on observing $\{ Y_{ij0} : {i = 1,\dots,n_j, j = 1,\dots, 3}\}$.
+Then, for 1000 trials, we randomly assigned treatment to half of the individuals within each stratum, generated random errors, and constructed new outcomes $Y_{ij1}$ using Equation~\ref{eqn:dgp}.
+We conducted all five tests on this new data, obtaining a two-sided $p$-value for each.
 
-In our first set of simulations, we assume that the baseline measurements $Y_{ij0}$ are standard normally distributed and the pairs $(Y_{ij0}, \varepsilon_{ij})$ are independent across $i$ and $j$.
-We will assume that $\beta_0 = 1$ and let the treatment effect be $\gamma = 1$
-For each of the two simulation designs that vary the treatment effect, we use two distributions of $\mathbf{\varepsilon}$.
-In the first case, we use $\varepsilon \sim N(0, 1)$ to fit with the parametric ANOVA assumption of normal errors.
-In the second case, $\mathbf{\varepsilon}$ comes from a $t$ distribution with 2 degrees of freedom so the errors are heavy-tailed.
-Thus, there are four total simulation designs.
+In our first set of simulations, we assumed that the baseline measurements $Y_{ij0}$ were standard normally distributed and the pairs $(Y_{ij0}, \varepsilon_{ij})$ were independent across $i$ and $j$.
+We let $\beta_0 = 1$ and the treatment effect be $\gamma = 1$.
+We used a balanced design with $n_j = 16$ individuals per stratum and balanced treatment assignment, i.e. 8 people received each treatment within each stratum.
+There were four total simulation designs:
+for each of the two possible treatment effects, we used two distributions of $\mathbf{\varepsilon}$.
+In the first case, we used $\varepsilon \sim N(0, 1)$ to fit with the parametric assumption of normal errors.
+In the second case, $\mathbf{\varepsilon}$ were drawn from a $t$ distribution with 2 degrees of freedom so the errors were heavy-tailed.
 
@@
 The best case is when the errors are Gaussian and the treatment effect is constant across strata, while the worst case is when the errors are heavy-tailed and the treatment effect only appears in one stratum.
-Intuitively, it makes sense that power decreases relative to the Gaussian, constant treatment effect case, as we increasingly obscure the treatment effect by modifying the design.
-A consistent pattern appears in each case: the stratified permutation test has the lowest power, while the other four tests tend to have very similar power.
+Intuitively, it makes sense that power decreased relative to the Gaussian, constant treatment effect case, as each violated assumption further obscured the treatment effect.
+A consistent pattern appears in each design: the stratified permutation test has the lowest power, while the other four tests tend to have similar power.
 
@@
 If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.}
-These numbers show that actually, for $\alpha=0.05$, the discrepancy between the stratified permutation test and the others is not so large when the effect is isolated in a single stratum.
-In this case, the permutation test may have slightly higher power than the parametric ANCOVA at small significance levels like those used in practice.
+These numbers show that actually, for $\alpha=0.05$, the stratified permutation test does not lose a large amount of power when the effect is isolated in a single stratum.
+In this case, the permutation test may have slightly higher power than the parametric ANCOVA at the small significance levels used in practice.
 \begin{figure}[h]
@@
 
-In our second set of simulations, we used the same design as the first simulations, but modified the sizes of the treatment groups.
-Within each stratum, 4 individuals received one treatment and 12 received the second treatment.
+Our second set of simulations used the same design as the previous ones, but modified the sizes of the treatment groups.
+Within each stratum, 4 individuals received one treatment and 12 received the other.
 This mimics real-world experiments, where the more expensive treatment is administered less frequently than the other.
-It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exaccerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
-Here, the data have the same variance, so the test size should not be a problem, but power may suffer.\footnote{
-See the supplementary simulation files for a treatment of heterogeneous variances.}
+It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exacerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
+Here, the data have the same variance so we would expect power, not the test level, to be an issue.\footnote{
+See the supplementary Gaussian simulation files for a treatment of heterogeneous variances.}
 We omit a plot of the power curves from these simulations, as the patterns appear similar to those in Figure~\ref{fig:normal_sim_power}.
@@
 
-In our third set of simulations, we assume that the baseline measurements $Y_{ij0}$ are discrete and skewed.
-This is representative of some survey data, where individuals are asked to rate their experience on a scale from 1 to 10, for instance.
-We generate the baseline measures using independent draws from a Poisson distribution with mean 4, and censoring values larger than 10, instead setting them to 10.
-We condition on these observed baseline measures after we have generated them once.
-Then, for each 1000 trials, we randomly assign treatment to half of the individuals in each stratum and generate errors taking on values $0.5$ and $-0.5$ with equal probability.
-We construct $Y_{ij1}$ using Equation~\ref{eqn:dgp} with a treatment effect of $\gamma = 0.5$.
-However we don't observe $Y_{ij1}$ but instead $\tilde{Y}_{ij1}$, which is defined as
+In our third set of simulations, we made the baseline measurements $Y_{ij0}$ discrete and skewed.
+This type of data occurs in some surveys where individuals are asked to rate their symptom severity or some experience on a scale from 1 to 10, for instance.
+We generated the baseline measures using independent draws from a Poisson distribution with mean 4 and censored values at 10.
+We conditioned on these observed baseline measures after we having generated them once.
+Then, for each 1000 trials, we randomly assigned treatment to half of the individuals in each stratum and generated errors taking on values $0.5$ and $-0.5$ with equal probability.
+We constructed $Y_{ij1}$ using Equation~\ref{eqn:dgp} with a treatment effect of $\gamma = 0.5$.
+However, we didn't observe $Y_{ij1}$ but instead integer values $\tilde{Y}_{ij1}$ defined as
 
@@
 The observed outcomes $\tilde{Y}_{ij1}$, the baseline measures $Y_{ij0}$, and the errors are discrete.
-The assumption of normality is clearly violated here.
+The normality assumption is clearly violated here.
 
-We examine the effect of stratum size, rather than error distribution, in this set of simulations.
-First, we consider balanced samples as before, where each stratum contains 16 individuals.
-Next, we consider imbalanced strata, where the smallest stratum has only 8 individuals, the next has 16, and the largest has 24.
-We again consider the four combinations of treatment effects and sample sizes.
-In the case where there is only a non-null treatment effect in one stratum and sample sizes are different, this effect occurs in the smallest stratum.
+We examined the effect of stratum size in this set of simulations.
+For each of the two treatment effect designs, we considered both equal and unequal stratum sizes.
+First, we considered balanced samples as before, where each stratum contained $n_j = 16$ individuals.
+Next, we considered imbalanced strata, where the smallest stratum had only $n_1=8$ individuals, the next had $n_2= 16$, and the largest had $n_3=24$.
+The design with a single stratum effect and unequal stratum sizes was particularly unfavorable for hypothesis testing:
+the nonzero effect only occured in the smallest stratum.
 
 Figure~\ref{fig:skewed_sim_power} shows the estimated power curves in these four designs.
-When the treatment effect is constant, sample sizes within each stratum don't matter; the power curves look the same up to chance variation.
-When the treatment effect only occurs in one stratum, there is a substantial loss in power because the effect only appears in the smallest stratum.
-In practice, one does not usually know a priori which individuals will be affected by treatment (otherwise, those unaffected by treatment would be excluded from the study altogether),
+When the treatment effect is constant, sample sizes within each stratum don't matter; 
+the two power curve plots in the first row of Figure~\ref{fig:skewed_sim_power} look the same up to chance variation.
+When the treatment effect only occurs in one stratum, there is a substantial loss in power.
+It is nearly impossible to detect a difference in treatments when the affected stratum is too small.
+In practice, one does not usually know a priori which individuals will be affected by treatment (otherwise, those not affected would be excluded from the study altogether).
 This result suggests that when stratifying, one must be cautious not to stratify too finely, otherwise treatment effects may be indistinguishable from random variability.
 
+All of the tests appear to be conservative and have lower than nominal level.
 As in the first set of simulations, the stratified permutation test using the raw outcomes has the lowest power of the five tests.
-Now, however, the stratified permutation test using the difference between outcome and baseline also has lower power than the three using linear models.
-Presumably, this is because the differences $\tilde{Y}_{ij1} - Y_{ij0}$ are also discrete and lie in a small range of values, so the test statistic does not vary greatly across permutations.
+Now, however, the stratified permutation test using the change scores also has lower power than the three linear model tests.
+Presumably, this is because the differences $\tilde{Y}_{ij1} - Y_{ij0}$ are discrete and lie in a small range of values, so the test statistic does not vary greatly across permutations.
 Table~\ref{tab:skewed_power} confirms this: power for the three linear model methods is high when the treatment is constant across strata, and power for all of the methods suffers when the effect is isolated in a single stratum.
-All of the tests have lower than nominal level.
 
@@
 
-It makes sense that in all of these simulations, the permutation test using the difference in outcome and baseline as the response measure is more powerful than the permutation test using the raw outcomes.
-In additional simulations\footnote{see supplementary files on GitHub}, we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had correlation of $0.25$.
-In this case, the result was reversed: the differenced test had low power while the test using the raw outcomes had a power curve closer to the tests that adjust for covariates using a linear model.
-This phenomenon has been studied before by \cite{frison_repeated_1992}, who recommended to use the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
+It makes sense that in all of these simulations, the permutation test using the change scores as the response measure is more powerful than the permutation test using the raw outcomes.
+The baseline and outcome are highly correlated, so the change score has lower variance the outcome alone.
+In additional simulations\footnote{
+See the simulation using Gaussian, balanced design for a more detailed treatment.}, 
+we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had correlation of $0.25$.
+In this case, the result was reversed: 
+the differenced test had low power while the test using the raw outcomes had a power curve closer to the linear model-based tests.
+This behavior has been studied before by \cite{frison_repeated_1992}, who recommended to use the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
 
 \subsection*{Clinical data results}
-We compare the methods we discussed on a dataset from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease (GERD).
+We compared the parametric ANCOVA, the stratified permutation test, and the two linear model-based permutation tests using a dataset from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease (GERD).
 Patients were treated at eight sites in two different countries.
@@
 Patients were observed for a week before receiving treatment and for a week after receiving treatment.
-On each of the fourteen days of observation, patients responded to a survey asking about their heartburn, regurgitation, and dyspepsia frequency and severity.
-These endpoints are measured on a discrete scale.
-There were several additional derived endpoints, such as daily regurgitation, daily ``hrdq'', and daily dyspepsia, which are functions of the survey measures.
+On each of the fourteen days of observation, patients responded to a survey about their heartburn, regurgitation, and dyspepsia frequency and severity.
+These endpoints were measured on a discrete scale.
+There were several additional endpoints, such as daily regurgitation, daily ``hrdq'', and daily dyspepsia, calculated from the survey measures.
 Daily ``hrdq'' was the primary endpoint.
-To reduce day-to-day variation, we averaged the measures in the pretreatment week and post-treatment week for each endpoint.
+To reduce day-to-day variation, we averaged the measures from each week to obtain two time points per patient, one pretreatment and one post-treatment.
 
-In our model, we treat site as the stratification variable, as this is the level at which treatment was randomized.
-We do not include country in the model, as a country-level effect should be captured in the site-level effects.
-The linear model for ANCOVA analysis is as in Equation~\ref{eqn:ancova}.
-We chose to use this form of the model, with the actual outcomes as the $\mathbf{Y}$ and the baseline measures as control variables, rather than testing the difference in outcome and baseline. 
-The outcome and baseline had a low correlation (for instance, $0.56$ for daily ``hrdq''), so we did not use their difference as the dependent variable. 
+We treated site as the stratification variable, as this is the level at which treatment was randomized.
+We did not include country in the model, as a country-level effect should be captured in the site-level effects.
+The model used for the linear regression-based tests was defined as in Equation~\ref{eqn:ancova}.
+This model allowed for the intercept $\alpha_j$ to vary across sites and used the pretreatment, baseline measurement as the control variable $\mathbf{X}$.
+The outcome and baseline had a low correlation (for instance, $0.56$ for daily ``hrdq''), so we used raw outcomes and not the change scores as the dependent variable. 
 As our simulations and previous work \cite{frison_repeated_1992} show, when the correlation between baseline and outcome is low, it is more powerful to control for baseline outcomes using a model.
 
-Figure~\ref{fig:clinical_distr} shows the distribution of each clinical endpoint, separated by treatment group.
-There is a clear visual difference in distributions for daily heartburn (``daily\_heart''), daily ``hrdq'', and heartburn frequency (``heart\_freq'').
+Figure~\ref{fig:clinical_distr} shows the distribution of each clinical endpoint for the two treatment groups.
+There is a clear difference in distributions for daily heartburn (``daily\_heart''), daily ``hrdq'', and heartburn frequency (``heart\_freq'').
 The difference is less clear for daily regurgitation (``daily\_regurg'') and regurgitation frequency (``regurg\_freq'').
@@
 The $p$-values for the ANCOVA tend to be smaller than the linear model and the residual permutation tests, while the $p$-values stratified, unadjusted permutation test have no consistent pattern.
-These results match our simulations: the parametric test has comparable or slightly higher power than the other permutation tests.
+The parametric test has comparable or slightly higher power than the other permutation tests.
 
@@
 Our results match those of \cite{vickers_parametric_2005, anderson_empirical_1999}, where parametric ANCOVA and regression-based $t$ tests performed the same or better than the comparable nonparametric tests.
-We simulated x y and z violations
+We simulated a variety of data-generating processes ranging from the ideal case when all assumptions are met 
+to the case when data are non-normal, discrete, or skewed,
+as well as data with other possible issues such as heterogeneous treatment effects and imbalanced sample sizes.
 When the ANCOVA assumptions were violated, the linear regression-based tests lost power relative to the best case data-generating process where all assumptions were satisfied.
-However, power did not change \textit{relative to each other}.
-It is a matter of philosophy which one chooses: while the parametric test may be robust to violations of its assumptions, it is somewhat reassuring that the permutation test can exactly match the randomization that was done and makes fewer distributional assumptions.
+However, the power of the tests remained the same \textit{relative to each other}.
+It is a matter of taste which test one chooses for their experiment: while the parametric test may be robust to violations of its assumptions, it is somewhat reassuring that the permutation test can exactly match the randomization that was done while making no distributional assumptions.
 
-Permutation tests don't come free of assumptions, though.  \cite{romano_behavior_1990} warns against using permutation tests naively if items are not truly exchangeable. 
+Permutation tests do not come entirely free of assumptions, though.  
+\cite{romano_behavior_1990} warns against using permutation tests naively if items are not truly exchangeable. 
 For instance, he points to the case where observations have unequal variances.  
-This is a problem as we cannot observe the errors; it is a leap of faith to suppose that they are homogeneous.
-\cite{boik_fisherpitman_1987} illustrates this phenomenon, comparing the normal theory $F$ test to its permutation counterpart and demonstrating by simulation that the latter has larger than nominal level when the error variances are heterogeneous.
+This is a problem as we cannot observe the errors; it is a leap of faith to assert that they are homogeneous and therefore exchangeable.
+\cite{boik_fisherpitman_1987} illustrates this phenomenon using the normal theory $F$ test and its permutation counterpart, 
+and demonstrates by simulation that the latter has larger than nominal level when the error variances are heterogeneous.
 
-It is important to note that the permutation tests described here are only approximately exact in the context of randomized experiments.
+It is important to note that the permutation tests described here are approximately exact only in the context of randomized experiments.
 In this case, treatment is assigned at random and is therefore statistically independent of the covariates $\mathbf{X}$ and the errors $\mathbf{\varepsilon}$.
@@
 When exchangeability doesn't hold, we cannot disentangle the effect of $\mathbf{Z}$ from the effect of $\mathbf{X}$.
-In observational studies, one must provide evidence that items are exchangeable in order to achieve approximate exchangeability.
-One must argue that the object being permuted is uncorrelated with the other variables being held fixed.
+One must provide evidence that items in observational studies are exchangeable in order to achieve an approximately exact test.
+The onus is on the researcher to make the case that the object being permuted is uncorrelated with the other variables being held fixed.
 This can be done visually using scatterplots and residual plots \cite{freedman_nonstochastic_1983} and from subject-matter knowledge.
 
-One should be mindful of which covariates are important to control for.
-Ideally, one knows all covariates related to the outcome.
+One should be mindful of which covariates to control for.
+Ideally, one knows and observes all covariates related to the outcome.
 In that case, the coefficient for treatment in a fully saturated linear model (including all covariates and their interactions) is an asymptotically consistent estimate of the average treatment effect \cite{lin_agnostic_2013}.
 It is rare that one knows \textit{all} relevant covariates in practice; failure to control for all relevant variables can introduce bias in certain permutation tests.
- \cite{gail_tests_1988} propose a randomization test based on residuals, using not a linear model but an exponential family model to construct residuals, and find that omitting relevant covariates leads to tests with higher than nominal level.
+ \cite{gail_tests_1988} propose a randomization test based on residuals from an exponential family model and find that omitting relevant covariates leads to tests with higher than nominal level.
 
 The method of controlling for baseline covariates matters, too.
-The naive thing to do is to consider change scores, subtracting baseline measures from the outcome and doing hypothesis testing using the change as the dependent variable.
+The naive way is to consider change scores, subtracting baseline measures from the outcome and doing hypothesis testing using the change as the dependent variable.
 Our simulations confirm the suggestion of \cite{frison_repeated_1992} to use change scores only when there is a strong correlation between baseline and outcome.
+Weak correlations between baseline and outcome occur often in practice, as was the case with our GERD dataset.
 If the correlation is weak, then the resulting test using the difference may be less powerful than ignoring the baseline altogether.
@@
 treating the change scores as the dependent variable is a special case of the linear regression, constraining the coefficient on the baseline measure to be $1$.
-Weak correlations between baseline and outcome occur often in practice, as was the case with our GERD dataset.
+The regression approach is more flexible and demonstrably more powerful.
 
@@
 % if your bibliography is in bibtex format, use those commands:
-\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
+\bibliographystyle{vancouver} % Style BST file (bmc-mathphys, vancouver, spbasic).
 \bibliography{references}      % Bibliography file (usually '*.bib' )

### 2017-01-31 ms/ms.tex
Incorporate Fraser's comments

@@
 \begin{abstract}
-\parttitle{Background}
-In clinical trials, the standard method of measuring the effectiveness of one treatment over another is parametric analysis of covariance (ANCOVA), a special case of linear regression which incorporates stratification.
-The hypothesis test is derived based on strong, untestable assumptions about the data-generating process such as a constant treatment effect across groups and Gaussian, homoskedastic random errors.
-The effectiveness of the treatment is decided by the $p$-value from this hypothesis test, which may not be valid if the assumptions for ANCOVA are not met.
-We would like to understand which assumptions may be relaxed or violated for the ANCOVA to remain valid (but potentially less powerful).
-
-\parttitle{Methods}
-Permutation tests are a nonparametric hypothesis test and require only minimal assumptions which are often guaranteed by the randomization that was conducted.
-In contrast to parametric tests, which rely up on strong assumptions to obtain a null distribution with a simple closed form, one uses the assumption of exchangeability or randomization to simulate approximate null distributions for permutation tests.
-These tests are exact, even in finite samples.
-We compare ANCOVA to several permutation tests based on linear models, which control for pretreatment covariates to increase power and precision.
-
-\parttitle{Results}
-We simulate randomized trials using a variety of data-generating processes, some of which violate the parametric ANCOVA assumptions.
+%\parttitle{Background}
+Statistical test of treatment effects based on the linear model, such as analysis of variance (ANOVA) and analysis of covariance (ANCOVA), are widely accepted by organizations that regulate clinical trials.
+In this paper, we focus on ANCOVA, a special case of linear regression which incorporates stratification.
+Traditional ANCOVA is derived using strong assumptions about the data-generating process so that the resulting inference is based on parametric distributions.
+Because these methods are well understood and robust, they are sometimes applied to data that depart from the method's assumptions, such as ordinal integer scores.
+Given the considerable costs and scientific importance of clinical trials, it is desirable to include an additional methodology which is potentially more robust to departures from assumptions as a check on the statistical inference for the main study endpoints.
+%\parttitle{Methods}
+Permutation tests are a nonparametric alternative that require only minimal assumptions which are often guaranteed by the randomization that was conducted.
+In contrast to parametric tests, which rely up on strong assumptions to obtain a null distribution with a simple closed form, permutation tests use the assumption of exchangeability or randomization to obtain null distributions directly by considering what would have been observed under different possible randomizations.
+We compare parametric ANCOVA to several permutation tests based on linear models, which control for pretreatment covariates to increase power and precision.
+%\parttitle{Results}
+We simulate randomized trials using a variety of data-generating processes, some of which violate the parametric assumptions.
 The permutation tests based on linear models maintain power comparable to the ANCOVA in every case.
@@
 each having a different numbers of patients.
-
-\parttitle{Conclusions}
-These results suggest that the parametric ANCOVA and linear model permutation tests perform similarly, even when not all assumptions for the parametric test are met.
+%\parttitle{Conclusions}
+These results suggest that a linear model permutation test may serve as a secondary analysis to give strength to the statistical inferences from the primary analyses of a clinical data study.
 
@@
 and an individual's assignment to treatment may affect others' outcomes.
-Techniques have been developed to deal with each of these issues, but solutions are rarely as simple as an asymptotic normality argument.
-In this paper, we focus on the first question of heterogeneity and assume that the others do not affect the study.
+Techniques have been developed to deal with each of these issues, invoking additional assumptions in order to carry out valid inference.
+%For any particular dataset, it is difficult to determine to what extent the assumptions for asymptotic normality are satisfied.
+In this paper, we focus on the issue of heterogeneity of pretreatment characteristics and assume that the other issues are not present.
 
@@
 However, the modeling assumptions are clearly violated in many cases:
-when the data are discrete or ordinal, when the treatment has a differential effect across strata, and when measurement error sizes differ across strata.
+when the data are discrete or ordinal, 
+when the treatment has a differential effect across subgroups of individuals and no interaction terms are included, 
+and when measurement error sizes differ across strata.
 Furthermore, the hypothesis test implicitly assumes that the data were sampled at random from some underlying population, when in fact,
@@
 Now, computational power is no longer a barrier to finding exact (or exact to pre-specified precision) randomization distributions.
-In most cases, a randomization test is preferable to a parametric one:
+In most cases, a randomization test is the ``gold standard'':
 ``[a] corresponding parametric test is valid only to the extent that it results in the same statistical decision [as the randomization test]'' (\cite{bradley_distribution_1968}).
+Therefore, when conducting creating estimates and confidence intervals using parametric methods, one may have greater confidence in the results when the decision to accept or reject the null hypothesis is the same for the parametric and permutation tests.
 

### 2017-01-31 ms/SBR-template.tex
Start putting MS into SBR template, making changes that the journal requests

--- 2017-01-31 ms/ms.tex
@@
-\documentclass[11pt]{article}
+\documentclass[12pt]{article}
 
-%\documentclass{bmcart}
-
-%%% Load packages
-%\usepackage{amsthm,amsmath}
-%\RequirePackage{natbib}
-%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
-%\RequirePackage{hyperref}
-\usepackage[utf8]{inputenc} %unicode support
-
-\usepackage[breaklinks=true]{hyperref}
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage{enumerate}
+\usepackage{natbib}
+\usepackage{url} % not crucial - just used below for the URL 
 \usepackage{color}
-\usepackage{amsmath,amssymb,amsthm}
-\usepackage[numbers]{natbib}
-\usepackage{array}
-%\usepackage{booktabs, multicol, multirow}
-\usepackage[nohead, margin=1in]{geometry}
-%\usepackage[singlespacing]{setspace}
-\usepackage{graphicx}
-
-%%% Put your definitions there:
-%\startlocaldefs
-
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{assumption}{Assumption}
-
-\newcommand{\beq}{\begin{equation}}
-\newcommand{\eeq}{\end{equation}}
-\newcommand{\bit}{\begin{itemize}}
-\newcommand{\eit}{\end{itemize}}
-
 \newcommand{\todo}[1]{{\color{red}{TO DO: \sc #1}}}
 
-\newcommand{\reals}{\mathbb{R}}
-\newcommand{\integers}{\mathbb{Z}}
-\newcommand{\naturals}{\mathbb{N}}
-\newcommand{\rationals}{\mathbb{Q}}
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{0}
 
-\newcommand{\ind}{\mathbb{I}} % Indicator function
-\newcommand{\pr}{\mathbb{P}} % Generic probability
-\newcommand{\ex}{\mathbb{E}} % Generic expectation
-\newcommand{\var}{\textrm{Var}}
-\newcommand{\cov}{\textrm{Cov}}
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.3in}%
+\addtolength{\topmargin}{-.8in}%
 
-\newcommand{\normal}{N} % for normal distribution (can probably skip this)
-\newcommand{\eps}{\varepsilon}
-\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
-\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
-\newcommand{\argmax}{\textrm{argmax}}
-\newcommand{\argmin}{\textrm{argmin}}
-\renewcommand{\baselinestretch}{1.5}
-
-%\endlocaldefs
-
-\newcommand{\parttitle}[1]{\textbf{#1}}
-
-%%% Begin ...
 \begin{document}
 
-%%% Start of article front matter
-%\begin{frontmatter}
+%\bibliographystyle{natbib}
 
-%\begin{fmbox}
-%\dochead{Research}
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the title of your article here     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\title{A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Control Trials}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors here                   %%
-%%                                          %%
-%% Specify information, if available,       %%
-%% in the form:                             %%
-%%   <key>={<id1>,<id2>}                    %%
-%%   <key>=                                 %%
-%% Comment or delete the keys which are     %%
-%% not used. Repeat \author command as much %%
-%% as required.                             %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%\author[
-%   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
-%   corref={aff1},                       % id of corresponding address, if any
-%   noteref={n1},                        % id's of article notes, if any
-%   email={jane.e.doe@cambridge.co.uk}   % email address
-%]{\inits{JE}\fnm{Jane E} \snm{Doe}}
-%\author[
-%   addressref={aff1,aff2},
-%   email={john.RS.Smith@cambridge.co.uk}
-%]{\inits{JRS}\fnm{John RS} \snm{Smith}}
-
+\if0\blind
+{
+  \title{\bf A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Control Trials}
 \author{Kellie Ottoboni \\
 Department of Statistics; Berkeley Institute for Data Science\\
-University of California, Berkeley \\ [.2in]
+University of California, Berkeley \\ 
 Fraser Lewis \\
-RB\\ [.2in]
+RB\\ 
 Luigi Salmaso\\
@@
 University of Padova
-}\date{Draft \today}
-\maketitle
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors' addresses here        %%
-%%                                          %%
-%% Repeat \address commands as much as      %%
-%% required.                                %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+}  \maketitle
+} \fi
 
-%\address[id=aff1]{%                           % unique id
-%  \orgname{Department of Zoology, Cambridge}, % university, etc
-%  \street{Waterloo Road},                     %
-%  %\postcode{}                                % post or zip code
-%  \city{London},                              % city
-%  \cny{UK}                                    % country
-%}
-%\address[id=aff2]{%
-%  \orgname{Marine Ecology Department, Institute of Marine Sciences Kiel},
-%  \street{D\"{u}sternbrooker Weg 20},
-%  \postcode{24105}
-%  \city{Kiel},
-%  \cny{Germany}
-%}
+\if1\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf  A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Control Trials}
+\end{center}
+  \medskip
+} \fi
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter short notes here                   %%
-%%                                          %%
-%% Short notes will be after addresses      %%
-%% on first page.                           %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%\begin{artnotes}
-%%\note{Sample of title note}     % note to the article
-%\note[id=n1]{Equal contributor} % note, connected to author
-%\end{artnotes}
-%
-%\end{fmbox}% comment this for two column layout
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Abstract begins here                 %%
-%%                                          %%
-%% Please refer to the Instructions for     %%
-%% authors on http://www.biomedcentral.com  %%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\begin{abstractbox}
-
+\bigskip
 \begin{abstract}
-%\parttitle{Background}
-Statistical test of treatment effects based on the linear model, such as analysis of variance (ANOVA) and analysis of covariance (ANCOVA), are widely accepted by organizations that regulate clinical trials.
-In this paper, we focus on ANCOVA, a special case of linear regression which incorporates stratification.
-Traditional ANCOVA is derived using strong assumptions about the data-generating process so that the resulting inference is based on parametric distributions.
-Because these methods are well understood and robust, they are sometimes applied to data that depart from the method's assumptions, such as ordinal integer scores.
-Given the considerable costs and scientific importance of clinical trials, it is desirable to include an additional methodology which is potentially more robust to departures from assumptions as a check on the statistical inference for the main study endpoints.
-%\parttitle{Methods}
-Permutation tests are a nonparametric alternative that require only minimal assumptions which are often guaranteed by the randomization that was conducted.
-In contrast to parametric tests, which rely up on strong assumptions to obtain a null distribution with a simple closed form, permutation tests use the assumption of exchangeability or randomization to obtain null distributions directly by considering what would have been observed under different possible randomizations.
-We compare parametric ANCOVA to several permutation tests based on linear models, which control for pretreatment covariates to increase power and precision.
-%\parttitle{Results}
-We simulate randomized trials using a variety of data-generating processes, some of which violate the parametric assumptions.
-The permutation tests based on linear models maintain power comparable to the ANCOVA in every case.
-We illustrate the use of these permutation tests alongside the parametric ANCOVA with data from a clinical trial conducted across eight sites,
-each having a different numbers of patients.
-%\parttitle{Conclusions}
-These results suggest that a linear model permutation test may serve as a secondary analysis to give strength to the statistical inferences from the primary analyses of a clinical data study.
-
+% 200 words or less
+Hypothesis tests based on the linear model are widely accepted by organizations that regulate clinical trials.
+We focus on analysis of covariance (ANCOVA), a special case of linear regression that incorporates stratification.
+ANCOVA is derived using strong assumptions about the data-generating process so that the resulting inference is based on parametric distributions.
+Because these methods are well understood and robust, they are sometimes applied to data that depart from assumptions, such as ordinal integer scores.
+Permutation tests are a nonparametric alternative that require minimal assumptions which are often guaranteed by the randomization that was conducted.
+We compare ANCOVA to several permutation tests based on linear models that control for pretreatment covariates.
+In simulated trials using a variety of data-generating processes, some of which violate the parametric assumptions,
+the permutation tests maintain power comparable to the ANCOVA.
+We illustrate the use of these permutation tests alongside ANCOVA with data from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease.
+Given the considerable costs and scientific importance of clinical trials, one may want to include an additional robust method, such as a linear model permutation test, as a check on the statistical inference for the main study endpoints.
 \end{abstract}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The keywords begin here                  %%
-%%                                          %%
-%% Put each keyword in separate \kwd{}.     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\noindent%
+{\it Keywords:}  nonparametric, linear model, ANCOVA, inference
+\vfill
 
-%\begin{keyword}
-%\kwd{permutation test}
-%\kwd{linear model}
-%\kwd{ANCOVA}
-%\end{keyword}
-
-% MSC classifications codes, if any
-%\begin{keyword}[class=AMS]
-%\kwd[Primary ]{}
-%\kwd{}
-%\kwd[; secondary ]{}
-%\end{keyword}
-%
-%\end{abstractbox}
-%%
-%%\end{fmbox}% uncomment this for twcolumn layout
-%
-%\end{frontmatter}
-
-\section*{Background}
+\newpage
+%\spacingset{1.45} % DON'T change the spacing!
+\section{Background}
 A hypothesis test is a statistical method for determining whether observed data is consistent with a belief about the process that generated the data.
@@
 
-\section*{Methods}
+\section{Methods}
 % The methods section should include:
@@
 
-\subsection*{Notation}
+\subsection{Notation}
 Suppose we randomly assign two treatments labelled $0$ and $1$ to a group of individuals.
@@
 This quantity represents the effect of treatment for individual $i$ in group $j$: 
-it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.\footnote{
+it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.
+A typical problem is to estimate the average of differences $Y_{ij}(1) - Y_{ij}(0)$ in some group, such as the study sample or in a target population.
 Which function of these differences will be considered depends on the chosen method of analysis.
-Any of these possible quantities may be of clinical interest; it is a matter of philosophy which one prefers.}
-A typical problem is to estimate the average of differences $Y_{ij}(1) - Y_{ij}(0)$ in some group, such as the study sample or in a target population.
-Instead of estimation, we will focus on hypothesis testing for whether these differences are nonzero.
+Any of these possible quantities may be of clinical interest; it is a matter of philosophy which one prefers.
+Rather than estimation, we will focus on hypothesis testing for whether these differences are nonzero.
 
-\subsection*{Parametric ANCOVA}
+\subsection{Parametric ANCOVA}
 
@@
 \begin{equation}\label{eqn:ancova}
-Y_{ij} = \alpha_j + \beta X_{ij} + \gamma Z_{ij} + \eps_{ij}
+Y_{ij} = \alpha_j + \beta X_{ij} + \gamma Z_{ij} + \varepsilon_{ij}
 \end{equation}
@@
 $\gamma$ is the coefficient for treatment,
-and $\eps_{ij}$ is an error term.
+and $\varepsilon_{ij}$ is an error term.
 The parameter of interest is $\gamma$. 
@@
 \item \textbf{Constant slopes:} Stratum only affects the intercept $\alpha_j$, not the slopes $\beta$ and $\gamma$.
-\item \textbf{IID Errors:} The $\eps_{ij}$ are independent and identically distributed with mean $0$ and variance $\sigma^2$.
+\item \textbf{IID Errors:} The $\varepsilon_{ij}$ are independent and identically distributed with mean $0$ and variance $\sigma^2$.
 \item \textbf{Independence:} If $\mathbf{X}$ is random, $\mathbf{\varepsilon}$ is statistically independent of $\mathbf{X}$.
@@
 
-The coefficients are estimated using least squares.\footnote{
-Note, only Assumptions (1) and (2) are needed to guarantee the existence of a solution.
-The solution will be unique as long as there is no linear relationship between $\mathbf{X}$, $\mathbf{Z}$, and stratum membership.
-}
-From this procedure, we may estimate $\hat{\sigma}_{\hat{\gamma}}^2$, the  standard error of $\hat{\gamma}$\footnote{
-$\hat{\sigma}_{\hat{\gamma}}^2$ is estimated by first computing the residual standard error $\hat{\sigma}^2$,
-then taking the diagonal element corresponding to the treatment variable in the inverse covariance matrix of covariates.
-}.
+The coefficients are estimated using least squares.
+(Note, only Assumptions (1) and (2) are needed to guarantee the existence of a solution.
+The solution will be unique as long as there is no linear relationship between $\mathbf{X}$, $\mathbf{Z}$, and stratum membership.)
+This procedure yields an estimate of $\hat{\sigma}_{\hat{\gamma}}^2$, the  standard error of $\hat{\gamma}$
+(namely, the diagonal element corresponding to the treatment variable in the inverse covariance matrix of regression covariates).
 Under the null hypothesis, the test statistic 
@@
 
-\subsection*{Stratified permutation test}
+\subsection{Stratified permutation test}
 Suppose we wish to test the null hypothesis that individual by individual, treatment has no effect.
@@
 In this framework, one would break down the ``global'' null hypothesis of no effect whatsoever into the intersection of ``partial'' null hypotheses of no treatment effect within each stratum.
-For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies\footnote{ 
-We considered using the difference in means within each stratum, aggregated by taking the sum of their absolute values over stratum.
-However this would not be ideal for imbalanced designs, which are common in real-world applications, so we did not consider it in the simulations.
-See the clinical data analysis file for a more detailed treatment.}.
+For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies.
 In this case, strata are randomized independently so there are no dependencies to preserve, making this method is equivalent to the first.
@@
 
-\subsection*{Permutation tests with the linear model}
+\subsection{Permutation tests with the linear model}
 The tests in the previous section only use information on the treatment and the outcome.
@@
 
-Freedman and Lane \cite{freedman_nonstochastic_1983} propose an alternative test based on the residuals of the linear regression.
+\citet{freedman_nonstochastic_1983} propose an alternative test based on the residuals of the linear regression.
 They take an alternative view of the problem, still writing the outcome in the form of Equation~\ref{eqn:ancova}, but instead of treating the $\varepsilon_{ij}$ as random they define the errors to be the difference between the observed outcome $Y_{ij}$ and the data's linear projection onto the plane $\alpha_j + \beta X_{ij}+ \gamma Z_{ij}$.
@@
 Then we estimate the errors $\mathbf{\hat{\varepsilon}}$ by $\mathbf{Y} - \mathbf{\hat{Y}}$, where $\mathbf{\hat{Y}}$ is the vector of fitted values $\hat{Y}_{ij} = \hat{\alpha}_j + \hat{\beta}X_{ij}$.
-The $\mathbf{\hat{\varepsilon}}$ approximate the true errors $\mathbf{\varepsilon}$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit\footnote{
-\cite{freedman_nonstochastic_1983} advise against using this method if there are large outlier values in the covariate $\mathbf{X}$ or when $\mathbf{X}$ and $\mathbf{Z}$ are highly colinear.
-}.
+The $\mathbf{\hat{\varepsilon}}$ approximate the true errors $\mathbf{\varepsilon}$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit.
+(\cite{freedman_nonstochastic_1983} advise against using this method if there are large outlier values in the covariate $\mathbf{X}$ or when $\mathbf{X}$ and $\mathbf{Z}$ are highly colinear.)
 Furthermore, under the null hypothesis, the $\mathbf{\varepsilon}$ are independent $\mathbf{Z}$ within strata. 
@@
 
-\section*{Results}
-\subsection*{Simulations}\label{simulations}
+\section{Results}
+\subsection{Simulations}\label{simulations}
 
@@
 and the Freedman-Lane permutation test.
+(We decided against using the NPC statistic computed by taking the difference in means within each stratum and summing their absolute values over strata. 
+However, this would not be ideal for imbalanced designs, which are common in real-world applications.  
+See the clinical data analysis file for a more detailed treatment.) \todo{supplementary ref}
 
@@
 Table~\ref{tab:normal_power} displays the size of the test and the power at level $0.05$.
-The three tests based on linear models have slightly higher than nominal level, though the margin is within two standard errors of $0.05$.\footnote{
-The number of tests rejected under the null in 1000 trials has a binomial distribution.
-If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.}
+The three tests based on linear models have slightly higher than nominal level, though the margin is within two standard errors of $0.05$.
+(The number of tests rejected under the null in 1000 trials has a binomial distribution.
+If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.)
 These numbers show that actually, for $\alpha=0.05$, the stratified permutation test does not lose a large amount of power when the effect is isolated in a single stratum.
@@
 \centering
-\includegraphics[width = \textwidth]{fig/normal_simulation_power}
+\includegraphics[width = \textwidth]{fig/normal_simulation_power.pdf}
 \caption{Empirical power curves for the Gaussian simulated data}
@@
 It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exacerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
-Here, the data have the same variance so we would expect power, not the test level, to be an issue.\footnote{
-See the supplementary Gaussian simulation files for a treatment of heterogeneous variances.}
+Here, the data have the same variance so we would expect power, not the test level, to be an issue.
+(See the supplementary Gaussian simulations for a discussion of heterogeneous variances.)
 We omit a plot of the power curves from these simulations, as the patterns appear similar to those in Figure~\ref{fig:normal_sim_power}.
@@
 \centering
-\includegraphics[width = \textwidth]{fig/skewed_simulation_power}
+\includegraphics[width = \textwidth]{fig/skewed_simulation_power.pdf}
 \caption{Empirical power curves for the skewed, discrete simulated data}
@@
 The baseline and outcome are highly correlated, so the change score has lower variance the outcome alone.
-In additional simulations\footnote{
-See the simulation using Gaussian, balanced design for a more detailed treatment.}, 
-we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had correlation of $0.25$.
+In additional simulations, 
+we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had correlation of $0.25$ (see the supplementary Gaussian simulations for a more detailed treatment).
 In this case, the result was reversed: 
@@
 
-\subsection*{Clinical data results}
+\subsection{Clinical data results}
 We compared the parametric ANCOVA, the stratified permutation test, and the two linear model-based permutation tests using a dataset from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease (GERD).
@@
 \centering
-\includegraphics[width = \textwidth]{fig/clinical_distr}
+\includegraphics[width = \textwidth]{fig/clinical_distr.pdf}
 \caption{Distribution of each endpoint for the two treatments A and B}
@@
 
-\section*{Discussion}
+\section{Discussion}
 % This section should discuss the implications of the findings in context of existing research and highlight limitations of the study.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Backmatter begins here                   %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bigskip
+\begin{center}
+{\large\bf SUPPLEMENTARY MATERIAL}
+\end{center}
 
-%\begin{backmatter}
-%
-%\section*{Competing interests}
-%  The authors declare that they have no competing interests.
-%
-%\section*{Author's contributions}
-%    Text for this section \ldots
-%
-%\section*{Acknowledgements}
-%  Text for this section \ldots
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  Bmc_mathpys.bst  will be used to                       %%
-%%  create a .BBL file for submission.                     %%
-%%  After submission of the .TEX file,                     %%
-%%  you will be prompted to submit your .BBL file.         %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{description}
 
-% if your bibliography is in bibtex format, use those commands:
-\bibliographystyle{vancouver} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{references}      % Bibliography file (usually '*.bib' )
-%% for author-year bibliography (bmc-mathphys or spbasic)
-%% a) write to bib file (bmc-mathphys only)
-%% @settings{label, options="nameyear"}
-%% b) uncomment next line
-%%\nocite{label}
-%
-%% or include bibliography directly:
-%% \begin{thebibliography}
-%% \bibitem{b1}
-%% \end{thebibliography}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                               %%
-%%% Figures                       %%
-%%%                               %%
-%%% NB: this is for captions and  %%
-%%% Titles. All graphics must be  %%
-%%% submitted separately and NOT  %%
-%%% included in the Tex document  %%
-%%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%%%
-%%% Do not use \listoffigures as most will included as separate files
-%
-%\section*{Figures}
-%  \begin{figure}[h!]
-%  \caption{\csentence{Sample figure title.}
-%      A short description of the figure content
-%      should go here.}
-%      \end{figure}
-%
-%\begin{figure}[h!]
-%  \caption{\csentence{Sample figure title.}
-%      Figure legend text.}
-%      \end{figure}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                               %%
-%%% Tables                        %%
-%%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%%% Use of \listoftables is discouraged.
-%%%
-%\section*{Tables}
-%\begin{table}[h!]
-%\caption{Sample table title. This is where the description of the table should go.}
-%      \begin{tabular}{cccc}
-%        \hline
-%           & B1  &B2   & B3\\ \hline
-%        A1 & 0.1 & 0.2 & 0.3\\
-%        A2 & ... & ..  & .\\
-%        A3 & ..  & .   & .\\ \hline
-%      \end{tabular}
-%\end{table}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                               %%
-%%% Additional Files              %%
-%%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%\section*{Additional Files}
-%  \subsection*{Additional file 1 --- Sample additional file title}
-%    Additional file descriptions text (including details of how to
-%    view the file, if it is in a non-standard format or the file extension).  This might
-%    refer to a multi-page table or a figure.
-%
-%  \subsection*{Additional file 2 --- Sample additional file title}
-%    Additional file descriptions text.
-%
-%
-%\end{backmatter}
+\item[Title:] Brief description. (file type)
 
+\end{description}
+
+\bibliographystyle{chicago}
+
+\bibliography{references}
 \end{document}

### 2017-01-31 ms/SBR-template.tex
Reformat tables nicely

@@
 \usepackage{color}
+ \usepackage{multirow}
 \newcommand{\todo}[1]{{\color{red}{TO DO: \sc #1}}}
@@
 \end{figure}
-\begin{center}
-\input{fig/normal_simulation_power_summary.tex}
-\end{center}
+%\begin{center}
+%\input{fig/normal_simulation_power_summary.tex}
+%\end{center}
+
+\begin{table}[]
+\centering
+\label{tab:normal_power}
+\begin{tabular}{ll|ccccc}
+\multicolumn{2}{c|}{Design}                                                                  & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
+Errors                        & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
+\multirow{3}{*}{Gaussian}     & None                                                        & 0.058                      & 0.042                                                                                & 0.038                                                                                 & 0.062                                                                        & 0.057                                                                        \\
+                              & Constant                                                    & 0.886                      & 0.545                                                                                 & 0.858                                                                                 & 0.891                                                                        & 0.892                                                                        \\
+                              & Single stratum                                              & 0.159                      & 0.089                                                                                & 0.160                                                                                 & 0.162                                                                        & 0.170                                                                        \\ \hline
+\multirow{3}{*}{Heavy-tailed} & None                                                        & 0.032                      & 0.040                                                                                & 0.045                                                                                 & 0.042                                                                        & 0.040                                                                        \\
+                              & Constant                                                    & 0.355                      & 0.240                                                                                & 0.378                                                                                 & 0.400                                                                        & 0.393                                                                        \\
+                              & Single stratum                                              & 0.066                      & 0.080                                                                                & 0.082                                                                                 & 0.076                                                                        & 0.080                                                                        \\ \hline
+\end{tabular}
+\caption{Empirical power at level $0.05$ for Gaussian simulated data.}
+\end{table}
 
@@
 %\end{figure}
-\begin{center}
-\input{fig/imbalanced_simulation_power_summary.tex}
-\end{center}
+%\begin{center}
+%\input{fig/imbalanced_simulation_power_summary.tex}
+%\end{center}
+
+\begin{table}[ht]
+\centering
+\label{tab:imbalanced_power}
+\begin{tabular}{ll|ccccc}
+\multicolumn{2}{c|}{Design}                                                                 & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
+Errors                        & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
+\multirow{3}{*}{Gaussian}     & None                                                        & 0.046                      & 0.031                                                                                & 0.028                                                                                 & 0.042                                                                        & 0.045                                                                        \\
+                              & Constant                                                    & 0.840                      & 0.490                                                                                & 0.810                                                                                 & 0.870                                                                        & 0.860                                                                        \\
+                              & Single stratum                                              & 0.120                      & 0.050                                                                                & 0.090                                                                                 & 0.100                                                                        & 0.120                                                                        \\ \hline
+\multirow{3}{*}{Heavy-tailed} & None                                                        & 0.034                      & 0.054                                                                                & 0.055                                                                                 & 0.047                                                                        & 0.042                                                                        \\
+                              & Constant                                                    & 0.230                      & 0.200                                                                                & 0.270                                                                                 & 0.260                                                                        & 0.270                                                                        \\
+                              & Single stratum                                              & 0.070                      & 0.080                                                                                & 0.090                                                                                 & 0.090                                                                        & 0.080                                                                        \\ \hline
+\end{tabular}
+\caption{Empirical power at level $0.05$ for Gaussian simulated data with imbalanced treatment groups} 
+\end{table}
 
@@
 
-\begin{center}
-\input{fig/skewed_simulation_power_summary.tex}
-\end{center}
-
+%\begin{center}
+%\input{fig/skewed_simulation_power_summary.tex}
+%\end{center}
+\begin{table}[]
+\centering
+\label{tab:skewed_power}
+\begin{tabular}{ll|ccccc}
+\multicolumn{2}{c|}{Design}                                                               & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
+Group sizes                 & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
+\multirow{3}{*}{Balanced}   & None                                                        & 0.036                      & 0.025                                                                                & 0.021                                                                                 & 0.035                                                                        & 0.037                                                                        \\
+                            & Constant                                                    & 0.974                      & 0.143                                                                                & 0.853                                                                                 & 0.972                                                                        & 0.977                                                                        \\
+                            & Single stratum                                              & 0.422                      & 0.050                                                                                & 0.275                                                                                 & 0.429                                                                        & 0.424                                                                        \\ \hline
+\multirow{3}{*}{Imbalanced} & None                                                        & 0.038                      & 0.024                                                                                & 0.019                                                                                 & 0.037                                                                        & 0.039                                                                        \\
+                            & Constant                                                    & 0.971                      & 0.147                                                                                & 0.860                                                                                 & 0.970                                                                        & 0.972                                                                        \\
+                            & Single stratum                                              & 0.190                      & 0.070                                                                                & 0.100                                                                                 & 0.190                                                                        & 0.190                                                                        \\ \hline
+\end{tabular}
+\caption{Empirical power at level $0.05$ for discrete, skewed simulated data} 
+\end{table}
 It makes sense that in all of these simulations, the permutation test using the change scores as the response measure is more powerful than the permutation test using the raw outcomes.
@@
 
-\begin{center}
-\input{fig/results_pvalue_summary.tex}
-\end{center}
+%
+%\begin{center}
+%\input{fig/results_pvalue_summary.tex}
+%\end{center}
+
+\begin{table}[]
+\centering
+\label{tab:clinical_pvalues}
+\begin{tabular}{l|cccc}
+Endpoint      & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
+heart\_freq   & 0.035                      & 0.006                                                                                & 0.080                                                                        & 0.082                                                                        \\
+regurg\_freq  & 0.136                      & 0.118                                                                                & 0.280                                                                        & 0.220                                                                        \\
+dysp\_freq    & 0.565                      & 0.948                                                                                & 0.616                                                                        & 0.592                                                                        \\
+daily\_heart & 0.032                      & 0.004                                                                                & 0.056                                                                        & 0.068                                                                        \\
+daily\_regur  & 0.142                      & 0.174                                                                                & 0.286                                                                        & 0.246                                                                        \\
+daily\_hrdq   & 0.043                      & 0.012                                                                                & 0.088                                                                        & 0.098                                                                        \\
+daily\_dysp   & 0.582                      & 0.810                                                                                & 0.756                                                                        & 0.722                                                                        \\
+\hline
+\end{tabular}
+\caption{Comparison of p-values from four tests, for each continuous endpoint.} 
+\end{table}
 

### 2017-01-31 ms/fig/normal_simulation_power_summary.tex
Reformat tables nicely

--- 2016-12-08 ms/fig/normal_simulation_power_summary.tex
@@
   \hline
-Treatment & Errors & ANOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
+Errors & Treatment & ANCOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
   \hline
-Constant Additive Effect & Gaussian & 0.89 & 0.55 & 0.86 & 0.89 & 0.89 \\ 
-  Single Stratum Effect & Gaussian & 0.16 & 0.09 & 0.16 & 0.16 & 0.17 \\ 
-  Constant Additive Effect & Heavy-tailed & 0.35 & 0.24 & 0.38 & 0.40 & 0.39 \\ 
-  Single Stratum Effect & Heavy-tailed & 0.07 & 0.08 & 0.08 & 0.08 & 0.08 \\ 
+Gaussian & No Effect & 0.058 & 0.042 & 0.038 & 0.062 & 0.057 \\ 
+  Gaussian & Constant Additive Effect & 0.886 & 0.545 & 0.858 & 0.891 & 0.892 \\ 
+  Gaussian & Single Stratum Effect & 0.159 & 0.089 & 0.160 & 0.162 & 0.170 \\ 
+   \hline
+Heavy-tailed & No Effect & 0.032 & 0.040 & 0.045 & 0.042 & 0.040 \\ 
+  Heavy-tailed & Constant Additive Effect & 0.355 & 0.240 & 0.378 & 0.400 & 0.393 \\ 
+  Heavy-tailed & Single Stratum Effect & 0.066 & 0.080 & 0.082 & 0.076 & 0.080 \\ 
    \hline

### 2017-01-31 results/results.Rmd
Make figures nicer with theme_bw

@@
   ggtitle("Number of Individuals per Treatment") +
-  labs(fill = "Treatment") 
+  labs(fill = "Treatment")+
+  theme_bw()
 ```
@@
         labs(x = variable, fill = "Treatment") +
+        theme_bw() +
         theme(
@@
         legend.title = element_text(size = 12),
-        legend.text = element_text(size = 12),
+        legend.text = element_text(size = 14),
         strip.text.x = element_text(size = 12)
@@
 dev.off()
+do.call(grid.arrange, c(plot_distrs, nrow = 2))
 ```
@@
   facet_grid(~model) +
-  geom_vline(aes(xintercept = xintercept))
+  geom_vline(aes(xintercept = xintercept)) +
+  theme_bw()
 
@@
   facet_grid(~model) +
-  geom_vline(aes(xintercept = xintercept))
+  geom_vline(aes(xintercept = xintercept)) +
+  theme_bw()
 ```
@@
   ylab("Frequency") +
-  ggtitle("Permutation Distributions")
+  ggtitle("Permutation Distributions") +
+  theme_bw()
 ```
@@
   xlab("Treatment") + ylab("Residual") + 
-  ggtitle("Outcome Model")
+  ggtitle("Outcome Model") +
+  theme_bw()
 p21 <- daily_hrdq %>% 
@@
   xlab("Site ID") + ylab("Residual") + 
-  ggtitle("Outcome Model")
+  ggtitle("Outcome Model") +
+  theme_bw()
 p12 <- daily_hrdq %>% 
@@
   xlab("Treatment") + ylab("Residual") +
-  ggtitle("Differences Model")
+  ggtitle("Differences Model") +
+  theme_bw()
 p22 <- daily_hrdq %>% 
@@
   xlab("Site ID") + ylab("Residual") +
-  ggtitle("Differences Model")
+  ggtitle("Differences Model") +
+  theme_bw()
 grid.arrange(p12, p11, p22, p21, nrow = 2, ncol = 2)

### 2017-01-31 simulations/simulation.Rmd
Make figures nicer with theme_bw

@@
 require(gridExtra)
+
+report_theme <- theme_bw() + theme(
+  axis.text.x = element_text(size = 14, angle = -45),
+  axis.text.y = element_text(size = 14),
+  axis.title = element_text(size = 20),
+  title = element_text(size = 20),
+  legend.title = element_text(size = 14),
+  legend.text = element_text(size = 14),
+  strip.text.x = element_text(size = 14)
+)
 ```
@@
   ggtitle(title)
+
 }
@@
 plot_power_curves(design0_power, "No Effect, Normal Errors")
-
 plot_pvalue_hist(design0_pvalues, "No Effect, Normal Errors")
@@
 plot_power_curves(design00_power, "No Effect, Heavy-tailed Errors")
-
 plot_pvalue_hist(design00_pvalues, "No Effect, Heavy-tailed Errors") 
@@
 plot_power_curves(design1_power, "Constant Additive Effect of 1, Normal Errors")
-
 plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 1, Normal Errors")
@@
 plot_power_curves(design2_power, "Effect of 1 at a Single stratum, Normal Errors")
-
 plot_pvalue_hist(design2_pvalues, "Effect of 1, Normal Errors")
@@
 plot_power_curves(design3_power, "Constant Additive Effect of 1, Heavy-tailed Errors")
-
 plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 1, Heavy-tailed Errors")
@@
 plot_power_curves(design4_power, "Effect of 1 at a Single stratum, Heavy-tailed Errors") 
-
 plot_pvalue_hist(design4_pvalues, "Effect of 1, Heavy-tailed Errors")
@@
 plot_power_curves(design5_power, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
-
 plot_pvalue_hist(design5_pvalues, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
@@
 plot_power_curves(design6_power, "Constant Additive Effect of 1, Normal Errors")
-
 plot_pvalue_hist(design6_pvalues, "Constant Additive Effect of 1, Normal Errors")
@@
 plot_power_curves(design7_power, "Effect of 1 at a Single stratum, Normal Errors")
-
 plot_pvalue_hist(design7_pvalues, "Effect of 1, Normal Errors")
@@
 plot_power_curves(design8_power, "Constant Additive Effect of 1, Heteroskedastic Errors")
-
 plot_pvalue_hist(design8_pvalues, "Constant Additive Effect of 1, Heteroskedastic Errors")
@@
 plot_power_curves(design9_power, "Effect of 1 at a Single stratum, Heteroskedastic Errors")
-
 plot_pvalue_hist(design9_pvalues, "Effect of 1, Heteroskedastic Errors")
@@
   geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
-  xlab("P-value") +
+  xlab("Significance Level") +
   ylab("Power") +
@@
   ggtitle("Gaussian Data") +
-  theme(
-    axis.text.x = element_text(size = 12),
-    axis.text.y = element_text(size = 12),
-    axis.title = element_text(size = 16),
-    title = element_text(size = 16),
-    legend.title = element_text(size = 12),
-    legend.text = element_text(size = 12),
-    strip.text.x = element_text(size = 12),
-    legend.position = "bottom"
-  ) + 
+  report_theme +
+  theme(legend.position = "bottom") + 
   guides(color=guide_legend(nrow=2,byrow=TRUE))
-
 twobytwo
@@
 ```{r power_summary_tab, results = "asis"}
-summary05 <- t(sapply(powers, function(x) x[5, c(6, 7, 1:5)]))
+summary05 <- t(sapply(powers, function(x) x[5, c(7, 6, 1:5)]))
 summary05 <- summary05[c(5, 1, 2, 6, 3, 4), ]

### 2017-01-31 simulations/simulation_skewed.Rmd
Make figures nicer with theme_bw

@@
 ---
-title: "ANOVA Comparison Simulations: Skewed, Poisson Data"
+title: "ANCOVA Comparison Simulations: Skewed, Poisson Data"
 author: "Kellie Ottoboni"
@@
 require(gtable)
+
+report_theme <- theme_bw() + theme(
+  axis.text.x = element_text(size = 14, angle = -45),
+  axis.text.y = element_text(size = 14),
+  axis.title = element_text(size = 20),
+  title = element_text(size = 20),
+  legend.title = element_text(size = 14),
+  legend.text = element_text(size = 14),
+  strip.text.x = element_text(size = 14)
+)
 ```
@@
 tmp <- generate_simulated_data(gamma = 0, effect = "same effect")
-design0_pvalues <- replicate(100, {
+design0_pvalues <- replicate(1000, {
   tmp$epsilon <- rnorm(nrow(tmp))
@@
 ```{r design_null_plots, fig.height = 4, fig.align="center"}
-plot_power_curves(design0_power, "No Effect, Normal Errors")
+plot_power_curves(design0_power, "No Effect, Balanced Design")
 
-plot_pvalue_hist(design0_pvalues, "No Effect, Normal Errors")
+plot_pvalue_hist(design0_pvalues, "No Effect, Balanced Design")
 ```
@@
 
-There is no discernable difference in power between the ANOVA, the differenced stratified permutation test, the LM permutations, or the Freedman-Lane test.
+There is no discernable difference in power between the ANCOVA, the differenced stratified permutation test, the LM permutations, or the Freedman-Lane test.
 However, the simple stratified permutation test of $Y_1$ has substantially less power than the other four.
@@
 
-# Design 4: Constant additive effect, imbalanced design
+# Test level: imbalanced design
 
@@
 stratum 1 has 8 patients, stratum 2 has 16 patients, and stratum 3 has 24 patients.
-This imbalance does not violate any assumptions of the ANOVA.
+This imbalance does not violate any assumptions of the ANCOVA.
 However, it may reduce power if the effect is concentrated in strata with fewer patients.
+
+```{r design_null_imbalanced, fig.height = 4}
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+
+tmp <- generate_simulated_data(gamma = 0, effect = "same effect", n = c(8, 16, 24))
+design00_pvalues <- replicate(1000, {
+  tmp$epsilon <- rnorm(nrow(tmp))
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <- generate_outcome(tmp$Y0, rep(0, nrow(tmp)), tmp$stratum_effect, tmp$epsilon)
+  generate_simulated_pvalues(tmp)
+})
+design00_pvalues <- t(design00_pvalues)
+colnames(design00_pvalues) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+design00_power <- apply(design00_pvalues, 2, compute_power)
+```
+
+```{r design_null_imbalanced_plots, fig.height = 4, fig.align="center"}
+plot_power_curves(design00_power, "No Effect, Imbalanced design")
+
+plot_pvalue_hist(design00_pvalues, "No Effect, Imbalanced design")
+```
+
+# Design 4: Constant additive effect, imbalanced design
 
@@
                design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design4_power)), "Design" = rep("Imbalanced", nrow(design4_power))), 
-               design5_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design5_power)), "Design" = rep("Imbalanced", nrow(design5_power))))
+               design5_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design5_power)), "Design" = rep("Imbalanced", nrow(design5_power))),
+               design0_power %>% as.data.frame() %>% mutate("Treatment" = rep("No Effect", nrow(design0_power)), "Design" = rep("Balanced", nrow(design0_power))),
+               design00_power %>% as.data.frame() %>% mutate("Treatment" = rep("No Effect", nrow(design00_power)), "Design" = rep("Imbalanced", nrow(design00_power))))
 
 all_power_curves <- do.call(rbind, powers)
-twobytwo <- melt(all_power_curves, id.vars = c("Treatment", "Design")) %>% 
+twobytwo <- all_power_curves %>%
+  filter(Treatment != "No Effect") %>%
+  melt(id.vars = c("Treatment", "Design"))  %>% 
   mutate("pvalue" = rep((1: 100)/100, 5*4)) %>%
@@
   ggtitle("Discrete, Skewed Data") +
-    theme(
-    axis.text.x = element_text(size = 12),
-    axis.text.y = element_text(size = 12),
-    axis.title = element_text(size = 16),
-    title = element_text(size = 16),
-    legend.title = element_text(size = 12),
-    legend.text = element_text(size = 12),
-    strip.text.x = element_text(size = 12),
-    legend.position = "bottom"
-  ) + 
+  report_theme +
+  theme(legend.position = "bottom") + 
   guides(color=guide_legend(nrow=2,byrow=TRUE))
@@
 summary05 <- t(sapply(powers, function(x) x[5, c(6, 7, 1:5)]))
-summarytab <- xtable(summary05, align = "rp{1.25in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}",
+summary05 <- summary05[c(5, 1, 2, 6, 3, 4), ]
+summarytab <- xtable(summary05, digits = 3, align = "rp{1.25in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}",
        caption = "Empirical power at level $0.05$ for discrete, skewed simulated data",
        label = "tab:skewed_power")
-print(summarytab, type="latex", file="../ms/fig/skewed_simulation_power_summary.tex") 
+print(summarytab, hline.after = c(-1, 0, 3, nrow(summarytab)), type="latex", file="../ms/fig/skewed_simulation_power_summary.tex") 
 summarytab 

### 2017-02-01 results/results.Rmd
Edits to supplementary results file

--- 2016-12-19 results/results.Rmd
@@
                       warning = FALSE,
-                      tidy = TRUE, tidy.opts = list(width.cutoff = 60),
+                      tidy = TRUE, 
+                      tidy.opts = list(width.cutoff = 60),
                       cache = TRUE)
@@
 
+* We renamed treatments, subject IDs, and site IDs to remove identifying information.
 * Initially, there were 188 subjects.
 52 of these were never assigned to a treatment group and were only observed at baseline.
-This ensures that censoring was independent of treatment, so removing them from the analysis does not introduce any selection bias.
-We are left with 66 subjects in group A and 70 subjects in group B.
-* We renamed treatments, subject IDs, and site IDs to remove identifying information.
+Because censoring occurred before treatment assignment, we are guaranteed that censoring and treatment are independent.
+Removing these individuals from the analysis does not introduce any selection bias.
+After removing the 52 people with missing data, we are left with 66 subjects in group A and 70 subjects in group B.
 * Each individual was observed for a total of 14 days, 7 during the baseline week and 7 after receiving treatment.
@@
 
-We notice several things patterns from Figure \ref{fig:count_by_site}.
+We notice several things in Figure \ref{fig:count_by_site}.
 Some of the sites have many more subjects than others: these are sites 1, 3, and 6, and 7.
@@
 
-```{r count_by_site, fig.align="center", fig.cap = "Number of subjects in each treatment group, by site ID."}
+```{r count_by_site, fig.align="center", fig.height=6, fig.width=12, fig.cap = "Number of subjects in each treatment group, by site ID."}
 clinical %>% 
@@
   ggtitle("Number of Individuals per Treatment") +
-  labs(fill = "Treatment") 
+  labs(fill = "Treatment")+
+  theme_bw()
 ```
 
-The data are not in the right shape to analyze a single variable.
+The dataset is not in the right format to analyze a single variable.
 Below, we include a function to take the raw data and reshape it to analyze one variable of interest.
@@
 reshape_data <- function(variable, data = data_by_subjid_visitnum){
+  # Reshape data to be analyzed with regression
+  # Inputs:
+  # variable = the clinical endpoint of interest, input as a string
+  # data     = dataset with the variable as a column name, grouped by subject id and visit number
+  #            default is the dataframe we just created, data_by_subjid_visitnum
+  # Output:
+  # A dataframe with a single row per subject and columns for treatment, site ID, baseline + outcome measures
   data <- data %>% mutate_(myvariable = variable)
@@
 
-```{r hist_continuous_vars, fig.height=6, fig.width=12, fig.cap = "Distribution of continuous outcomes"}
+```{r hist_continuous_vars, fig.align="center", fig.height=6, fig.width=12, fig.cap = "Comparison of the distributions of the seven continuous clinical endpoints during week 2 for the two treatment groups."}
+
+# Create a plot for each variable, store in a list
 continuous_vars <- c("daily_heart", "daily_regurg", "daily_dysp", 
@@
         geom_density(alpha = 0.6, aes(fill = tr))+ 
-#        geom_histogram(binwidth = 0.5, position = "dodge", aes(fill = tr)) +
-#        scale_fill_discrete(name="Experimental\nCondition",
-#                       breaks=c("0", "1"),
-#                       labels=c("A", "B")) + 
         labs(x = variable, fill = "Treatment") +
+        theme_bw() +
         theme(
@@
 
+# Move the legend
 tmp <- ggplot_gtable(ggplot_build(plot_distrs[[1]]))
@@
 
+# Save figure for paper
 pdf("../ms/fig/clinical_distr.pdf", width = 8, height = 4)
@@
 dev.off()
+
+# Print figure
+do.call(grid.arrange, c(plot_distrs, nrow = 2))
 ```
 
-# Analysis - Primary endpoint
-To illustrate the different methods, let's first restrict attention to the primary endpoint, `daily_hrdq`.
-Throughout, we compare two models: 
-the first model considers the final outcome measure, adjusting for site ID and perhaps also the baseline score.
-The second model considers the difference between outcome and baseline score as its dependent variable, adjusting for site ID.
-
-We compare four testing methods: 
-the usual parametric ANCOVA,
-a permutation test considering the difference in means as test statistic, stratifying by site,
-the stratified permutation test, using the $t$-statistic from the linear regression instead,
-and a permutation test in the style of Freedman and Lane (1983), using a linear regression to adjust for covariates.
-
+The primary endpoint is `daily_hrdq`, a measure derived from other variables that were measured in the survey.
+We show summary statistics for this outcome for each treatment group below.
 ```{r hrdq, results = "asis"}
@@
 print(xtable(summary_stats_hrdq,
-             caption = "Summary statistics for change in primary endpoint (daily hrdq) for the two treatment groups."),
+             caption = "Summary statistics for change in primary 
+                        endpoint (daily hrdq) for the two treatment groups."),
       include.rownames = TRUE)
 ```
+
+# Analysis - Primary endpoint
+To illustrate the different methods, let's first restrict attention to the primary endpoint, `daily_hrdq`.
+Throughout this section, we compare two models which both control for site ID.
+The first model uses the outcome in week two as the dependent variable, controlling for the baseline score in the models with covariates.
+The second model uses the difference between outcome and baseline score as its dependent variable.
+The difference in the two models is how we control for the baseline measure.
+
+We compare four testing methods:
+
+1. the usual parametric ANCOVA,
+2. a permutation test, where the difference in means is the test statistic and permutations are stratified by site,
+3. a variation of this stratified permutation test, using the $t$-statistic from linear regression instead, and 
+4. a stratified permutation test in the style of Freedman and Lane (1983), using a linear regression to adjust for covariates.
 
@@
 Under model 2, the significance goes away.
-(Note that treatment is still coded as A and B, which are characters.
-The `lm` function is smart enough to know that it should be coded as a factor, where we compare the effect of B relative to A.
-In general, be very careful when you're doing this!)
+(Note that treatment is still coded as A and B with the character type. 
+The `lm` function is smart enough to know that it should be coded as a factor, where we compare the effect of B relative to A,
+and coerces the variable type from string to factor behind the scenes. 
+Be very careful when you're doing this!
+It is generally safer to change types yourself.)
 ```{r hrdq_parametric, results = "asis"}
@@
 Treatment was completely randomized at each site, independently across sites.
-Assuming that drop-out from the study was independent of treatment assignment (which is true if treatment was blinded and assigned at random), then we may condition on the number of individuals who received A and B at each site.
-Any assignment of treatments which preserves the number of treated and controls at each site is valid and has equal probability of occurring.
-Therefore, by using this principle of equal probabilities and by imputing the unobserved potential outcomes assuming that the null hypothesis is true, we may obtain the permutation distribution of any statistic under the null hypothesis.
+We assume that drop-out from the study was independent of treatment assignment (which is true if treatment was blinded and assigned at random).
+This will be a conditional hypothesis test: we condition on the number of individuals who received A and B at each site.
+This conditioning asserts that any assignment of treatments that preserves the number of treated and controls at each site is valid and has an equal probability of occurring.
+Therefore, using this principle of equal probabilities and by imputing the unobserved potential outcomes assuming that the null hypothesis is true, we can compute the permutation distribution of any statistic under the null hypothesis.
 
-We compare two test statistics: the usual difference in means between group B and group A, and the difference in means within each site, aggregated by taking the sum of their absolute values over sites.
+We compare two test statistics: the difference in mean outcomes for group B and mean outcomes for group A, and this difference of means within each individual site, aggregated over sites by taking the sum of their absolute values.
 The first statistic is more comparable to what one would obtain from an ANCOVA and it is readily interpretable.
 It does not directly account for the stratification by site, so variation between sites may be hidden.
-The second statistic is useful for testing the two-sided alternative hypothesis that treatment has some effect. 
+The second statistic is useful for testing the two-sided alternative hypothesis that treatment has _some_ nonzero effect. 
 It may be more powerful than the simple difference in means if the effect of the drug varies across sites.
-For instance, if the drug has a positive effect at one site but a negative effect at another site, this test statistic would be large; 
-if we used the simple difference in means without accounting for sites, then the positive effect and negative effect may cancel each other out and appear as no effect.
+For instance, if the drug has a positive effect at one site but a negative effect at another site, this test statistic would be large.
+If we used the simple difference in means without accounting for sites, then the positive effect and negative effect may cancel each other out;
+the difference in means will be attenuated toward zero.
 
-Figure \ref{fig:hrdq_unadjusted_permutation} shows the permutation distributions of the simple difference in means.
+Figure \ref{fig:hrdq_unadjusted_permutation} shows the permutation distributions of the simple difference in means for the two models.
 Figure \ref{fig:hrdq_unadj_perm_withinstrata} shows the permutation distribution of the absolute value of the difference of means within strata, summed across strata.
+These empirical permutation distributions are computed by randomly permuting treatment assignments within each site $10^4$ times and computing the statistic for each permutation.
+The vertical line shows the observed value of the test statistic.
+The $p$-value is two times the area of the permutation distribution to the right of the observed value;
+it is the probability of observing a value at least as extreme (falling that far away from the center of the distribution) as the observed value.
 There are several points to notice:
+
 * The second test is far less powerful than the first.
@@
 
-Since the summed difference in means statistic has such lower power, we abandon it for the rest of the analyses.
+Since the summed difference in means statistic has such low power, we abandon it in the rest of the analyses.
 
-```{r hrdq_unadjusted_permutation, fig.height = 3, fig.cap = "Permutation distribution of the difference of means for daily_hrdq"}
+```{r hrdq_unadjusted_permutation, fig.align="center", fig.height = 3, fig.cap = "Permutation distribution of the difference of means for daily hrdq"}
 # method 1 : do permutation of differences
@@
 data.frame("perm" = c(diff_means_distr1, diff_means_distr2),
-           "model" = c(rep("Outcome", length(diff_means_distr1)), 
-                       rep("Differenced", length(diff_means_distr2))),
+           "model" = c(rep("Raw Outcome Model", length(diff_means_distr1)), 
+                       rep("Differenced Model", length(diff_means_distr2))),
            "xintercept" = c(rep(observed_diff_means1, length(diff_means_distr1)),
@@
   facet_grid(~model) +
-  geom_vline(aes(xintercept = xintercept))
+  geom_vline(aes(xintercept = xintercept)) +
+  theme_bw() +
+  labs(x = "Difference in Means Statistic", y = "Frequency") +
+  theme(legend.position = "none")
 
@@
 
-```{r hrdq_unadj_perm_withinstrata, fig.height = 3, fig.cap = "Permutation distribution of the difference of means within strata, summed across strata, for daily_hrdq"}
+```{r hrdq_unadj_perm_withinstrata, fig.align="center", fig.height = 3, fig.cap = "Permutation distribution of the difference of means within strata, summed across strata, for daily hrdq"}
 
@@
 data.frame("perm" = c(diff_means_distr_bystrata1, diff_means_distr_bystrata2),
-           "model" = c(rep("Outcome", length(diff_means_distr_bystrata1)), 
-                       rep("Differenced", length(diff_means_distr_bystrata2))),
+           "model" = c(rep("Raw Outcome Model", length(diff_means_distr_bystrata1)), 
+                       rep("Differenced Model", length(diff_means_distr_bystrata2))),
            "xintercept" = c(rep(obs_diff_means_bystrata1, length(diff_means_distr_bystrata1)),
@@
   facet_grid(~model) +
-  geom_vline(aes(xintercept = xintercept))
+  geom_vline(aes(xintercept = xintercept)) +
+  theme_bw() +
+  labs(x = "Summed Difference in Means Statistic", y = "Frequency") +
+  theme(legend.position = "none")
 ```
@@
 ## Covariate-adjusted permutation test
-We would like to test for a difference in outcomes between the treatment groups group, but control for other covariates.
+We would like to test for a difference in outcomes between the treatment groups, but control for other covariates.
 In particular, when the outcome is the average response during the second week of follow-up, we would like to control for the average response during the first week of follow-up and the location.
@@
 In other words, the null hypothesis is that $\beta_2 = 0$.
-In a randomized experiment such as this, there are two ways we may test this null hypothesis.
+In a randomized experiment like this, several variables are exchangeable.
+We will show two different permutation tests that exploit exchangeability of different variables.
 
 ### Standard linear model
-First, we may permute the treatment assignments $Z$ directly.
+
 Treatment was assigned at random within each site.
@@
 
-```{r tr_permutation, fig.height = 3, fig.cap = "Permutation distribution of the stratified permutation test between daily_hrdq and treatment after controlling for covariates in a linear model."}
+```{r tr_permutation, fig.align = "center", fig.height = 3, fig.cap = "Permutation distribution of the stratified permutation test between daily hrdq and treatment after controlling for covariates in a linear model."}
 # method 2 : linear model
@@
 data.frame("perm" = c(lm1_t_distr, lm2_t_distr),
-           "model" = c(rep("Outcome", length(lm1_t_distr)), 
-                       rep("Differenced", length(lm2_t_distr))),
+           "model" = c(rep("Raw Outcome Model", length(lm1_t_distr)), 
+                       rep("Differenced Model", length(lm2_t_distr))),
            "xintercept" = c(rep(observed_t1, length(lm1_t_distr)),
@@
   geom_vline(aes(xintercept = xintercept)) +
-  xlab("t Statistic") +
-  ylab("Frequency") +
-  ggtitle("Permutation Distributions")
+  theme_bw() +
+  labs(x = "t Statistic", y = "Frequency") +
+  theme(legend.position = "none")
 ```
@@
 However, now, we do not treat the $\varepsilon$ as random.
-They are simply defined to be the diffence between $Y$ and the data's linear projection onto the plane $\beta_0 + \beta_1 X + \beta_2 Z$.
+They are simply defined to be the difference between $Y$ and the data's linear projection onto the plane $\beta_0 + \beta_1 X + \beta_2 Z$.
 
@@
 Therefore, we may estimate the errors $\hat{\varepsilon}$ by $Y - \hat{Y}$, where $\hat{Y}$ is obtained by regressing $Y$ on $X$ but not $Z$.
-They approximate the true errors $\varepsilon$ from the equation $Y = \beta_0 + \beta_1 X + \beta_2 Z + \varepsilon$.
 The $\varepsilon$ are orthogonal to $X$ and $Z$.
-Therefore, within sites, these estimated $\hat{\varepsilon}$ are approximately exchangeable.
+Assuming that the $\hat{\varepsilon}$ are ``close'' to $\varepsilon$, thenthese estimated $\hat{\varepsilon}$ are approximately exchangeable within sites.
 
@@
 For this dataset, we are guaranteed that treatment $Z$ is independent of $\varepsilon$, as it is randomized within site.
-We should also check that site ID is uncorrelated with $\hat{varepsilon}$.
-If not, then the $\hat{\varepsilon}$ are not exchangeable across sites.
-However, our permutations are done within sites, so even if this is violated, there is not a big issue.
 We check these associations using residual plots in Figure \ref{fig:residual_plots}.
@@
 Sites 4 and 5 have larger variance in the differenced model.
-As mentioned, this should not be an issue since we permute treatments within sites, independently across sites.
+This should not be an issue since we permute treatments within sites, independently across sites.
 (In general, when treatment is not randomly assigned, this condition is necessary for the permutation test to be valid.)
 
-```{r residual_plots, echo = FALSE, message = FALSE, fig.height = 5, fig.cap = "Residual plots of the linear regression of daily_hrdq on treatment and covariates."}
-# Residual plots to check that treatment and covariates are uncorrelated with errors
-
+```{r residual_plots, echo = FALSE, message = FALSE, fig.align = "center", fig.height = 5, fig.cap = "Residual plots of the linear regression of daily hrdq on treatment and covariates.
+These are done to check that treatment and covariates are uncorrelated with the regression errors for the Freedman and Lane style permutation test."}
 p11 <- daily_hrdq %>% 
@@
   xlab("Treatment") + ylab("Residual") + 
-  ggtitle("Outcome Model")
+  ggtitle("Raw Outcome Model") +
+  theme_bw()
 p21 <- daily_hrdq %>% 
@@
   xlab("Site ID") + ylab("Residual") + 
-  ggtitle("Outcome Model")
+  ggtitle("Raw Outcome Model") +
+  theme_bw()
 p12 <- daily_hrdq %>% 
@@
   xlab("Treatment") + ylab("Residual") +
-  ggtitle("Differences Model")
+  ggtitle("Differenced Model") +
+  theme_bw()
 p22 <- daily_hrdq %>% 
@@
   xlab("Site ID") + ylab("Residual") +
-  ggtitle("Differences Model")
+  ggtitle("Differenced Model") +
+  theme_bw()
 grid.arrange(p12, p11, p22, p21, nrow = 2, ncol = 2)
@@
 
-```{r fl_permutation, fig.height = 3, fig.cap = "Permutation distribution of the Freedman-Lane tests of correlation between daily_hrdq and treatment after controlling for covariates."}
+```{r fl_permutation, fig.align = "center", fig.height = 3, fig.cap = "Permutation distribution of the Freedman-Lane tests of correlation between daily hrdq and treatment after controlling for covariates."}
 # method 3 : freedman lane perm residuals
@@
   facet_grid(~model) +
-  geom_vline(aes(xintercept = xintercept))
+  geom_vline(aes(xintercept = xintercept)) +
+  theme_bw() +
+  labs(x = "t Statistic", y = "Frequency") +
+  theme(legend.position = "none")
 
@@
 
-Table \ref{primary_endpoint_res} shows the p-values for each of the four methods, using the two models (treatment controlling for baseline, and the difference of treatment minus baseline).
-For the models using the treatment measure as the dependent variable, the effect of `daily_hrdq` is significant at the $0.05$ level in two of the five models
-and is significant at the $0.1$ level in four of the five models.
-It is never significant in the models using the difference from baseline to treatment.
+Table 2 shows the $p$-values for each of the four methods and the two models.
+For the models using the raw outcome measure as the dependent variable, the effect of `daily_hrdq` is significant at the $0.05$ level in two of the five tests
+and is significant at the $0.1$ level in four of the five tests.
+It is never significant in the models using the difference from baseline to outcome.
 As expected, the stratified permutation test using the sum of differences in means across sites has low power.
-We restrict our attention to the model using treatment measurement and controlling for the baseline, as this is what RB's original analysis does.
-Now, we run this procedure for all continuous variables.
-Table \ref{primary_endpoint_res} shows the results.
 
-```{r primary_endpoint_res, results = "asis", fig.cap = "Comparison of p-values for two measures (average outcome during treatment vs.
+```{r primary_endpoint_res, results = "asis", fig.align = "center", fig.cap = "Comparison of $p$-values for two measures (average outcome during treatment vs.
 difference of average outcome and average baseline) of the primary endpoint."}
@@
 
-tests <- c("Parametric ANCOVA", "Unadjusted permutation", "Unadjusted permutation (summed across strata)", "Linear regression permutation", "Residual permutation")
+tests <- c("Parametric ANCOVA", "Unadjusted permutation", 
+           "Unadjusted permutation (summed across strata)", 
+           "Linear regression permutation", "Residual permutation")
 rownames(pvalues_table) <- tests
@@
 print(xtable(pvalues_table, digits = 3,
-             caption = "Comparison of p-values for two measures (average outcome during treatment vs. difference of average outcome and average baseline) of the primary endpoint."),
+             caption = "Comparison of $p$-values for two 
+             measures (average outcome during treatment 
+             vs. difference of average outcome and average
+             baseline) of the primary endpoint."),
       include.rownames = TRUE)
 ```
+
+# Analysis - Secondary endpoints
+
+We restrict our attention to the model using treatment measurement and controlling for the baseline, as this is what RB's original analysis does.
+We run this procedure for all seven continuous endpoints.
+Table 3 shows the results.
 
@@
 
-tests <- c("Parametric ANCOVA", "Stratified permutation", "LM permutation", "Residual permutation")
+tests <- c("ANCOVA", "Stratified perm", "LM perm", "Residual perm")
 pvalues_table_contin <- as.data.frame(matrix(NA, nrow = length(continuous_vars), ncol = 4))
@@
 ```
-```{r results_table_print, results = "asis", fig.cap = "Comparison of p-values from four tests, for each continuous endpoint."}
+```{r results_table_print, results = "asis", fig.align = "center", fig.cap = "Comparison of $p$-values from four tests, for each continuous endpoint."}
 rownames(pvalues_table_contin) <- continuous_vars
@@
              digits = 3,
-             align = paste0(c("r|", rep("p{1.2in}", ncol(pvalues_table_contin))), 
+             align = paste0(c("r|", rep("p{1in}", ncol(pvalues_table_contin))), 
                             collapse = ""),
-             caption = "Comparison of p-values from four tests, for each continuous endpoint.",
+             caption = "Comparison of $p$-values from four tests, for each continuous endpoint.",
              label = "tab:clinical_pvalues")
@@
 
-print(summarytab, include.rownames = TRUE, type="latex", file="../ms/fig/results_pvalue_summary.tex") 
+print(summarytab, include.rownames = TRUE, type="latex", 
+      file="../ms/fig/results_pvalue_summary.tex") 
 ```

### 2017-02-05 ms/SBR-template.tex
Second round of major paper edits by KO

@@
 % 200 words or less
-Hypothesis tests based on the linear model are widely accepted by organizations that regulate clinical trials.
-We focus on analysis of covariance (ANCOVA), a special case of linear regression that incorporates stratification.
-ANCOVA is derived using strong assumptions about the data-generating process so that the resulting inference is based on parametric distributions.
+Hypothesis tests based on linear models are widely accepted by organizations that regulate clinical trials.
+These tests are derived using strong assumptions about the data-generating process so that the resulting inference is based on parametric distributions.
 Because these methods are well understood and robust, they are sometimes applied to data that depart from assumptions, such as ordinal integer scores.
 Permutation tests are a nonparametric alternative that require minimal assumptions which are often guaranteed by the randomization that was conducted.
-We compare ANCOVA to several permutation tests based on linear models that control for pretreatment covariates.
-In simulated trials using a variety of data-generating processes, some of which violate the parametric assumptions,
-the permutation tests maintain power comparable to the ANCOVA.
+We compare analysis of covariance (ANCOVA), a special case of linear regression that incorporates stratification, to several permutation tests based on linear models that control for pretreatment covariates.
+In simulations using a variety of data-generating processes, some of which violate the parametric assumptions,
+the permutation tests maintain power comparable to ANCOVA.
 We illustrate the use of these permutation tests alongside ANCOVA with data from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease.
@@
 Stratification is one method to control for covariates that are known a priori to be associated with the outcome.
-Strata are groups of individuals with similar levels of a covariate, and random assignment of treatments is conducted within each group, independently across groups.
-The stratification is done \textit{before any outcome data is collected}.
+Strata are groups of individuals with similar levels of a covariate.
+These groups are defined \textit{before any outcome data is collected}.
+Random assignment of treatments is conducted within each stratum, independently across strata.
 This guarantees that the stratification variable is balanced between treatment groups.
@@
 Controlling for stratum can have the same beneficial effects as controlling for other pretreatment covariates, 
-as outcomes may vary across strata due to factors unrelated to individuals.
+as outcomes may vary across strata due to factors unrelated to individuals' treatment and covariates.
 
 Linear regression is another way to control for baseline covariates.
-It is done \textit{after} data are collected and outcomes have been observed.
+It is done \textit{after outcomes have been observed}.
 While stratification guarantees balance on the stratifying variable, other covariates may still be imbalanced.
-Stronger modeling assumptions may be necessary to balance them after the fact.
+Stronger modeling assumptions may be necessary to balance them after the experiment.
 Linear regression is a method which projects the outcomes onto the plane that best summarizes each variable's relationship to the outcome.
@@
 When controlling for strata, such as location, it is standard to use analysis of covariance (ANCOVA).
-ANCOVA is a particular version of linear regression that allows each stratum to have its own mean outcome level.
-This amounts to fitting a plane to each stratum, allowing each plane to have a different intercept but requiring them to have a common slope.
+ANCOVA is a particular case of linear regression that allows the mean outcome to vary from stratum to stratum.
+This amounts to fitting a plane to each stratum, with the constraint that they have a common slope.
 
@@
 When this holds, a regression coefficient scaled by its estimated variance follows Student's $t$ distribution.
-Thus, if we believe the linear model is correctly specified, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the ANCOVA model can be evaluated analytically and quickly.
+Thus, if we believe the linear model is correctly specified, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the ANCOVA model, and can be evaluated analytically and quickly.
 However, the modeling assumptions are clearly violated in many cases:
@@
 
-In the past, parametric tests were necessary because asymptotic approximations were the only computationally feasible way to estimate distributions. 
-Now, computational power is no longer a barrier to finding exact (or exact to pre-specified precision) randomization distributions.
+In the past, parametric methods were necessary because asymptotic approximations were the only computationally feasible way to estimate distributions and construct confidence intervals
+Now, computational power is no longer a barrier to finding exact (or exact to pre-specified precision) randomization distributions, though they remain the most common way of constructing confidence intervals.
 In most cases, a randomization test is the ``gold standard'':
 ``[a] corresponding parametric test is valid only to the extent that it results in the same statistical decision [as the randomization test]'' (\cite{bradley_distribution_1968}).
-Therefore, when conducting creating estimates and confidence intervals using parametric methods, one may have greater confidence in the results when the decision to accept or reject the null hypothesis is the same for the parametric and permutation tests.
+If the permutation test agrees with the parametric test, one may have a greater degree of confidence in the estimates and confidence intervals that come from the parametric method.
 
@@
 However, parametric and nonparametric tests seem to perform similarly when compared side-by-side in simulations, even when data violate the assumptions of the parametric method.
-Medical trials studying pain or perceived severity scores often use Likert scale data, which is discrete and does not match the normality assumptions of parametric tests.
+Medical trials often use Likert scales to score pain or symptom severity, resulting in discrete data that does not match the normality assumptions of parametric tests.
 However, \cite{winter_five-point_2010} found that the two sample $t$ test and Mann-Whitney test had comparable Type I and II error rates for five-point Likert scale data, suggesting that the violation of normality does not entirely invalidate the parametric test.
@@
 \section{Methods}
-% The methods section should include:
-
-%the aim, design and setting of the study
-%the characteristics of participants or description of materials
-%a clear description of all processes, interventions and comparisons. Generic drug names should generally be used. When proprietary brands are used in research, include the brand names in parentheses
-%the type of statistical analysis used, including a power calculation if appropriate
 
@@
 We are interested in comparing the relative effectiveness of treatment 1 to treatment 0.
-Let $\mathbf{Z}$ indicate treatment assignment, 0 or 1, 
-$\mathbf{Y}$ denote the observed outcome of interest,
-and $\mathbf{X}$ be a pretreatment covariate that is observed and associated with the outcome.
-For expository clarity, we suppose that $\mathbf{X}$ is univariate, but all results are easily extended to the case when $\mathbf{X}$ is multivariate.
+Let $Z$ indicate treatment assignment,
+$Y$ denote the observed outcome of interest,
+and $X$ be a pretreatment covariate that is observed and associated with the outcome.
+For expository clarity, we suppose that $X$ is univariate, but all results are easily extended to the case when $X$ is multivariate.
 Suppose further that there is a categorical pretreatment variable with $J$ levels used to stratify individuals.
 
-We observe $\{Y_{ij}, Z_{ij}, X_{ij}\}$ for individuals $i = 1, \dots, n_j$, in strata $j = 1, \dots, J$.
+Strata are indexed by subscript $j = 1, \dots, J$.
+Let $n_j$ be the number of individuals in stratum $j$.
+Individuals within stratum $j$ are indexed by subscript $i = 1, \dots, n_j$.
+We observe $\{Y_{ij}, Z_{ij}, X_{ij}\}$ for $i = 1, \dots, n_j$, $j = 1, \dots, J$.
 All individuals have two potential outcomes, $Y_{ij}(1)$ and $Y_{ij}(0)$, representing their responses to treatments $1$ and $0$, respectively.
@@
 Throughout, we assume that there is no interference between individuals 
-(in other words, $Y_{ij}$ is a function of $(Z_{ij}, Y_{ij}(1), Y_{ij}(0))$ and not any other $Z_{i', j'}$)
-and that there is no non-compliance or study dropout 
-(we actually observe $Y_{ij} = Y_{ij}(Z_{ij})$ for all $i, j$).
+(in other words, $Y_{ij}$ is a function of $(Z_{ij}, Y_{ij}(1), Y_{ij}(0))$ and not any other $Z_{i', j'}$ for $(i', j') \neq (i, j)$.)
+and that there is no censoring or non-compliance
+(we actually observe $Y_{ij} = Y_{ij}(Z_{ij})$ for all $(i, j)$).
 
@@
 it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.
-A typical problem is to estimate the average of differences $Y_{ij}(1) - Y_{ij}(0)$ in some group, such as the study sample or in a target population.
-Which function of these differences will be considered depends on the chosen method of analysis.
-Any of these possible quantities may be of clinical interest; it is a matter of philosophy which one prefers.
-Rather than estimation, we will focus on hypothesis testing for whether these differences are nonzero.
+A typical problem is to estimate the mean difference $Y_{ij}(1) - Y_{ij}(0)$ in some group, such as the study sample or in a target population.
+The chosen method of analysis determines which function of differences is considered.
+Various functions of potential outcomes may be of clinical interest; which one prefers depends on the study's goal and the researcher's personal preference.
+Rather than estimation, we focus on hypothesis testing for whether these differences are nonzero.
 
-\subsection{Parametric ANCOVA}
+\subsection{Parametric ANCOVA}\label{subsec:ancova}
 
+ANCOVA is based on a linear model with an indicator variable for membership in each stratum.
 The model is
@@
 
-To carry out the standard parametric hypothesis test, we make the following assumptions (\cite{freedman_statistical_2005}):
+To carry out the standard parametric hypothesis test for a linear model, we make the following assumptions (\cite{freedman_statistical_2005}):
 
 \begin{enumerate}
-\item \textbf{Linearity:} The data $\mathbf{Y}$ are related to $\mathbf{X}$ and $\mathbf{Z}$ linearly.
-\item \textbf{Constant slopes:} Stratum only affects the intercept $\alpha_j$, not the slopes $\beta$ and $\gamma$.
-\item \textbf{IID Errors:} The $\varepsilon_{ij}$ are independent and identically distributed with mean $0$ and variance $\sigma^2$.
-\item \textbf{Independence:} If $\mathbf{X}$ is random, $\mathbf{\varepsilon}$ is statistically independent of $\mathbf{X}$.
+\item \textbf{Linearity:} The data $Y$ are related to $X$ and $Z$ linearly.
+\item \textbf{Constant slopes:} Stratum membership only affects the intercept $\alpha_j$, not the slopes $\beta$ and $\gamma$.
+\item \textbf{IID Errors:} The $\varepsilon_{ij}$ are independent and identically distributed with mean $0$ and common variance $\sigma^2$.
+\item \textbf{Independence:} If $X$ is random, $\varepsilon$ is statistically independent of $X$.
 \item \textbf{Normality:} The errors are normally distributed.
@@
 (Note, only Assumptions (1) and (2) are needed to guarantee the existence of a solution.
-The solution will be unique as long as there is no linear relationship between $\mathbf{X}$, $\mathbf{Z}$, and stratum membership.)
+The solution will be unique as long as there is no linear relationship between $X$, $Z$, and stratum membership.)
 This procedure yields an estimate of $\hat{\sigma}_{\hat{\gamma}}^2$, the  standard error of $\hat{\gamma}$
-(namely, the diagonal element corresponding to the treatment variable in the inverse covariance matrix of regression covariates).
+(namely, the corresponding diagonal element of the inverse covariance matrix of regression covariates).
 Under the null hypothesis, the test statistic 
@@
 
-The ANCOVA model allows for fixed stratum effects that are additive in the outcome.  
-The model does not account for variation in the treatment effect across strata.  
+The ANCOVA model includes a fixed effect for each stratum, but 
+the model does not account for variation in the treatment effect across strata.  
 The estimated linear model may not accurately capture details of the true data-generating process:
@@
 A test of the null hypothesis that $\gamma = 0$ will be valid, but will not reflect the true magnitude of treatment effects among individuals.
+One can account for differential treatment effects across strata by including interaction terms for treatment and stratum membership, but this complicates the hypothesis testing problem.
 
-\subsection{Stratified permutation test}
+\subsection{Stratified permutation test}\label{subsec:strat_perm_test}
 Suppose we wish to test the null hypothesis that individual by individual, treatment has no effect.
@@
 For the permutation test, we condition on the number of individuals who received each treatment within each stratum.
-Any assignment of treatments which preserves the number of treated units within each stratum is valid and was just as likely to occur as the randomization that was actually observed.
-We can construct the permutation distribution of any statistic under the null hypothesis by using this principle of equal probabilities and imputing the unobserved potential outcomes under the sharp null hypothesis.
+Any assignment of treatments that preserves the number of treated units within each stratum is valid and was just as likely to occur as the randomization that was actually observed.
+We can construct the permutation distribution of any statistic under the null hypothesis using this principle of equal probabilities and by imputing the unobserved potential outcomes using the sharp null hypothesis.
 
@@
 However, the difference in means may not be optimal if we want to be sensitive to heterogeneous effects.  
-For an extreme example, imagine that there an equal number of males and females in the sample, and each treatment is assigned to half of males and half of females.  
+For an extreme example, imagine that the sample contains an equal number of males and females, and each treatment is assigned to half of males and half of females.  
 Everybody who receives treatment 0 has an outcome of 0, but males who receive treatment 1 have an outcome of $1$ and females who receive treatment 1 have an outcome of $-1$.  
@@
 To avoid this, one may want to stratify the sample according to important confounding variables, then compute a statistic for each stratum separately. 
-There is a great degree of freedom in choosing how exactly to construct such a statistic.
-Two factors must be decided: how to stratify and how to combine the statistics across strata.
+There is a great degree of freedom in deciding how to construct such a statistic.
+Two choices must be made: how to stratify and how to combine the statistics across strata.
 There is a tradeoff in how finely we choose to stratify.
-On the one hand, the strata must be sufficiently large to be informative,
+On the one hand, each stratum must be sufficiently large to be informative,
  but on the other hand, must be fine enough to capture variation in treatment effects.
-Our simulations in Section~\ref{simulations} show a loss of power when effects are concentrated in small strata.
+Our simulations in Section~\ref{simulations} show that power decreases when effects are concentrated in small strata.
 
-One way to construct a test statistic is to directly combine the statistics into a single statistic on the data, for instance taking the sum of their absolute values.
+One way to construct a test statistic is to directly combine the stratum statistics into a single value, for instance taking the sum of their absolute values.
 Taking the absolute value before summing ensures that effects with different signs do not cancel each other out.
-The permutation test is conducted identically, using this statistic in place of the difference in means.
-Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by Pesarin and Salmaso \cite{pesarin_permutation_2010}.
+Permutations for this test are conducted identically to those in the previous section, but this combined statistic is used in place of the difference in means.
+Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by \citet{pesarin_permutation_2010}.
 In this framework, one would break down the ``global'' null hypothesis of no effect whatsoever into the intersection of ``partial'' null hypotheses of no treatment effect within each stratum.
 For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies.
-In this case, strata are randomized independently so there are no dependencies to preserve, making this method is equivalent to the first.
-In general, these tests based will not be equivalent.
+In randomized experiments, strata are randomized independently, making this method equivalent to combining stratum statistics directly.
 
-\subsection{Permutation tests with the linear model}
-The tests in the previous section only use information on the treatment and the outcome.
+\subsection{Permutation tests with the linear model}\label{subsec:lm_perm_tests}
+The tests in Section~\ref{subsec:strat_perm_test} only use information on the treatment and the outcome.
 However, experimenters record additional covariates, some of which may be predictive of the outcome.
-One would like to construct a more powerful test by incorporating these covariates in a way to reduce the variance of the statistic.
-In particular, we will continue to model the outcome as in Equation~\ref{eqn:ancova}, but use fewer assumptions about the data to conduct a nonparametric permutation test.
+One would like to construct a more powerful test by incorporating these covariates to reduce the variance of the statistic.
+The permutation tests in this section continue to model the outcome as in Equation~\ref{eqn:ancova}, but require fewer assumptions about the data.
 In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes,
-making several things ``permutable'' and allowing us to construct several different tests.
+making several variables ``permutable''.
+We show two permutation tests developed in this framework.
 
 First, we may do a simple variation on the stratified permutation test described above.
-Instead of using the difference in means as the statistic, we use the $t$ statistic for the coefficient on $\mathbf{Z}$ in the linear regression.
-If treatment was assigned at random within each stratum, then it is guaranteed that $\mathbf{Z}$ and $\mathbf{\varepsilon}$ are statistically independent, conditional on stratum.
-Therefore, we may permute treatment assignments $\mathbf{Z}$ within strata, independently across strata, and calculate a test statistic for each such permutation to obtain a null distribution.
+Instead of using the difference in means as the statistic, we use the $t$ statistic for the coefficient on $Z$ in the linear regression.
+If treatment was assigned at random within each stratum, then it is guaranteed that $Z$ and $\varepsilon$ are statistically independent, conditional on stratum.
+Therefore, we can construct a permutation distribution by repeatedly permuting treatment assignments $Z$ within strata, independently across strata, and regressing the outcomes on the covariate and permuted treatment vectors.
 
 \citet{freedman_nonstochastic_1983} propose an alternative test based on the residuals of the linear regression.
-They take an alternative view of the problem, still writing the outcome in the form of Equation~\ref{eqn:ancova}, but instead of treating the $\varepsilon_{ij}$ as random they define the errors to be the difference between the observed outcome $Y_{ij}$ and the data's linear projection onto the plane $\alpha_j + \beta X_{ij}+ \gamma Z_{ij}$.
-The $\mathbf{\varepsilon}$ are fixed approximation errors in this framework.
+They take an alternative view of the problem, still writing the outcome in the form of Equation~\ref{eqn:ancova}, but they \textit{define} the errors $\varepsilon_{ij}$ to be the difference between the observed outcome $Y_{ij}$ and the data's linear projection onto the plane $\alpha_j + \beta X_{ij}+ \gamma Z_{ij}$.
+The $\varepsilon$ are fixed approximation errors in this framework, not independent and identically distributed random errors.
 
 If the null hypothesis is true, then $\gamma = 0$ and $\varepsilon_{ij} = Y_{ij} - \alpha_{ij} - \beta X_{ij}$ for all $i = 1, \dots, n_j, j = 1, \dots, J$.
-We may regress $\mathbf{Y}$ on $\mathbf{X}$ \textit{and not $\mathbf{Z}$} to obtain coefficient estimates $\hat{\alpha}_j$ and $\hat{\beta}$.
-Then we estimate the errors $\mathbf{\hat{\varepsilon}}$ by $\mathbf{Y} - \mathbf{\hat{Y}}$, where $\mathbf{\hat{Y}}$ is the vector of fitted values $\hat{Y}_{ij} = \hat{\alpha}_j + \hat{\beta}X_{ij}$.
-The $\mathbf{\hat{\varepsilon}}$ approximate the true errors $\mathbf{\varepsilon}$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit.
-(\cite{freedman_nonstochastic_1983} advise against using this method if there are large outlier values in the covariate $\mathbf{X}$ or when $\mathbf{X}$ and $\mathbf{Z}$ are highly colinear.)
-Furthermore, under the null hypothesis, the $\mathbf{\varepsilon}$ are independent $\mathbf{Z}$ within strata. 
-If we make the additional assumption that $\mathbf{\varepsilon}$ are exchangeable against $\mathbf{X}$ within strata, then 
+Therefore, we may estimate the errors $\hat{\varepsilon}$ by $Y - \hat{Y}$, where $\hat{Y}$ is the vector of fitted values from the regression of $Y$ on $X$ but not $Z$.
+The $\hat{\varepsilon}$ approximate the true errors $\varepsilon$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit.
+(\cite{freedman_nonstochastic_1983} advise against using this method if there are large outlier values in the covariate $X$ or when $X$ and $Z$ are highly colinear.)
+Furthermore, under the null hypothesis, the $\varepsilon$ are independent $Z$ within strata. 
+If we make the additional assumption that $\varepsilon$ are exchangeable against $X$ within strata, then 
 we may permute the estimated $\hat{\varepsilon}$ within strata, independently across strata.
+This additional exchangeability assumption should be checked using the residuals $\hat{\varepsilon}$;
+for instance, visual inspection of a scatterplot of the residuals against $X$ should not reveal any patterns if they are exchangeable.
 
@@
 \begin{enumerate}
-\item Estimate $\mathbf{\hat{\varepsilon}}$ by $Y_{ij} - \hat{\alpha}_j - \hat{\beta} X_{ij}$ for all $i$ and $j$, where $\hat{\alpha}_j$ and $\hat{\beta}$ are obtained by regressing $\mathbf{Y}$ on $\mathbf{X}$ and stratum assignment, \textit{but not $\mathbf{Z}$}.
-\item Obtain permuted errors $\mathbf{\hat{\varepsilon}}^\pi$ by permuting the $\mathbf{\hat{\varepsilon}}$ within strata.
+\item Estimate $\hat{\varepsilon}$ by $Y_{ij} - \hat{\alpha}_j - \hat{\beta} X_{ij}$ for all $i$ and $j$, where $\hat{\alpha}_j$ and $\hat{\beta}$ are obtained by regressing $Y$ on $X$ and stratum assignment, \textit{but not $Z$}.
+\item Obtain permuted errors $\hat{\varepsilon}^\pi$ by permuting the $\hat{\varepsilon}$ within strata.
 \item Compute permuted responses $Y_{ij}^\pi = \hat{\alpha}_j + \hat{\beta} X_{ij}+ \hat{\varepsilon}_{ij}^\pi$.
-\item Regress $\mathbf{Y}^\pi$ on $\mathbf{X}$, $\mathbf{Z}$, and stratum. The test statistic is the $t$ statistic for the coefficient of $\mathbf{Z}$.
+\item Regress $Y^\pi$ on $X$, $Z$, and stratum. The test statistic is the $t$ statistic for the coefficient of $Z$.
 \end{enumerate}
@@
 only the estimates $\hat{\alpha}_j$ and $\hat{\beta}$.
-\cite{manly_randomization_2006} proposes randomizing the outcomes $\mathbf{Y}$, treating them as though they were randomly assigned to pairs $(X_{ij}, Z_{ij})$ under the null hypothesis.
-Permutations under the null treat all coefficients as 0, which may not reflect the true data-generating process.
+\cite{manly_randomization_2006} proposed randomizing the outcomes $Y$, treating them as though they were randomly assigned to pairs $(X_{ij}, Z_{ij})$ under the null hypothesis.
+Permutations under the null treat assume that all coefficients are 0 in the true data-generating process, which may not reflect the true relationship between variables.
 \cite{kennedy_randomization_1995} proposed a permutation method similar in spirit to \cite{freedman_nonstochastic_1983}, but which differs procedurally.
 Both methods attempt to measure the correlation between treatment and unexplained variation in outcomes, but
-instead of regressing pseudo-outcomes $\mathbf{Y}^\pi$ on covariates to obtain a permutation distribution, \cite{kennedy_randomization_1995} recommends using the $t$ statistic from regressing $\mathbf{Z}$ on the permuted residuals $\mathbf{\hat{\varepsilon}}^\pi$.
+instead of regressing pseudo-outcomes $Y^\pi$ on covariates to obtain a permutation distribution, \cite{kennedy_randomization_1995} recommends using the $t$ statistic from regressing $Z$ on the permuted residuals $\hat{\varepsilon}^\pi$.
 
-Several authors have compared these tests theoretically and through simulations \cite{anderson_empirical_1999, anderson_permutation_2001, kennedy_randomization_1996}.
-They find that the Freedman-Lane test of residuals is asymptotically equivalent to the ``oracle'' exact hypothesis test which we could conduct if we knew which permutations of $\mathbf{Y}$ given $\mathbf{X}$ were equiprobable under the null hypothesis \cite{anderson_permutation_2001}.
-This confirms empirical results, which show that the Freedman-Lane test performs better than other linear regression based tests in simulations \cite{anderson_empirical_1999}, though its advantage is small.
-Therefore, throughout the rest of the paper we focus on the two linear regression permutations we described in detail: linear regression after permuting treatment assignments and the Freedman-Lane method of permuting residuals.
+Several authors have compared these tests theoretically and through simulations (\cite{anderson_empirical_1999, anderson_permutation_2001, kennedy_randomization_1996}).
+They find that the Freedman-Lane test of residuals is asymptotically equivalent to the ``oracle'' exact hypothesis test which we could conduct if we knew which permutations of $Y$ given $X$ were equiprobable under the null hypothesis (\cite{anderson_permutation_2001}).
+This confirms empirical results, which show that the Freedman-Lane test performs better than other linear regression based tests in simulations (\cite{anderson_empirical_1999}), though its advantage is small.
+Therefore, throughout the rest of the paper we focus on the two linear regression permutation tests we described in detail: linear regression with permuted treatment assignments and the Freedman-Lane method of permuting residuals.
 
@@
 
-We simulated data coming from a randomized experiment using several different data-generating processes.
+We simulated data from a randomized experiment using several different data-generating processes.
 We applied the tests to the data and compared their empirical power over repeated random treatment assignments and random errors.
@@
 and the Freedman-Lane permutation test.
-(We decided against using the NPC statistic computed by taking the difference in means within each stratum and summing their absolute values over strata. 
-However, this would not be ideal for imbalanced designs, which are common in real-world applications.  
-See the clinical data analysis file for a more detailed treatment.) \todo{supplementary ref}
+ \todo{supplementary ref}
 
@@
 \noindent We drew the baseline measurements once, treating them as fixed, and conditioned on observing $\{ Y_{ij0} : {i = 1,\dots,n_j, j = 1,\dots, 3}\}$.
-Then, for 1000 trials, we randomly assigned treatment to half of the individuals within each stratum, generated random errors, and constructed new outcomes $Y_{ij1}$ using Equation~\ref{eqn:dgp}.
+We randomly assigned treatment to half of the individuals within each stratum, generated random errors, and constructed new outcomes $Y_{ij1}$ using Equation~\ref{eqn:dgp}.
 We conducted all five tests on this new data, obtaining a two-sided $p$-value for each.
+We repeated these steps using the same $\{ Y_{ij0} \}$ 1000 times.
 
@@
 There were four total simulation designs:
-for each of the two possible treatment effects, we used two distributions of $\mathbf{\varepsilon}$.
-In the first case, we used $\varepsilon \sim N(0, 1)$ to fit with the parametric assumption of normal errors.
-In the second case, $\mathbf{\varepsilon}$ were drawn from a $t$ distribution with 2 degrees of freedom so the errors were heavy-tailed.
+for each of the two possible treatment effects, we used two distributions for $\varepsilon$.
+In the first case, we used $\varepsilon \sim N(0, 1)$ to match the parametric assumption of normal errors.
+In the second case, $\varepsilon$ were drawn from a $t$ distribution with 2 degrees of freedom so the errors were heavy-tailed.
 
 Figure~\ref{fig:normal_sim_power} shows the estimated power curves in these four designs.
-The best case is when the errors are Gaussian and the treatment effect is constant across strata, while the worst case is when the errors are heavy-tailed and the treatment effect only appears in one stratum.
+The best case was when the errors were Gaussian and the treatment effect was constant across strata, while the worst case was when the errors are heavy-tailed and the treatment effect only appeared in one stratum.
 Intuitively, it makes sense that power decreased relative to the Gaussian, constant treatment effect case, as each violated assumption further obscured the treatment effect.
-A consistent pattern appears in each design: the stratified permutation test has the lowest power, while the other four tests tend to have similar power.
+A consistent pattern appeared in each design: the stratified permutation test had the lowest power, while the other four tests tended to have similar power.
 
 Table~\ref{tab:normal_power} displays the size of the test and the power at level $0.05$.
-The three tests based on linear models have slightly higher than nominal level, though the margin is within two standard errors of $0.05$.
+The three tests based on linear models had slightly higher than nominal level, though the margin was within two standard errors of $0.05$.
 (The number of tests rejected under the null in 1000 trials has a binomial distribution.
 If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.)
-These numbers show that actually, for $\alpha=0.05$, the stratified permutation test does not lose a large amount of power when the effect is isolated in a single stratum.
-In this case, the permutation test may have slightly higher power than the parametric ANCOVA at the small significance levels used in practice.
+These numbers show that actually, for $\alpha=0.05$, the stratified permutation test did not lose a large amount of power when the effect was isolated in a single stratum.
+In this case, the stratified permutation test may have slightly higher power than ANCOVA at the small significance levels typically used in practice.
 \begin{figure}[h]
@@
 Within each stratum, 4 individuals received one treatment and 12 received the other.
-This mimics real-world experiments, where the more expensive treatment is administered less frequently than the other.
+This mimics some real-world experiments, where the more expensive treatment is administered less frequently than the placebo or standard of care.
 It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exacerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
-Here, the data have the same variance so we would expect power, not the test level, to be an issue.
-(See the supplementary Gaussian simulations for a discussion of heterogeneous variances.)
-We omit a plot of the power curves from these simulations, as the patterns appear similar to those in Figure~\ref{fig:normal_sim_power}.
+Here, the data had equal variances so we would expect power, not the test level, to be an issue.
+(See the supplementary Gaussian simulations for simulations using data with heterogeneous variances.)
 Table~\ref{tab:imbalanced_power} displays the value of these power curves and the size of the test at level $0.05$.
-In this case, the permutation tests (except for the simple stratified permutation) have slightly higher power than the ANCOVA, but the difference is not substantial.
+In this case, the permutation tests (except for the simple stratified permutation) had slightly higher power than ANCOVA, but the difference was not substantial.
 %\begin{figure}
@@
 In our third set of simulations, we made the baseline measurements $Y_{ij0}$ discrete and skewed.
-This type of data occurs in some surveys where individuals are asked to rate their symptom severity or some experience on a scale from 1 to 10, for instance.
+This type of data may occur in surveys where individuals are asked to rate their symptom severity on a discrete ordinal scale.
 We generated the baseline measures using independent draws from a Poisson distribution with mean 4 and censored values at 10.
-We conditioned on these observed baseline measures after we having generated them once.
-Then, for each 1000 trials, we randomly assigned treatment to half of the individuals in each stratum and generated errors taking on values $0.5$ and $-0.5$ with equal probability.
+We conditioned on these observed baseline measures after generating them once.
+Then, for each of 1000 trials, we randomly assigned treatment to half of the individuals in each stratum and generated errors taking on values $0.5$ and $-0.5$ with equal probability.
 We constructed $Y_{ij1}$ using Equation~\ref{eqn:dgp} with a treatment effect of $\gamma = 0.5$.
-However, we didn't observe $Y_{ij1}$ but instead integer values $\tilde{Y}_{ij1}$ defined as
+To ensure that both baseline and outcome were integer scores, we let the $Y_{ij1}$ be unobservable and defined the observed outcome as
 
@@
      \begin{array}{ll}
-       1 & \text{if } \tilde{Y}_{ij1} < 1\\
-       10 & \text{if } \tilde{Y}_{ij1} > 10 \\
-       \lfloor \tilde{Y}_{ij1} \rfloor & \text{otherwise}
+       1 & \text{if } {Y}_{ij1} < 1\\
+       10 & \text{if } {Y}_{ij1} > 10 \\
+       \lfloor {Y}_{ij1} \rfloor & \text{otherwise}
      \end{array}
@@
 
-The observed outcomes $\tilde{Y}_{ij1}$, the baseline measures $Y_{ij0}$, and the errors are discrete.
-The normality assumption is clearly violated here.
+The observed outcomes $\tilde{Y}_{ij1}$, the baseline measures $Y_{ij0}$, and the errors were discrete, violating normality assumptions.
 
@@
 For each of the two treatment effect designs, we considered both equal and unequal stratum sizes.
-First, we considered balanced samples as before, where each stratum contained $n_j = 16$ individuals.
-Next, we considered imbalanced strata, where the smallest stratum had only $n_1=8$ individuals, the next had $n_2= 16$, and the largest had $n_3=24$.
-The design with a single stratum effect and unequal stratum sizes was particularly unfavorable for hypothesis testing:
-the nonzero effect only occured in the smallest stratum.
+The simulations with equal stratum sizes were generated as before, where each stratum contained $n_j = 16$ individuals.
+The simulations with unequal stratum sizes were set up so the smallest stratum had only $n_1=8$ individuals, the next had $n_2= 16$, and the largest had $n_3=24$.
+The design with a nonzero treatment effect in a single stratum and unequal stratum sizes was particularly unfavorable:
+the nonzero effect only occurred in the smallest stratum.
 
 Figure~\ref{fig:skewed_sim_power} shows the estimated power curves in these four designs.
-When the treatment effect is constant, sample sizes within each stratum don't matter; 
+When the treatment effect was constant, sample sizes within each stratum didn't matter; 
 the two power curve plots in the first row of Figure~\ref{fig:skewed_sim_power} look the same up to chance variation.
-When the treatment effect only occurs in one stratum, there is a substantial loss in power.
-It is nearly impossible to detect a difference in treatments when the affected stratum is too small.
+When the treatment effect only occurred in one stratum, there was a substantial loss in power.
+It was nearly impossible to detect a difference in treatments when the affected stratum was too small.
 In practice, one does not usually know a priori which individuals will be affected by treatment (otherwise, those not affected would be excluded from the study altogether).
-This result suggests that when stratifying, one must be cautious not to stratify too finely, otherwise treatment effects may be indistinguishable from random variability.
+This result suggests that when stratifying, one must be cautious not to stratify too finely, as it is impossible to measure an effect with an insufficient sample size.
 
-All of the tests appear to be conservative and have lower than nominal level.
-As in the first set of simulations, the stratified permutation test using the raw outcomes has the lowest power of the five tests.
-Now, however, the stratified permutation test using the change scores also has lower power than the three linear model tests.
-Presumably, this is because the differences $\tilde{Y}_{ij1} - Y_{ij0}$ are discrete and lie in a small range of values, so the test statistic does not vary greatly across permutations.
-Table~\ref{tab:skewed_power} confirms this: power for the three linear model methods is high when the treatment is constant across strata, and power for all of the methods suffers when the effect is isolated in a single stratum.
+All of the tests appeared to be conservative and to have lower than nominal level.
+As in the first set of simulations, the stratified permutation test using the raw outcomes had the lowest power of the five tests.
+The stratified permutation test using the change scores also had lower power than the three linear model tests.
+Presumably, this was because the differences $\tilde{Y}_{ij1} - Y_{ij0}$ were discrete and limited to a small number of values, so the test statistic could not vary greatly across permutations.
+Table~\ref{tab:skewed_power} confirms this: power for the three linear model methods was high when the treatment was constant across strata, and power for all of the methods suffered when the effect was isolated in a single stratum.
 
@@
 It makes sense that in all of these simulations, the permutation test using the change scores as the response measure is more powerful than the permutation test using the raw outcomes.
-The baseline and outcome are highly correlated, so the change score has lower variance the outcome alone.
+The baseline and outcome were highly correlated, so the change score had lower variance the outcome alone. \todo{introduce the term change score earlier}
 In additional simulations, 
-we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had correlation of $0.25$ (see the supplementary Gaussian simulations for a more detailed treatment).
+we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had a correlation of $0.25$.
 In this case, the result was reversed: 
 the differenced test had low power while the test using the raw outcomes had a power curve closer to the linear model-based tests.
-This behavior has been studied before by \cite{frison_repeated_1992}, who recommended to use the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
+(See the supplementary Gaussian simulations for a more detailed treatment.)
+This behavior has been studied before by \cite{frison_repeated_1992}, who recommended using the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
 
@@
 We compared the parametric ANCOVA, the stratified permutation test, and the two linear model-based permutation tests using a dataset from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease (GERD).
+A detailed discussion of the data and analysis is provided in a supplementary file.
+We summarize the analysis here.
 Patients were treated at eight sites in two different countries.
@@
 Daily ``hrdq'' was the primary endpoint.
-To reduce day-to-day variation, we averaged the measures from each week to obtain two time points per patient, one pretreatment and one post-treatment.
+To reduce day-to-day variation, we averaged the measures from each week to obtain two observations per patient, one pre-treatment and one post-treatment.
 
-We treated site as the stratification variable, as this is the level at which treatment was randomized.
+We used site as the stratification variable, as this is the level at which treatment was randomized.
 We did not include country in the model, as a country-level effect should be captured in the site-level effects.
 The model used for the linear regression-based tests was defined as in Equation~\ref{eqn:ancova}.
-This model allowed for the intercept $\alpha_j$ to vary across sites and used the pretreatment, baseline measurement as the control variable $\mathbf{X}$.
-The outcome and baseline had a low correlation (for instance, $0.56$ for daily ``hrdq''), so we used raw outcomes and not the change scores as the dependent variable. 
-As our simulations and previous work \cite{frison_repeated_1992} show, when the correlation between baseline and outcome is low, it is more powerful to control for baseline outcomes using a model.
+This model allowed the intercept $\alpha_j$ to vary across sites and used the pretreatment, baseline measurement as the control variable $X$.
+The outcome and baseline had a low correlation (for instance, the correlation between pre- and post-treatment daily ``hrdq'' was $0.56$), so we used raw outcomes and not the change scores as the dependent variable.
+As our simulations and previous work (\cite{frison_repeated_1992}) show, when the correlation between baseline and outcome is low, it is more powerful to control for baseline outcomes using a model.
 
 Figure~\ref{fig:clinical_distr} shows the distribution of each clinical endpoint for the two treatment groups.
-There is a clear difference in distributions for daily heartburn (``daily\_heart''), daily ``hrdq'', and heartburn frequency (``heart\_freq'').
+There is a clear difference in distributions for daily heartburn (``daily\_heart''), daily ``hrdq'' (``daily\_hrdq''), and heartburn frequency (``heart\_freq'').
 The difference is less clear for daily regurgitation (``daily\_regurg'') and regurgitation frequency (``regurg\_freq'').
 
-Table~\ref{tab:clinical_pvalues} shows the $p$-values for the four tests.
+Table~\ref{tab:clinical_pvalues} shows the $p$-values for the four tests and the seven continuous study endpoints.
 Overall, the results confirm our expectations based on visual comparison in Figure~\ref{fig:clinical_distr}:
@@
 The three tests based on the linear model all give similar results here.
-The $p$-values for the ANCOVA tend to be smaller than the linear model and the residual permutation tests, while the $p$-values stratified, unadjusted permutation test have no consistent pattern.
+The $p$-values from ANCOVA tend to be smaller than the linear model and the residual permutation tests, while the $p$-values stratified, unadjusted permutation test have no consistent pattern.
 The parametric test has comparable or slightly higher power than the other permutation tests.
+
+\todo{need a conclusion}
 
@@
 \section{Discussion}
-% This section should discuss the implications of the findings in context of existing research and highlight limitations of the study.
 
 This paper adds to the literature comparing parametric and nonparametric tests.
-Our results match those of \cite{vickers_parametric_2005, anderson_empirical_1999}, where parametric ANCOVA and regression-based $t$ tests performed the same or better than the comparable nonparametric tests.
-We simulated a variety of data-generating processes ranging from the ideal case when all assumptions are met 
-to the case when data are non-normal, discrete, or skewed,
-as well as data with other possible issues such as heterogeneous treatment effects and imbalanced sample sizes.
-When the ANCOVA assumptions were violated, the linear regression-based tests lost power relative to the best case data-generating process where all assumptions were satisfied.
-However, the power of the tests remained the same \textit{relative to each other}.
+Our results match those of \citet{vickers_parametric_2005} and \citet{anderson_empirical_1999}, where parametric ANCOVA and regression-based $t$ tests performed the same or better than the comparable nonparametric tests.
+We simulated a variety of data-generating processes, ranging from the ideal case when all assumptions are met, 
+to the case when data are non-normal, discrete, or skewed.
+In the supplementary simulation files, we also considered data with other possible issues such as heterogeneous treatment effects and unequal sample sizes.
+The linear regression-based tests, both the parametric ANCOVA and the permutation tests, suffered a loss of power when the ANCOVA assumptions were violated.
+However, the tests remained comparable to each other.
 It is a matter of taste which test one chooses for their experiment: while the parametric test may be robust to violations of its assumptions, it is somewhat reassuring that the permutation test can exactly match the randomization that was done while making no distributional assumptions.
+\todo{reiterate secondary analysis point}
 
@@
 For instance, he points to the case where observations have unequal variances.  
-This is a problem as we cannot observe the errors; it is a leap of faith to assert that they are homogeneous and therefore exchangeable.
-\cite{boik_fisherpitman_1987} illustrates this phenomenon using the normal theory $F$ test and its permutation counterpart, 
-and demonstrates by simulation that the latter has larger than nominal level when the error variances are heterogeneous.
+This is a problem in practice as one cannot observe errors; it is a leap of faith to assert that they are homogeneous and therefore exchangeable.
+\cite{boik_fisherpitman_1987} illustrates this phenomenon using the traditional $F$ test and its permutation counterpart, 
+and demonstrates by simulation that the latter has larger than nominal level when the error variances are unequal.
 
-It is important to note that the permutation tests described here are approximately exact only in the context of randomized experiments.
-In this case, treatment is assigned at random and is therefore statistically independent of the covariates $\mathbf{X}$ and the errors $\mathbf{\varepsilon}$.
-Conversely, in observational studies, treatment may be associated with $\mathbf{X}$ and/or $\mathbf{\varepsilon}$, often in a way that is difficult or impossible to model.
-Independence, or the weaker condition of exchangeability, is what enables us to construct permutation distributions while holding $\mathbf{X}$ fixed.
-When exchangeability doesn't hold, we cannot disentangle the effect of $\mathbf{Z}$ from the effect of $\mathbf{X}$.
-One must provide evidence that items in observational studies are exchangeable in order to achieve an approximately exact test.
-The onus is on the researcher to make the case that the object being permuted is uncorrelated with the other variables being held fixed.
-This can be done visually using scatterplots and residual plots \cite{freedman_nonstochastic_1983} and from subject-matter knowledge.
+It is important to note that the permutation tests described here are exact only in the context of randomized experiments.
+Treatment is assigned at random and is therefore statistically independent of the covariates $X$ and the errors $\varepsilon$.
+In observational studies, treatment may be associated with $X$, $\varepsilon$, or both, often in a way that is difficult or impossible to model.
+The statistical independence guaranteed by randomized experiments enables us to construct permutation distributions while holding $X$ fixed.
+When exchangeability doesn't hold, we cannot disentangle the effect of $Z$ from the effect of $X$.
+One must provide evidence that variables in observational studies are exchangeable in order to achieve an approximately exact test.
+The onus is on the researcher to make the case that the variable being permuted is uncorrelated with the other variables being held fixed.
+This can be done visually using scatterplots and residual plots \cite{freedman_nonstochastic_1983} and from knowledge of how the data arose and were collected.
 
-One should be mindful of which covariates to control for.
-Ideally, one knows and observes all covariates related to the outcome.
-In that case, the coefficient for treatment in a fully saturated linear model (including all covariates and their interactions) is an asymptotically consistent estimate of the average treatment effect \cite{lin_agnostic_2013}.
-It is rare that one knows \textit{all} relevant covariates in practice; failure to control for all relevant variables can introduce bias in certain permutation tests.
+One must be mindful of which covariates to include as controls.
+Ideally, all covariates related to the outcome are known and observed.
+In that case, the coefficient for treatment in a fully saturated linear model (including all covariates and their interactions) is an asymptotically consistent estimate of the average treatment effect (\cite{lin_agnostic_2013}).
+It is rare that one knows \textit{all} relevant covariates in practice; failure to control for all relevant variables can introduce bias in some permutation tests.
  \cite{gail_tests_1988} propose a randomization test based on residuals from an exponential family model and find that omitting relevant covariates leads to tests with higher than nominal level.
@@
 Weak correlations between baseline and outcome occur often in practice, as was the case with our GERD dataset.
-If the correlation is weak, then the resulting test using the difference may be less powerful than ignoring the baseline altogether.
+When the correlation is weak, the test of change scores may be less powerful than ignoring the baseline altogether.
 Instead, we suggest incorporating the baseline in a regression model.
 This is more general than using differences;
-treating the change scores as the dependent variable is a special case of the linear regression, constraining the coefficient on the baseline measure to be $1$.
+treating the change scores as the dependent variable is a special case of the linear regression that constrains the coefficient of the baseline measure to be $1$.
 The regression approach is more flexible and demonstrably more powerful.
@@
 \begin{center}
+\todo{}
 {\large\bf SUPPLEMENTARY MATERIAL}

### 2017-02-05 results/results.Rmd
Minor changes to results Rmd

@@
 4. a stratified permutation test in the style of Freedman and Lane (1983), using a linear regression to adjust for covariates.
+
+The permutation tests rely on the `permuter` R package.  If you haven't installed it, you should run the following command:
+
+```{r download_permuter, eval = FALSE}
+devtools::install_github("statlab/permuter")
+```
 
@@
 Therefore, we may estimate the errors $\hat{\varepsilon}$ by $Y - \hat{Y}$, where $\hat{Y}$ is obtained by regressing $Y$ on $X$ but not $Z$.
-The $\varepsilon$ are orthogonal to $X$ and $Z$.
-Assuming that the $\hat{\varepsilon}$ are ``close'' to $\varepsilon$, thenthese estimated $\hat{\varepsilon}$ are approximately exchangeable within sites.
+Assuming that the $\hat{\varepsilon}$ are ``close'' to $\varepsilon$, then these estimated $\hat{\varepsilon}$ are approximately exchangeable within sites.
 

### 2017-02-14 ms/SBR-template.tex
More edits to ms

@@
 {
-  \title{\bf A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Control Trials}
+  \title{\bf A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Experiments}
 \author{Kellie Ottoboni \\
 Department of Statistics; Berkeley Institute for Data Science\\
-University of California, Berkeley \\ 
+University of California, Berkeley \\ [.1in]
 Fraser Lewis \\
-RB\\ 
+Medical Affairs and Evidence Generation \\
+Reckitt Benckiser\\  [.1in]
 Luigi Salmaso\\
@@
 \newpage
-%\spacingset{1.45} % DON'T change the spacing!
+\spacingset{1.45} % DON'T change the spacing!
 \section{Background}
@@
 Techniques have been developed to deal with each of these issues, invoking additional assumptions in order to carry out valid inference.
-%For any particular dataset, it is difficult to determine to what extent the assumptions for asymptotic normality are satisfied.
 In this paper, we focus on the issue of heterogeneity of pretreatment characteristics and assume that the other issues are not present.
@@
 ``[a] corresponding parametric test is valid only to the extent that it results in the same statistical decision [as the randomization test]'' (\cite{bradley_distribution_1968}).
-If the permutation test agrees with the parametric test, one may have a greater degree of confidence in the estimates and confidence intervals that come from the parametric method.
+If the permutation test agrees with the parametric test, one may have a greater degree of confidence in the estimates and confidence intervals constructed using the parametric method.
 
@@
 Most similar to our question of study, \cite{anderson_empirical_1999} found little difference between several permutation tests for coefficients in a linear model alongside the parametric $t$ test.
+In these situations, the permutation test strengthens conclusions by giving evidence that the parametric test is robust to departures from its assumptions.
 
@@
 We focus on ANCOVA and its permutation counterparts, comparing their performance in different scenarios and illustrating their application with a clinical dataset.
-We find in simulations that even when assumptions are not satisfied, the parametric and permutation tests have comparable power to detect a treatment effect.
-We apply each of the methods to data from a clinical trial which assessed the performance of two treatments for gastroesophageal reflux disease (GERD).
-We conclude by discussing implications of these results for practitioners.
+Section~\ref{sec:methods} describes the various tests mathematically and states their assumptions explicitly.
+Section~\ref{sec:simulations} presents simulations that suggest that even when assumptions are not satisfied, the parametric and permutation tests have comparable power to detect a treatment effect.
+In Section~\ref{sec:results}, we apply each of the tests to data from a clinical trial comparing the performance of two treatments for gastroesophageal reflux disease (GERD).
+We conclude in Section~\ref{sec:discussion} with implications of these results for practitioners.
 
-\section{Methods}
+\section{Methods}\label{sec:methods}
 
@@
 In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes,
-making several variables ``permutable''.
+making several variables ``permutable.''
 We show two permutation tests developed in this framework.
@@
 
-Several authors have compared these tests theoretically and through simulations (\cite{anderson_empirical_1999, anderson_permutation_2001, kennedy_randomization_1996}).
+Several authors have compared these tests theoretically and empirically (\cite{anderson_empirical_1999, anderson_permutation_2001, kennedy_randomization_1996}).
 They find that the Freedman-Lane test of residuals is asymptotically equivalent to the ``oracle'' exact hypothesis test which we could conduct if we knew which permutations of $Y$ given $X$ were equiprobable under the null hypothesis (\cite{anderson_permutation_2001}).
@@
 
-\section{Results}
-\subsection{Simulations}\label{simulations}
+\section{Simulations}\label{sec:simulations}
 
@@
 and the Freedman-Lane permutation test.
- \todo{supplementary ref}
+The simulations are included in three supplementary files.
 
@@
 It makes sense that in all of these simulations, the permutation test using the change scores as the response measure is more powerful than the permutation test using the raw outcomes.
-The baseline and outcome were highly correlated, so the change score had lower variance the outcome alone. \todo{introduce the term change score earlier}
+The baseline and outcome were highly correlated, so the change score had lower variance the outcome alone. 
 In additional simulations, 
@@
 
-\subsection{Clinical data results}
+\section{Clinical data results}\label{sec:results}
 We compared the parametric ANCOVA, the stratified permutation test, and the two linear model-based permutation tests using a dataset from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease (GERD).
@@
 The difference is less clear for daily regurgitation (``daily\_regurg'') and regurgitation frequency (``regurg\_freq'').
+The distribution of outcomes for each endpoint is extremely right skewed, calling the assumption of normally distributed error terms into question.
+It is possible that the
 
@@
 but not for any of the other endpoints.
-The three tests based on the linear model all give similar results here.
-The $p$-values from ANCOVA tend to be smaller than the linear model and the residual permutation tests, while the $p$-values stratified, unadjusted permutation test have no consistent pattern.
-The parametric test has comparable or slightly higher power than the other permutation tests.
+The $p$-values for the stratified, unadjusted permutation test have no consistent pattern: sometimes they are smaller than the $p$-values from the other tests and sometimes they are larger.
 
-\todo{need a conclusion}
+The three tests based on the linear model give qualitatively similar results here.
+The ANCOVA $p$-values tend to be smaller than the permutation linear model and residual permutation test $p$-values.
+The permutation tests are somewhat more conservative, accounting for the highly skewed outcome distributions.
+Our conclusions for the heartburn frequency, daily heartburn, and daily ``hrdq'' endpoints differ between ANCOVA and the permutation tests at significance level $0.05$, but not at level $0.1$.
+None of the three tests based on linear models is significant at level $0.1$ for the other four endpoints;
+there is no endpoint which would be deemed significant using a permutation test but insiginificant with ANCOVA.
+This suggests that the parametric test correctly discriminates between endpoints that are significantly different and endpoints that are not different between treatment A and treatment B.
 
@@
 
-\section{Discussion}
+\section{Discussion}\label{sec:discussion}
 
@@
 It is a matter of taste which test one chooses for their experiment: while the parametric test may be robust to violations of its assumptions, it is somewhat reassuring that the permutation test can exactly match the randomization that was done while making no distributional assumptions.
-\todo{reiterate secondary analysis point}
+Applying a linear model based permutation test as a secondary analysis can give insight into how ANCOVA results depend on the method's assumptions
+(making it conservative, anti-conservative, or not systematically affecting the $p$-value in a particular direction),
+and give evidence that inferences based on ANCOVA are reliable.
 
@@
 and demonstrates by simulation that the latter has larger than nominal level when the error variances are unequal.
+Randomized experiments mitigate this problem, but is important to conduct permutation tests by exactly mimicking the randomization that was done.
+For instance, in stratified experiments, it is conceivable that the variance of outcomes differs across strata for reasons unrelated to the treatment.
+Treatment assignment is exchangeable within strata, but not across strata, and a permutation test that ignores this fact may be invalid.
 

### 2017-02-15 ms/SBR-template.tex
Progress on suppl files

@@
   \begin{center}
-    {\LARGE\bf  A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Control Trials}
+    {\LARGE\bf  A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Experiments}
 \end{center}
@@
 Hypothesis tests based on linear models are widely accepted by organizations that regulate clinical trials.
-These tests are derived using strong assumptions about the data-generating process so that the resulting inference is based on parametric distributions.
+These tests are derived using strong assumptions about the data-generating process so that the resulting inference can be based on parametric distributions.
 Because these methods are well understood and robust, they are sometimes applied to data that depart from assumptions, such as ordinal integer scores.
@@
 We illustrate the use of these permutation tests alongside ANCOVA with data from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease.
-Given the considerable costs and scientific importance of clinical trials, one may want to include an additional robust method, such as a linear model permutation test, as a check on the statistical inference for the main study endpoints.
+Given the considerable costs and scientific importance of clinical trials, one may want to include an additional nonparametric method, such as a linear model permutation test, as a robustness check on the statistical inference for the main study endpoints.
 \end{abstract}
@@
 \noindent%
-{\it Keywords:}  nonparametric, linear model, ANCOVA, inference
+{\it Keywords:}  Nonparametric methods, linear model, analysis of covariance, analysis of designed experiments
 \vfill
@@
 In this case, the permutation tests (except for the simple stratified permutation) had slightly higher power than ANCOVA, but the difference was not substantial.
-%\begin{figure}
-%\centering
-%\includegraphics[width = \textwidth]{fig/imbalanced_simulation_power}
-%\caption{Empirical power curves for the Gaussian simulated data with imbalanced group sizes}
-%\label{fig:imbalanced_sim_power}
-%\end{figure}
+
 %\begin{center}
@@
 \begin{center}
-\todo{}
 {\large\bf SUPPLEMENTARY MATERIAL}
 \end{center}
+All files are also available at \url{https://github.com/kellieotto/ancova-permutations}.
 
@@
 
-\item[Title:] Brief description. (file type)
+\item[Gaussian data simulations:] Simulations using Gaussian data in Section~\ref{sec:simulations}, including code and results. (PDF)
+\item[Imbalanced design simulations:] Simulations using data with different stratum sizes in Section~\ref{sec:simulations}, including code and results. (PDF)
+\item[Skewed data simulations:] Simulations using discrete, skewed data in Section~\ref{sec:simulations}, including code and results. (PDF)
+\item[Clinical trial data:] Dataset used in Section~\ref{sec:results} to compare methods. (csv)
+\item[README:] Data descriptor file. (txt)
+\item[Results:] Detailed explanation, code, and results of comparing methods using the clinical trial dataset in Section~\ref{sec:results}. (PDF)
 

### 2017-02-15 simulations/simulation_gaussian.Rmd
Progress on suppl files

--- 2017-01-31 simulations/simulation.Rmd
@@
   ylab("Power") +
-  ggtitle(title)
-
-}
-
-plot_pvalue_hist <- function(pvalue_mat, title){
-  melt(pvalue_mat) %>%
-  mutate("Method" = Var2) %>%
-  ggplot(aes(x = value, fill = Method)) +
-  geom_histogram() +
-  facet_wrap(~Method) +
-  ggtitle(title)
-}
-
-plot_pvalue_scatter <- function(pvalue_mat, title){
-  pvalue_mat %>%
-  as.data.frame() %>%
-  select(ANCOVA, strat = starts_with("Stratified")) %>%
-  ggplot(aes(x = ANCOVA, y = strat)) + 
-  geom_point() +
-  xlim(0, 1) + ylim(0, 1) +
-  ylab("Stratified Permutation") +
-  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
-  ggtitle(title)
+  ggtitle(title) +
+  theme_bw() +
+  theme(
+    axis.text.x = element_text(size = 12),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 16),
+    title = element_text(size = 16),
+    legend.title = element_text(size = 12),
+    legend.text = element_text(size = 14),
+    strip.text.x = element_text(size = 12)
+  ) 
 }
@@
 We follow the procedure described above, using an effect size of $\gamma = 0$ at all strata and using standard normal errors.
-To have the correct level means that the test rate of rejection at level $\alpha$ is $\alpha 100\%$.
+To have the correct level means that the test rate of rejection at level $\alpha$ is $100\alpha\%$.
 In other words, the p-values are uniformly distributed and the power curve should coincide with the line with slope $1$ through the origin.
-Figure \ref{fig:design_null_plots} demonstrates that this is the case.
-If anything, the differenced stratified permutation test has fewer than $\alpha 100\%$ false positives when using level $\alpha$.
+The figure below demonstrates that this is the case.
+If anything, the differenced stratified permutation test has fewer than $100\alpha\%$ false positives when using level $\alpha$.
 
@@
 plot_power_curves(design0_power, "No Effect, Normal Errors")
-plot_pvalue_hist(design0_pvalues, "No Effect, Normal Errors")
 ```
@@
 plot_power_curves(design00_power, "No Effect, Heavy-tailed Errors")
-plot_pvalue_hist(design00_pvalues, "No Effect, Heavy-tailed Errors") 
 ```
@@
 ```{r design1_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design1_power, "Constant Additive Effect of 1, Normal Errors")
-plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 1, Normal Errors")
+plot_power_curves(design1_power, "Constant Additive Effect of 1 \n Normal Errors")
 ```
@@
 ```{r design2_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design2_power, "Effect of 1 at a Single stratum, Normal Errors")
-plot_pvalue_hist(design2_pvalues, "Effect of 1, Normal Errors")
+plot_power_curves(design2_power, "Effect of 1 at a Single stratum \n Normal Errors")
 ```
@@
 ```{r design3_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design3_power, "Constant Additive Effect of 1, Heavy-tailed Errors")
-plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 1, Heavy-tailed Errors")
+plot_power_curves(design3_power, "Constant Additive Effect of 1 \n Heavy-tailed Errors")
 ```
@@
 ```{r design4_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design4_power, "Effect of 1 at a Single stratum, Heavy-tailed Errors") 
-plot_pvalue_hist(design4_pvalues, "Effect of 1, Heavy-tailed Errors")
+plot_power_curves(design4_power, "Effect of 1 at a Single stratum \n Heavy-tailed Errors") 
 ```
@@
 ```{r design5_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design5_power, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
-plot_pvalue_hist(design5_pvalues, "Constant Additive Effect of 1, Normal Errors \n Unpredictive baseline")
+plot_power_curves(design5_power, "Constant Additive Effect of 1 \n Normal Errors \n Unpredictive baseline")
 ```
@@
 ```{r design6_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design6_power, "Constant Additive Effect of 1, Normal Errors")
-plot_pvalue_hist(design6_pvalues, "Constant Additive Effect of 1, Normal Errors")
+plot_power_curves(design6_power, "Constant Additive Effect of 1 \n Normal Errors")
 ```
@@
 ```{r design7_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design7_power, "Effect of 1 at a Single stratum, Normal Errors")
-plot_pvalue_hist(design7_pvalues, "Effect of 1, Normal Errors")
+plot_power_curves(design7_power, "Effect of 1 at a Single stratum \n Normal Errors")
 ```
@@
 ```{r design8_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design8_power, "Constant Additive Effect of 1, Heteroskedastic Errors")
-plot_pvalue_hist(design8_pvalues, "Constant Additive Effect of 1, Heteroskedastic Errors")
+plot_power_curves(design8_power, "Constant Additive Effect of 1 \n Heteroskedastic Errors")
 ```
@@
 ```{r design9_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design9_power, "Effect of 1 at a Single stratum, Heteroskedastic Errors")
-plot_pvalue_hist(design9_pvalues, "Effect of 1, Heteroskedastic Errors")
+plot_power_curves(design9_power, "Effect of 1 at a Single stratum \n Heteroskedastic Errors")
 ```

### 2017-02-15 simulations/simulation_imbalanced.Rmd
Suppl file progress for imbalanced sims

@@
 
-We use two designs:
-
-* Design 1: $\gamma_1 = \gamma_2 = \gamma_3 = 1$. This is the standard assumption of a constant, additive treatment effect.
-* Design 2: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$.
-This is a constant, additive treatment effect in stratum 1, but no treatment effect in stratums 2 and 3.
-This is a simplistic case of heterogeneous treatment effects.
-
-With these two designs, we vary the distribution of $\varepsilon$.
-In the first case, we use $\varepsilon \sim N(0, 1)$ to mimic the usual ANCOVA assumptions.
-In the second case, $\varepsilon \sim t(2)$ so the errors are heavy-tailed.
-Thus, there are four total simulation designs.
-
-We compare five tests:
-
-* ANCOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for stratum.
-* Stratified permutation: we permute treatment assignment within stratum, then take the difference in means between treated and control outcomes $Y_1$
-* Differenced permutation: we do the same permutation procedure as the stratified permutation test, except we use the difference between outcome and baseline, $Y_1 - Y_0$
-* Linear model (LM) permutation: we use the same stratified permutation procedure as above, except use the $t$-statistic for the coefficient on treatment in the linear regression of $Y_1$ on $Y_0$, $Z$, and stratum dummies
-* Freedman-Lane test: see the other Rmd document for a full description of this procedure
-
-Throughout our simulations, we first fix $Y_0$ and stratum ID.
-Treatment $Z$ and the errors $\varepsilon$ are randomly drawn according to their respective distributions.
-Then, $Y_1$ is constructed using the linear data-generating process above.
-We regenerate $Z$, $\varepsilon$, and $Y_1$ 100 times for each design, then compute the empirical power of the five tests.
+The same simulation guidelines as detailed in the Gaussian simulation file.
 
@@
   ylab("Power") +
-  ggtitle(title)
+  ggtitle(title) +
+  theme_bw() +
+  theme(
+    axis.text.x = element_text(size = 12),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 16),
+    title = element_text(size = 16),
+    legend.title = element_text(size = 12),
+    legend.text = element_text(size = 14),
+    strip.text.x = element_text(size = 12)
+  ) 
 }
 
-plot_pvalue_hist <- function(pvalue_mat, title){
-  melt(pvalue_mat) %>%
-  mutate("Method" = Var2) %>%
-  ggplot(aes(x = value, fill = Method)) +
-  geom_histogram() +
-  facet_wrap(~Method) +
-  ggtitle(title)
-}
-
-plot_pvalue_scatter <- function(pvalue_mat, title){
-  pvalue_mat %>%
-  as.data.frame() %>%
-  select(ANCOVA, strat = starts_with("Stratified")) %>%
-  ggplot(aes(x = ANCOVA, y = strat)) + 
-  geom_point() +
-  xlim(0, 1) + ylim(0, 1) +
-  ylab("Stratified Permutation") +
-  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
-  ggtitle(title)
-}
 ```
@@
 
-Before testing for different kinds of effects, we begin checking that the tests have the correct level.
-We follow the procedure described above, using an effect size of $\gamma = 0$ at all strata and using standard normal errors.
-To have the correct level means that the test rate of rejection at level $\alpha$ is $\alpha 100\%$.
-In other words, the p-values are uniformly distributed and the power curve should coincide with the line with slope $1$ through the origin.
-Figure \ref{fig:design_null_plots} demonstrates that this is the case.
-If anything, the differenced stratified permutation test has fewer than $\alpha 100\%$ false positives when using level $\alpha$.
+The figures below show that the test has the correct level under the null for both error distributions.
 
@@
 ```{r design_null_plots, fig.height = 4, fig.align="center"}
-plot_power_curves(design0_power, "No Effect, Normal Errors")
+plot_power_curves(design0_power, "No Effect \n Normal Errors")
 
-plot_pvalue_hist(design0_pvalues, "No Effect, Normal Errors")
 ```
@@
 ```{r design_null_heavy_plots, fig.height = 4, fig.align="center"}
-plot_power_curves(design00_power, "No Effect, Heavy-tailed Errors")
+plot_power_curves(design00_power, "No Effect \n Heavy-tailed Errors")
 
-plot_pvalue_hist(design00_pvalues, "No Effect, Heavy-tailed Errors") 
 ```
@@
 There is no discernable difference in power between the ANCOVA, the differenced stratified permutation test, the LM permutations, or the Freedman-Lane test.
-However, the simple stratified permutation test of $Y_1$ has substantially less power than the other four.
+Only the simple stratified permutation test of $Y_1$ has substantially less power than the other four.
 Without controlling for the baseline values, the variance in $Y_1$ masks the treatment effect.
@@
 ```{r design1_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design1_power, "Constant Additive Effect of 1, Normal Errors")
+plot_power_curves(design1_power, "Constant Additive Effect of 1 \n Normal Errors")
 
-plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 1, Normal Errors")
 ```
@@
 
-A similar pattern emerges: the simple stratified permutation test has low power, while the other four power curves roughly coincide.
-The Freedman-Lane test may have the highest power for small p-values, but this could also just be noise.
-All five power curves are closer to the line passing through the origin: since the effect is only present at one stratum, it is more difficult to detect.
+The results here are similar to the Gaussian simulations.
+We omit comments for the rest: these is simply a loss in power for all tests relative to the balanced design.
 
@@
 ```{r design2_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design2_power, "Effect of 1 at a Single stratum, Normal Errors")
+plot_power_curves(design2_power, "Effect of 1 at a Single stratum \n Normal Errors")
 
-plot_pvalue_hist(design2_pvalues, "Effect of 1, Normal Errors")
 ```
@@
 ## Design 3: Constant additive effect, heavy-tailed errors
-
-Again, we find that the four power curves coincide while the curve for the stratified permutation test is below the others.
-Here, the difference between the curves is not large.
-Controlling for baseline $Y_0$ does not substantially help reduce variance.
 
@@
 ```{r design3_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design3_power, "Constant Additive Effect of 1, Heavy-tailed Errors")
+plot_power_curves(design3_power, "Constant Additive Effect of 1 \n Heavy-tailed Errors")
 
-plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 1, Heavy-tailed Errors")
 ```
@@
 ## Design 4: single stratum effect, heavy-tailed errors
-
-The noise from the heavy-tailed errors masks the treatment effect so much that controlling for baseline measures makes no difference.
-There is almost no power to detect an effect using any of the five tests.
 
@@
 ```{r design4_plot, fig.height = 4, fig.align="center"}
-plot_power_curves(design4_power, "Effect of 1 at a Single stratum, Heavy-tailed Errors")
+plot_power_curves(design4_power, "Effect of 1 at a Single stratum \n Heavy-tailed Errors")
 
-plot_pvalue_hist(design4_pvalues, "Effect of 1, Heavy-tailed Errors")
 ```

### 2017-02-15 ms/ms.tex
Fix typo in ms

@@
-\documentclass[11pt]{article}
+\documentclass[12pt]{article}
 
-%\documentclass{bmcart}
-
-%%% Load packages
-%\usepackage{amsthm,amsmath}
-%\RequirePackage{natbib}
-%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
-%\RequirePackage{hyperref}
-\usepackage[utf8]{inputenc} %unicode support
-
-\usepackage[breaklinks=true]{hyperref}
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage{enumerate}
+\usepackage{natbib}
+\usepackage{url} % not crucial - just used below for the URL 
 \usepackage{color}
-\usepackage{amsmath,amssymb,amsthm}
-\usepackage[numbers]{natbib}
-\usepackage{array}
-%\usepackage{booktabs, multicol, multirow}
-\usepackage[nohead, margin=1in]{geometry}
-%\usepackage[singlespacing]{setspace}
-\usepackage{graphicx}
-
-%%% Put your definitions there:
-%\startlocaldefs
-
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{assumption}{Assumption}
-
-\newcommand{\beq}{\begin{equation}}
-\newcommand{\eeq}{\end{equation}}
-\newcommand{\bit}{\begin{itemize}}
-\newcommand{\eit}{\end{itemize}}
-
+ \usepackage{multirow}
 \newcommand{\todo}[1]{{\color{red}{TO DO: \sc #1}}}
 
-\newcommand{\reals}{\mathbb{R}}
-\newcommand{\integers}{\mathbb{Z}}
-\newcommand{\naturals}{\mathbb{N}}
-\newcommand{\rationals}{\mathbb{Q}}
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{0}
 
-\newcommand{\ind}{\mathbb{I}} % Indicator function
-\newcommand{\pr}{\mathbb{P}} % Generic probability
-\newcommand{\ex}{\mathbb{E}} % Generic expectation
-\newcommand{\var}{\textrm{Var}}
-\newcommand{\cov}{\textrm{Cov}}
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.3in}%
+\addtolength{\topmargin}{-.8in}%
 
-\newcommand{\normal}{N} % for normal distribution (can probably skip this)
-\newcommand{\eps}{\varepsilon}
-\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
-\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
-\newcommand{\argmax}{\textrm{argmax}}
-\newcommand{\argmin}{\textrm{argmin}}
-\renewcommand{\baselinestretch}{1.5}
-
-%\endlocaldefs
-
-\newcommand{\parttitle}[1]{\textbf{#1}}
-
-%%% Begin ...
 \begin{document}
 
-%%% Start of article front matter
-%\begin{frontmatter}
+%\bibliographystyle{natbib}
 
-%\begin{fmbox}
-%\dochead{Research}
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the title of your article here     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\title{A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Control Trials}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors here                   %%
-%%                                          %%
-%% Specify information, if available,       %%
-%% in the form:                             %%
-%%   <key>={<id1>,<id2>}                    %%
-%%   <key>=                                 %%
-%% Comment or delete the keys which are     %%
-%% not used. Repeat \author command as much %%
-%% as required.                             %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%\author[
-%   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
-%   corref={aff1},                       % id of corresponding address, if any
-%   noteref={n1},                        % id's of article notes, if any
-%   email={jane.e.doe@cambridge.co.uk}   % email address
-%]{\inits{JE}\fnm{Jane E} \snm{Doe}}
-%\author[
-%   addressref={aff1,aff2},
-%   email={john.RS.Smith@cambridge.co.uk}
-%]{\inits{JRS}\fnm{John RS} \snm{Smith}}
-
+\if0\blind
+{
+  \title{\bf A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Experiments}
 \author{Kellie Ottoboni \\
 Department of Statistics; Berkeley Institute for Data Science\\
-University of California, Berkeley \\ [.2in]
+University of California, Berkeley \\ [.1in]
 Fraser Lewis \\
-RB\\ [.2in]
+Medical Affairs and Evidence Generation \\
+Reckitt Benckiser\\  [.1in]
 Luigi Salmaso\\
@@
 University of Padova
-}\date{Draft \today}
-\maketitle
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors' addresses here        %%
-%%                                          %%
-%% Repeat \address commands as much as      %%
-%% required.                                %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+}  \maketitle
+} \fi
 
-%\address[id=aff1]{%                           % unique id
-%  \orgname{Department of Zoology, Cambridge}, % university, etc
-%  \street{Waterloo Road},                     %
-%  %\postcode{}                                % post or zip code
-%  \city{London},                              % city
-%  \cny{UK}                                    % country
-%}
-%\address[id=aff2]{%
-%  \orgname{Marine Ecology Department, Institute of Marine Sciences Kiel},
-%  \street{D\"{u}sternbrooker Weg 20},
-%  \postcode{24105}
-%  \city{Kiel},
-%  \cny{Germany}
-%}
+\if1\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf  A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Experiments}
+\end{center}
+  \medskip
+} \fi
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter short notes here                   %%
-%%                                          %%
-%% Short notes will be after addresses      %%
-%% on first page.                           %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%\begin{artnotes}
-%%\note{Sample of title note}     % note to the article
-%\note[id=n1]{Equal contributor} % note, connected to author
-%\end{artnotes}
-%
-%\end{fmbox}% comment this for two column layout
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Abstract begins here                 %%
-%%                                          %%
-%% Please refer to the Instructions for     %%
-%% authors on http://www.biomedcentral.com  %%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\begin{abstractbox}
-
+\bigskip
 \begin{abstract}
-%\parttitle{Background}
-Statistical test of treatment effects based on the linear model, such as analysis of variance (ANOVA) and analysis of covariance (ANCOVA), are widely accepted by organizations that regulate clinical trials.
-In this paper, we focus on ANCOVA, a special case of linear regression which incorporates stratification.
-Traditional ANCOVA is derived using strong assumptions about the data-generating process so that the resulting inference is based on parametric distributions.
-Because these methods are well understood and robust, they are sometimes applied to data that depart from the method's assumptions, such as ordinal integer scores.
-Given the considerable costs and scientific importance of clinical trials, it is desirable to include an additional methodology which is potentially more robust to departures from assumptions as a check on the statistical inference for the main study endpoints.
-%\parttitle{Methods}
-Permutation tests are a nonparametric alternative that require only minimal assumptions which are often guaranteed by the randomization that was conducted.
-In contrast to parametric tests, which rely up on strong assumptions to obtain a null distribution with a simple closed form, permutation tests use the assumption of exchangeability or randomization to obtain null distributions directly by considering what would have been observed under different possible randomizations.
-We compare parametric ANCOVA to several permutation tests based on linear models, which control for pretreatment covariates to increase power and precision.
-%\parttitle{Results}
-We simulate randomized trials using a variety of data-generating processes, some of which violate the parametric assumptions.
-The permutation tests based on linear models maintain power comparable to the ANCOVA in every case.
-We illustrate the use of these permutation tests alongside the parametric ANCOVA with data from a clinical trial conducted across eight sites,
-each having a different numbers of patients.
-%\parttitle{Conclusions}
-These results suggest that a linear model permutation test may serve as a secondary analysis to give strength to the statistical inferences from the primary analyses of a clinical data study.
-
+% 200 words or less
+Hypothesis tests based on linear models are widely accepted by organizations that regulate clinical trials.
+These tests are derived using strong assumptions about the data-generating process so that the resulting inference can be based on parametric distributions.
+Because these methods are well understood and robust, they are sometimes applied to data that depart from assumptions, such as ordinal integer scores.
+Permutation tests are a nonparametric alternative that require minimal assumptions which are often guaranteed by the randomization that was conducted.
+We compare analysis of covariance (ANCOVA), a special case of linear regression that incorporates stratification, to several permutation tests based on linear models that control for pretreatment covariates.
+In simulations using a variety of data-generating processes, some of which violate the parametric assumptions,
+the permutation tests maintain power comparable to ANCOVA.
+We illustrate the use of these permutation tests alongside ANCOVA with data from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease.
+Given the considerable costs and scientific importance of clinical trials, one may want to include an additional nonparametric method, such as a linear model permutation test, as a robustness check on the statistical inference for the main study endpoints.
 \end{abstract}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The keywords begin here                  %%
-%%                                          %%
-%% Put each keyword in separate \kwd{}.     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\noindent%
+{\it Keywords:}  Nonparametric methods, linear model, analysis of covariance, analysis of designed experiments
+\vfill
 
-%\begin{keyword}
-%\kwd{permutation test}
-%\kwd{linear model}
-%\kwd{ANCOVA}
-%\end{keyword}
-
-% MSC classifications codes, if any
-%\begin{keyword}[class=AMS]
-%\kwd[Primary ]{}
-%\kwd{}
-%\kwd[; secondary ]{}
-%\end{keyword}
-%
-%\end{abstractbox}
-%%
-%%\end{fmbox}% uncomment this for twcolumn layout
-%
-%\end{frontmatter}
-
-\section*{Background}
+\newpage
+\spacingset{1.45} % DON'T change the spacing!
+\section{Background}
 A hypothesis test is a statistical method for determining whether observed data is consistent with a belief about the process that generated the data.
@@
 
-In the past, parametric tests were necessary because asymptotic approximations were the only computationally feasible way to estimate distributions. 
-Now, computational power is no longer a barrier to finding exact (or exact to pre-specified precision) randomization distributions.
+In the past, parametric methods were necessary because asymptotic approximations were the only computationally feasible way to estimate distributions and construct confidence intervals
+Now, computational power is no longer a barrier to finding exact (or exact to pre-specified precision) randomization distributions, though they remain the most common way of constructing confidence intervals.
 In most cases, a randomization test is the ``gold standard'':
 ``[a] corresponding parametric test is valid only to the extent that it results in the same statistical decision [as the randomization test]'' (\cite{bradley_distribution_1968}).
-Therefore, when conducting creating estimates and confidence intervals using parametric methods, one may have greater confidence in the results when the decision to accept or reject the null hypothesis is the same for the parametric and permutation tests.
+If the permutation test agrees with the parametric test, one may have a greater degree of confidence in the estimates and confidence intervals constructed using the parametric method.
 
@@
 We focus on ANCOVA and its permutation counterparts, comparing their performance in different scenarios and illustrating their application with a clinical dataset.
-We find in simulations that even when assumptions are not satisfied, the parametric and permutation tests have comparable power to detect a treatment effect.
-We apply each of the methods to data from a clinical trial which assessed the performance of two treatments for gastroesophageal reflux disease (GERD).
-We conclude by discussing implications of these results for practitioners.
+Section~\ref{sec:methods} describes the various tests mathematically and states their assumptions explicitly.
+Section~\ref{sec:simulations} presents simulations that suggest that even when assumptions are not satisfied, the parametric and permutation tests have comparable power to detect a treatment effect.
+In Section~\ref{sec:results}, we apply each of the tests to data from a clinical trial comparing the performance of two treatments for gastroesophageal reflux disease (GERD).
+We conclude in Section~\ref{sec:discussion} with implications of these results for practitioners.
 
-\section*{Methods}
-% The methods section should include:
+\section{Methods}\label{sec:methods}
 
-%the aim, design and setting of the study
-%the characteristics of participants or description of materials
-%a clear description of all processes, interventions and comparisons. Generic drug names should generally be used. When proprietary brands are used in research, include the brand names in parentheses
-%the type of statistical analysis used, including a power calculation if appropriate
-
-\subsection*{Notation}
+\subsection{Notation}
 Suppose we randomly assign two treatments labelled $0$ and $1$ to a group of individuals.
 We are interested in comparing the relative effectiveness of treatment 1 to treatment 0.
-Let $\mathbf{Z}$ indicate treatment assignment, 0 or 1, 
-$\mathbf{Y}$ denote the observed outcome of interest,
-and $\mathbf{X}$ be a pretreatment covariate that is observed and associated with the outcome.
-For expository clarity, we suppose that $\mathbf{X}$ is univariate, but all results are easily extended to the case when $\mathbf{X}$ is multivariate.
+Let $Z$ indicate treatment assignment,
+$Y$ denote the observed outcome of interest,
+and $X$ be a pretreatment covariate that is observed and associated with the outcome.
+For expository clarity, we suppose that $X$ is univariate, but all results are easily extended to the case when $X$ is multivariate.
 Suppose further that there is a categorical pretreatment variable with $J$ levels used to stratify individuals.
 
-We observe $\{Y_{ij}, Z_{ij}, X_{ij}\}$ for individuals $i = 1, \dots, n_j$, in strata $j = 1, \dots, J$.
+Strata are indexed by subscript $j = 1, \dots, J$.
+Let $n_j$ be the number of individuals in stratum $j$.
+Individuals within stratum $j$ are indexed by subscript $i = 1, \dots, n_j$.
+We observe $\{Y_{ij}, Z_{ij}, X_{ij}\}$ for $i = 1, \dots, n_j$, $j = 1, \dots, J$.
 All individuals have two potential outcomes, $Y_{ij}(1)$ and $Y_{ij}(0)$, representing their responses to treatments $1$ and $0$, respectively.
@@
 This quantity represents the effect of treatment for individual $i$ in group $j$: 
-it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.\footnote{
-Which function of these differences will be considered depends on the chosen method of analysis.
-Any of these possible quantities may be of clinical interest; it is a matter of philosophy which one prefers.}
-A typical problem is to estimate the average of differences $Y_{ij}(1) - Y_{ij}(0)$ in some group, such as the study sample or in a target population.
-Instead of estimation, we will focus on hypothesis testing for whether these differences are nonzero.
+it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.
+A typical problem is to estimate the mean difference $Y_{ij}(1) - Y_{ij}(0)$ in some group, such as the study sample or in a target population.
+The chosen method of analysis determines which function of differences is considered.
+Various functions of potential outcomes may be of clinical interest; which one prefers depends on the study's goal and the researcher's personal preference.
+Rather than estimation, we focus on hypothesis testing for whether these differences are nonzero.
 
-\subsection*{Parametric ANCOVA}
+\subsection{Parametric ANCOVA}\label{subsec:ancova}
 
+ANCOVA is based on a linear model with an indicator variable for membership in each stratum.
 The model is
@@
 
-To carry out the standard parametric hypothesis test, we make the following assumptions (\cite{freedman_statistical_2005}):
+To carry out the standard parametric hypothesis test for a linear model, we make the following assumptions (\cite{freedman_statistical_2005}):
 
 \begin{enumerate}
-\item \textbf{Linearity:} The data $\mathbf{Y}$ are related to $\mathbf{X}$ and $\mathbf{Z}$ linearly.
-\item \textbf{Constant slopes:} Stratum only affects the intercept $\alpha_j$, not the slopes $\beta$ and $\gamma$.
-\item \textbf{IID Errors:} The $\eps_{ij}$ are independent and identically distributed with mean $0$ and variance $\sigma^2$.
-\item \textbf{Independence:} If $\mathbf{X}$ is random, $\mathbf{\varepsilon}$ is statistically independent of $\mathbf{X}$.
+\item \textbf{Linearity:} The data $Y$ are related to $X$ and $Z$ linearly.
+\item \textbf{Constant slopes:} Stratum membership only affects the intercept $\alpha_j$, not the slopes $\beta$ and $\gamma$.
+\item \textbf{IID Errors:} The $\varepsilon_{ij}$ are independent and identically distributed with mean $0$ and common variance $\sigma^2$.
+\item \textbf{Independence:} If $X$ is random, $\varepsilon$ is statistically independent of $X$.
 \item \textbf{Normality:} The errors are normally distributed.
@@
 
-The coefficients are estimated using least squares.\footnote{
-Note, only Assumptions (1) and (2) are needed to guarantee the existence of a solution.
-The solution will be unique as long as there is no linear relationship between $\mathbf{X}$, $\mathbf{Z}$, and stratum membership.
-}
-From this procedure, we may estimate $\hat{\sigma}_{\hat{\gamma}}^2$, the  standard error of $\hat{\gamma}$\footnote{
-$\hat{\sigma}_{\hat{\gamma}}^2$ is estimated by first computing the residual standard error $\hat{\sigma}^2$,
-then taking the diagonal element corresponding to the treatment variable in the inverse covariance matrix of covariates.
-}.
+The coefficients are estimated using least squares.
+(Note, only Assumptions (1) and (2) are needed to guarantee the existence of a solution.
+The solution will be unique as long as there is no linear relationship between $X$, $Z$, and stratum membership.)
+This procedure yields an estimate of $\hat{\sigma}_{\hat{\gamma}}^2$, the  standard error of $\hat{\gamma}$
+(namely, the corresponding diagonal element of the inverse covariance matrix of regression covariates).
 Under the null hypothesis, the test statistic 
@@
 A test of the null hypothesis that $\gamma = 0$ will be valid, but will not reflect the true magnitude of treatment effects among individuals.
+One can account for differential treatment effects across strata by including interaction terms for treatment and stratum membership, but this complicates the hypothesis testing problem.
 
-\subsection*{Stratified permutation test}
+\subsection{Stratified permutation test}\label{subsec:strat_perm_test}
 Suppose we wish to test the null hypothesis that individual by individual, treatment has no effect.
@@
 To avoid this, one may want to stratify the sample according to important confounding variables, then compute a statistic for each stratum separately. 
-There is a great degree of freedom in choosing how exactly to construct such a statistic.
-Two factors must be decided: how to stratify and how to combine the statistics across strata.
+There is a great degree of freedom in deciding how to construct such a statistic.
+Two choices must be made: how to stratify and how to combine the statistics across strata.
 There is a tradeoff in how finely we choose to stratify.
-On the one hand, the strata must be sufficiently large to be informative,
+On the one hand, each stratum must be sufficiently large to be informative,
  but on the other hand, must be fine enough to capture variation in treatment effects.
-Our simulations in Section~\ref{simulations} show a loss of power when effects are concentrated in small strata.
+Our simulations in Section~\ref{simulations} show that power decreases when effects are concentrated in small strata.
 
-One way to construct a test statistic is to directly combine the statistics into a single statistic on the data, for instance taking the sum of their absolute values.
+One way to construct a test statistic is to directly combine the stratum statistics into a single value, for instance taking the sum of their absolute values.
 Taking the absolute value before summing ensures that effects with different signs do not cancel each other out.
-The permutation test is conducted identically, using this statistic in place of the difference in means.
-Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by Pesarin and Salmaso \cite{pesarin_permutation_2010}.
+Permutations for this test are conducted identically to those in the previous section, but this combined statistic is used in place of the difference in means.
+Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by \citet{pesarin_permutation_2010}.
 In this framework, one would break down the ``global'' null hypothesis of no effect whatsoever into the intersection of ``partial'' null hypotheses of no treatment effect within each stratum.
-For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies\footnote{ 
-We considered using the difference in means within each stratum, aggregated by taking the sum of their absolute values over stratum.
-However this would not be ideal for imbalanced designs, which are common in real-world applications, so we did not consider it in the simulations.
-See the clinical data analysis file for a more detailed treatment.}.
-In this case, strata are randomized independently so there are no dependencies to preserve, making this method is equivalent to the first.
-In general, these tests based will not be equivalent.
+For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies.
+In randomized experiments, strata are randomized independently, making this method equivalent to combining stratum statistics directly.
 
-\subsection*{Permutation tests with the linear model}
-The tests in the previous section only use information on the treatment and the outcome.
+\subsection{Permutation tests with the linear model}\label{subsec:lm_perm_tests}
+The tests in Section~\ref{subsec:strat_perm_test} only use information on the treatment and the outcome.
 However, experimenters record additional covariates, some of which may be predictive of the outcome.
-One would like to construct a more powerful test by incorporating these covariates in a way to reduce the variance of the statistic.
-In particular, we will continue to model the outcome as in Equation~\ref{eqn:ancova}, but use fewer assumptions about the data to conduct a nonparametric permutation test.
+One would like to construct a more powerful test by incorporating these covariates to reduce the variance of the statistic.
+The permutation tests in this section continue to model the outcome as in Equation~\ref{eqn:ancova}, but require fewer assumptions about the data.
 In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes,
-making several things ``permutable'' and allowing us to construct several different tests.
+making several variables ``permutable.''
+We show two permutation tests developed in this framework.
 
 First, we may do a simple variation on the stratified permutation test described above.
-Instead of using the difference in means as the statistic, we use the $t$ statistic for the coefficient on $\mathbf{Z}$ in the linear regression.
-If treatment was assigned at random within each stratum, then it is guaranteed that $\mathbf{Z}$ and $\mathbf{\varepsilon}$ are statistically independent, conditional on stratum.
-Therefore, we may permute treatment assignments $\mathbf{Z}$ within strata, independently across strata, and calculate a test statistic for each such permutation to obtain a null distribution.
+Instead of using the difference in means as the statistic, we use the $t$ statistic for the coefficient on $Z$ in the linear regression.
+If treatment was assigned at random within each stratum, then it is guaranteed that $Z$ and $\varepsilon$ are statistically independent, conditional on stratum.
+Therefore, we can construct a permutation distribution by repeatedly permuting treatment assignments $Z$ within strata, independently across strata, and regressing the outcomes on the covariate and permuted treatment vectors.
 
-Freedman and Lane \cite{freedman_nonstochastic_1983} propose an alternative test based on the residuals of the linear regression.
-They take an alternative view of the problem, still writing the outcome in the form of Equation~\ref{eqn:ancova}, but instead of treating the $\varepsilon_{ij}$ as random they define the errors to be the difference between the observed outcome $Y_{ij}$ and the data's linear projection onto the plane $\alpha_j + \beta X_{ij}+ \gamma Z_{ij}$.
-The $\mathbf{\varepsilon}$ are fixed approximation errors in this framework.
+\citet{freedman_nonstochastic_1983} propose an alternative test based on the residuals of the linear regression.
+They take an alternative view of the problem, still writing the outcome in the form of Equation~\ref{eqn:ancova}, but they \textit{define} the errors $\varepsilon_{ij}$ to be the difference between the observed outcome $Y_{ij}$ and the data's linear projection onto the plane $\alpha_j + \beta X_{ij}+ \gamma Z_{ij}$.
+The $\varepsilon$ are fixed approximation errors in this framework, not independent and identically distributed random errors.
 
 If the null hypothesis is true, then $\gamma = 0$ and $\varepsilon_{ij} = Y_{ij} - \alpha_{ij} - \beta X_{ij}$ for all $i = 1, \dots, n_j, j = 1, \dots, J$.
-We may regress $\mathbf{Y}$ on $\mathbf{X}$ \textit{and not $\mathbf{Z}$} to obtain coefficient estimates $\hat{\alpha}_j$ and $\hat{\beta}$.
-Then we estimate the errors $\mathbf{\hat{\varepsilon}}$ by $\mathbf{Y} - \mathbf{\hat{Y}}$, where $\mathbf{\hat{Y}}$ is the vector of fitted values $\hat{Y}_{ij} = \hat{\alpha}_j + \hat{\beta}X_{ij}$.
-The $\mathbf{\hat{\varepsilon}}$ approximate the true errors $\mathbf{\varepsilon}$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit\footnote{
-\cite{freedman_nonstochastic_1983} advise against using this method if there are large outlier values in the covariate $\mathbf{X}$ or when $\mathbf{X}$ and $\mathbf{Z}$ are highly colinear.
-}.
-Furthermore, under the null hypothesis, the $\mathbf{\varepsilon}$ are independent $\mathbf{Z}$ within strata. 
-If we make the additional assumption that $\mathbf{\varepsilon}$ are exchangeable against $\mathbf{X}$ within strata, then 
+Therefore, we may estimate the errors $\hat{\varepsilon}$ by $Y - \hat{Y}$, where $\hat{Y}$ is the vector of fitted values from the regression of $Y$ on $X$ but not $Z$.
+The $\hat{\varepsilon}$ approximate the true errors $\varepsilon$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit.
+(\cite{freedman_nonstochastic_1983} advise against using this method if there are large outlier values in the covariate $X$ or when $X$ and $Z$ are highly colinear.)
+Furthermore, under the null hypothesis, the $\varepsilon$ are independent $Z$ within strata. 
+If we make the additional assumption that $\varepsilon$ are exchangeable against $X$ within strata, then 
 we may permute the estimated $\hat{\varepsilon}$ within strata, independently across strata.
+This additional exchangeability assumption should be checked using the residuals $\hat{\varepsilon}$;
+for instance, visual inspection of a scatterplot of the residuals against $X$ should not reveal any patterns if they are exchangeable.
 
@@
 only the estimates $\hat{\alpha}_j$ and $\hat{\beta}$.
-\cite{manly_randomization_2006} proposes randomizing the outcomes $\mathbf{Y}$, treating them as though they were randomly assigned to pairs $(X_{ij}, Z_{ij})$ under the null hypothesis.
-Permutations under the null treat all coefficients as 0, which may not reflect the true data-generating process.
+\cite{manly_randomization_2006} proposed randomizing the outcomes $Y$, treating them as though they were randomly assigned to pairs $(X_{ij}, Z_{ij})$ under the null hypothesis.
+Permutations under the null treat assume that all coefficients are 0 in the true data-generating process, which may not reflect the true relationship between variables.
 \cite{kennedy_randomization_1995} proposed a permutation method similar in spirit to \cite{freedman_nonstochastic_1983}, but which differs procedurally.
 Both methods attempt to measure the correlation between treatment and unexplained variation in outcomes, but
-instead of regressing pseudo-outcomes $\mathbf{Y}^\pi$ on covariates to obtain a permutation distribution, \cite{kennedy_randomization_1995} recommends using the $t$ statistic from regressing $\mathbf{Z}$ on the permuted residuals $\mathbf{\hat{\varepsilon}}^\pi$.
+instead of regressing pseudo-outcomes $Y^\pi$ on covariates to obtain a permutation distribution, \cite{kennedy_randomization_1995} recommends using the $t$ statistic from regressing $Z$ on the permuted residuals $\hat{\varepsilon}^\pi$.
 
-Several authors have compared these tests theoretically and through simulations \cite{anderson_empirical_1999, anderson_permutation_2001, kennedy_randomization_1996}.
-They find that the Freedman-Lane test of residuals is asymptotically equivalent to the ``oracle'' exact hypothesis test which we could conduct if we knew which permutations of $\mathbf{Y}$ given $\mathbf{X}$ were equiprobable under the null hypothesis \cite{anderson_permutation_2001}.
-This confirms empirical results, which show that the Freedman-Lane test performs better than other linear regression based tests in simulations \cite{anderson_empirical_1999}, though its advantage is small.
-Therefore, throughout the rest of the paper we focus on the two linear regression permutations we described in detail: linear regression after permuting treatment assignments and the Freedman-Lane method of permuting residuals.
+Several authors have compared these tests theoretically and empirically (\cite{anderson_empirical_1999, anderson_permutation_2001, kennedy_randomization_1996}).
+They find that the Freedman-Lane test of residuals is asymptotically equivalent to the ``oracle'' exact hypothesis test which we could conduct if we knew which permutations of $Y$ given $X$ were equiprobable under the null hypothesis (\cite{anderson_permutation_2001}).
+This confirms empirical results, which show that the Freedman-Lane test performs better than other linear regression based tests in simulations (\cite{anderson_empirical_1999}), though its advantage is small.
+Therefore, throughout the rest of the paper we focus on the two linear regression permutation tests we described in detail: linear regression with permuted treatment assignments and the Freedman-Lane method of permuting residuals.
 
-\section*{Results}
-\subsection*{Simulations}\label{simulations}
+\section{Simulations}\label{sec:simulations}
 
-We simulated data coming from a randomized experiment using several different data-generating processes.
+We simulated data from a randomized experiment using several different data-generating processes.
 We applied the tests to the data and compared their empirical power over repeated random treatment assignments and random errors.
@@
 and the Freedman-Lane permutation test.
+The simulations are included in three supplementary files.
 
@@
 There were four total simulation designs:
-for each of the two possible treatment effects, we used two distributions of $\mathbf{\varepsilon}$.
-In the first case, we used $\varepsilon \sim N(0, 1)$ to fit with the parametric assumption of normal errors.
-In the second case, $\mathbf{\varepsilon}$ were drawn from a $t$ distribution with 2 degrees of freedom so the errors were heavy-tailed.
+for each of the two possible treatment effects, we used two distributions for $\varepsilon$.
+In the first case, we used $\varepsilon \sim N(0, 1)$ to match the parametric assumption of normal errors.
+In the second case, $\varepsilon$ were drawn from a $t$ distribution with 2 degrees of freedom so the errors were heavy-tailed.
 
 Figure~\ref{fig:normal_sim_power} shows the estimated power curves in these four designs.
-The best case is when the errors are Gaussian and the treatment effect is constant across strata, while the worst case is when the errors are heavy-tailed and the treatment effect only appears in one stratum.
+The best case was when the errors were Gaussian and the treatment effect was constant across strata, while the worst case was when the errors are heavy-tailed and the treatment effect only appeared in one stratum.
 Intuitively, it makes sense that power decreased relative to the Gaussian, constant treatment effect case, as each violated assumption further obscured the treatment effect.
-A consistent pattern appears in each design: the stratified permutation test has the lowest power, while the other four tests tend to have similar power.
+A consistent pattern appeared in each design: the stratified permutation test had the lowest power, while the other four tests tended to have similar power.
 
 Table~\ref{tab:normal_power} displays the size of the test and the power at level $0.05$.
-The three tests based on linear models have slightly higher than nominal level, though the margin is within two standard errors of $0.05$.\footnote{
-The number of tests rejected under the null in 1000 trials has a binomial distribution.
-If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.}
-These numbers show that actually, for $\alpha=0.05$, the stratified permutation test does not lose a large amount of power when the effect is isolated in a single stratum.
-In this case, the permutation test may have slightly higher power than the parametric ANCOVA at the small significance levels used in practice.
+The three tests based on linear models had slightly higher than nominal level, though the margin was within two standard errors of $0.05$.
+(The number of tests rejected under the null in 1000 trials has a binomial distribution.
+If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.)
+These numbers show that actually, for $\alpha=0.05$, the stratified permutation test did not lose a large amount of power when the effect was isolated in a single stratum.
+In this case, the stratified permutation test may have slightly higher power than ANCOVA at the small significance levels typically used in practice.
 \begin{figure}[h]
 \centering
-\includegraphics[width = \textwidth]{fig/normal_simulation_power}
+\includegraphics[width = \textwidth]{fig/normal_simulation_power.pdf}
 \caption{Empirical power curves for the Gaussian simulated data}
@@
 Within each stratum, 4 individuals received one treatment and 12 received the other.
-This mimics real-world experiments, where the more expensive treatment is administered less frequently than the other.
+This mimics some real-world experiments, where the more expensive treatment is administered less frequently than the placebo or standard of care.
 It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exacerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
-Here, the data have the same variance so we would expect power, not the test level, to be an issue.\footnote{
-See the supplementary Gaussian simulation files for a treatment of heterogeneous variances.}
-We omit a plot of the power curves from these simulations, as the patterns appear similar to those in Figure~\ref{fig:normal_sim_power}.
+Here, the data had equal variances so we would expect power, not the test level, to be an issue.
+(See the supplementary Gaussian simulations for simulations using data with heterogeneous variances.)
 Table~\ref{tab:imbalanced_power} displays the value of these power curves and the size of the test at level $0.05$.
-In this case, the permutation tests (except for the simple stratified permutation) have slightly higher power than the ANCOVA, but the difference is not substantial.
-%\begin{figure}
-%\centering
-%\includegraphics[width = \textwidth]{fig/imbalanced_simulation_power}
-%\caption{Empirical power curves for the Gaussian simulated data with imbalanced group sizes}
-%\label{fig:imbalanced_sim_power}
-%\end{figure}
-\begin{center}
-\input{fig/imbalanced_simulation_power_summary.tex}
-\end{center}
+In this case, the permutation tests (except for the simple stratified permutation) had slightly higher power than ANCOVA, but the difference was not substantial.
+
+%\begin{center}
+%\input{fig/imbalanced_simulation_power_summary.tex}
+%\end{center}
+
+\begin{table}[ht]
+\centering
+\label{tab:imbalanced_power}
+\begin{tabular}{ll|ccccc}
+\multicolumn{2}{c|}{Design}                                                                 & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
+Errors                        & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
+\multirow{3}{*}{Gaussian}     & None                                                        & 0.046                      & 0.031                                                                                & 0.028                                                                                 & 0.042                                                                        & 0.045                                                                        \\
+                              & Constant                                                    & 0.840                      & 0.490                                                                                & 0.810                                                                                 & 0.870                                                                        & 0.860                                                                        \\
+                              & Single stratum                                              & 0.120                      & 0.050                                                                                & 0.090                                                                                 & 0.100                                                                        & 0.120                                                                        \\ \hline
+\multirow{3}{*}{Heavy-tailed} & None                                                        & 0.034                      & 0.054                                                                                & 0.055                                                                                 & 0.047                                                                        & 0.042                                                                        \\
+                              & Constant                                                    & 0.230                      & 0.200                                                                                & 0.270                                                                                 & 0.260                                                                        & 0.270                                                                        \\
+                              & Single stratum                                              & 0.070                      & 0.080                                                                                & 0.090                                                                                 & 0.090                                                                        & 0.080                                                                        \\ \hline
+\end{tabular}
+\caption{Empirical power at level $0.05$ for Gaussian simulated data with imbalanced treatment groups} 
+\end{table}
 
 In our third set of simulations, we made the baseline measurements $Y_{ij0}$ discrete and skewed.
-This type of data occurs in some surveys where individuals are asked to rate their symptom severity or some experience on a scale from 1 to 10, for instance.
+This type of data may occur in surveys where individuals are asked to rate their symptom severity on a discrete ordinal scale.
 We generated the baseline measures using independent draws from a Poisson distribution with mean 4 and censored values at 10.
-We conditioned on these observed baseline measures after we having generated them once.
-Then, for each 1000 trials, we randomly assigned treatment to half of the individuals in each stratum and generated errors taking on values $0.5$ and $-0.5$ with equal probability.
+We conditioned on these observed baseline measures after generating them once.
+Then, for each of 1000 trials, we randomly assigned treatment to half of the individuals in each stratum and generated errors taking on values $0.5$ and $-0.5$ with equal probability.
 We constructed $Y_{ij1}$ using Equation~\ref{eqn:dgp} with a treatment effect of $\gamma = 0.5$.
-However, we didn't observe $Y_{ij1}$ but instead integer values $\tilde{Y}_{ij1}$ defined as
+To ensure that both baseline and outcome were integer scores, we let the $Y_{ij1}$ be unobservable and defined the observed outcome as
 
@@
 
-\begin{center}
-\input{fig/skewed_simulation_power_summary.tex}
-\end{center}
-
+%\begin{center}
+%\input{fig/skewed_simulation_power_summary.tex}
+%\end{center}
+\begin{table}[]
+\centering
+\label{tab:skewed_power}
+\begin{tabular}{ll|ccccc}
+\multicolumn{2}{c|}{Design}                                                               & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
+Group sizes                 & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
+\multirow{3}{*}{Balanced}   & None                                                        & 0.036                      & 0.025                                                                                & 0.021                                                                                 & 0.035                                                                        & 0.037                                                                        \\
+                            & Constant                                                    & 0.974                      & 0.143                                                                                & 0.853                                                                                 & 0.972                                                                        & 0.977                                                                        \\
+                            & Single stratum                                              & 0.422                      & 0.050                                                                                & 0.275                                                                                 & 0.429                                                                        & 0.424                                                                        \\ \hline
+\multirow{3}{*}{Imbalanced} & None                                                        & 0.038                      & 0.024                                                                                & 0.019                                                                                 & 0.037                                                                        & 0.039                                                                        \\
+                            & Constant                                                    & 0.971                      & 0.147                                                                                & 0.860                                                                                 & 0.970                                                                        & 0.972                                                                        \\
+                            & Single stratum                                              & 0.190                      & 0.070                                                                                & 0.100                                                                                 & 0.190                                                                        & 0.190                                                                        \\ \hline
+\end{tabular}
+\caption{Empirical power at level $0.05$ for discrete, skewed simulated data} 
+\end{table}
 It makes sense that in all of these simulations, the permutation test using the change scores as the response measure is more powerful than the permutation test using the raw outcomes.
-The baseline and outcome are highly correlated, so the change score has lower variance the outcome alone.
-In additional simulations\footnote{
-See the simulation using Gaussian, balanced design for a more detailed treatment.}, 
-we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had correlation of $0.25$.
+The baseline and outcome were highly correlated, so the change score had lower variance the outcome alone. 
+In additional simulations, 
+we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had a correlation of $0.25$.
 In this case, the result was reversed: 
 the differenced test had low power while the test using the raw outcomes had a power curve closer to the linear model-based tests.
-This behavior has been studied before by \cite{frison_repeated_1992}, who recommended to use the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
+(See the supplementary Gaussian simulations for a more detailed treatment.)
+This behavior has been studied before by \cite{frison_repeated_1992}, who recommended using the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
 
-\subsection*{Clinical data results}
+\section{Clinical data results}\label{sec:results}
 We compared the parametric ANCOVA, the stratified permutation test, and the two linear model-based permutation tests using a dataset from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease (GERD).
+A detailed discussion of the data and analysis is provided in a supplementary file.
+We summarize the analysis here.
 Patients were treated at eight sites in two different countries.
@@
 Daily ``hrdq'' was the primary endpoint.
-To reduce day-to-day variation, we averaged the measures from each week to obtain two time points per patient, one pretreatment and one post-treatment.
+To reduce day-to-day variation, we averaged the measures from each week to obtain two observations per patient, one pre-treatment and one post-treatment.
 
-We treated site as the stratification variable, as this is the level at which treatment was randomized.
+We used site as the stratification variable, as this is the level at which treatment was randomized.
 We did not include country in the model, as a country-level effect should be captured in the site-level effects.
 The model used for the linear regression-based tests was defined as in Equation~\ref{eqn:ancova}.
-This model allowed for the intercept $\alpha_j$ to vary across sites and used the pretreatment, baseline measurement as the control variable $\mathbf{X}$.
-The outcome and baseline had a low correlation (for instance, $0.56$ for daily ``hrdq''), so we used raw outcomes and not the change scores as the dependent variable. 
-As our simulations and previous work \cite{frison_repeated_1992} show, when the correlation between baseline and outcome is low, it is more powerful to control for baseline outcomes using a model.
+This model allowed the intercept $\alpha_j$ to vary across sites and used the pretreatment, baseline measurement as the control variable $X$.
+The outcome and baseline had a low correlation (for instance, the correlation between pre- and post-treatment daily ``hrdq'' was $0.56$), so we used raw outcomes and not the change scores as the dependent variable.
+As our simulations and previous work (\cite{frison_repeated_1992}) show, when the correlation between baseline and outcome is low, it is more powerful to control for baseline outcomes using a model.
 
 Figure~\ref{fig:clinical_distr} shows the distribution of each clinical endpoint for the two treatment groups.
-There is a clear difference in distributions for daily heartburn (``daily\_heart''), daily ``hrdq'', and heartburn frequency (``heart\_freq'').
+There is a clear difference in distributions for daily heartburn (``daily\_heart''), daily ``hrdq'' (``daily\_hrdq''), and heartburn frequency (``heart\_freq'').
 The difference is less clear for daily regurgitation (``daily\_regurg'') and regurgitation frequency (``regurg\_freq'').
+The distribution of outcomes for each endpoint is extremely right skewed, calling the assumption of normally distributed error terms into question.
+It is possible that the
 
-Table~\ref{tab:clinical_pvalues} shows the $p$-values for the four tests.
+Table~\ref{tab:clinical_pvalues} shows the $p$-values for the four tests and the seven continuous study endpoints.
 Overall, the results confirm our expectations based on visual comparison in Figure~\ref{fig:clinical_distr}:
@@
 but not for any of the other endpoints.
-The three tests based on the linear model all give similar results here.
-The $p$-values for the ANCOVA tend to be smaller than the linear model and the residual permutation tests, while the $p$-values stratified, unadjusted permutation test have no consistent pattern.
-The parametric test has comparable or slightly higher power than the other permutation tests.
+The $p$-values for the stratified, unadjusted permutation test have no consistent pattern: sometimes they are smaller than the $p$-values from the other tests and sometimes they are larger.
+
+The three tests based on the linear model give qualitatively similar results here.
+The ANCOVA $p$-values tend to be smaller than the permutation linear model and residual permutation test $p$-values.
+The permutation tests are somewhat more conservative, accounting for the highly skewed outcome distributions.
+Our conclusions for the heartburn frequency, daily heartburn, and daily ``hrdq'' endpoints differ between ANCOVA and the permutation tests at significance level $0.05$, but not at level $0.1$.
+None of the three tests based on linear models is significant at level $0.1$ for the other four endpoints;
+there is no endpoint which would be deemed significant using a permutation test but insignificant with ANCOVA.
+This suggests that the parametric test correctly discriminates between endpoints that are significantly different and endpoints that are not different between treatment A and treatment B.
 
@@
 
-\begin{center}
-\input{fig/results_pvalue_summary.tex}
-\end{center}
+%
+%\begin{center}
+%\input{fig/results_pvalue_summary.tex}
+%\end{center}
 
-\section*{Discussion}
-% This section should discuss the implications of the findings in context of existing research and highlight limitations of the study.
+\begin{table}[]
+\centering
+\label{tab:clinical_pvalues}
+\begin{tabular}{l|cccc}
+Endpoint      & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
+heart\_freq   & 0.035                      & 0.006                                                                                & 0.080                                                                        & 0.082                                                                        \\
+regurg\_freq  & 0.136                      & 0.118                                                                                & 0.280                                                                        & 0.220                                                                        \\
+dysp\_freq    & 0.565                      & 0.948                                                                                & 0.616                                                                        & 0.592                                                                        \\
+daily\_heart & 0.032                      & 0.004                                                                                & 0.056                                                                        & 0.068                                                                        \\
+daily\_regur  & 0.142                      & 0.174                                                                                & 0.286                                                                        & 0.246                                                                        \\
+daily\_hrdq   & 0.043                      & 0.012                                                                                & 0.088                                                                        & 0.098                                                                        \\
+daily\_dysp   & 0.582                      & 0.810                                                                                & 0.756                                                                        & 0.722                                                                        \\
+\hline
+\end{tabular}
+\caption{Comparison of p-values from four tests, for each continuous endpoint.} 
+\end{table}
+
+\section{Discussion}\label{sec:discussion}
 
 This paper adds to the literature comparing parametric and nonparametric tests.
-Our results match those of \cite{vickers_parametric_2005, anderson_empirical_1999}, where parametric ANCOVA and regression-based $t$ tests performed the same or better than the comparable nonparametric tests.
-We simulated a variety of data-generating processes ranging from the ideal case when all assumptions are met 
-to the case when data are non-normal, discrete, or skewed,
-as well as data with other possible issues such as heterogeneous treatment effects and imbalanced sample sizes.
-When the ANCOVA assumptions were violated, the linear regression-based tests lost power relative to the best case data-generating process where all assumptions were satisfied.
-However, the power of the tests remained the same \textit{relative to each other}.
+Our results match those of \citet{vickers_parametric_2005} and \citet{anderson_empirical_1999}, where parametric ANCOVA and regression-based $t$ tests performed the same or better than the comparable nonparametric tests.
+We simulated a variety of data-generating processes, ranging from the ideal case when all assumptions are met, 
+to the case when data are non-normal, discrete, or skewed.
+In the supplementary simulation files, we also considered data with other possible issues such as heterogeneous treatment effects and unequal sample sizes.
+The linear regression-based tests, both the parametric ANCOVA and the permutation tests, suffered a loss of power when the ANCOVA assumptions were violated.
+However, the tests remained comparable to each other.
 It is a matter of taste which test one chooses for their experiment: while the parametric test may be robust to violations of its assumptions, it is somewhat reassuring that the permutation test can exactly match the randomization that was done while making no distributional assumptions.
+Applying a linear model based permutation test as a secondary analysis can give insight into how ANCOVA results depend on the method's assumptions
+(making it conservative, anti-conservative, or not systematically affecting the $p$-value in a particular direction),
+and give evidence that inferences based on ANCOVA are reliable.
 
@@
 For instance, he points to the case where observations have unequal variances.  
-This is a problem as we cannot observe the errors; it is a leap of faith to assert that they are homogeneous and therefore exchangeable.
-\cite{boik_fisherpitman_1987} illustrates this phenomenon using the normal theory $F$ test and its permutation counterpart, 
-and demonstrates by simulation that the latter has larger than nominal level when the error variances are heterogeneous.
+This is a problem in practice as one cannot observe errors; it is a leap of faith to assert that they are homogeneous and therefore exchangeable.
+\cite{boik_fisherpitman_1987} illustrates this phenomenon using the traditional $F$ test and its permutation counterpart, 
+and demonstrates by simulation that the latter has larger than nominal level when the error variances are unequal.
+Randomized experiments mitigate this problem, but is important to conduct permutation tests by exactly mimicking the randomization that was done.
+For instance, in stratified experiments, it is conceivable that the variance of outcomes differs across strata for reasons unrelated to the treatment.
+Treatment assignment is exchangeable within strata, but not across strata, and a permutation test that ignores this fact may be invalid.
 
-It is important to note that the permutation tests described here are approximately exact only in the context of randomized experiments.
-In this case, treatment is assigned at random and is therefore statistically independent of the covariates $\mathbf{X}$ and the errors $\mathbf{\varepsilon}$.
-Conversely, in observational studies, treatment may be associated with $\mathbf{X}$ and/or $\mathbf{\varepsilon}$, often in a way that is difficult or impossible to model.
-Independence, or the weaker condition of exchangeability, is what enables us to construct permutation distributions while holding $\mathbf{X}$ fixed.
-When exchangeability doesn't hold, we cannot disentangle the effect of $\mathbf{Z}$ from the effect of $\mathbf{X}$.
-One must provide evidence that items in observational studies are exchangeable in order to achieve an approximately exact test.
-The onus is on the researcher to make the case that the object being permuted is uncorrelated with the other variables being held fixed.
-This can be done visually using scatterplots and residual plots \cite{freedman_nonstochastic_1983} and from subject-matter knowledge.
+It is important to note that the permutation tests described here are exact only in the context of randomized experiments.
+Treatment is assigned at random and is therefore statistically independent of the covariates $X$ and the errors $\varepsilon$.
+In observational studies, treatment may be associated with $X$, $\varepsilon$, or both, often in a way that is difficult or impossible to model.
+The statistical independence guaranteed by randomized experiments enables us to construct permutation distributions while holding $X$ fixed.
+When exchangeability doesn't hold, we cannot disentangle the effect of $Z$ from the effect of $X$.
+One must provide evidence that variables in observational studies are exchangeable in order to achieve an approximately exact test.
+The onus is on the researcher to make the case that the variable being permuted is uncorrelated with the other variables being held fixed.
+This can be done visually using scatterplots and residual plots \cite{freedman_nonstochastic_1983} and from knowledge of how the data arose and were collected.
 
-One should be mindful of which covariates to control for.
-Ideally, one knows and observes all covariates related to the outcome.
-In that case, the coefficient for treatment in a fully saturated linear model (including all covariates and their interactions) is an asymptotically consistent estimate of the average treatment effect \cite{lin_agnostic_2013}.
-It is rare that one knows \textit{all} relevant covariates in practice; failure to control for all relevant variables can introduce bias in certain permutation tests.
+One must be mindful of which covariates to include as controls.
+Ideally, all covariates related to the outcome are known and observed.
+In that case, the coefficient for treatment in a fully saturated linear model (including all covariates and their interactions) is an asymptotically consistent estimate of the average treatment effect (\cite{lin_agnostic_2013}).
+It is rare that one knows \textit{all} relevant covariates in practice; failure to control for all relevant variables can introduce bias in some permutation tests.
  \cite{gail_tests_1988} propose a randomization test based on residuals from an exponential family model and find that omitting relevant covariates leads to tests with higher than nominal level.
@@
 Weak correlations between baseline and outcome occur often in practice, as was the case with our GERD dataset.
-If the correlation is weak, then the resulting test using the difference may be less powerful than ignoring the baseline altogether.
+When the correlation is weak, the test of change scores may be less powerful than ignoring the baseline altogether.
 Instead, we suggest incorporating the baseline in a regression model.
 This is more general than using differences;
-treating the change scores as the dependent variable is a special case of the linear regression, constraining the coefficient on the baseline measure to be $1$.
+treating the change scores as the dependent variable is a special case of the linear regression that constrains the coefficient of the baseline measure to be $1$.
 The regression approach is more flexible and demonstrably more powerful.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Backmatter begins here                   %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bigskip
+\begin{center}
+{\large\bf SUPPLEMENTARY MATERIAL}
+\end{center}
+All files are also available at \url{https://github.com/kellieotto/ancova-permutations}.
 
-%\begin{backmatter}
-%
-%\section*{Competing interests}
-%  The authors declare that they have no competing interests.
-%
-%\section*{Author's contributions}
-%    Text for this section \ldots
-%
-%\section*{Acknowledgements}
-%  Text for this section \ldots
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  Bmc_mathpys.bst  will be used to                       %%
-%%  create a .BBL file for submission.                     %%
-%%  After submission of the .TEX file,                     %%
-%%  you will be prompted to submit your .BBL file.         %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{description}
 
-% if your bibliography is in bibtex format, use those commands:
-\bibliographystyle{vancouver} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{references}      % Bibliography file (usually '*.bib' )
-%% for author-year bibliography (bmc-mathphys or spbasic)
-%% a) write to bib file (bmc-mathphys only)
-%% @settings{label, options="nameyear"}
-%% b) uncomment next line
-%%\nocite{label}
-%
-%% or include bibliography directly:
-%% \begin{thebibliography}
-%% \bibitem{b1}
-%% \end{thebibliography}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                               %%
-%%% Figures                       %%
-%%%                               %%
-%%% NB: this is for captions and  %%
-%%% Titles. All graphics must be  %%
-%%% submitted separately and NOT  %%
-%%% included in the Tex document  %%
-%%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%%%
-%%% Do not use \listoffigures as most will included as separate files
-%
-%\section*{Figures}
-%  \begin{figure}[h!]
-%  \caption{\csentence{Sample figure title.}
-%      A short description of the figure content
-%      should go here.}
-%      \end{figure}
-%
-%\begin{figure}[h!]
-%  \caption{\csentence{Sample figure title.}
-%      Figure legend text.}
-%      \end{figure}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                               %%
-%%% Tables                        %%
-%%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%%% Use of \listoftables is discouraged.
-%%%
-%\section*{Tables}
-%\begin{table}[h!]
-%\caption{Sample table title. This is where the description of the table should go.}
-%      \begin{tabular}{cccc}
-%        \hline
-%           & B1  &B2   & B3\\ \hline
-%        A1 & 0.1 & 0.2 & 0.3\\
-%        A2 & ... & ..  & .\\
-%        A3 & ..  & .   & .\\ \hline
-%      \end{tabular}
-%\end{table}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%                               %%
-%%% Additional Files              %%
-%%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%\section*{Additional Files}
-%  \subsection*{Additional file 1 --- Sample additional file title}
-%    Additional file descriptions text (including details of how to
-%    view the file, if it is in a non-standard format or the file extension).  This might
-%    refer to a multi-page table or a figure.
-%
-%  \subsection*{Additional file 2 --- Sample additional file title}
-%    Additional file descriptions text.
-%
-%
-%\end{backmatter}
+\item[Gaussian data simulations:] Simulations using Gaussian data in Section~\ref{sec:simulations}, including code and results. (PDF)
+\item[Imbalanced design simulations:] Simulations using data with different stratum sizes in Section~\ref{sec:simulations}, including code and results. (PDF)
+\item[Skewed data simulations:] Simulations using discrete, skewed data in Section~\ref{sec:simulations}, including code and results. (PDF)
+\item[Clinical trial data:] Dataset used in Section~\ref{sec:results} to compare methods. (csv)
+\item[README:] Data descriptor file. (txt)
+\item[Results:] Detailed explanation, code, and results of comparing methods using the clinical trial dataset in Section~\ref{sec:results}. (PDF)
 
+\end{description}
+
+\bibliographystyle{chicago}
+
+\bibliography{references}
 \end{document}

### 2017-02-16 simulations/simulation_skewed.Rmd
Clean up suppl simulation files

@@
 
-We use two designs:
-
-* Design 1: $\gamma_1 = \gamma_2 = \gamma_3 = 1$. This is the standard assumption of a constant, additive treatment effect.
-* Design 2: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$.
-This is a constant, additive treatment effect at stratum 1, but no treatment effect at strata 2 and 3.
-This is a simplistic case of heterogeneous treatment effects.
-
-With these two designs, we vary the distribution of $\varepsilon$.
-In the first case, we use $\varepsilon \sim N(0, 1)$ to mimic the usual ANCOVA assumptions.
-In the second case, $\varepsilon \sim t(2)$ so the errors are heavy-tailed.
-Thus, there are four total simulation designs.
-
-We compare five tests:
-
-* ANCOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for stratum.
-* Stratified permutation: we permute treatment assignment within stratum, then take the difference in means between treated and control outcomes $Y_1$
-* Differenced permutation: we do the same permutation procedure as the stratified permutation test, except we use the difference between outcome and baseline, $Y_1 - Y_0$
-* Linear model (LM) permutation: we use the same stratified permutation procedure as above, except use the $t$-statistic for the coefficient on treatment in the linear regression of $Y_1$ on $Y_0$, $Z$, and stratum dummies
-* Freedman-Lane test: see the other Rmd document for a full description of this procedure
-
-Throughout our simulations, we first fix $Y_0$ and stratum ID.
-Treatment $Z$ and the errors $\varepsilon$ are randomly drawn according to their respective distributions.
-Then, $Y_1$ is constructed using the linear data-generating process above.
-We regenerate $Z$, $\varepsilon$, and $Y_1$ 100 times for each design, then compute the empirical power of the five tests.
+We use the same simulation guidelines as detailed in the Gaussian simulation file.
+We omit comments.
 
@@
   ylab("Power") +
-  ggtitle(title)
-}
-
-plot_pvalue_hist <- function(pvalue_mat, title){
-  melt(pvalue_mat) %>%
-  mutate("Method" = Var2) %>%
-  ggplot(aes(x = value, fill = Method)) +
-  geom_histogram() +
-  facet_wrap(~Method) +
-  ggtitle(title)
-}
-
-plot_pvalue_scatter <- function(pvalue_mat, title){
-  pvalue_mat %>%
-  as.data.frame() %>%
-  select(ANCOVA, strat = starts_with("Stratified")) %>%
-  ggplot(aes(x = ANCOVA, y = strat)) + 
-  geom_point() +
-  xlim(0, 1) + ylim(0, 1) +
-  ylab("Stratified Permutation") +
-  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
-  ggtitle(title)
+  ggtitle(title) +
+  theme_bw() +
+  theme(
+    axis.text.x = element_text(size = 12),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 16),
+    title = element_text(size = 16),
+    legend.title = element_text(size = 12),
+    legend.text = element_text(size = 14),
+    strip.text.x = element_text(size = 12)
+  ) 
 }
@@
 ## Test level: simulation under the null
-
-Before testing for different kinds of effects, we begin checking that the tests have the correct level.
-We follow the procedure described above, using an effect size of $\gamma = 0$ at all strata and using standard normal errors.
-To have the correct level means that the test rate of rejection at level $\alpha$ is $\alpha 100\%$.
-In other words, the p-values are uniformly distributed and the power curve should coincide with the line with slope $1$ through the origin.
-Figure \ref{fig:design_null_plots} demonstrates that this is the case.
-If anything, the differenced stratified permutation test has fewer than $\alpha 100\%$ false positives when using level $\alpha$.
 
@@
 ```{r design_null_plots, fig.height = 4, fig.align="center"}
-plot_power_curves(design0_power, "No Effect, Balanced Design")
-
-plot_pvalue_hist(design0_pvalues, "No Effect, Balanced Design")
+plot_power_curves(design0_power, "No Effect \n Balanced Design")
 ```
@@
 ## Design 1: Constant additive effect
-
-There is no discernable difference in power between the ANCOVA, the differenced stratified permutation test, the LM permutations, or the Freedman-Lane test.
-However, the simple stratified permutation test of $Y_1$ has substantially less power than the other four.
-Without controlling for the baseline values, the variance in $Y_1$ masks the treatment effect.
 
@@
 ```{r design1_plot, fig.height = 4, fig.align="center"}
-p1 <- plot_power_curves(design1_power, "Constant Additive Effect of 0.5")
-p1
-plot_pvalue_hist(design1_pvalues, "Constant Additive Effect of 0.5")
+plot_power_curves(design1_power, "Constant Additive Effect of 0.5")
+
 ```
@@
 ## Design 2: Single stratum effect
-
-A similar pattern emerges: the simple stratified permutation test has low power, while the other four power curves roughly coincide.
-The Freedman-Lane test may have the highest power for small p-values, but this could also just be noise.
-All five power curves are closer to the line passing through the origin: since the effect is only present at one stratum, it is more difficult to detect.
 
@@
 ```{r design2_plot, fig.height = 4, fig.align="center"}
-p2 <- plot_power_curves(design2_power, "Effect of 0.5 in a Single Stratum")
-p2
-plot_pvalue_hist(design2_pvalues, "Effect of 0.5 in a Single Stratum")
+plot_power_curves(design2_power, "Effect of 0.5 in a Single Stratum")
 ```
@@
 ```{r design3_plot, fig.height = 4, fig.align="center"}
-p3 <- plot_power_curves(design3_power, "Constant Additive Effect of 0.5 \n Unpredictive baseline")
-p3
-plot_pvalue_hist(design3_pvalues, "Constant Additive Effect of 0.5 \n Unpredictive baseline")
+plot_power_curves(design3_power, "Constant Additive Effect of 0.5 \n Unpredictive baseline")
 ```
@@
 ```{r design_null_imbalanced_plots, fig.height = 4, fig.align="center"}
-plot_power_curves(design00_power, "No Effect, Imbalanced design")
-
-plot_pvalue_hist(design00_pvalues, "No Effect, Imbalanced design")
+plot_power_curves(design00_power, "No Effect \n Imbalanced design")
 ```
@@
 ```{r design4_plot, fig.height = 4, fig.align="center"}
-p4 <- plot_power_curves(design4_power, "Constant Additive Effect of 0.5 \n Imbalanced Design")
-p4
-plot_pvalue_hist(design4_pvalues, "Constant Additive Effect of 0.5 \n Imbalanced Design")
+plot_power_curves(design4_power, "Constant Additive Effect of 0.5 \n Imbalanced Design")
 ```
@@
 # Design 5: Single stratum effect, imbalanced design
-
-There is very low power to detect the effect here, since in the simulated data, the stratum with only 8 patients was the only stratum with a nonzero effect of treatment.
-The treatment effect is on the same order of magnitude as the error variance.
 
@@
 ```{r design5_plot, fig.height = 4, fig.align="center"}
-p5 <- plot_power_curves(design5_power, "Effect of 0.5 at a Single stratum, Imbalanced Design")
-p5 
-plot_pvalue_hist(design5_pvalues, "Effect of 0.5 at a Single stratum, Imbalanced Design")
+plot_power_curves(design5_power, "Effect of 0.5 at a Single stratum \n Imbalanced Design")
 ```
@@
 ```{r final_summary, fig.align = "center"}
-#plots4 <- lapply(list(p1, p2, p4, p5), function(pl) pl + theme(legend.position = "none") + ggtitle(""))
-#do.call(grid.arrange, plots4)
-
 powers <- list(design1_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design1_power)), "Design" = rep("Balanced", nrow(design1_power))), 

### 2017-08-28 ms/ms.tex
first pass at edits

@@
 \addtolength{\topmargin}{-.8in}%
+
+\pdfminorversion=4
 
@@
 Medical experiments use hypothesis testing to assess the evidence that a treatment affects one or more clinically relevant outcomes.
-The simplest version of this experiment, which involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome, has been studied for nearly a century (see \cite{fisher_design_1935} and \cite{neyman_application_1923} for early references).
+The simplest version of this experiment, which involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome, has been studied for nearly a century (see \cite{fisher_design_1935} and \cite[1990 translation]{neyman_application_1923} for early references).
 One can conduct hypothesis tests and construct confidence intervals for the estimated treatment effect
@@
 patients may not take the treatment they are assigned or may drop out of the study;
-and an individual's assignment to treatment may affect others' outcomes.
+and peer effects, the effect of being in the same treatment group as others, may affect outcomes on top of the treatment effect alone.
 Techniques have been developed to deal with each of these issues, invoking additional assumptions in order to carry out valid inference.
@@
 Strata are groups of individuals with similar levels of a covariate.
-These groups are defined \textit{before any outcome data is collected}.
+These groups are defined during the design stage (i.e. before outcome data are collected).
 Random assignment of treatments is conducted within each stratum, independently across strata.
@@
 Linear regression is another way to control for baseline covariates.
-It is done \textit{after outcomes have been observed}.
+It is done during the data analysis stage.
 While stratification guarantees balance on the stratifying variable, other covariates may still be imbalanced.
@@
 When this holds, a regression coefficient scaled by its estimated variance follows Student's $t$ distribution.
-Thus, if we believe the linear model is correctly specified, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the ANCOVA model, and can be evaluated analytically and quickly.
+Thus, if we believe the linear model is correctly specified, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the ANCOVA model, and can be evaluated analytically and efficiently.
 However, the modeling assumptions are clearly violated in many cases:
@@
 when the treatment has a differential effect across subgroups of individuals and no interaction terms are included, 
-and when measurement error sizes differ across strata.
+and when the variance of outcomes differs across strata.
 Furthermore, the hypothesis test implicitly assumes that the data were sampled at random from some underlying population, when in fact,
@@
 Such a test is exact, meaning that it controls the type I error rate at the pre-specified level even in finite samples, whereas parametric hypothesis tests based on asymptotic approximations do not always guarantee good finite sample properties.
-Permutation tests achieve this by operating conditional on the observed sample; the resulting inference applies to the sample at hand, but does not necessarily generalize to other groups of individuals.
-This may be acceptable when the individuals who make up the sample are not representative of any well-defined population.
+Permutation tests condition on the observed sample and do not require any assumptions about the way individuals were sampled from a larger population.
+This is useful when the sampling frame is difficult to specify, such as when individuals are a convenience sample.
+While external validity may be an issue, we may still be interested in whether the treatment had an effect on the observed individuals.
+\todo{clear?}
 
-In the past, parametric methods were necessary because asymptotic approximations were the only computationally feasible way to estimate distributions and construct confidence intervals
-Now, computational power is no longer a barrier to finding exact (or exact to pre-specified precision) randomization distributions, though they remain the most common way of constructing confidence intervals.
+In the past, statisticians relied on parametric methods because asymptotic approximations were a computationally feasible way to estimate distributions and construct confidence intervals.
+Now, computational power is no longer a barrier to finding exact (or exact to pre-specified precision) randomization distributions and confidence intervals.
 In most cases, a randomization test is the ``gold standard'':
@@
 Many people recommend using permutation tests in place of the usual parametric tests for typical analyses, such as ANOVA and generalized linear models (\cite{still_approximate_1981, winkler_permutation_2014}).
-They argue that there are a myriad of ways that the data may violate the necessary assumptions for the test, and so it is more robust to use a permutation test.
+They argue that there are a myriad of ways that the data may violate the necessary assumptions for the test, and so permutation tests are more robust.
 
@@
 it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.
-A typical problem is to estimate the mean difference $Y_{ij}(1) - Y_{ij}(0)$ in some group, such as the study sample or in a target population.
+We may never learn this difference for any particular individual, so
+a typical problem is to estimate the mean difference in some group, such as the study sample or in a target population.
 The chosen method of analysis determines which function of differences is considered.
-Various functions of potential outcomes may be of clinical interest; which one prefers depends on the study's goal and the researcher's personal preference.
-Rather than estimation, we focus on hypothesis testing for whether these differences are nonzero.
+Various functions of potential outcomes may be of clinical interest; which one prefers depends on the goal of the study.
+
+We study hypothesis testing for whether these differences are nonzero using parametric ANCOVA and its permutation counterparts, assuming this potential outcomes framework throughout.
+\todo{make sure the parametric section actually reflects the potential outcomes framework. See Jas's email and the Winston Lim paper}
+Other valid methods for comparing two groups include using a two-sample $t$-test to test the difference in two means from normal distributions,
+the Wilcoxon rank sum test to test for differences in the medians of two independent groups, 
+and the Kolmogorov-Smirnov test and receiver operating characteristic curve analyses to test whether two groups have different distribution functions (\cite{lehmann_nonparametrics_1975,  vexler_statistical_2016}).
+We focus on testing using the linear model as this is standard in clinical trials, requires fewer distributional assumptions on the data when using the potential outcomes framework, and incorporates control variables to increase power.
 
@@
 
+\todo{are the estimator, standard error, variance clear here?}
 The coefficients are estimated using least squares.
+The estimated coefficient $\hat{\gamma}$ is the estimated average treatment effect. 
 (Note, only Assumptions (1) and (2) are needed to guarantee the existence of a solution.
 The solution will be unique as long as there is no linear relationship between $X$, $Z$, and stratum membership.)
-This procedure yields an estimate of $\hat{\sigma}_{\hat{\gamma}}^2$, the  standard error of $\hat{\gamma}$
-(namely, the corresponding diagonal element of the inverse covariance matrix of regression covariates).
+This procedure also yields an estimate $\hat{\sigma}_{\hat{\gamma}}^2$ of the variance of $\hat{\gamma}$
+(namely, the corresponding diagonal element of the inverse covariance matrix of regression covariates scaled by the residual sum of squares over the degrees of freedom).
 Under the null hypothesis, the test statistic 
@@
 follows the Student $t$ distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
-The $p$-value for this hypothesis test is the probability that a value drawn from the $t$ distribution is larger in magnitude than $T$.
+The $p$-value for this hypothesis test is the probability, assuming the null hypothesis of zero coefficient is true, that a value drawn from the $t$ distribution is larger in magnitude than $T$.
+This test is equivalent to the $F$-test and the likelihood ratio test assuming the Gaussian model,
+and \todo{ cite this? is therefore locally most powerful ????}.
 
@@
 
-The most commonly used statistic is the difference in mean outcomes of people who received treatment 1 and the mean outcomes of people who received treatment 0.  
-This statistic is unbiased and readily interpretable in randomized experiments (``on average, taking treatment changes the outcome by x amount''), and has nice theoretical properties owing to it being the difference of two sums.
+The most commonly used statistic is the difference in mean outcomes of subjects who received treatment 1 and the mean outcomes of subjects who received treatment 0.  
+This statistic is unbiased over all possible random assignments of treatment which preserve the number of treated individuals,
+is interpretable (``on average, taking treatment changes the outcome by x amount''), 
+and has convenient theoretical properties owing to it being the difference of two means.
 However, the difference in means may not be optimal if we want to be sensitive to heterogeneous effects.  
@@
  but on the other hand, must be fine enough to capture variation in treatment effects.
-Our simulations in Section~\ref{simulations} show that power decreases when effects are concentrated in small strata.
+Our simulations in Section~\ref{sec:simulations} show that power decreases when effects are concentrated in small strata.
 
@@
 (\cite{freedman_nonstochastic_1983} advise against using this method if there are large outlier values in the covariate $X$ or when $X$ and $Z$ are highly colinear.)
+\todo{not clear how to formally test the conditions for the permutation test. visual inspection is not satisfactory}
 Furthermore, under the null hypothesis, the $\varepsilon$ are independent $Z$ within strata. 
@@
 \cite{manly_randomization_2006} proposed randomizing the outcomes $Y$, treating them as though they were randomly assigned to pairs $(X_{ij}, Z_{ij})$ under the null hypothesis.
-Permutations under the null treat assume that all coefficients are 0 in the true data-generating process, which may not reflect the true relationship between variables.
+\todo{is this clear?}
+Permutations under the null hypothesis assume that all coefficients are 0 in the true data-generating process, which may not reflect the true relationship between variables.
 \cite{kennedy_randomization_1995} proposed a permutation method similar in spirit to \cite{freedman_nonstochastic_1983}, but which differs procedurally.
@@
 We let $\beta_0 = 1$ and the treatment effect be $\gamma = 1$.
-We used a balanced design with $n_j = 16$ individuals per stratum and balanced treatment assignment, i.e. 8 people received each treatment within each stratum.
+We used a balanced design with $n_j = 16$ individuals per stratum and balanced treatment assignment, i.e. 8 subjects received each treatment within each stratum.
 There were four total simulation designs:
@@
 \centering
-\label{tab:normal_power}
 \begin{tabular}{ll|ccccc}
@@
 \end{tabular}
-\caption{Empirical power at level $0.05$ for Gaussian simulated data.}
+\caption{Empirical power at level $0.05$ for Gaussian simulated data with balanced treatment groups.}
+\label{tab:normal_power}
 \end{table}
@@
 \centering
-\label{tab:imbalanced_power}
 \begin{tabular}{ll|ccccc}
@@
 \end{tabular}
-\caption{Empirical power at level $0.05$ for Gaussian simulated data with imbalanced treatment groups} 
+\caption{Empirical power at level $0.05$ for Gaussian simulated data with imbalanced treatment groups.} 
+\label{tab:imbalanced_power}
 \end{table}
@@
 \centering
-\label{tab:skewed_power}
 \begin{tabular}{ll|ccccc}
@@
 \caption{Empirical power at level $0.05$ for discrete, skewed simulated data} 
+\label{tab:skewed_power}
 \end{table}
 It makes sense that in all of these simulations, the permutation test using the change scores as the response measure is more powerful than the permutation test using the raw outcomes.
-The baseline and outcome were highly correlated, so the change score had lower variance the outcome alone. 
+The baseline and outcome were highly correlated, so the change score had lower variance than the outcome alone. 
 In additional simulations, 
@@
 This model allowed the intercept $\alpha_j$ to vary across sites and used the pretreatment, baseline measurement as the control variable $X$.
-The outcome and baseline had a low correlation (for instance, the correlation between pre- and post-treatment daily ``hrdq'' was $0.56$), so we used raw outcomes and not the change scores as the dependent variable.
+The outcome and baseline had a low correlation (for instance, the correlation between pre- and post-treatment daily ``hrdq'' was $0.56$ \todo{were other outcomes even lower?
+the cutoff for correlation was 0.5}), so we used raw outcomes and not the change scores as the dependent variable.
 As our simulations and previous work (\cite{frison_repeated_1992}) show, when the correlation between baseline and outcome is low, it is more powerful to control for baseline outcomes using a model.
@@
 There is a clear difference in distributions for daily heartburn (``daily\_heart''), daily ``hrdq'' (``daily\_hrdq''), and heartburn frequency (``heart\_freq'').
-The difference is less clear for daily regurgitation (``daily\_regurg'') and regurgitation frequency (``regurg\_freq'').
+The difference is less clear for daily regurgitation (``daily\_regurg''), daily dyspepsia (``daily\_dysp''), regurgitation frequency (``regurg\_freq''), and dyspepsia frequency (``dysp\_freq'').
 The distribution of outcomes for each endpoint is extremely right skewed, calling the assumption of normally distributed error terms into question.
-It is possible that the
 
@@
 The ANCOVA $p$-values tend to be smaller than the permutation linear model and residual permutation test $p$-values.
-The permutation tests are somewhat more conservative, accounting for the highly skewed outcome distributions.
+The permutation tests are somewhat more conservative, as they give higher $p$-values in Table~\ref{tab:clinical_pvalues}, accounting for the highly skewed outcome distributions.
+\todo{is this a correct statement?
+conservative means reject less often}
 Our conclusions for the heartburn frequency, daily heartburn, and daily ``hrdq'' endpoints differ between ANCOVA and the permutation tests at significance level $0.05$, but not at level $0.1$.
@@
 
-Permutation tests do not come entirely free of assumptions, though.  
+Permutation tests do not come entirely free of assumptions, though: 
+subjects must be exchangeable if we generate the permutation distribution assuming that all possible allocations of treatment which preserve the number of treated units are equally likely.
 \cite{romano_behavior_1990} warns against using permutation tests naively if items are not truly exchangeable. 
@@
 and demonstrates by simulation that the latter has larger than nominal level when the error variances are unequal.
-Randomized experiments mitigate this problem, but is important to conduct permutation tests by exactly mimicking the randomization that was done.
-For instance, in stratified experiments, it is conceivable that the variance of outcomes differs across strata for reasons unrelated to the treatment.
-Treatment assignment is exchangeable within strata, but not across strata, and a permutation test that ignores this fact may be invalid.
+Randomized experiments mitigate this problem: by definition, treatment assignment is statistically independent of all other variables (possibly conditioning on strata).
 
@@
 The onus is on the researcher to make the case that the variable being permuted is uncorrelated with the other variables being held fixed.
-This can be done visually using scatterplots and residual plots \cite{freedman_nonstochastic_1983} and from knowledge of how the data arose and were collected.
+This can be done visually using scatterplots and residual plots (see e.g. \cite{freedman_nonstochastic_1983}) and from knowledge of how the data arose and were collected.
 

### 2017-09-05 ms/ms.tex
Edits to methods section

@@
 Medical experiments use hypothesis testing to assess the evidence that a treatment affects one or more clinically relevant outcomes.
-The simplest version of this experiment, which involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome, has been studied for nearly a century (see \cite{fisher_design_1935} and \cite[1990 translation]{neyman_application_1923} for early references).
+The simplest version of this experiment involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome, and has been studied for nearly a century (see \cite{fisher_design_1935} and \cite[1990 translation]{neyman_application_1923} for early references).
 One can conduct hypothesis tests and construct confidence intervals for the estimated treatment effect
@@
 patients may not take the treatment they are assigned or may drop out of the study;
-and peer effects, the effect of being in the same treatment group as others, may affect outcomes on top of the treatment effect alone.
-Techniques have been developed to deal with each of these issues, invoking additional assumptions in order to carry out valid inference.
-In this paper, we focus on the issue of heterogeneity of pretreatment characteristics and assume that the other issues are not present.
+and peer effects, the influence of other members of a treatment group, may affect outcomes beyond the treatment itself.
+Techniques have been developed to deal with each of these issues, invoking additional assumptions to make valid inference possible.
+In this paper, we focus on the issue of heterogeneous pretreatment characteristics and assume that the other issues are not present.
 
@@
 
-Stratification is one method to control for covariates that are known a priori to be associated with the outcome.
+Stratification, sometimes called blocking, is one method to control for covariates that are known a priori to be associated with the outcome.
 Strata are groups of individuals with similar levels of a covariate.
@@
 especially when the object of study is a rare disease or a rare outcome.
-Controlling for stratum can have the same beneficial effects as controlling for other pretreatment covariates, 
+Stratification can have the same beneficial effects as controlling for other pretreatment covariates, 
 as outcomes may vary across strata due to factors unrelated to individuals' treatment and covariates.
 
-Linear regression is another way to control for baseline covariates.
+Linear regression is another way to control for imbalanced baseline covariates.
 It is done during the data analysis stage.
-While stratification guarantees balance on the stratifying variable, other covariates may still be imbalanced.
-Stronger modeling assumptions may be necessary to balance them after the experiment.
-Linear regression is a method which projects the outcomes onto the plane that best summarizes each variable's relationship to the outcome.
+In its simplest form, linear regression projects the outcomes onto the plane that best summarizes each variable's relationship to the outcome.
 The coefficient of any particular covariate answers the question, if we were to hold fixed all other variables and increase this variable by one unit, how much would we expect the outcome to change?
@@
 if the true relationship is not linear, then linear regression gives the best linear approximation to the conditional expectation of the outcome.
-When controlling for strata, such as location, it is standard to use analysis of covariance (ANCOVA).
+It is standard to use analysis of covariance (ANCOVA) to account for stratification in a linear model.
 ANCOVA is a particular case of linear regression that allows the mean outcome to vary from stratum to stratum.
@@
 Permutation tests condition on the observed sample and do not require any assumptions about the way individuals were sampled from a larger population.
-This is useful when the sampling frame is difficult to specify, such as when individuals are a convenience sample.
-While external validity may be an issue, we may still be interested in whether the treatment had an effect on the observed individuals.
-\todo{clear?}
+This is useful when the sampling frame is difficult to specify, such as when the study uses a convenience sample.
 
@@
 ``[a] corresponding parametric test is valid only to the extent that it results in the same statistical decision [as the randomization test]'' (\cite{bradley_distribution_1968}).
-If the permutation test agrees with the parametric test, one may have a greater degree of confidence in the estimates and confidence intervals constructed using the parametric method.
+There is no hard and fast rule describing the rate at which parametric tests approach the exact permutation solution, as they are both highly dependent on the particular data observed.
+However, if the permutation test agrees with the parametric test, one may have a greater degree of confidence in the estimates and confidence intervals constructed using the parametric method.
 
-There is no hard and fast rule describing the rate at which parametric tests approach the exact permutation solution, as they are both highly dependent on the particular data observed.
-On the one hand, some argue that violations of parametric test assumptions necessitate the use of permutation methods.
-\cite{ludbrook_why_1998} point out that medical trials rarely follow the population sampling model implicit in parametric methods.
-Many people recommend using permutation tests in place of the usual parametric tests for typical analyses, such as ANOVA and generalized linear models (\cite{still_approximate_1981, winkler_permutation_2014}).
-They argue that there are a myriad of ways that the data may violate the necessary assumptions for the test, and so permutation tests are more robust.
-
-However, parametric and nonparametric tests seem to perform similarly when compared side-by-side in simulations, even when data violate the assumptions of the parametric method.
-Medical trials often use Likert scales to score pain or symptom severity, resulting in discrete data that does not match the normality assumptions of parametric tests.
-However, \cite{winter_five-point_2010} found that the two sample $t$ test and Mann-Whitney test had comparable Type I and II error rates for five-point Likert scale data, suggesting that the violation of normality does not entirely invalidate the parametric test.
-\cite{vickers_parametric_2005} compared the parametric ANCOVA to the Mann-Whitney rank test in the context of randomized experiments, finding that except in extreme situations, ANCOVA was more powerful than the nonparametric test.
-Most similar to our question of study, \cite{anderson_empirical_1999} found little difference between several permutation tests for coefficients in a linear model alongside the parametric $t$ test.
-In these situations, the permutation test strengthens conclusions by giving evidence that the parametric test is robust to departures from its assumptions.
-
-In this paper, we review several hypothesis tests for a treatment effect which adjust for pretreatment covariates to increase power.  
+We review several hypothesis tests for randomized experiments which adjust for pretreatment covariates to increase power to detect a nonzero treatment effect.  
 We focus on ANCOVA and its permutation counterparts, comparing their performance in different scenarios and illustrating their application with a clinical dataset.
-Section~\ref{sec:methods} describes the various tests mathematically and states their assumptions explicitly.
+Section~\ref{sec:methods} describes the tests and compares their assumptions.
 Section~\ref{sec:simulations} presents simulations that suggest that even when assumptions are not satisfied, the parametric and permutation tests have comparable power to detect a treatment effect.
@@
 \subsection{Notation}
-Suppose we randomly assign two treatments labelled $0$ and $1$ to a group of individuals.
-We are interested in comparing the relative effectiveness of treatment 1 to treatment 0.
-Let $Z$ indicate treatment assignment,
-$Y$ denote the observed outcome of interest,
-and $X$ be a pretreatment covariate that is observed and associated with the outcome.
+Suppose we have a finite population of $N$ individuals.
+Individuals are grouped into strata indexed by $j = 1, \dots, J$, with $n_j$ individuals in stratum $j$ and $\sum_{j=1}^J n_j = N$.
+Of the $n_j$ subjects in stratum $j$, $n_{tj}$ are assigned treatment 1, while the remaining $n_j - n_{tj}$ are assigned treatment 0.
+Let $Z_{ij}$ indicate the treatment assigned to individual $i$ in stratum $j$.
+
+All individuals have two potential outcomes, $Y_{ij}(1)$ and $Y_{ij}(0)$, representing their responses to treatments $1$ and $0$, respectively.
+We can never observe both; random assignment of treatment reveals $Y_{ij} = Z_{ij}Y_{ij}(1) + (1-Z_{ij})Y_{ij}(0)$.
+Throughout, we assume that there is no interference between individuals 
+(in other words, $Y_{ij}$ is a function of $(Z_{ij}, Y_{ij}(1), Y_{ij}(0))$ and not any other $Z_{i', j'}$ for $(i', j') \neq (i, j)$)
+and that there is no censoring or non-compliance
+(we actually observe $Y_{ij} = Y_{ij}(Z_{ij})$ for all $(i, j)$).
+
+Furthermore, we observe a covariate $X_{ij}$ that may be associated with the outcome.
 For expository clarity, we suppose that $X$ is univariate, but all results are easily extended to the case when $X$ is multivariate.
@@
 
-Strata are indexed by subscript $j = 1, \dots, J$.
-Let $n_j$ be the number of individuals in stratum $j$.
-Individuals within stratum $j$ are indexed by subscript $i = 1, \dots, n_j$.
-We observe $\{Y_{ij}, Z_{ij}, X_{ij}\}$ for $i = 1, \dots, n_j$, $j = 1, \dots, J$.
-All individuals have two potential outcomes, $Y_{ij}(1)$ and $Y_{ij}(0)$, representing their responses to treatments $1$ and $0$, respectively.
-We can never observe both; random assignment of treatment reveals $Y_{ij} = Z_{ij}Y_{ij}(1) + (1-Z_{ij})Y_{ij}(0)$.
-Throughout, we assume that there is no interference between individuals 
-(in other words, $Y_{ij}$ is a function of $(Z_{ij}, Y_{ij}(1), Y_{ij}(0))$ and not any other $Z_{i', j'}$ for $(i', j') \neq (i, j)$.)
-and that there is no censoring or non-compliance
-(we actually observe $Y_{ij} = Y_{ij}(Z_{ij})$ for all $(i, j)$).
-
-We are interested in the differences in potential outcomes $Y_{ij}(1) - Y_{ij}(0)$.
-This quantity represents the effect of treatment for individual $i$ in group $j$: 
-it is the difference between what we would have observed under treatment 1 and what we would have observed under treatment $0$.
-We may never learn this difference for any particular individual, so
-a typical problem is to estimate the mean difference in some group, such as the study sample or in a target population.
+We are interested in the effect of treatment, measured as differences in potential outcomes $Y_{ij}(1) - Y_{ij}(0)$.
+We may never learn this difference for any particular individual.
+However, a tractable problem is to estimate the mean difference in some group, such as the study sample or in a target population.
 The chosen method of analysis determines which function of differences is considered.
@@
 We study hypothesis testing for whether these differences are nonzero using parametric ANCOVA and its permutation counterparts, assuming this potential outcomes framework throughout.
-\todo{make sure the parametric section actually reflects the potential outcomes framework. See Jas's email and the Winston Lim paper}
 Other valid methods for comparing two groups include using a two-sample $t$-test to test the difference in two means from normal distributions,
@@
 and $\varepsilon_{ij}$ is an error term.
-The parameter of interest is $\gamma$. 
-If we believe the linear model is the true data-generating process, it asserts that for each individual, $Y(1) = Y(0) + \gamma$.
+The parameter of interest is $\gamma$, and the parametric ANCOVA tests the null hypothesis $H_0: \gamma = 0$ against
+the two-sided alternative hypothesis $H_1: \gamma \neq 0$.
+If the linear model is the true data-generating process, then $Y_{ij}(1) = Y_{ij}(0) + \gamma$ for all $(i, j)$.
 However, we needn't take this perspective for $\gamma$ to be a useful quantity; it represents the average treatment effect, holding the other variables fixed.
-We would like to test the null hypothesis $H_0: \gamma = 0$ against
-the two-sided alternative hypothesis $H_1: \gamma \neq 0$.
 
-To carry out the standard parametric hypothesis test for a linear model, we make the following assumptions (\cite{freedman_statistical_2005}):
-
+To carry out the standard parametric hypothesis test for a linear model, the following assumptions are needed (\cite{freedman_statistical_2005}):
+\todo{CHECK}
 \begin{enumerate}
@@
 \todo{are the estimator, standard error, variance clear here?}
-The coefficients are estimated using least squares.
+The coefficients are estimated using least squares (or equivalently, by maximizing the likelihood).
 The estimated coefficient $\hat{\gamma}$ is the estimated average treatment effect. 
@@
 The solution will be unique as long as there is no linear relationship between $X$, $Z$, and stratum membership.)
-This procedure also yields an estimate $\hat{\sigma}_{\hat{\gamma}}^2$ of the variance of $\hat{\gamma}$
-(namely, the corresponding diagonal element of the inverse covariance matrix of regression covariates scaled by the residual sum of squares over the degrees of freedom).
+This procedure also yields an estimate $\hat{\sigma}_{\hat{\gamma}}^2$ of the variance of $\hat{\gamma}$.
 Under the null hypothesis, the test statistic 
@@
 
-The ANCOVA model includes a fixed effect for each stratum, but 
-the model does not account for variation in the treatment effect across strata.  
-The estimated linear model may not accurately capture details of the true data-generating process:
-if effects are not constant across strata, then the coefficient $\gamma$ may be attenuated towards 0.
-A test of the null hypothesis that $\gamma = 0$ will be valid, but will not reflect the true magnitude of treatment effects among individuals.
-One can account for differential treatment effects across strata by including interaction terms for treatment and stratum membership, but this complicates the hypothesis testing problem.
+However, in the context of a randomized experiment, these assumptions will not be satisfied.
+The estimated treatment effect $\hat{\gamma}$ and nominal standard errors $\hat{\sigma}_{\hat{\gamma}}$ can be severely biased (\todo{cite freedman 2008, lin 2013}).
+\todo{Lin 2013} shows that running a regression with a full set of treatment and covariate interaction terms cannot hurt asymptotic precision, and using the Huber-White sandwich standard errors can yield asymptotically valid confidence intervals.
+In small samples, the bias may still be substantial.
+\todo{cite Sekhon post-stratification} show that post-stratification, estimating treatment effects within groups of similar individuals defined after data are collected, can ameliorate this bias.
+This is equivalent to estimating a fully saturated linear model with interaction terms for treatment and stratum membership.
+The linear ANCOVA model does not account for variation in the treatment effect across strata in this way;
+if the difference in potential outcomes is heterogeneous across strata, then the coefficient $\gamma$ may be attenuated towards 0.
 
 \subsection{Stratified permutation test}\label{subsec:strat_perm_test}
+In randomization inference, the only random quantities are those involving $Z$, the vector of treatment assignments.
+The exact inference is derived from the randomization distribution of statistics involving $Z$.
+In particular, the potential outcomes and covariates $(Y_{ij}(1), Y_{ij}(0), X_{ij})$ are fixed in the finite population under study, while the observed response $Y_{ij}$ changes with $Z_{ij}$.
+\todo{cite rosenberg 2002}
+
 Suppose we wish to test the null hypothesis that individual by individual, treatment has no effect.
@@
 
-Then, which treatment an individual received amounts to an arbitrary label.
+Under the null hypothesis, which treatment an individual received amounts to an arbitrary label.
 Once we observe one response under a particular treatment, we can impute the other potential outcome; namely, it would have been the same.
-This null hypothesis is stronger than the null hypothesis for the parametric ANCOVA, which only asserts that the treatment has no effect \textit{on average}.
+Thus, the observed $Y_{ij}$ form a sufficient statistic for the data. \todo{cite P\&S}
+This null hypothesis is stronger than the null hypothesis for the parametric ANCOVA, which only asserts that differences in potential outcomes are zero \textit{on average}.
 
-For the permutation test, we condition on the number of individuals who received each treatment within each stratum.
-Any assignment of treatments that preserves the number of treated units within each stratum is valid and was just as likely to occur as the randomization that was actually observed.
-We can construct the permutation distribution of any statistic under the null hypothesis using this principle of equal probabilities and by imputing the unobserved potential outcomes using the sharp null hypothesis.
+In a randomized experiment with complete randomization, any randomization that assigns $n_j$ individuals to treatment 1 in stratum $j$ for each stratum has equal probability.
+We can construct the permutation distribution of any statistic under the null hypothesis by imputing the unobserved potential outcomes using the sharp null hypothesis and by re-randomizing treatment assignments using this principle of equal probabilities.
+The experimental design specifies the permutation scheme; no additional assumptions are necessary.
 
@@
 This statistic is unbiased over all possible random assignments of treatment which preserve the number of treated individuals,
-is interpretable (``on average, taking treatment changes the outcome by x amount''), 
+is interpretable, 
 and has convenient theoretical properties owing to it being the difference of two means.
-However, the difference in means may not be optimal if we want to be sensitive to heterogeneous effects.  
+However, the difference in means may not be optimal if we want to detect heterogeneous effects.  
 For an extreme example, imagine that the sample contains an equal number of males and females, and each treatment is assigned to half of males and half of females.  
@@
 
-To avoid this, one may want to stratify the sample according to important confounding variables, then compute a statistic for each stratum separately. 
-There is a great degree of freedom in deciding how to construct such a statistic.
-Two choices must be made: how to stratify and how to combine the statistics across strata.
-There is a tradeoff in how finely we choose to stratify.
-On the one hand, each stratum must be sufficiently large to be informative,
- but on the other hand, must be fine enough to capture variation in treatment effects.
-Our simulations in Section~\ref{sec:simulations} show that power decreases when effects are concentrated in small strata.
-
-One way to construct a test statistic is to directly combine the stratum statistics into a single value, for instance taking the sum of their absolute values.
+To avoid this, one may want to account for the stratification. 
+One way to construct a stratified test is to use the same stratified permutation scheme but a different statistic which combines the stratum-specific statistics into a single value, for instance taking the sum of their absolute values.
 Taking the absolute value before summing ensures that effects with different signs do not cancel each other out.
-Permutations for this test are conducted identically to those in the previous section, but this combined statistic is used in place of the difference in means.
 Another way to construct a test is to use the nonparametric combination (NPC) framework proposed by \citet{pesarin_permutation_2010}.
-In this framework, one would break down the ``global'' null hypothesis of no effect whatsoever into the intersection of ``partial'' null hypotheses of no treatment effect within each stratum.
+In this framework, the ``global'' null hypothesis of no effect whatsoever is decomposed into the intersection of ``partial'' null hypotheses of no treatment effect within each stratum.
 For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies.
-In randomized experiments, strata are randomized independently, making this method equivalent to combining stratum statistics directly.
+This method is equivalent to combining stratum statistics directly when data are from a randomized experiment, because treatment is independent across strata.
 
@@
 The tests in Section~\ref{subsec:strat_perm_test} only use information on the treatment and the outcome.
-However, experimenters record additional covariates, some of which may be predictive of the outcome.
-One would like to construct a more powerful test by incorporating these covariates to reduce the variance of the statistic.
-The permutation tests in this section continue to model the outcome as in Equation~\ref{eqn:ancova}, but require fewer assumptions about the data.
+However, experimenters record additional covariates that may be predictive of the outcome.
+A test that incorporates these covariates may reduce the variance of the statistic and have more power.
+The permutation tests in this section use the linear model in Equation~\ref{eqn:ancova}, but make different assumptions about the data.
 In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes,
-making several variables ``permutable.''
+making several variables exchangeable.
 We show two permutation tests developed in this framework.
 
-First, we may do a simple variation on the stratified permutation test described above.
-Instead of using the difference in means as the statistic, we use the $t$ statistic for the coefficient on $Z$ in the linear regression.
-If treatment was assigned at random within each stratum, then it is guaranteed that $Z$ and $\varepsilon$ are statistically independent, conditional on stratum.
-Therefore, we can construct a permutation distribution by repeatedly permuting treatment assignments $Z$ within strata, independently across strata, and regressing the outcomes on the covariate and permuted treatment vectors.
+The first linear model permutation test is a variation on the stratified permutation test described in Section~\ref{subsec:strat_perm_test}.
+The only difference is the test statistic:
+this test uses the $t$ statistic for the regression coefficient of $Z$ in the linear model in Equation~\ref{eqn:ancova}.
+Rather than using the $t$ distribution to calculate the significance of the observed $t$ statistic, we construct the permutation distribution of the $t$ statistic.
+Again, the experimental design specifies the randomization distribution and no additional assumptions are needed:
+$Z$ is independent of $(Y(1), Y(0), X)$ conditional on strata.
 
 \citet{freedman_nonstochastic_1983} propose an alternative test based on the residuals of the linear regression.
-They take an alternative view of the problem, still writing the outcome in the form of Equation~\ref{eqn:ancova}, but they \textit{define} the errors $\varepsilon_{ij}$ to be the difference between the observed outcome $Y_{ij}$ and the data's linear projection onto the plane $\alpha_j + \beta X_{ij}+ \gamma Z_{ij}$.
+They take an alternative view of the problem:
+instead of assuming the data are generated according to the linear model Equation~\ref{eqn:ancova}, they \textit{define} the errors $\varepsilon_{ij}$ to be the difference between the observed outcome $Y_{ij}$ and the data's linear projection onto the plane $\alpha_j + \beta X_{ij}+ \gamma Z_{ij}$.
 The $\varepsilon$ are fixed approximation errors in this framework, not independent and identically distributed random errors.
@@
 If the null hypothesis is true, then $\gamma = 0$ and $\varepsilon_{ij} = Y_{ij} - \alpha_{ij} - \beta X_{ij}$ for all $i = 1, \dots, n_j, j = 1, \dots, J$.
-Therefore, we may estimate the errors $\hat{\varepsilon}$ by $Y - \hat{Y}$, where $\hat{Y}$ is the vector of fitted values from the regression of $Y$ on $X$ but not $Z$.
-The $\hat{\varepsilon}$ approximate the true errors $\varepsilon$ from the true data-generating process, assuming that the null hypothesis is true and the linear model has a reasonable in-sample fit.
-(\cite{freedman_nonstochastic_1983} advise against using this method if there are large outlier values in the covariate $X$ or when $X$ and $Z$ are highly colinear.)
-\todo{not clear how to formally test the conditions for the permutation test. visual inspection is not satisfactory}
-Furthermore, under the null hypothesis, the $\varepsilon$ are independent $Z$ within strata. 
-If we make the additional assumption that $\varepsilon$ are exchangeable against $X$ within strata, then 
-we may permute the estimated $\hat{\varepsilon}$ within strata, independently across strata.
-This additional exchangeability assumption should be checked using the residuals $\hat{\varepsilon}$;
-for instance, visual inspection of a scatterplot of the residuals against $X$ should not reveal any patterns if they are exchangeable.
+Therefore, we may estimate the errors by $\hat{\varepsilon} =Y - \hat{Y}$, where $\hat{Y}$ is the vector of fitted values from the regression of $Y$ on $X$ but not $Z$.
+The $\hat{\varepsilon}$ approximate the true errors $\varepsilon$ from the true data-generating process, assuming that the null hypothesis is true.
+Furthermore, under the null hypothesis, the $\varepsilon$ are independent of $Z$ within strata. 
+Thus, we can carry out an approximate permutation test by permuting the estimated $\hat{\varepsilon}$ within strata, independently across strata.
+
+The test has two important assumptions:
+the relationship between $Y$ and $X$ is linear and errors are IID within strata.
+Together, these conditions ensure that $\varepsilon$ are exchangeable against $X$ within strata.
+In practice, we must estimate $\alpha$; using $\hat{\alpha}$ to estimate $\hat{Y}$ yields an approximate permutation test.
+The $\hat{\varepsilon}$ may not even be approximately exchangeable within strata if the regression does not describe the relation between $Y$ and $X$ well.
+For this reason, \cite{freedman_nonstochastic_1983} advise against using the method if there are large outlier values in the covariate $X$ or when $X$ and $Z$ are highly colinear.
+Randomization does not guarantee this additional exchangeability assumption, so it should be checked.
+One way, suggested by \cite{freedman_nonstochastic_1983}, is to visually inspect a scatterplot of the residuals against $X$.
+One may also run a hypothesis test for association, such as a stratified permutation test using the correlation as a statistic.
+\todo{Do this test in the empirical section}
 
@@
 \begin{enumerate}
-\item Estimate $\hat{\varepsilon}$ by $Y_{ij} - \hat{\alpha}_j - \hat{\beta} X_{ij}$ for all $i$ and $j$, where $\hat{\alpha}_j$ and $\hat{\beta}$ are obtained by regressing $Y$ on $X$ and stratum assignment, \textit{but not $Z$}.
-\item Obtain permuted errors $\hat{\varepsilon}^\pi$ by permuting the $\hat{\varepsilon}$ within strata.
-\item Compute permuted responses $Y_{ij}^\pi = \hat{\alpha}_j + \hat{\beta} X_{ij}+ \hat{\varepsilon}_{ij}^\pi$.
-\item Regress $Y^\pi$ on $X$, $Z$, and stratum. The test statistic is the $t$ statistic for the coefficient of $Z$.
+\item Regress $Y$ on $X$ and stratum indicators, \textit{but not $Z$} to obtain $\hat{Y}$. Estimate the errors by $\hat{\varepsilon} = Y - \hat{Y}$.
+\item Permute the $\hat{\varepsilon}$ within strata to obtain permuted errors $\hat{\varepsilon}^\pi$.
+\item Construct permuted responses $Y^\pi = \hat{Y}+ \hat{\varepsilon}^\pi$.
+\item Regress $Y^\pi$ on $X$, $Z$, and stratum indicators. The test statistic is the $t$ statistic for the coefficient of $Z$.
 \end{enumerate}
@@
 Others have developed variants on these approximate regression-based permutation tests. 
-There is some disagreement on what constitutes an appropriate permutation scheme;
-everything we have described so far is approximate insofar as we do not know the ``true'' data-generating process described by the $\alpha_j$ and $\beta$,
-only the estimates $\hat{\alpha}_j$ and $\hat{\beta}$.
-\cite{manly_randomization_2006} proposed randomizing the outcomes $Y$, treating them as though they were randomly assigned to pairs $(X_{ij}, Z_{ij})$ under the null hypothesis.
-\todo{is this clear?}
-Permutations under the null hypothesis assume that all coefficients are 0 in the true data-generating process, which may not reflect the true relationship between variables.
+There is some disagreement on what constitutes an appropriate permutation scheme.
+\cite{manly_randomization_2006} proposed randomizing the outcomes $Y$, treating them as though they were randomly assigned to pairs $(X, Z)$ under the null hypothesis.
+Permuting data this way corresponds to the null hypothesis that all coefficients are 0 in the linear model, which may not reflect the true relationship between variables.
 \cite{kennedy_randomization_1995} proposed a permutation method similar in spirit to \cite{freedman_nonstochastic_1983}, but which differs procedurally.
@@
 They find that the Freedman-Lane test of residuals is asymptotically equivalent to the ``oracle'' exact hypothesis test which we could conduct if we knew which permutations of $Y$ given $X$ were equiprobable under the null hypothesis (\cite{anderson_permutation_2001}).
-This confirms empirical results, which show that the Freedman-Lane test performs better than other linear regression based tests in simulations (\cite{anderson_empirical_1999}), though its advantage is small.
+This agrees with empirical results, which show that the Freedman-Lane test performs better than other linear model permutation tests in simulations, though its advantage is small (\cite{anderson_empirical_1999}).
 Therefore, throughout the rest of the paper we focus on the two linear regression permutation tests we described in detail: linear regression with permuted treatment assignments and the Freedman-Lane method of permuting residuals.
@@
 
-Outcomes were generated according to the linear equation 
-\begin{equation}\label{eqn:dgp}
-Y_{ij1} =\alpha_j + \beta_0Y_{ij0} + \gamma_j Z_{ij} + \varepsilon_{ij}
-\end{equation}
+\subsection{Continuous outcomes}
 
-\noindent for individuals $i = 1, \dots, n_j$, $j = 1, 2, 3$.
-$\alpha_j$ is the mean effect of being in stratum $j$, 
-$\beta_0$ is the coefficient for the baseline measurement $Y_{ij0}$, 
-$Z_{ij}$ is the treatment level, 
-$\gamma_j$ is the effect of treatment in stratum $j$, 
-and $\varepsilon_{ij}$ is an error term.
-We used three strata with $\alpha_1 = 1, \alpha_2 = 1.5,$ and $\alpha_3 = 2$.
-
-We used two designs:
-\begin{itemize}
-\item Constant additive treatment effect: $\gamma_1 = \gamma_2 = \gamma_3 = \gamma$. This is the implicit assumption when using a linear model.
-\item Treatment effect in a single stratum: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$.
-This is a constant, additive treatment effect in stratum 1, but no treatment effect in strata 2 and 3.
-This is a simplistic example of a heterogeneous treatment effect.
-The standard ANCOVA model does not account for this scenario.
-\end{itemize}
-
-\noindent We drew the baseline measurements once, treating them as fixed, and conditioned on observing $\{ Y_{ij0} : {i = 1,\dots,n_j, j = 1,\dots, 3}\}$.
-We randomly assigned treatment to half of the individuals within each stratum, generated random errors, and constructed new outcomes $Y_{ij1}$ using Equation~\ref{eqn:dgp}.
-We conducted all five tests on this new data, obtaining a two-sided $p$-value for each.
-We repeated these steps using the same $\{ Y_{ij0} \}$ 1000 times.
-
-In our first set of simulations, we assumed that the baseline measurements $Y_{ij0}$ were standard normally distributed and the pairs $(Y_{ij0}, \varepsilon_{ij})$ were independent across $i$ and $j$.
-We let $\beta_0 = 1$ and the treatment effect be $\gamma = 1$.
-We used a balanced design with $n_j = 16$ individuals per stratum and balanced treatment assignment, i.e. 8 subjects received each treatment within each stratum.
-There were four total simulation designs:
-for each of the two possible treatment effects, we used two distributions for $\varepsilon$.
-In the first case, we used $\varepsilon \sim N(0, 1)$ to match the parametric assumption of normal errors.
-In the second case, $\varepsilon$ were drawn from a $t$ distribution with 2 degrees of freedom so the errors were heavy-tailed.
-
-Figure~\ref{fig:normal_sim_power} shows the estimated power curves in these four designs.
-The best case was when the errors were Gaussian and the treatment effect was constant across strata, while the worst case was when the errors are heavy-tailed and the treatment effect only appeared in one stratum.
-Intuitively, it makes sense that power decreased relative to the Gaussian, constant treatment effect case, as each violated assumption further obscured the treatment effect.
-A consistent pattern appeared in each design: the stratified permutation test had the lowest power, while the other four tests tended to have similar power.
-
-Table~\ref{tab:normal_power} displays the size of the test and the power at level $0.05$.
-The three tests based on linear models had slightly higher than nominal level, though the margin was within two standard errors of $0.05$.
-(The number of tests rejected under the null in 1000 trials has a binomial distribution.
-If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.)
-These numbers show that actually, for $\alpha=0.05$, the stratified permutation test did not lose a large amount of power when the effect was isolated in a single stratum.
-In this case, the stratified permutation test may have slightly higher power than ANCOVA at the small significance levels typically used in practice.
-\begin{figure}[h]
-\centering
-\includegraphics[width = \textwidth]{fig/normal_simulation_power.pdf}
-\caption{Empirical power curves for the Gaussian simulated data}
-\label{fig:normal_sim_power}
-\end{figure}
-%\begin{center}
-%\input{fig/normal_simulation_power_summary.tex}
-%\end{center}
-
-\begin{table}[]
-\centering
-\begin{tabular}{ll|ccccc}
-\multicolumn{2}{c|}{Design}                                                                  & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
-Errors                        & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
-\multirow{3}{*}{Gaussian}     & None                                                        & 0.058                      & 0.042                                                                                & 0.038                                                                                 & 0.062                                                                        & 0.057                                                                        \\
-                              & Constant                                                    & 0.886                      & 0.545                                                                                 & 0.858                                                                                 & 0.891                                                                        & 0.892                                                                        \\
-                              & Single stratum                                              & 0.159                      & 0.089                                                                                & 0.160                                                                                 & 0.162                                                                        & 0.170                                                                        \\ \hline
-\multirow{3}{*}{Heavy-tailed} & None                                                        & 0.032                      & 0.040                                                                                & 0.045                                                                                 & 0.042                                                                        & 0.040                                                                        \\
-                              & Constant                                                    & 0.355                      & 0.240                                                                                & 0.378                                                                                 & 0.400                                                                        & 0.393                                                                        \\
-                              & Single stratum                                              & 0.066                      & 0.080                                                                                & 0.082                                                                                 & 0.076                                                                        & 0.080                                                                        \\ \hline
-\end{tabular}
-\caption{Empirical power at level $0.05$ for Gaussian simulated data with balanced treatment groups.}
-\label{tab:normal_power}
-\end{table}
-
-Our second set of simulations used the same design as the previous ones, but modified the sizes of the treatment groups.
-Within each stratum, 4 individuals received one treatment and 12 received the other.
-This mimics some real-world experiments, where the more expensive treatment is administered less frequently than the placebo or standard of care.
-It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exacerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
-Here, the data had equal variances so we would expect power, not the test level, to be an issue.
-(See the supplementary Gaussian simulations for simulations using data with heterogeneous variances.)
-Table~\ref{tab:imbalanced_power} displays the value of these power curves and the size of the test at level $0.05$.
-In this case, the permutation tests (except for the simple stratified permutation) had slightly higher power than ANCOVA, but the difference was not substantial.
-
-%\begin{center}
-%\input{fig/imbalanced_simulation_power_summary.tex}
-%\end{center}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ll|ccccc}
-\multicolumn{2}{c|}{Design}                                                                 & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
-Errors                        & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
-\multirow{3}{*}{Gaussian}     & None                                                        & 0.046                      & 0.031                                                                                & 0.028                                                                                 & 0.042                                                                        & 0.045                                                                        \\
-                              & Constant                                                    & 0.840                      & 0.490                                                                                & 0.810                                                                                 & 0.870                                                                        & 0.860                                                                        \\
-                              & Single stratum                                              & 0.120                      & 0.050                                                                                & 0.090                                                                                 & 0.100                                                                        & 0.120                                                                        \\ \hline
-\multirow{3}{*}{Heavy-tailed} & None                                                        & 0.034                      & 0.054                                                                                & 0.055                                                                                 & 0.047                                                                        & 0.042                                                                        \\
-                              & Constant                                                    & 0.230                      & 0.200                                                                                & 0.270                                                                                 & 0.260                                                                        & 0.270                                                                        \\
-                              & Single stratum                                              & 0.070                      & 0.080                                                                                & 0.090                                                                                 & 0.090                                                                        & 0.080                                                                        \\ \hline
-\end{tabular}
-\caption{Empirical power at level $0.05$ for Gaussian simulated data with imbalanced treatment groups.} 
-\label{tab:imbalanced_power}
-\end{table}
-
-In our third set of simulations, we made the baseline measurements $Y_{ij0}$ discrete and skewed.
-This type of data may occur in surveys where individuals are asked to rate their symptom severity on a discrete ordinal scale.
-We generated the baseline measures using independent draws from a Poisson distribution with mean 4 and censored values at 10.
-We conditioned on these observed baseline measures after generating them once.
-Then, for each of 1000 trials, we randomly assigned treatment to half of the individuals in each stratum and generated errors taking on values $0.5$ and $-0.5$ with equal probability.
-We constructed $Y_{ij1}$ using Equation~\ref{eqn:dgp} with a treatment effect of $\gamma = 0.5$.
-To ensure that both baseline and outcome were integer scores, we let the $Y_{ij1}$ be unobservable and defined the observed outcome as
-
+In the first set of simulations, we generated continuous potential outcomes.
+We used two scenarios.
+In the first, the treatment effect was constant across strata.
+We drew a latent random variate $v_{ij}$ from the uniform distribution on $[-4, 4]$.
+In the second scenario, the treatment effect varied across strata, and we drew the latent random variable according to
 \begin{displaymath}
-   \tilde{Y}_{ij1} = \left\{
-     \begin{array}{ll}
-       1 & \text{if } {Y}_{ij1} < 1\\
-       10 & \text{if } {Y}_{ij1} > 10 \\
-       \lfloor {Y}_{ij1} \rfloor & \text{otherwise}
+   v_{ij} \sim \left\{
+     \begin{array}{lr}
+       \text{Unif}[-4, -1] & : j=1\\
+       \text{Unif}[-1, 1] & : j=2\\
+       \text{Unif}[1, 4] & : j=3\\
      \end{array}
@@
 
-The observed outcomes $\tilde{Y}_{ij1}$, the baseline measures $Y_{ij0}$, and the errors were discrete, violating normality assumptions.
+We generated independent and identically distributed errors $\varepsilon_{ij}$ and $\delta_{ij}$ and varied the error distribution. 
+The errors were either standard normal (to mimic the usual ANCOVA assumptions),
+$t$ distributed with two degrees of freedom,
+standard lognormal,
+or exponentially distributed, with scale parameter $1$ and shifted to have mean zero.
+The observed $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$ were independent across $i$ and $j$.
 
-We examined the effect of stratum size in this set of simulations.
-For each of the two treatment effect designs, we considered both equal and unequal stratum sizes.
-The simulations with equal stratum sizes were generated as before, where each stratum contained $n_j = 16$ individuals.
-The simulations with unequal stratum sizes were set up so the smallest stratum had only $n_1=8$ individuals, the next had $n_2= 16$, and the largest had $n_3=24$.
-The design with a nonzero treatment effect in a single stratum and unequal stratum sizes was particularly unfavorable:
-the nonzero effect only occurred in the smallest stratum.
+The baseline value for individual $i,j$ was 
+$$ X_{ij} = \frac{\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}$$.
 
-Figure~\ref{fig:skewed_sim_power} shows the estimated power curves in these four designs.
-When the treatment effect was constant, sample sizes within each stratum didn't matter; 
-the two power curve plots in the first row of Figure~\ref{fig:skewed_sim_power} look the same up to chance variation.
-When the treatment effect only occurred in one stratum, there was a substantial loss in power.
-It was nearly impossible to detect a difference in treatments when the affected stratum was too small.
-In practice, one does not usually know a priori which individuals will be affected by treatment (otherwise, those not affected would be excluded from the study altogether).
-This result suggests that when stratifying, one must be cautious not to stratify too finely, as it is impossible to measure an effect with an insufficient sample size.
+Then we generated potential outcomes as
 
-All of the tests appeared to be conservative and to have lower than nominal level.
-As in the first set of simulations, the stratified permutation test using the raw outcomes had the lowest power of the five tests.
-The stratified permutation test using the change scores also had lower power than the three linear model tests.
-Presumably, this was because the differences $\tilde{Y}_{ij1} - Y_{ij0}$ were discrete and limited to a small number of values, so the test statistic could not vary greatly across permutations.
-Table~\ref{tab:skewed_power} confirms this: power for the three linear model methods was high when the treatment was constant across strata, and power for all of the methods suffered when the effect was isolated in a single stratum.
+$$Y_{ij}(Z_{ij}) = \frac{(2Z_{ij} - 1)\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \delta_{ij}.$$
 
-\begin{figure}
-\centering
-\includegraphics[width = \textwidth]{fig/skewed_simulation_power.pdf}
-\caption{Empirical power curves for the skewed, discrete simulated data}
-\label{fig:skewed_sim_power}
-\end{figure}
+The treatment effect for individual $(i, j)$ is $\gamma e^{v_{ij}}$ and we would like to estimate the average treatment effect in the sample.
 
-%\begin{center}
-%\input{fig/skewed_simulation_power_summary.tex}
-%\end{center}
-\begin{table}[]
-\centering
-\begin{tabular}{ll|ccccc}
-\multicolumn{2}{c|}{Design}                                                               & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
-Group sizes                 & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
-\multirow{3}{*}{Balanced}   & None                                                        & 0.036                      & 0.025                                                                                & 0.021                                                                                 & 0.035                                                                        & 0.037                                                                        \\
-                            & Constant                                                    & 0.974                      & 0.143                                                                                & 0.853                                                                                 & 0.972                                                                        & 0.977                                                                        \\
-                            & Single stratum                                              & 0.422                      & 0.050                                                                                & 0.275                                                                                 & 0.429                                                                        & 0.424                                                                        \\ \hline
-\multirow{3}{*}{Imbalanced} & None                                                        & 0.038                      & 0.024                                                                                & 0.019                                                                                 & 0.037                                                                        & 0.039                                                                        \\
-                            & Constant                                                    & 0.971                      & 0.147                                                                                & 0.860                                                                                 & 0.970                                                                        & 0.972                                                                        \\
-                            & Single stratum                                              & 0.190                      & 0.070                                                                                & 0.100                                                                                 & 0.190                                                                        & 0.190                                                                        \\ \hline
-\end{tabular}
-\caption{Empirical power at level $0.05$ for discrete, skewed simulated data} 
-\label{tab:skewed_power}
-\end{table}
-It makes sense that in all of these simulations, the permutation test using the change scores as the response measure is more powerful than the permutation test using the raw outcomes.
-The baseline and outcome were highly correlated, so the change score had lower variance than the outcome alone. 
-In additional simulations, 
-we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had a correlation of $0.25$.
-In this case, the result was reversed: 
-the differenced test had low power while the test using the raw outcomes had a power curve closer to the linear model-based tests.
-(See the supplementary Gaussian simulations for a more detailed treatment.)
-This behavior has been studied before by \cite{frison_repeated_1992}, who recommended using the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
+Assume that there are $n_j=16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
+After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerate $Z$ and the potential outcomes $Y(Z)$ 10,000 times.
+We repeat this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.
+
+\todo{insert figures and table}
+
+\todo{what's the correlation between $X$ and $Y$?}
+\subsection{Discrete outcomes}
+
+\subsection{Imbalanced treatment groups}
+%Outcomes were generated according to the linear equation 
+%\begin{equation}\label{eqn:dgp}
+%Y_{ij1} =\alpha_j + \beta_0Y_{ij0} + \gamma_j Z_{ij} + \varepsilon_{ij}
+%\end{equation}
+%
+%\noindent for individuals $i = 1, \dots, n_j$, $j = 1, 2, 3$.
+%$\alpha_j$ is the mean effect of being in stratum $j$, 
+%$\beta_0$ is the coefficient for the baseline measurement $Y_{ij0}$, 
+%$Z_{ij}$ is the treatment level, 
+%$\gamma_j$ is the effect of treatment in stratum $j$, 
+%and $\varepsilon_{ij}$ is an error term.
+%We used three strata with $\alpha_1 = 1, \alpha_2 = 1.5,$ and $\alpha_3 = 2$.
+%
+%We used two designs:
+%\begin{itemize}
+%\item Constant additive treatment effect: $\gamma_1 = \gamma_2 = \gamma_3 = \gamma$. This is the implicit assumption when using a linear model.
+%\item Treatment effect in a single stratum: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$.
+This is a constant, additive treatment effect in stratum 1, but no treatment effect in strata 2 and 3.
+This is a simplistic example of a heterogeneous treatment effect.
+The standard ANCOVA model does not account for this scenario.
+%\end{itemize}
+%
+%\noindent We drew the baseline measurements once, treating them as fixed, and conditioned on observing $\{ Y_{ij0} : {i = 1,\dots,n_j, j = 1,\dots, 3}\}$.
+%We randomly assigned treatment to half of the individuals within each stratum, generated random errors, and constructed new outcomes $Y_{ij1}$ using Equation~\ref{eqn:dgp}.
+%We conducted all five tests on this new data, obtaining a two-sided $p$-value for each.
+%We repeated these steps using the same $\{ Y_{ij0} \}$ 1000 times.
+%
+%In our first set of simulations, we assumed that the baseline measurements $Y_{ij0}$ were standard normally distributed and the pairs $(Y_{ij0}, \varepsilon_{ij})$ were independent across $i$ and $j$.
+%We let $\beta_0 = 1$ and the treatment effect be $\gamma = 1$.
+%We used a balanced design with $n_j = 16$ individuals per stratum and balanced treatment assignment, i.e. 8 subjects received each treatment within each stratum.
+%There were four total simulation designs:
+%for each of the two possible treatment effects, we used two distributions for $\varepsilon$.
+%In the first case, we used $\varepsilon \sim N(0, 1)$ to match the parametric assumption of normal errors.
+%In the second case, $\varepsilon$ were drawn from a $t$ distribution with 2 degrees of freedom so the errors were heavy-tailed.
+%
+%Figure~\ref{fig:normal_sim_power} shows the estimated power curves in these four designs.
+%The best case was when the errors were Gaussian and the treatment effect was constant across strata, while the worst case was when the errors are heavy-tailed and the treatment effect only appeared in one stratum.
+%Intuitively, it makes sense that power decreased relative to the Gaussian, constant treatment effect case, as each violated assumption further obscured the treatment effect.
+%A consistent pattern appeared in each design: the stratified permutation test had the lowest power, while the other four tests tended to have similar power.
+%
+%Table~\ref{tab:normal_power} displays the size of the test and the power at level $0.05$.
+%The three tests based on linear models had slightly higher than nominal level, though the margin was within two standard errors of $0.05$.
+%(The number of tests rejected under the null in 1000 trials has a binomial distribution.
+%If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.)
+%These numbers show that actually, for $\alpha=0.05$, the stratified permutation test did not lose a large amount of power when the effect was isolated in a single stratum.
+%In this case, the stratified permutation test may have slightly higher power than ANCOVA at the small significance levels typically used in practice.
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width = \textwidth]{fig/normal_simulation_power.pdf}
+%\caption{Empirical power curves for the Gaussian simulated data}
+%\label{fig:normal_sim_power}
+%\end{figure}
+%%\begin{center}
+%%\input{fig/normal_simulation_power_summary.tex}
+%%\end{center}
+%
+%
+%
+%\begin{table}[]
+%\centering
+%\begin{tabular}{ll|ccccc}
+%\multicolumn{2}{c|}{Design}                                                                  & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
+%Errors                        & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
+%\multirow{3}{*}{Gaussian}     & None                                                        & 0.058                      & 0.042                                                                                & 0.038                                                                                 & 0.062                                                                        & 0.057                                                                        \\
+%                              & Constant                                                    & 0.886                      & 0.545                                                                                 & 0.858                                                                                 & 0.891                                                                        & 0.892                                                                        \\
+%                              & Single stratum                                              & 0.159                      & 0.089                                                                                & 0.160                                                                                 & 0.162                                                                        & 0.170                                                                        \\ \hline
+%\multirow{3}{*}{Heavy-tailed} & None                                                        & 0.032                      & 0.040                                                                                & 0.045                                                                                 & 0.042                                                                        & 0.040                                                                        \\
+%                              & Constant                                                    & 0.355                      & 0.240                                                                                & 0.378                                                                                 & 0.400                                                                        & 0.393                                                                        \\
+%                              & Single stratum                                              & 0.066                      & 0.080                                                                                & 0.082                                                                                 & 0.076                                                                        & 0.080                                                                        \\ \hline
+%\end{tabular}
+%\caption{Empirical power at level $0.05$ for Gaussian simulated data with balanced treatment groups.}
+%\label{tab:normal_power}
+%\end{table}
+%
+%Our second set of simulations used the same design as the previous ones, but modified the sizes of the treatment groups.
+%Within each stratum, 4 individuals received one treatment and 12 received the other.
+%This mimics some real-world experiments, where the more expensive treatment is administered less frequently than the placebo or standard of care.
+%It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exacerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
+%Here, the data had equal variances so we would expect power, not the test level, to be an issue.
+%(See the supplementary Gaussian simulations for simulations using data with heterogeneous variances.)
+%Table~\ref{tab:imbalanced_power} displays the value of these power curves and the size of the test at level $0.05$.
+%In this case, the permutation tests (except for the simple stratified permutation) had slightly higher power than ANCOVA, but the difference was not substantial.
+%
+%%\begin{center}
+%%\input{fig/imbalanced_simulation_power_summary.tex}
+%%\end{center}
+%
+%
+%\begin{table}[ht]
+%\centering
+%\begin{tabular}{ll|ccccc}
+%\multicolumn{2}{c|}{Design}                                                                 & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
+%Errors                        & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
+%\multirow{3}{*}{Gaussian}     & None                                                        & 0.046                      & 0.031                                                                                & 0.028                                                                                 & 0.042                                                                        & 0.045                                                                        \\
+%                              & Constant                                                    & 0.840                      & 0.490                                                                                & 0.810                                                                                 & 0.870                                                                        & 0.860                                                                        \\
+%                              & Single stratum                                              & 0.120                      & 0.050                                                                                & 0.090                                                                                 & 0.100                                                                        & 0.120                                                                        \\ \hline
+%\multirow{3}{*}{Heavy-tailed} & None                                                        & 0.034                      & 0.054                                                                                & 0.055                                                                                 & 0.047                                                                        & 0.042                                                                        \\
+%                              & Constant                                                    & 0.230                      & 0.200                                                                                & 0.270                                                                                 & 0.260                                                                        & 0.270                                                                        \\
+%                              & Single stratum                                              & 0.070                      & 0.080                                                                                & 0.090                                                                                 & 0.090                                                                        & 0.080                                                                        \\ \hline
+%\end{tabular}
+%\caption{Empirical power at level $0.05$ for Gaussian simulated data with imbalanced treatment groups.} 
+%\label{tab:imbalanced_power}
+%\end{table}
+%
+%In our third set of simulations, we made the baseline measurements $Y_{ij0}$ discrete and skewed.
+%This type of data may occur in surveys where individuals are asked to rate their symptom severity on a discrete ordinal scale.
+%We generated the baseline measures using independent draws from a Poisson distribution with mean 4 and censored values at 10.
+%We conditioned on these observed baseline measures after generating them once.
+%Then, for each of 1000 trials, we randomly assigned treatment to half of the individuals in each stratum and generated errors taking on values $0.5$ and $-0.5$ with equal probability.
+%We constructed $Y_{ij1}$ using Equation~\ref{eqn:dgp} with a treatment effect of $\gamma = 0.5$.
+%To ensure that both baseline and outcome were integer scores, we let the $Y_{ij1}$ be unobservable and defined the observed outcome as
+%
+%\begin{displaymath}
+%   \tilde{Y}_{ij1} = \left\{
+%     \begin{array}{ll}
+%       1 & \text{if } {Y}_{ij1} < 1\\
+%       10 & \text{if } {Y}_{ij1} > 10 \\
+%       \lfloor {Y}_{ij1} \rfloor & \text{otherwise}
+%     \end{array}
+%   \right.
+%\end{displaymath}
+%
+%
+%The observed outcomes $\tilde{Y}_{ij1}$, the baseline measures $Y_{ij0}$, and the errors were discrete, violating normality assumptions.
+%
+%We examined the effect of stratum size in this set of simulations.
+%For each of the two treatment effect designs, we considered both equal and unequal stratum sizes.
+%The simulations with equal stratum sizes were generated as before, where each stratum contained $n_j = 16$ individuals.
+%The simulations with unequal stratum sizes were set up so the smallest stratum had only $n_1=8$ individuals, the next had $n_2= 16$, and the largest had $n_3=24$.
+%The design with a nonzero treatment effect in a single stratum and unequal stratum sizes was particularly unfavorable:
+%the nonzero effect only occurred in the smallest stratum.
+%
+%Figure~\ref{fig:skewed_sim_power} shows the estimated power curves in these four designs.
+%When the treatment effect was constant, sample sizes within each stratum didn't matter; 
+%the two power curve plots in the first row of Figure~\ref{fig:skewed_sim_power} look the same up to chance variation.
+%When the treatment effect only occurred in one stratum, there was a substantial loss in power.
+%It was nearly impossible to detect a difference in treatments when the affected stratum was too small.
+%In practice, one does not usually know a priori which individuals will be affected by treatment (otherwise, those not affected would be excluded from the study altogether).
+%This result suggests that when stratifying, one must be cautious not to stratify too finely, as it is impossible to measure an effect with an insufficient sample size.
+%
+%All of the tests appeared to be conservative and to have lower than nominal level.
+%As in the first set of simulations, the stratified permutation test using the raw outcomes had the lowest power of the five tests.
+%The stratified permutation test using the change scores also had lower power than the three linear model tests.
+%Presumably, this was because the differences $\tilde{Y}_{ij1} - Y_{ij0}$ were discrete and limited to a small number of values, so the test statistic could not vary greatly across permutations.
+%Table~\ref{tab:skewed_power} confirms this: power for the three linear model methods was high when the treatment was constant across strata, and power for all of the methods suffered when the effect was isolated in a single stratum.
+%
+%
+%\begin{figure}
+%\centering
+%\includegraphics[width = \textwidth]{fig/skewed_simulation_power.pdf}
+%\caption{Empirical power curves for the skewed, discrete simulated data}
+%\label{fig:skewed_sim_power}
+%\end{figure}
+%
+%%\begin{center}
+%%\input{fig/skewed_simulation_power_summary.tex}
+%%\end{center}
+%\begin{table}[]
+%\centering
+%\begin{tabular}{ll|ccccc}
+%\multicolumn{2}{c|}{Design}                                                               & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
+%Group sizes                 & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
+%\multirow{3}{*}{Balanced}   & None                                                        & 0.036                      & 0.025                                                                                & 0.021                                                                                 & 0.035                                                                        & 0.037                                                                        \\
+%                            & Constant                                                    & 0.974                      & 0.143                                                                                & 0.853                                                                                 & 0.972                                                                        & 0.977                                                                        \\
+%                            & Single stratum                                              & 0.422                      & 0.050                                                                                & 0.275                                                                                 & 0.429                                                                        & 0.424                                                                        \\ \hline
+%\multirow{3}{*}{Imbalanced} & None                                                        & 0.038                      & 0.024                                                                                & 0.019                                                                                 & 0.037                                                                        & 0.039                                                                        \\
+%                            & Constant                                                    & 0.971                      & 0.147                                                                                & 0.860                                                                                 & 0.970                                                                        & 0.972                                                                        \\
+%                            & Single stratum                                              & 0.190                      & 0.070                                                                                & 0.100                                                                                 & 0.190                                                                        & 0.190                                                                        \\ \hline
+%\end{tabular}
+%\caption{Empirical power at level $0.05$ for discrete, skewed simulated data} 
+%\label{tab:skewed_power}
+%\end{table}
+%It makes sense that in all of these simulations, the permutation test using the change scores as the response measure is more powerful than the permutation test using the raw outcomes.
+%The baseline and outcome were highly correlated, so the change score had lower variance than the outcome alone. 
+%In additional simulations, 
+%we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had a correlation of $0.25$.
+%In this case, the result was reversed: 
+%the differenced test had low power while the test using the raw outcomes had a power curve closer to the linear model-based tests.
+%(See the supplementary Gaussian simulations for a more detailed treatment.)
+%This behavior has been studied before by \cite{frison_repeated_1992}, who recommended using the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
+%
 
@@
 This paper adds to the literature comparing parametric and nonparametric tests.
+On the one hand, some argue that violations of parametric test assumptions necessitate the use of permutation methods.
+\cite{ludbrook_why_1998} point out that medical trials rarely follow the population sampling model implicit in parametric methods.
+Many people recommend using permutation tests in place of the usual parametric tests for typical analyses, such as ANOVA and generalized linear models (\cite{still_approximate_1981, winkler_permutation_2014}).
+They argue that there are a myriad of ways that the data may violate the necessary assumptions for the test, and so permutation tests are more robust.
+
+However, parametric and nonparametric tests seem to perform similarly when compared side-by-side in simulations, even when data violate the assumptions of the parametric method.
+Medical trials often use Likert scales to score pain or symptom severity, resulting in discrete data that does not match the normality assumptions of parametric tests.
+However, \cite{winter_five-point_2010} found that the two sample $t$ test and Mann-Whitney test had comparable Type I and II error rates for five-point Likert scale data, suggesting that the violation of normality does not entirely invalidate the parametric test.
+\cite{vickers_parametric_2005} compared the parametric ANCOVA to the Mann-Whitney rank test in the context of randomized experiments, finding that except in extreme situations, ANCOVA was more powerful than the nonparametric test.
+Most similar to our question of study, \cite{anderson_empirical_1999} found little difference between several permutation tests for coefficients in a linear model alongside the parametric $t$ test.
+In these situations, the permutation test strengthens conclusions by giving evidence that the parametric test is robust to departures from its assumptions.
+
 Our results match those of \citet{vickers_parametric_2005} and \citet{anderson_empirical_1999}, where parametric ANCOVA and regression-based $t$ tests performed the same or better than the comparable nonparametric tests.

### 2017-09-05 simulations/continuous_outcomes.Rmd
Create new simulations with continuous outcomes in potential outcome framework

---
title: "ANCOVA Comparison Simulations: Continuous Outcomes"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE,
                      tidy = TRUE, tidy.opts = list(width.cutoff = 60),
                      cache = TRUE)
#devtools::install_github("statlab/permuter")
require(permuter)
require(dplyr)
require(reshape2)
require(ggplot2)
require(xtable)
options(xtable.comment = FALSE, xtable.include.rownames = FALSE, xtable.align = "p{3.5cm}")
require(grid)
require(gridExtra)

report_theme <- theme_bw() + theme(
  axis.text.x = element_text(size = 14, angle = -45),
  axis.text.y = element_text(size = 14),
  axis.title = element_text(size = 20),
  title = element_text(size = 20),
  legend.title = element_text(size = 14),
  legend.text = element_text(size = 14),
  strip.text.x = element_text(size = 14)
)
```

# Continuous data

Step 1:
We generated continuous potential outcomes.
We used two scenarios.
In the first, the treatment effect was constant across strata.
We drew a latent random variate $v_{ij}$ from the uniform distribution on $[-4, 4]$.
In the second scenario, the treatment effect varied across strata, and we drew the latent random variable according to
\begin{displaymath}
   v_{ij} \sim \left\{
     \begin{array}{lr}
       \text{Unif}[-4, -1] & : j=1\\
       \text{Unif}[-1, 1] & : j=2\\
       \text{Unif}[1, 4] & : j=3\\
     \end{array}
   \right.
\end{displaymath}

We generated independent and identically distributed errors $\varepsilon_{ij}$ and $\delta_{ij}$ and varied the error distribution. 
The errors were either standard normal (to mimic the usual ANCOVA assumptions),
$t$ distributed with two degrees of freedom,
standard lognormal,
or exponentially distributed, with scale parameter $1$ and shifted to have mean zero.
The observed $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$ were independent across $i$ and $j$.

The baseline value for individual $i,j$ was 
$$ X_{ij} = \frac{\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}$$.

Then we generated potential outcomes as

$$Y_{ij}(Z_{ij}) = \frac{(2Z_{ij} - 1)\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \delta_{ij}.$$

The treatment effect for individual $(i, j)$ is $\gamma e^{v_{ij}}$ and we would like to estimate the average treatment effect in the sample.

Assume that there are $n_j=16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerate $Z$ 10,000 times.
We repeat this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.

Step 2:
We regenerate $Z$ and recompute $Y_(Z)$ 10,000 times for each design.

Step 3:
We repeat step 1 and 2 for different distributions on $\varepsilon$ and $\delta$.

In expectation, the average treatment effect is $\gamma$.
We compare the empirical power of five tests to detect this treatment effect:

* ANCOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for stratum.
* Stratified permutation: we permute treatment assignment within stratum, then take the difference in means between treated and control outcomes $Y_1$
* Differenced permutation: we do the same permutation procedure as the stratified permutation test, except we use the difference between outcome and baseline, $Y_1 - Y_0$
* Linear model (LM) permutation: we use the same stratified permutation procedure as above, except use the $t$-statistic for the coefficient on treatment in the linear regression of $Y_1$ on $Y_0$, $Z$, and stratum dummies
* Freedman-Lane test: see the other Rmd document for a full description of this procedure

# Data-generation, tests, and plotting functions

```{r generate_sample}
gen_y1 <- function(gamma, v, error, Z){
  y1 <- 0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error
  return(y1)
}

gen_y0 <- function(gamma, v, error){
  y0 <- 0.5*(-1*gamma*exp(v) + exp(v/2)) + error
  return(y0)
}

generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
  # Input:
  # gamma = multiplier for the magnitude of the treatment effect
  # effect = "same effect" or "heterogeneous"
  # errors = "normal" or "heavy"
  # n = number of individuals at each stratum
  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), stratumID (stratum), stratum_effect (beta coefficient per individual), and epsilon (errors)

  stratumID <- rep(1:3, times = n)
  N <- sum(n)

  # What is the treatment effect?
  if(effect == "same effect"){
    v <- runif(N, min=-4, max=4)
  } else if(effect == "heterogeneous"){
    v <- rep(0, N)
    v[stratumID == 1] <- runif(n[1], min=-4, max=-1)
    v[stratumID == 2] <- runif(n[2], min=-1, max=1)
    v[stratumID == 3] <- runif(n[3], min=1, max=4)
  } else {
    stop("invalid parameter effect")
  }

  # Generate errors
  if(errors == "normal"){
    epsilon <- rnorm(N)
    delta <- rnorm(N)
  } else if (errors == "t"){
    epsilon <- rt(N, df = 2)
    delta <- rt(N, df = 2)
  } else if (errors == "lognormal"){
    epsilon <- rlnorm(N)
    delta <- rlnorm(N)
  } else if (errors == "exponential"){
    epsilon <- rexp(N) - 1
    delta <- rexp(N) - 1
  } else {
    stop("invalid errors parameter")
  }

  # Generate covariates
  Z <- rep(0:1, length.out = N)
  Y0 <- gen_y0(gamma, v, epsilon)
  Y1 <- gen_y1(gamma, v, delta, Z)
  return(data.frame(Y1, Y0, Z, v, stratumID, epsilon, delta))
}

generate_simulated_pvalues <- function(dataset, reps = 1e3){
  # Inputs:
  # dataset = a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), and stratumID (stratum)
  # Returns: a vector of p-values
  # first element is the p-value from the ANCOVA
  # second element is the p-value from the stratified two-sample permutation test
  # third element is the p-value from the linear model test, permuting treatment
  # fourth element is the p-value from the Freedman-Lane linear model test, permuting residuals

  # ANCOVA
  modelfit <- lm(Y1 ~ Y0 + Z + factor(stratumID), data = dataset)
  resanova <- summary(aov(modelfit))
  anova_pvalue <- resanova[[1]]["Z", "Pr(>F)"]

  # Stratified permutation test of Y1
  observed_diff_means <- mean(dataset$Y1[dataset$Z == 1]) - mean(dataset$Y1[dataset$Z == 0])
  diff_means_distr <- stratified_two_sample(group = dataset$Z, response = dataset$Y1, stratum = dataset$stratumID, reps = reps)
  perm_pvalue <- t2p(observed_diff_means, diff_means_distr, alternative = "two-sided")

  # Diffed permutation test of Y1-Y0
  dataset$diff <- dataset$Y1 - dataset$Y0
  observed_diff_means2 <- mean(dataset$diff[dataset$Z == 1]) - mean(dataset$diff[dataset$Z == 0])
  diff_means_distr2 <- stratified_two_sample(group = dataset$Z, response = dataset$diff, stratum = dataset$stratumID, reps = reps)
  perm_pvalue2 <- t2p(observed_diff_means2, diff_means_distr2, alternative = "two-sided")

  # Permutation of treatment in linear model
  observed_t1 <- summary(modelfit)[["coefficients"]]["Z", "t value"]

  # Freedman-Lane linear model residual permutation
  lm2_no_tr <- lm(Y1~Y0 + factor(stratumID), data = dataset)
  dataset$lm2_resid <- residuals(lm2_no_tr)
  lm2_yhat <- fitted(lm2_no_tr)

  lm1and2_t_distr <-  replicate(reps, {
    dataset[, c("Z_perm", "lm2_resid_perm")] <- permute_within_groups(dataset[, c("Z", "lm2_resid")], dataset$stratumID)
    lm1_perm <- lm(Y1 ~ Y0 + Z_perm + factor(stratumID), data = dataset)

    dataset$response_fl <-  lm2_yhat + dataset$lm2_resid_perm
    lm2_perm <- lm(response_fl ~ Y0 + Z + factor(stratumID), data = dataset)

    c(summary(lm1_perm)[["coefficients"]]["Z_perm", "t value"],
      summary(lm2_perm)[["coefficients"]]["Z", "t value"])
  })
  lm_pvalue <- t2p(observed_t1, lm1and2_t_distr[1, ], alternative = "two-sided")
  fl_pvalue <- t2p(observed_t1, lm1and2_t_distr[2, ], alternative = "two-sided")

  return(c("ANCOVA" = anova_pvalue, 
           "Stratified Permutation" = perm_pvalue, 
           "Differenced Permutation" = perm_pvalue2,
           "LM Permutation" = lm_pvalue, 
           "Freedman-Lane" = fl_pvalue))
}
```

```{r function_power_curves}
compute_power <- function(pvalues){
  sapply((0:99)/100, function(p) mean(pvalues <= p, na.rm = TRUE))
}

plot_power_curves <- function(power_mat, title){
  melt(power_mat) %>% 
  mutate("pvalue" = Var1/100) %>%
  mutate("Method" = Var2) %>%
  ggplot(aes_string(x = "pvalue", y = "value", color = "Method")) +
  geom_line() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  xlab("P-value") +
  ylab("Power") +
  ggtitle(title) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 16),
    title = element_text(size = 16),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 14),
    strip.text.x = element_text(size = 12)
  ) 
}

plot_power_gamma <- function(powermat_list, gamma_vec, alpha, title){
  gamma_power_alpha <- t(sapply(powermat_list, function(x) x[floor(alpha*nrow(x)),]))
  colnames(gamma_power_alpha) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
  melt(gamma_power_alpha, value.name="Power") %>%
  mutate("Method" = Var2) %>%
  mutate("Effect" = rep(gamma_vec, 5)) %>% 
  ggplot(aes(x = Effect, y = Power)) +
  geom_line(aes(color = Method))+
  xlab("Effect size") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 16),
    title = element_text(size = 16),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 14),
    strip.text.x = element_text(size = 12)
  ) 
}

```

```{r vary_effect_size_plot, fig.height=4}
load("continuous_outcomes/vary_gamma_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power, gamma_vec, 0.05, "Power at level 5%, Constant Treatment Effect") 
```

```{r vary_effect_size_het_plot, fig.height=4}
load("continuous_outcomes/vary_gamma_het_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power_het <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power_het, gamma_vec, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
```
```{r gaussian_homogeneous}
gaussian_homogeneous <- gamma_power[[which(abs(gamma_vec - 0.2) < 1e-6)]]
```
```{r gaussian_heterogeneous}
load("continuous_outcomes/gaussian_heterogeneous.Rda")
```
```{r t_homogeneous}
load("continuous_outcomes/t_homogeneous.Rda")
```
```{r t_heterogeneous}
load("continuous_outcomes/t_heterogeneous.Rda")

```
```{r ln_homogeneous}
load("continuous_outcomes/lognormal_homogeneous.Rda")

```
```{r ln_heterogeneous}
load("continuous_outcomes/lognormal_heterogeneous.Rda")
```
```{r exponential_homogeneous}
load("continuous_outcomes/exponential_homogeneous.Rda")
```
```{r exponential_heterogeneous}
load("continuous_outcomes/exponential_heterogeneous.Rda")

```

```{r power_summary, results = "asis"}
powers <- list(gaussian_homogeneous,
               t_homogeneous,
               lognormal_homogeneous,
               exponential_homogeneous,
               gaussian_heterogeneous,
               t_heterogeneous,
               lognormal_heterogeneous,
               exponential_heterogeneous)
summary05 <- t(sapply(powers, function(x) x[5,]))
summary05 <- data.frame(rep(c("Normal", "t(2)", "Log Normal", "Exponential"), 2), summary05)
colnames(summary05) <-  c("Errors", "ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane") 
```

```{r power_summary_const_tab, results = "asis"}
summarytab_constant <- xtable(summary05[1:4,], digits = 3,
       caption = "Empirical power at level $0.05$ for simulated data with constant additive treatment effects",
       label = "tab:power_grid1") 
print(summarytab_constant, include.rownames=FALSE) 
```

```{r power_summary_het_tab, results = "asis"}
summarytab_het <- xtable(summary05[5:8, ], digits = 3,
       caption = "Empirical power at level $0.05$ for simulated data with heterogeneous treatment effects",
       label = "tab:power_grid2") 
print(summarytab_het, include.rownames=FALSE) 
```

### 2017-09-05 simulations/continuous_outcomes/exponential_heterogeneous.R

@@ -1,0 +1,14 @@
+source("simulation_utils.R")
+library(doParallel)
+registerDoParallel(cores=5)
+
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "exponential")
+pvalues <- foreach(i = 1:5) %dopar% {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  generate_simulated_pvalues(tmp)
+}
+pvalues <- do.call(rbind, pvalues)
+exponential_heterogeneous <- apply(pvalues, 2, compute_power)
+save(exponential_heterogeneous, file="exponential_heterogeneous.Rda")

### 2017-09-05 simulations/continuous_outcomes/exponential_homogeneous.R

@@ -1,0 +1,14 @@
+source("simulation_utils.R")
+library(doParallel)
+registerDoParallel(cores=5)
+
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+tmp <- generate_simulated_data(gamma = 0.2, effect = "same effect", errors = "exponential")
+pvalues <- foreach(i = 1:10000) %dopar% {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  generate_simulated_pvalues(tmp)
+}
+pvalues <- do.call(rbind, pvalues)
+exponential_homogeneous <- apply(pvalues, 2, compute_power)
+save(exponential_homogeneous, file="exponential_homogeneous.Rda")

### 2017-09-05 simulations/continuous_outcomes/gaussian_heterogeneous.R

@@ -1,0 +1,14 @@
+source("simulation_utils.R")
+library(doParallel)
+registerDoParallel(cores=5)
+
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "normal")
+pvalues <- foreach(i = 1:10000) %dopar% {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  generate_simulated_pvalues(tmp)
+}
+pvalues <- do.call(rbind, pvalues)
+gaussian_heterogeneous <- apply(pvalues, 2, compute_power)
+save(gaussian_heterogeneous, file="gaussian_heterogeneous.Rda")

### 2017-09-05 simulations/continuous_outcomes/lognormal_heterogeneous.R

@@ -1,0 +1,14 @@
+source("simulation_utils.R")
+library(doParallel)
+registerDoParallel(cores=5)
+
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "lognormal")
+pvalues <- foreach(i = 1:10000) %dopar% {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  generate_simulated_pvalues(tmp)
+}
+pvalues <- do.call(rbind, pvalues)
+lognormal_heterogeneous <- apply(pvalues, 2, compute_power)
+save(lognormal_heterogeneous, file="lognormal_heterogeneous.Rda")

### 2017-09-05 simulations/continuous_outcomes/lognormal_homogeneous.R

@@ -1,0 +1,15 @@
+source("simulation_utils.R")
+library(doParallel)
+
+registerDoParallel(cores=5)
+
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+tmp <- generate_simulated_data(gamma = 0.2, effect = "same effect", errors = "lognormal")
+pvalues <- foreach(i = 1:10000) %dopar% {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  generate_simulated_pvalues(tmp)
+}
+pvalues <- do.call(rbind, pvalues)
+lognormal_homogeneous <- apply(pvalues, 2, compute_power)
+save(lognormal_homogeneous, file="lognormal_homogeneous.Rda")

### 2017-09-05 simulations/continuous_outcomes/simulation_utils.R

@@ -1,0 +1,271 @@
+#devtools::install_github("statlab/permuter")
+#require(permuter)
+require(dplyr)
+require(reshape2)
+require(ggplot2)
+require(xtable)
+options(xtable.comment = FALSE, xtable.include.rownames = FALSE, xtable.align = "p{3.5cm}")
+require(grid)
+require(gridExtra)
+
+report_theme <- theme_bw() + theme(
+  axis.text.x = element_text(size = 14, angle = -45),
+  axis.text.y = element_text(size = 14),
+  axis.title = element_text(size = 20),
+  title = element_text(size = 20),
+  legend.title = element_text(size = 14),
+  legend.text = element_text(size = 14),
+  strip.text.x = element_text(size = 14)
+)
+
+permute_within_groups <- function(x, group) {
+  if(is.vector(x)){
+    for(g in unique(group)){
+      gg <- (group == g)
+      x[gg] <- sample(x[gg])
+    }
+  } else if(is.data.frame(x) | is.matrix(x)){
+    for(g in unique(group)){
+      gg <- which(group == g)
+      x[gg, ] <- x[sample(gg), ]
+    }
+  } else {
+    stop("x is an invalid data type")
+  }
+  return(x)
+}
+
+
+
+stratified_two_sample <- function(group, response, stratum, stat = "mean", reps = 1000) {
+  if (!is.vector(group) | !is.vector(response) | !is.vector(stratum)) {
+    stop("inputs must be vectors")
+  }
+  if (!is.numeric(response)) {
+    stop("response must be numeric")
+  }
+  
+  if (length(unique(group)) > 2) {
+    stop("two samples only")
+  }
+  
+  groups <- unique(group)
+  strata <- unique(stratum)
+  
+  ordering <- order(group)
+  response <- response[ordering]
+  stratum <- stratum[ordering]
+  group <- group[ordering]
+  
+  ntreat <- table(group)[1]
+  N <- length(group)
+  
+  # If stat is callable, use it as the test function. Otherwise, look in the
+  # dictionary
+  stats <- list(mean = function(u) {
+    mean(u[1:ntreat], na.rm = TRUE) - mean(u[(ntreat + 1):N], na.rm = TRUE)
+  }, t = function(u) {
+    t.test(u[1:ntreat], u[(ntreat + 1):N], var.equal = TRUE)$statistic
[…]

### 2017-09-05 simulations/continuous_outcomes/t_heterogeneous.R

@@ -1,0 +1,14 @@
+source("simulation_utils.R")
+library(doParallel)
+registerDoParallel(cores=5)
+
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "t")
+pvalues <- foreach(i = 1:10000) %dopar% {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  generate_simulated_pvalues(tmp)
+}
+pvalues <- do.call(rbind, pvalues)
+t_heterogeneous <- apply(pvalues, 2, compute_power)
+save(t_heterogeneous, file="t_heterogeneous.Rda")

### 2017-09-05 simulations/continuous_outcomes/t_homogeneous.R

@@ -1,0 +1,14 @@
+source("simulation_utils.R")
+library(doParallel)
+registerDoParallel(cores=5)
+
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+tmp <- generate_simulated_data(gamma = 0.2, effect = "same effect", errors = "t")
+pvalues <- foreach(i = 1:10000) %dopar% {
+  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  generate_simulated_pvalues(tmp)
+}
+pvalues <- do.call(rbind, pvalues)
+t_homogeneous <- apply(pvalues, 2, compute_power)
+save(t_homogeneous, file="t_homogeneous.Rda")

### 2017-09-05 simulations/continuous_outcomes/vary_gamma.R

@@ -1,0 +1,19 @@
+source("simulation_utils.R")
+library(doParallel)
+
+registerDoParallel(cores=5)
+
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+gamma_vec <- seq(0, 0.2, by=0.05)
+gamma_res <- foreach(i = seq_along(gamma_vec)) %dopar% {
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "same effect", errors = "normal")
+  pvalues <- replicate(10000, {
+    tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+    tmp$Y1 <-gen_y1(gamma = gamma_vec[i], v=tmp$v, error=tmp$delta, tmp$Z)
+    generate_simulated_pvalues(tmp, reps=1e3)
+  })
+  return(pvalues)
+}
+
+gamma_res <- lapply(gamma_res, t)
+save(gamma_res, file="vary_gamma_results.Rda")

### 2017-09-05 simulations/continuous_outcomes/vary_gamma_het.R

@@ -1,0 +1,19 @@
+source("simulation_utils.R")
+library(doParallel)
+
+registerDoParallel(cores=25)
+
+set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+gamma_vec <- seq(0, 0.2, by=0.05)
+gamma_res <- foreach(i = seq_along(gamma_vec)) %dopar% {
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "heterogeneous", errors = "normal")
+  pvalues <- replicate(10000, {
+    tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+    tmp$Y1 <-gen_y1(gamma = gamma_vec[i], v=tmp$v, error=tmp$delta, tmp$Z)
+    generate_simulated_pvalues(tmp, reps=1e3)
+  })
+  return(pvalues)
+}
+
+gamma_res_het <- lapply(gamma_res, t)
+save(gamma_res_het, file="vary_gamma_het_results.Rda")

### 2017-09-11 ms/ms.tex
Edits to text, add refs

@@
 To carry out the standard parametric hypothesis test for a linear model, the following assumptions are needed (\cite{freedman_statistical_2005}):
-\todo{CHECK}
 \begin{enumerate}
@@
 
-\todo{are the estimator, standard error, variance clear here?}
 The coefficients are estimated using least squares (or equivalently, by maximizing the likelihood).
 The estimated coefficient $\hat{\gamma}$ is the estimated average treatment effect. 
-(Note, only Assumptions (1) and (2) are needed to guarantee the existence of a solution.
-The solution will be unique as long as there is no linear relationship between $X$, $Z$, and stratum membership.)
 This procedure also yields an estimate $\hat{\sigma}_{\hat{\gamma}}^2$ of the variance of $\hat{\gamma}$.
@@
 The $p$-value for this hypothesis test is the probability, assuming the null hypothesis of zero coefficient is true, that a value drawn from the $t$ distribution is larger in magnitude than $T$.
-This test is equivalent to the $F$-test and the likelihood ratio test assuming the Gaussian model,
-and \todo{ cite this? is therefore locally most powerful ????}.
+This test is equivalent to the $F$-test.
+When the model assumptions are true, it is uniformly most powerful.
 
 However, in the context of a randomized experiment, these assumptions will not be satisfied.
-The estimated treatment effect $\hat{\gamma}$ and nominal standard errors $\hat{\sigma}_{\hat{\gamma}}$ can be severely biased (\todo{cite freedman 2008, lin 2013}).
-\todo{Lin 2013} shows that running a regression with a full set of treatment and covariate interaction terms cannot hurt asymptotic precision, and using the Huber-White sandwich standard errors can yield asymptotically valid confidence intervals.
+The estimated treatment effect $\hat{\gamma}$ and nominal standard errors $\hat{\sigma}_{\hat{\gamma}}$ can be severely biased (\cite{freedman_regression_2008, lin_agnostic_2013}).
+\cite{lin_agnostic_2013} shows that running a regression with a full set of treatment and covariate interaction terms cannot hurt asymptotic precision, and using the Huber-White sandwich standard errors can yield asymptotically valid confidence intervals.
 In small samples, the bias may still be substantial.
-\todo{cite Sekhon post-stratification} show that post-stratification, estimating treatment effects within groups of similar individuals defined after data are collected, can ameliorate this bias.
+\cite{miratrix_adjusting_2013} show that post-stratification, estimating treatment effects within groups of similar individuals defined after data are collected, can ameliorate this bias.
 This is equivalent to estimating a fully saturated linear model with interaction terms for treatment and stratum membership.
@@
 In particular, the potential outcomes and covariates $(Y_{ij}(1), Y_{ij}(0), X_{ij})$ are fixed in the finite population under study, while the observed response $Y_{ij}$ changes with $Z_{ij}$.
-\todo{cite rosenberg 2002}
 
@@
 Under the null hypothesis, which treatment an individual received amounts to an arbitrary label.
-Once we observe one response under a particular treatment, we can impute the other potential outcome; namely, it would have been the same.
-Thus, the observed $Y_{ij}$ form a sufficient statistic for the data. \todo{cite P\&S}
+Once we observe one response under a particular treatment, we can impute the other potential outcome; namely, it would have been the same  (\cite{rosenbaum_covariance_2002}).
+Thus, the observed $Y_{ij}$ form a sufficient statistic for the data (\cite{pesarin_permutation_2010}).
 This null hypothesis is stronger than the null hypothesis for the parametric ANCOVA, which only asserts that differences in potential outcomes are zero \textit{on average}.
@@
 
-The treatment effect for individual $(i, j)$ is $\gamma e^{v_{ij}}$ and we would like to estimate the average treatment effect in the sample.
+We included $n_j=16$ individuals per stratum and treatment assignment was balanced, i.e. 8 people received each treatment in each stratum.
+After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerated $Z$ and the potential outcomes $Y(Z)$ 10,000 times.
+We repeated this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.
 
-Assume that there are $n_j=16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
-After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerate $Z$ and the potential outcomes $Y(Z)$ 10,000 times.
-We repeat this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.
+The treatment effect for individual $(i, j)$ was $\gamma e^{v_{ij}}$.
+We tested whether the treatment had a nonzero effect in the sample using each of the tests described.
 
@@
 \subsection{Discrete outcomes}
+
+Many outcomes are measured on a discrete, ordinal scale but the same regression-based methods are used to test for a nonzero treatment effect.
+We repeated the procedure above, instead discretizing the covariate $X$ and potential outcomes $Y(Z)$ by taking their floor.
+We used the same distributions for the latent variables $v$.
+We drew $\varepsilon$ and $\delta$ from independent standard normal distributions.
 

### 2017-09-13 results/results.Rmd
Edits to text in clinical results, add correlation test, up the number of replications to 1e4

@@
 ```
+
+# Summary
+
+We reanalyzed a dataset from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease (GERD). 
+Patients were treated at eight sites in two different countries. 
+At each site, patients were randomly assigned one of two treatments. 
+Patients were observed for a week before receiving treatment and for a week after receiving treatment. 
+On each of the fourteen days of observation, patients responded to a survey about their heartburn, regurgitation, and dyspepsia frequency and severity. 
+These endpoints were measured on a discrete scale. 
+There were several additional endpoints, such as daily regurgitation, daily “hrdq”, and daily dyspepsia, calculated from the survey measures. 
+Daily “hrdq” was the primary endpoint. 
+To reduce day-to-day variation, we averaged the measures from each week to obtain two observations per patient, one pre-treatment and one post-treatment.
+
+In what follows, we compare the parametric ANCOVA, the stratified permutation test, and the two linear model-based permutation tests.
+We begin by walking through the tests in detail using the primary endpoint.
+We illustrate different ways of controlling for the baseline measure: including it in the model and using change scores as the dependent variable.
+The original analysis of the dataset controls for baseline by including it in the model, so for consistency we choose that method for our final analysis.
+We conclude by running the tests for each endpoint.
 
@@
 
-The primary endpoint is `daily_hrdq`, a measure derived from other variables that were measured in the survey.
-We show summary statistics for this outcome for each treatment group below.
-```{r hrdq, results = "asis"}
-daily_hrdq <- reshape_data("daily_hrdq")
-summary_stats_hrdq <- cbind(summary(daily_hrdq$difference[daily_hrdq$tr == "B"]),
-                            summary(daily_hrdq$difference[daily_hrdq$tr == "A"]))
-colnames(summary_stats_hrdq) <- c("B", "A")
-print(xtable(summary_stats_hrdq,
-             caption = "Summary statistics for change in primary 
-                        endpoint (daily hrdq) for the two treatment groups."),
-      include.rownames = TRUE)
+Table 1 shows the correlation between baseline and outcome for each variable.
+Correlations range between 0.548 and 0.696.
+This suggests that using the difference in outcome and baseline as the dependent variable in the following models might yield higher power than controlling for baseline as a covariate.
+
+```{r corr_table_print, results = "asis", fig.align = "center", fig.cap = "Correlation between baseline and outcome, for each continuous endpoint."}
+
+baseline_outcome_corr <- rep(0, length(continuous_vars))
+for(i in seq_along(continuous_vars)){
+  col <- continuous_vars[i]
+  tmpdata <- reshape_data(col)
+  baseline_outcome_corr[i] <- cor(tmpdata$Baseline, tmpdata$Outcome)
+}
+
+baseline_outcome_corr <- data.frame(baseline_outcome_corr)
+colnames(baseline_outcome_corr) <- "Correlation"
+rownames(baseline_outcome_corr) <- continuous_vars
+
+summarytabc <- xtable(baseline_outcome_corr, 
+             digits = 3,
+             align = paste0(c("r|", rep("p{1in}", 1)), 
+                            collapse = ""),
+             caption = "Correlation between baseline and outcome, for each continuous endpoint.",
+             label = "tab:clinical_corr")
+print(summarytabc, include.rownames = TRUE)
 ```
@@
 The second model uses the difference between outcome and baseline score as its dependent variable.
-The difference in the two models is how we control for the baseline measure.
+The difference in the two models is the way in which we control for the baseline measure.
 
@@
 
+The primary endpoint is `daily_hrdq`, a measure derived from other variables that were measured in the survey.
+We show summary statistics for the change from baseline for each treatment group below.
+Based on these statistics, it appears that members of group B had both larger outcomes and larger changes from baseline than group A.
+We will do two-sided hypothesis tests to assess whether the difference between groups A and B is meaningfully different from zero.
+```{r hrdq, results = "asis"}
+daily_hrdq <- reshape_data("daily_hrdq")
+summary_stats_hrdq <- cbind(summary(daily_hrdq$Outcome[daily_hrdq$tr == "B"]),
+                            summary(daily_hrdq$Outcome[daily_hrdq$tr == "A"]),
+                            summary(daily_hrdq$difference[daily_hrdq$tr == "B"]),
+                            summary(daily_hrdq$difference[daily_hrdq$tr == "A"]))
+colnames(summary_stats_hrdq) <- c("B Outcome", "A Outcome", "B Difference", "A Difference")
+print(xtable(summary_stats_hrdq,
+             caption = "Summary statistics for change in primary 
+                        endpoint (daily hrdq) for the two treatment groups."),
+      include.rownames = TRUE)
+```
+
 ## Parametric ANCOVA
-First, we do a standard parametric ANCOVA using the two models.
-Under model 1, using the outcome as response and adjusting for the baseline value, treatment has a significant effect.
-Under model 2, the significance goes away.
-(Note that treatment is still coded as A and B with the character type. 
-The `lm` function is smart enough to know that it should be coded as a factor, where we compare the effect of B relative to A,
-and coerces the variable type from string to factor behind the scenes. 
-Be very careful when you're doing this!
-It is generally safer to change types yourself.)
+
 ```{r hrdq_parametric, results = "asis"}
@@
 ```
+
+First, we do a standard parametric ANCOVA using the two models.
+Under model 1, using the outcome as response and adjusting for the baseline value, treatment has a significant effect with p-value `r round(summary(aov(lm1))[[1]]["tr", "Pr(>F)"], 3)`.
+Under model 2, the significance goes away (p-value of `r round(summary(aov(lm2))[[1]]["tr", "Pr(>F)"], 3)`).
+(Note that treatment is still coded as A and B with the character type. 
+The `lm` function is smart enough to know that it should be coded as a factor, where we compare the effect of B relative to A,
+and coerces the variable type from string to factor behind the scenes. 
+Be very careful when you're doing this!
+It is generally safer to change types yourself.)
 
@@
 the difference in means will be attenuated toward zero.
-
-Figure \ref{fig:hrdq_unadjusted_permutation} shows the permutation distributions of the simple difference in means for the two models.
-Figure \ref{fig:hrdq_unadj_perm_withinstrata} shows the permutation distribution of the absolute value of the difference of means within strata, summed across strata.
-These empirical permutation distributions are computed by randomly permuting treatment assignments within each site $10^4$ times and computing the statistic for each permutation.
-The vertical line shows the observed value of the test statistic.
-The $p$-value is two times the area of the permutation distribution to the right of the observed value;
-it is the probability of observing a value at least as extreme (falling that far away from the center of the distribution) as the observed value.
-There are several points to notice:
-
-* The second test is far less powerful than the first.
-This is likely because some sites had very few patients.
-These sites have correspondingly few unique permutations, and so they often contribute the same amount to the summed difference in means statistic.
-It is impossible to detect a difference at small significance levels when there are only 3 patients.
-* The summed difference of means only makes sense for testing against a two-sided alternative, since taking the absolute value removes the sign of the effect.
-The permutation distribution is strictly non-negative and skewed to the right.
-* This statistic is less interpretable than the difference in means because it doesn't correspond directly to any feature of the distribution of outcomes.
-
-Since the summed difference in means statistic has such low power, we abandon it in the rest of the analyses.
 
@@
 
+The empirical permutation distributions are computed by randomly permuting treatment assignments within each site $10^4$ times and computing the statistic for each permutation.
+A vertical line indicates the observed value of the test statistic.
+The $p$-value is two times the area of the permutation distribution to the right of the observed value;
+it is the probability of observing a value at least as extreme (falling that far away from the center of the distribution) as the observed value.
+Figure \ref{fig:hrdq_unadjusted_permutation} shows the permutation distributions of the simple difference in means for the two models.
+The p-values for the differenced model and raw outcome model are `r round(diff_means_pvalue2, 3)` and `r round(diff_means_pvalue1, 3)`, respectively.
+Figure \ref{fig:hrdq_unadj_perm_withinstrata} shows the permutation distribution of the absolute value of the difference of means within strata, summed across strata.
+The p-values for the differenced model and raw outcome model are `r round(diff_means_bystrata_pvalue2, 3)` and `r round(diff_means_bystrata_pvalue1, 3)`, respectively.
+
+There are several points to notice:
+
+* The second test is far less powerful than the first.
+This is likely because some sites had very few patients.
+These sites have correspondingly few unique permutations, and so they often contribute the same amount to the summed difference in means statistic.
+It is impossible to detect a difference at small significance levels when there are only 3 patients.
+* The summed difference of means only makes sense for testing against a two-sided alternative, since taking the absolute value removes the sign of the effect.
+The permutation distribution is strictly non-negative and skewed to the right.
+* This statistic is less interpretable than the difference in means because it doesn't correspond directly to any feature of the distribution of outcomes.
+
+Since the summed difference in means statistic has such low power, we abandon it in the rest of the analyses.
+
 ## Covariate-adjusted permutation test
@@
 
+The p-values for the differenced model and raw outcome model are `r round(lm_pvalue_2, 3)` and `r round(lm_pvalue_1, 3)`, respectively.
+
 ### Linear model residuals
@@
 For this dataset, we are guaranteed that treatment $Z$ is independent of $\varepsilon$, as it is randomized within site.
-We check these associations using residual plots in Figure \ref{fig:residual_plots}.
-Indeed, the distribution of residuals looks nearly equal between treatment groups in both models.
+However, the randomization does not guarantee that $\varepsilon$ is independent baseline measure, conditional on site.
+We check this associations using residual plots in Figure \ref{fig:residual_plots}.
 The distribution of residuals varies a bit across sites, but they all appear roughly centered around 0.
-Sites 4 and 5 have larger variance in the differenced model.
-This should not be an issue since we permute treatments within sites, independently across sites.
-(In general, when treatment is not randomly assigned, this condition is necessary for the permutation test to be valid.)
+
+```{r test_correlation_resid, results='asis'}
+obs_corr <- daily_hrdq %>% group_by(SITEID) %>% summarise("n"=n(), "Correlation"=cor(lm1_resid, Baseline))
+corr_pvalues <- rep(0, 8)
+
+for(i in unique(daily_hrdq$SITEID)){
+  tmp <- daily_hrdq %>% filter(SITEID == i)
+  corr_distr <- replicate(1e4,{
+    cor(permute(tmp$lm1_resid), tmp$Baseline)
+  })
+  corr_pvalues[i] <- t2p(unlist(obs_corr[i, "Correlation"]), corr_distr, alternative = "two-sided")
+}
+
+fisher_combined_corr <- pchisq(fisher(corr_pvalues), df = 2*length(corr_pvalues), lower.tail = FALSE)
+
+obs_corr <- obs_corr %>% mutate("P-value" = corr_pvalues)
+print(xtable(obs_corr, digits = 2,
+             caption = "Permutation test $p$-values for the correlation between baseline daily hrdq and residuals."))
+```
+
+By construction of the linear model, the correlation between the baseline measure and the residuals is zero.
+However, we would like to check formally that $\varepsilon$ is independent baseline measure for each site, using the residuals as a proxy for $\varepsilon$.
+To do so, we conduct independent permutation tests within each stratum using the Pearson correlation as the statistic.
+Table 3 shows the results of this test: while the correlation between baseline and residuals appears high, it is not statistically significant at the 5\% level in any stratum.
+The Fisher's combined $p$-value for the 8 strata is `r round(fisher_combined_corr, 2)`, so there is no evidence that the independence assumption is violated.
 
@@
 p11 <- daily_hrdq %>% 
-  select(-Baseline, -Outcome, -difference) %>% 
-  melt(id.vars = c("SUBJID", "tr", "SITEID")) %>%
+  select(-Outcome, -difference) %>% 
+  melt(id.vars = c("SUBJID", "Baseline", "SITEID")) %>%
   filter(variable == "lm1_resid") %>% 
-  ggplot(aes(x = factor(tr), y = value)) +
-  geom_boxplot() + 
+  ggplot(aes(x = Baseline, y = as.numeric(value))) +
+  geom_point() + 
   ylim(-2.25, 2.25)+
   geom_hline(yintercept = 0, linetype = "dashed") +
-  xlab("Treatment") + ylab("Residual") + 
+  xlab("Baseline") + ylab("Residual") + 
   ggtitle("Raw Outcome Model") +
@@
   filter(variable == "lm1_resid") %>% 
-  ggplot(aes(x = factor(SITEID), y = value)) +
+  ggplot(aes(x = factor(SITEID), y = as.numeric(value))) +
   geom_boxplot() +  
@@
   theme_bw()
-p12 <- daily_hrdq %>% 
-  select(-Baseline, -Outcome, -difference) %>% 
-  melt(id.vars = c("SUBJID", "tr", "SITEID")) %>%
-  filter(variable == "lm2_resid") %>% 
-  ggplot(aes(x = factor(tr), y = value)) +
-  geom_boxplot() +  
-  ylim(-2.25, 2.25)+
-  geom_hline(yintercept = 0, linetype = "dashed") +
-  xlab("Treatment") + ylab("Residual") +
-  ggtitle("Differenced Model") +
-  theme_bw()
 p22 <- daily_hrdq %>% 
@@
   filter(variable == "lm2_resid") %>% 
-  ggplot(aes(x = factor(SITEID), y = value)) +
+  ggplot(aes(x = factor(SITEID), y = as.numeric(value))) +
   geom_boxplot() +  
@@
   theme_bw()
-grid.arrange(p12, p11, p22, p21, nrow = 2, ncol = 2)
+grid.arrange( p21, p11, p22, layout_matrix = rbind(c(1, 2), c(3, NA)))
 ```
@@
 ```
+The p-values for the differenced model and raw outcome model are `r round(fl_pvalue_2, 3)` and `r round(fl_pvalue_1, 3)`, respectively.
 
-Table 2 shows the $p$-values for each of the four methods and the two models.
+## Summary
+
+Table 4 shows the $p$-values for each of the five methods and the two methods of controlling for the baseline.
 For the models using the raw outcome measure as the dependent variable, the effect of `daily_hrdq` is significant at the $0.05$ level in two of the five tests
@@
 
-```{r primary_endpoint_res, results = "asis", fig.align = "center", fig.cap = "Comparison of $p$-values for two measures (average outcome during treatment vs.
+```{r primary_endpoint_res, echo=FALSE, results = "asis", fig.align = "center", fig.cap = "Comparison of $p$-values for two measures (average outcome during treatment vs.
 difference of average outcome and average baseline) of the primary endpoint."}
 
-outcome_pvalues <- c(summary(aov(lm1))[[1]]["trB", "Pr(>F)"],
+outcome_pvalues <- c(summary(aov(lm1))[[1]]["tr", "Pr(>F)"],
                      diff_means_pvalue1,
@@
                      fl_pvalue_1)
-differenced_pvalues <- c(summary(aov(lm2))[[1]]["trB", "Pr(>F)"],
+differenced_pvalues <- c(summary(aov(lm2))[[1]]["tr", "Pr(>F)"],
                      diff_means_pvalue2,
@@
 
+\newpage
 # Analysis - Secondary endpoints
@@
 We run this procedure for all seven continuous endpoints.
-Table 3 shows the results.
+Table 5 shows the results.
 
@@
 pvalues_table_contin <- as.data.frame(matrix(NA, nrow = length(continuous_vars), ncol = 4))
-i <- 0
+baseline_outcome_corr <- rep(0, length(continuous_vars))
 
-for(col in continuous_vars){
-  i <- i+1
+for(i in seq_along(continuous_vars)){
+  col <- continuous_vars[i]
   tmpdata <- reshape_data(col)
+  baseline_outcome_corr[i] <- cor(tmpdata$Baseline, tmpdata$Outcome)
+
   lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = tmpdata)
@@
   observed_diff_means <- mean(tmpdata[tmpdata$tr == "B", ]$Outcome) - mean(tmpdata[tmpdata$tr == "A", ]$Outcome)
-  diff_means_distr <- stratified_two_sample(group = tmpdata$tr, response = tmpdata$Outcome, stratum = tmpdata$SITEID, reps = 1e3)
+  diff_means_distr <- stratified_two_sample(group = tmpdata$tr, response = tmpdata$Outcome, stratum = tmpdata$SITEID, reps = 1e4)
   pvalues_table_contin[i, 2] <- t2p(observed_diff_means, diff_means_distr, alternative = "two-sided")
@@
   observed_t <- summary(lm1)[["coefficients"]]["trB", "t value"]
-  lm1_t_distr <-  replicate(1e3, {
+  lm1_t_distr <-  replicate(1e4, {
     tmpdata$tr_perm <- permute_within_groups(tmpdata$tr, tmpdata$SITEID)
@@
   observed_t <- summary(lm1)[["coefficients"]]["trB", "t value"]
-  lm_t_distr <-  replicate(1e3, {
+  lm_t_distr <-  replicate(1e4, {
       lm_resid_perm <- permute_within_groups(lm_resid, tmpdata$SITEID)
@@
 ```
-```{r results_table_print, results = "asis", fig.align = "center", fig.cap = "Comparison of $p$-values from four tests, for each continuous endpoint."}
+
+```{r results_table_print, echo=FALSE, results = "asis", fig.align = "center", fig.cap = "Comparison of $p$-values from four tests, for each continuous endpoint."}
 rownames(pvalues_table_contin) <- continuous_vars

### 2017-09-13 simulations/continuous_outcomes.Rmd
Add continuous outcome simulations and results

@@
   axis.text.y = element_text(size = 14),
-  axis.title = element_text(size = 20),
-  title = element_text(size = 20),
+  axis.title = element_text(size = 14),
+  title = element_text(size = 14),
   legend.title = element_text(size = 14),
@@
 
+# Gaussian error terms, constant additive treatment effect
+
 ```{r vary_effect_size_plot, fig.height=4}
@@
 gamma_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
-plot_power_gamma(gamma_power, gamma_vec, 0.05, "Power at level 5%, Constant Treatment Effect") 
+plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
 ```
 
+First, we let the $v_{ij}$ have the same distribution across strata and generated normally distributed errors.
+We varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
+In the random populations that were generated, this corresponded to population average treatment effects of `r round(tr_effect, 2)[1:4]`, and `r round(tr_effect, 2)[5]`, respectively.
+Figure 1 shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
+When the effect size was zero, all five tests had the correct level of 5\%.
+Power for all tests increased with the effect size, though the differenced permutation test lagged behind the rest.
+This is unsurprising, as the correlation between baseline $X$ and outcome $Y$ was low:
+it was highest at $0.33$ when $\gamma=0$ and decreased to approximately $0$ when $\gamma=0.2$.
+The parametric and permutation linear model tests perform comparably and are plotted on top of each other, while the stratified permutation test has slightly more power.
+
+# Gaussian error terms, heterogeneous treatment effect
 ```{r vary_effect_size_het_plot, fig.height=4}
@@
 gamma_power_het <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
-plot_power_gamma(gamma_power_het, gamma_vec, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
+plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
 ```
+Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
+Once again, we varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
+The corresponding population average treatment effects were of `r round(tr_effect_het, 2)[1:4]`, and `r round(tr_effect_het, 2)[5]`, respectively.
+Figure 2 shows the empirical power at level 5\% for these increasing effect sizes.
+Four of the five tests had the correct level of 5\%, while the stratified permutation test rejected only 1\% of the time.
+Both the stratified and differenced permutation tests had lower power than the linear model tests as the effect size increased.
+The correlation between baseline $X$ and outcome $Y$ was low but nontrivial (between 0.04 and 0.44), so this result makes some sense:
+it is beneficial for power to control for the baseline in the linear model, but the correlation is too low to make the differencing method useful.
+
+# Comparison of error distributions
+
+We varied the distributions of the errors $\varepsilon$ and $\delta$ and examined the power of the five tests.
+Recall that these are not errors in the linear model framework, but more like disturbances on the potential outcomes and baseline measures for each individual.
+These disturbances affect the distributions of $X$ and $Y$, which may impact the power of the parametric ANCOVA, and affect the variances of $X$ and $Y$, which may make one method of controlling for the baseline more effective than another.
+We fixed $\gamma=0.2$ and let the errors have a standard normal distribution, a $t$ distribution with 2 degrees of freedom, a standard log normal distribution, or an exponential distribution with parameter $1$, shifted to have mean zero.
+
+Table 1 shows the empirical power (rejection rate among 10,000 simulations) for each of these error distributions and tests, where the $v_{ij}$ came from a single distribution across strata.
+In all cases, the correlation between baseline and outcome was lower than 0.05 in magnitude.
+Thus, the stratified permutation test, which omits the baseline measurement, had the highest power for all error distributions.
+All three linear model based tests had roughly the same rejection rate.
+The differenced permutation test had substantially less power than the other methods, due to the low correlation between baseline and outcome.
+Power was highest for normally distributed errors and was lowest for $t$ distributed errors.
+
+Table 2 shows the same results, but where the distribution of $v_{ij}$ varied across strata.
+In this case, the three linear model based tests had the highest power for each error distribution.
+While the correlation between baseline and outcome was still nearly zero, controlling for it in the linear model tended to increase precision.
+The stratified permutation test lost power for this reason.
+The differenced permutation test continued to have low power.
+Interestingly, when the errors were $t$ or log normally distributed, the power of each test did not change much whether the treatment effects were constant or heterogeneous across strata.
+However, power was lower for normally distributed errors when treatment effects varied across strata.
+In this scenario, the linear model does not fully describe the relationship between baseline and outcome: 
+interaction terms between baseline and stratum ID are needed in the model to capture this variation across strata.
+
 ```{r gaussian_homogeneous}
@@
 ```
+
+```{r plot_for_ms, include=FALSE}
+p1 <- plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
+  ylim(0, 0.9) +
+  theme(legend.position="bottom") + 
+  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
+  report_theme +
+  ggtitle("Constant Treatment \n Effect")
+
+p2 <- plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
+    ylim(0, 0.9) +
+    report_theme +
+    ggtitle("Heterogeneous Treatment \n Effect")
+
+tmp <- ggplot_gtable(ggplot_build(p1))
+leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
+mylegend <- tmp$grobs[[leg]]
+
+p3 <- grid.arrange(arrangeGrob(p1 + theme(legend.position="none"),
+                         p2 + theme(legend.position="none"),
+                         nrow=1),
+             mylegend, nrow=2,heights=c(10, 2))
+
+p3
+ggsave("../ms/fig/continuous_outcomes_simulation_power.pdf", p3, device="pdf", width=8)
+
+```

### 2017-09-13 simulations/continuous_outcomes/exponential_heterogeneous.R

@@ -6,7 +6,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "exponential")
 pvalues <- foreach(i = 1:5) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-09-13 simulations/continuous_outcomes/exponential_homogeneous.R

@@ -6,7 +6,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "same effect", errors = "exponential")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-09-13 simulations/continuous_outcomes/gaussian_heterogeneous.R

@@ -6,7 +6,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "normal")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-09-13 simulations/continuous_outcomes/lognormal_heterogeneous.R

@@ -6,7 +6,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "lognormal")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-09-13 simulations/continuous_outcomes/lognormal_homogeneous.R

@@ -7,7 +7,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "same effect", errors = "lognormal")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-09-13 simulations/continuous_outcomes/simulation_utils.R

@@ -113,14 +113,14 @@ t2p <- function(tst, distr, alternative = c("greater", "less", "two-sided")) {
 
 
 
-gen_y1 <- function(gamma, v, error, Z){
-  y1 <- 0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error
-  return(y1)
+gen_y <- function(gamma, v, error, Z){
+  y <- 0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error
+  return(y)
 }
 
-gen_y0 <- function(gamma, v, error){
-  y0 <- 0.5*(-1*gamma*exp(v) + exp(v/2)) + error
-  return(y0)
+gen_x <- function(gamma, v, error){
+  x <- 0.5*(-1*gamma*exp(v) + exp(v/2)) + error
+  return(x)
 }
 
 generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
@@ -165,8 +165,8 @@ generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
   
   # Generate covariates
   Z <- rep(0:1, length.out = N)
-  Y0 <- gen_y0(gamma, v, epsilon)
-  Y1 <- gen_y1(gamma, v, delta, Z)
+  Y0 <- gen_x(gamma, v, epsilon)
+  Y1 <- gen_y(gamma, v, delta, Z)
   return(data.frame(Y1, Y0, Z, v, stratumID, epsilon, delta))
 }
 

### 2017-09-13 simulations/continuous_outcomes/t_heterogeneous.R

@@ -6,7 +6,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "t")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-09-13 simulations/continuous_outcomes/t_homogeneous.R

@@ -6,7 +6,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "same effect", errors = "t")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y1(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-09-13 simulations/continuous_outcomes/vary_gamma.R

@@ -3,17 +3,25 @@ library(doParallel)
 
 registerDoParallel(cores=5)
 
-set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
 gamma_vec <- seq(0, 0.2, by=0.05)
 gamma_res <- foreach(i = seq_along(gamma_vec)) %dopar% {
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
   tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "same effect", errors = "normal")
   pvalues <- replicate(10000, {
     tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-    tmp$Y1 <-gen_y1(gamma = gamma_vec[i], v=tmp$v, error=tmp$delta, tmp$Z)
+    tmp$Y1 <-gen_y(gamma = gamma_vec[i], v=tmp$v, error=tmp$delta, tmp$Z)
     generate_simulated_pvalues(tmp, reps=1e3)
   })
   return(pvalues)
 }
 
 gamma_res <- lapply(gamma_res, t)
-save(gamma_res, file="vary_gamma_results.Rda")
+
+tr_effect <- rep(0, 5)
+for( i in seq_along(gamma_vec)){
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "same effect", errors = "normal")
+  tr_effect[i] <- mean(gamma_vec[i] * exp(tmp$v))
+}
+
+save(tr_effect, gamma_res, file="vary_gamma_results.Rda")

### 2017-09-13 simulations/continuous_outcomes/vary_gamma_het.R

@@ -3,17 +3,26 @@ library(doParallel)
 
 registerDoParallel(cores=25)
 
-set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
 gamma_vec <- seq(0, 0.2, by=0.05)
 gamma_res <- foreach(i = seq_along(gamma_vec)) %dopar% {
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
   tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "heterogeneous", errors = "normal")
   pvalues <- replicate(10000, {
     tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-    tmp$Y1 <-gen_y1(gamma = gamma_vec[i], v=tmp$v, error=tmp$delta, tmp$Z)
+    tmp$Y1 <-gen_y(gamma = gamma_vec[i], v=tmp$v, error=tmp$delta, tmp$Z)
     generate_simulated_pvalues(tmp, reps=1e3)
   })
   return(pvalues)
 }
 
 gamma_res_het <- lapply(gamma_res, t)
-save(gamma_res_het, file="vary_gamma_het_results.Rda")
+
+
+tr_effect_het <- rep(0, 5)
+for( i in seq_along(gamma_vec)){
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "same effect", errors = "normal")
+  tr_effect_het[i] <- mean(gamma_vec[i] * exp(tmp$v))
+}
+
+save(tr_effect_het, gamma_res_het, file="vary_gamma_het_results.Rda")

### 2017-09-13 simulations/discrete_outcomes.Rmd
Add discrete outcome simulations and results

--- 2017-09-05 simulations/continuous_outcomes.Rmd
@@
 ---
-title: "ANCOVA Comparison Simulations: Continuous Outcomes"
+title: "ANCOVA Comparison Simulations: Discrete Outcomes"
 author: "Kellie Ottoboni"
@@
 
-# Continuous data
+# Discrete data
 
 Step 1:
-We generated continuous potential outcomes.
-We used two scenarios.
-In the first, the treatment effect was constant across strata.
-We drew a latent random variate $v_{ij}$ from the uniform distribution on $[-4, 4]$.
-In the second scenario, the treatment effect varied across strata, and we drew the latent random variable according to
-\begin{displaymath}
-   v_{ij} \sim \left\{
-     \begin{array}{lr}
-       \text{Unif}[-4, -1] & : j=1\\
-       \text{Unif}[-1, 1] & : j=2\\
-       \text{Unif}[1, 4] & : j=3\\
-     \end{array}
-   \right.
-\end{displaymath}
-
-We generated independent and identically distributed errors $\varepsilon_{ij}$ and $\delta_{ij}$ and varied the error distribution. 
-The errors were either standard normal (to mimic the usual ANCOVA assumptions),
-$t$ distributed with two degrees of freedom,
-standard lognormal,
-or exponentially distributed, with scale parameter $1$ and shifted to have mean zero.
-The observed $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$ were independent across $i$ and $j$.
-
-The baseline value for individual $i,j$ was 
-$$ X_{ij} = \frac{\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}$$.
-
-Then we generated potential outcomes as
-
-$$Y_{ij}(Z_{ij}) = \frac{(2Z_{ij} - 1)\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \delta_{ij}.$$
-
-The treatment effect for individual $(i, j)$ is $\gamma e^{v_{ij}}$ and we would like to estimate the average treatment effect in the sample.
-
+We generated the covariate sand potential outcomes according to the methods in the continuous outcome section but discretized them by taking the floor.
 Assume that there are $n_j=16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
@@
 We regenerate $Z$ and recompute $Y_(Z)$ 10,000 times for each design.
-
-Step 3:
-We repeat step 1 and 2 for different distributions on $\varepsilon$ and $\delta$.
 
@@
 ```{r generate_sample}
-gen_y1 <- function(gamma, v, error, Z){
-  y1 <- 0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error
-  return(y1)
+gen_y <- function(gamma, v, error, Z){
+  y <- floor(0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error)
+  return(y)
 }
 
-gen_y0 <- function(gamma, v, error){
-  y0 <- 0.5*(-1*gamma*exp(v) + exp(v/2)) + error
-  return(y0)
+gen_x <- function(gamma, v, error){
+  x <- floor(0.5*(-1*gamma*exp(v) + exp(v/2)) + error)
+  return(x)
 }
@@
   Z <- rep(0:1, length.out = N)
-  Y0 <- gen_y0(gamma, v, epsilon)
-  Y1 <- gen_y1(gamma, v, delta, Z)
+  Y0 <- gen_x(gamma, v, epsilon)
+  Y1 <- gen_y(gamma, v, delta, Z)
   return(data.frame(Y1, Y0, Z, v, stratumID, epsilon, delta))
@@
 
+# Constant additive treatment effect
+
 ```{r vary_effect_size_plot, fig.height=4}
-load("continuous_outcomes/vary_gamma_results.Rda")
+load("discrete_outcomes/vary_gamma_results.Rda")
 gamma_vec <- seq(0, 0.2, by=0.05)
 gamma_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
-plot_power_gamma(gamma_power, gamma_vec, 0.05, "Power at level 5%, Constant Treatment Effect") 
+plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
 ```
 
+First, we let the $v_{ij}$ have the same distribution across strata and generated normally distributed errors.
+We varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
+In the random populations that were generated, this corresponded to population average treatment effects of `r round(tr_effect, 2)[1:4]`, and `r round(tr_effect, 2)[5]`, respectively.
+Figure 1 shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
+When the effect size was zero, all five tests had the correct level of 5\%, though the stratified and differenced permutation tests rejected 3.5\% of the time.
+The differenced permutation test had substantially lower power than the rest,
+again because the correlation between baseline $X$ and outcome $Y$ was low.
+In this case, the unadjusted stratified test had comparable power to the linear model based tests.
+
+# Heterogeneous treatment effect
+
 ```{r vary_effect_size_het_plot, fig.height=4}
-load("continuous_outcomes/vary_gamma_het_results.Rda")
+load("discrete_outcomes/vary_gamma_het_results.Rda")
 gamma_vec <- seq(0, 0.2, by=0.05)
 gamma_power_het <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
-plot_power_gamma(gamma_power_het, gamma_vec, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
+plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
 ```
-```{r gaussian_homogeneous}
-gaussian_homogeneous <- gamma_power[[which(abs(gamma_vec - 0.2) < 1e-6)]]
-```
-```{r gaussian_heterogeneous}
-load("continuous_outcomes/gaussian_heterogeneous.Rda")
-```
-```{r t_homogeneous}
-load("continuous_outcomes/t_homogeneous.Rda")
-```
-```{r t_heterogeneous}
-load("continuous_outcomes/t_heterogeneous.Rda")
+Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
+Once again, we varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
+The corresponding population average treatment effects were of `r round(tr_effect_het, 2)[1:4]`, and `r round(tr_effect_het, 2)[5]`, respectively.
+Figure 2 shows the empirical power at level 5\% for these increasing effect sizes.
+The pattern here is nearly identical to Figure 2 in the continuous outcome simulations, so we leave out further comments.
+
+```{r plot_for_ms, include=FALSE}
+p1 <- plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
+  ylim(0, 0.95 ) +
+  theme(legend.position="bottom") + 
+  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
+  report_theme +
+  ggtitle("Constant Treatment \n Effect")
+
+p2 <- plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
+    ylim(0, 0.95) +
+    report_theme +
+    ggtitle("Heterogeneous Treatment \n Effect")
+
+tmp <- ggplot_gtable(ggplot_build(p1))
+leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
+mylegend <- tmp$grobs[[leg]]
+
+p3 <- grid.arrange(arrangeGrob(p1 + theme(legend.position="none"),
+                         p2 + theme(legend.position="none"),
+                         nrow=1),
+             mylegend, nrow=2,heights=c(10, 2))
+
+p3
+ggsave("../ms/fig/discrete_outcomes_simulation_power.pdf", p3, device="pdf", width=8)
 
 ```
-```{r ln_homogeneous}
-load("continuous_outcomes/lognormal_homogeneous.Rda")
-
-```
-```{r ln_heterogeneous}
-load("continuous_outcomes/lognormal_heterogeneous.Rda")
-```
-```{r exponential_homogeneous}
-load("continuous_outcomes/exponential_homogeneous.Rda")
-```
-```{r exponential_heterogeneous}
-load("continuous_outcomes/exponential_heterogeneous.Rda")
-
-```
-
-```{r power_summary, results = "asis"}
-powers <- list(gaussian_homogeneous,
-               t_homogeneous,
-               lognormal_homogeneous,
-               exponential_homogeneous,
-               gaussian_heterogeneous,
-               t_heterogeneous,
-               lognormal_heterogeneous,
-               exponential_heterogeneous)
-summary05 <- t(sapply(powers, function(x) x[5,]))
-summary05 <- data.frame(rep(c("Normal", "t(2)", "Log Normal", "Exponential"), 2), summary05)
-colnames(summary05) <-  c("Errors", "ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane") 
-```
-
-```{r power_summary_const_tab, results = "asis"}
-summarytab_constant <- xtable(summary05[1:4,], digits = 3,
-       caption = "Empirical power at level $0.05$ for simulated data with constant additive treatment effects",
-       label = "tab:power_grid1") 
-print(summarytab_constant, include.rownames=FALSE) 
-```
-
-```{r power_summary_het_tab, results = "asis"}
-summarytab_het <- xtable(summary05[5:8, ], digits = 3,
-       caption = "Empirical power at level $0.05$ for simulated data with heterogeneous treatment effects",
-       label = "tab:power_grid2") 
-print(summarytab_het, include.rownames=FALSE) 
-```

### 2017-09-13 simulations/discrete_outcomes/vary_gamma.R

@@ -1,0 +1,27 @@
+source("simulation_utils.R")
+library(doParallel)
+
+registerDoParallel(cores=30)
+
+gamma_vec <- seq(0, 0.2, by=0.05)
+gamma_res <- foreach(i = seq_along(gamma_vec)) %dopar% {
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "same effect", errors = "normal")
+  pvalues <- replicate(10000, {
+    tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+    tmp$Y1 <-gen_y(gamma = gamma_vec[i], v=tmp$v, error=tmp$delta, tmp$Z)
+    generate_simulated_pvalues(tmp, reps=1e3)
+  })
+  return(pvalues)
+}
+
+gamma_res <- lapply(gamma_res, t)
+
+tr_effect <- rep(0, 5)
+for( i in seq_along(gamma_vec)){
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "same effect", errors = "normal")
+  tr_effect[i] <- mean(gamma_vec[i] * exp(tmp$v))
+}
+
+save(tr_effect, gamma_res, file="vary_gamma_results.Rda")

### 2017-09-13 simulations/discrete_outcomes/vary_gamma_het.R

@@ -1,0 +1,27 @@
+source("simulation_utils.R")
+library(doParallel)
+
+registerDoParallel(cores=30)
+
+gamma_vec <- seq(0, 0.2, by=0.05)
+gamma_res <- foreach(i = seq_along(gamma_vec)) %dopar% {
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "heterogeneous", errors = "normal")
+  pvalues <- replicate(10000, {
+    tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+    tmp$Y1 <-gen_y(gamma = gamma_vec[i], v=tmp$v, error=tmp$delta, tmp$Z)
+    generate_simulated_pvalues(tmp, reps=1e3)
+  })
+  return(pvalues)
+}
+
+gamma_res_het <- lapply(gamma_res, t)
+
+tr_effect_het <- rep(0, 5)
+for( i in seq_along(gamma_vec)){
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "same effect", errors = "normal")
+  tr_effect_het[i] <- mean(gamma_vec[i] * exp(tmp$v))
+}
+
+save(tr_effect_het, gamma_res_het, file="vary_gamma_het_results.Rda")

### 2017-09-13 ms/ms.tex
Rewrite simulations section and misc ms edits

@@
  \usepackage{multirow}
+
+\usepackage{adjustbox} 
+\usepackage{array}
+\newcolumntype{L}{>{\centering\arraybackslash}m{3cm}}
+\setlength{\tabcolsep}{3pt}     % General space between columns (6pt standard)
+\renewcommand{\arraystretch}{1} % General space between rows (1 standard)
+
 \newcommand{\todo}[1]{{\color{red}{TO DO: \sc #1}}}
@@
 We focus on ANCOVA and its permutation counterparts, comparing their performance in different scenarios and illustrating their application with a clinical dataset.
-Section~\ref{sec:methods} describes the tests and compares their assumptions.
-Section~\ref{sec:simulations} presents simulations that suggest that even when assumptions are not satisfied, the parametric and permutation tests have comparable power to detect a treatment effect.
+Section~\ref{sec:methods} introduces the potential outcomes model, shows how this model is inconsistent with the assumptions of the parametric ANCOVA, and describes permutation tests whose assumptions match this model.
+Section~\ref{sec:simulations} presents simulations that suggest that even in this potential outcomes framework, the parametric and permutation tests have comparable power to detect a treatment effect.
 In Section~\ref{sec:results}, we apply each of the tests to data from a clinical trial comparing the performance of two treatments for gastroesophageal reflux disease (GERD).
@@
 We can never observe both; random assignment of treatment reveals $Y_{ij} = Z_{ij}Y_{ij}(1) + (1-Z_{ij})Y_{ij}(0)$.
+The potential outcomes are fixed, but the observed outcome $Y_{ij}$ is random.
 Throughout, we assume that there is no interference between individuals 
@@
 follows the Student $t$ distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
-The $p$-value for this hypothesis test is the probability, assuming the null hypothesis of zero coefficient is true, that a value drawn from the $t$ distribution is larger in magnitude than $T$.
+The $p$-value for this hypothesis test is the probability, assuming the null hypothesis of zero coefficient is true, that a value drawn from the $t$ distribution is larger in magnitude than the observed $T$.
 This test is equivalent to the $F$-test.
@@
 One way, suggested by \cite{freedman_nonstochastic_1983}, is to visually inspect a scatterplot of the residuals against $X$.
-One may also run a hypothesis test for association, such as a stratified permutation test using the correlation as a statistic.
-\todo{Do this test in the empirical section}
+One may also run a hypothesis test for association, such as a stratified permutation test using the correlation as a statistic, independently within each stratum and use the nonparametric combination of tests to obtain a single $p$-value (\cite{pesarin_permutation_2010}).
 
@@
 
-\todo{insert figures and table}
+\begin{figure}[h]
+\centering
+\includegraphics[width = \textwidth]{fig/continuous_outcomes_simulation_power.pdf}
+\caption{Empirical power curves for the continuous simulated data.
+The left panel shows power when the treatment effect was constant across strata and the right panel shows power when the treatment effect was heterogeneous.}
+\label{fig:continuous_outcomes_sim_power}
+\end{figure}
 
-\todo{what's the correlation between $X$ and $Y$?}
+First, we let the $v_{ij}$ have the same distribution across strata, generated normally distributed errors, and varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
+In the random populations that were generated, this corresponded to population average treatment effects of $0, 0.34, 0.68, 1.02$, and $1.36$ respectively.
+The left panel of Figure~\ref{fig:continuous_outcomes_sim_power} shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
+When the effect size was zero, all five tests had the correct level of 5\%.
+Power for all tests increased with the effect size, though the differenced permutation test lagged behind the rest.
+This is unsurprising, as the correlation between baseline $X$ and outcome $Y$ was low:
+it was highest at $0.33$ when $\gamma=0$ and decreased to approximately $0$ when $\gamma=0.2$.
+The parametric and permutation linear model tests perform comparably and are plotted on top of each other, while the stratified permutation test has slightly more power.
+
+Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
+Once again, we varied $\gamma$ and the corresponding population average treatment effects.
+The right panel of Figure~\ref{fig:continuous_outcomes_sim_power} shows the empirical power at level 5\% for these increasing effect sizes.
+Four of the five tests had the correct level of 5\%, while the stratified permutation test rejected only 1\% of the time.
+Both the stratified and differenced permutation tests had lower power than the linear model tests as the effect size increased.
+The correlation between baseline $X$ and outcome $Y$ was low but nontrivial (between $0.04$ and $0.44$), so this result makes some sense:
+it is beneficial for power to control for the baseline in the linear model, but the correlation is too low to make the differencing method useful.
+Finally, we varied the distributions of the errors $\varepsilon$ and $\delta$ and examined the power of the five tests.
+Recall that these are not errors in the linear model framework, but more like disturbances on the potential outcomes and baseline measures for each individual.
+These disturbances affect the distributions of $X$ and $Y$, which may impact the power of the parametric ANCOVA, and affect the variances of $X$ and $Y$, which may make one method of controlling for the baseline more effective than another.
+We fixed $\gamma=0.2$ and let the errors have a standard normal distribution, a $t$ distribution with 2 degrees of freedom, a standard log normal distribution, or an exponential distribution with parameter $1$, shifted to have mean zero.
+
+The top panel in Table~\ref{tab:power_grid1} shows the empirical power for each of these error distributions and tests, where the $v_{ij}$ came from a single distribution across strata.
+In all cases, the correlation between baseline and outcome was lower than 0.05 in magnitude.
+Thus, the stratified permutation test, which omits the baseline measurement, had the highest power for all error distributions.
+All three linear model based tests had roughly the same rejection rate.
+The differenced permutation test had substantially less power than the other methods, due to the low correlation between baseline and outcome.
+Power was highest for normally distributed errors and was lowest for $t$ distributed errors.
+
+The bottom panel of Table~\ref{tab:power_grid1} shows the results when the distribution of $v_{ij}$ varied across strata.
+In this case, the three linear model based tests had the highest power for each error distribution.
+While the correlation between baseline and outcome was still nearly zero, controlling for it in the linear model tended to increase precision.
+The stratified permutation test lost power for this reason.
+The differenced permutation test continued to have low power.
+Interestingly, when the errors were $t$ or log normally distributed, the power of each test did not change much whether the treatment effects were constant or heterogeneous across strata.
+However, power was lower for normally distributed errors when treatment effects varied across strata.
+In this scenario, the linear model does not fully describe the relationship between baseline and outcome: 
+interaction terms between baseline and stratum ID are needed in the model to capture this variation across strata.
+
+\begin{table}[ht]
+\centering
+\begin{adjustbox}{width=\textwidth}
+\begin{tabular}{L | l |LLLLL}
+  \hline
+Treatment Effect & Errors & ANCOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
+  \hline
+\multirow{4}{*}{Constant} & Normal & 0.835 & 0.874 & 0.596 & 0.842 & 0.841 \\ 
+ & t(2) & 0.286 & 0.316 & 0.146 & 0.292 & 0.291 \\ 
+ & Log Normal & 0.450 & 0.472 & 0.178 & 0.463 & 0.465 \\ 
+ & Exponential & 0.540 & 0.588 & 0.468 & 0.552 & 0.552 \\ 
+   \hline
+   \hline
+\multirow{4}{*}{Heterogeneous} & Normal & 0.658 & 0.548 & 0.317 & 0.669 & 0.669 \\ 
+  & t(2) & 0.359 & 0.315 & 0.139 & 0.365 & 0.362 \\ 
+  & Log Normal & 0.487 & 0.361 & 0.155 & 0.507 & 0.509 \\ 
+  & Exponential & 0.631 & 0.543 & 0.407 & 0.636 & 0.633 \\ 
+   \hline
+\end{tabular}
+\end{adjustbox}
+\caption{Empirical power at level $0.05$ for simulated data with constant additive treatment effects (top panel) and heterogeneous treatment effects (bottom panel).} 
+\label{tab:power_grid1}
+\end{table}
+
 \subsection{Discrete outcomes}
@@
 We used the same distributions for the latent variables $v$.
-We drew $\varepsilon$ and $\delta$ from independent standard normal distributions.
+We drew $\varepsilon$ and $\delta$ from independent standard normal distributions, and again we varied $\gamma$ and the population average treatment effect in the same way.
+
+The left panel of Figure~\ref{fig:discrete_outcomes_sim_power} shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes when treatment effects were constant.
+When the effect size was zero, all five tests had the correct level of 5\%, though the stratified and differenced permutation tests rejected 3.5\% of the time.
+The differenced permutation test had substantially lower power than the rest,
+again because the correlation between baseline $X$ and outcome $Y$ was low.
+In this case, the unadjusted stratified test had comparable power to the linear model based tests.
+The right panel of Figure~\ref{fig:discrete_outcomes_sim_power} shows the empirical power when treatment effects varied across strata.
+The patterns are similar to when outcomes were continuous.
+It is interesting to note that for the discrete outcomes, all of the tests had more power when treatment effects were heterogeneous.
+\begin{figure}[h]
+\centering
+\includegraphics[width = \textwidth]{fig/discrete_outcomes_simulation_power.pdf}
+\caption{Empirical power curves for the discrete simulated data.
+The left panel shows power when the treatment effect was constant across strata and the right panel shows power when the treatment effect was heterogeneous.}
+\label{fig:discrete_outcomes_sim_power}
+\end{figure}
 
@@
 This model allowed the intercept $\alpha_j$ to vary across sites and used the pretreatment, baseline measurement as the control variable $X$.
-The outcome and baseline had a low correlation (for instance, the correlation between pre- and post-treatment daily ``hrdq'' was $0.56$ \todo{were other outcomes even lower?
-the cutoff for correlation was 0.5}), so we used raw outcomes and not the change scores as the dependent variable.
-As our simulations and previous work (\cite{frison_repeated_1992}) show, when the correlation between baseline and outcome is low, it is more powerful to control for baseline outcomes using a model.
+The outcome and baseline had a moderate correlation (for instance, the correlation ranged from $0.56$ for daily ``hrdq'' and $0.70$ for heartburn frequency).
+\cite{frison_repeated_1992} recommend using change scores only if the correlation is at least $0.5$.
+However, we choose to include the baseline as a covariate in the linear regression rather than using change scores as the dependent variables for illustration and to be consistent with the original analysis of the data.
 
@@
 The distribution of outcomes for each endpoint is extremely right skewed, calling the assumption of normally distributed error terms into question.
+\todo{discuss other assumptions}
 
@@
 The ANCOVA $p$-values tend to be smaller than the permutation linear model and residual permutation test $p$-values.
-The permutation tests are somewhat more conservative, as they give higher $p$-values in Table~\ref{tab:clinical_pvalues}, accounting for the highly skewed outcome distributions.
-\todo{is this a correct statement?
-conservative means reject less often}
+This may be due to the skewed outcome distributions: the parametric test is more sensitive to large values than the permutation tests.
 Our conclusions for the heartburn frequency, daily heartburn, and daily ``hrdq'' endpoints differ between ANCOVA and the permutation tests at significance level $0.05$, but not at level $0.1$.
@@
 \label{tab:clinical_pvalues}
-\begin{tabular}{l|cccc}
-Endpoint      & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
+\begin{tabular}{l|LLLL}
+Endpoint      & ANCOVA & Stratified Permutation & LM Permutation & Freedman-Lane \\ \hline
 heart\_freq   & 0.035                      & 0.006                                                                                & 0.080                                                                        & 0.082                                                                        \\

### 2017-10-02 ms/ms.tex
More edits to MS addressing reviewer comments

@@
 
-\usepackage{amsmath}
+\usepackage{amsmath, amsthm}
 \usepackage{graphicx}
@@
 We compare analysis of covariance (ANCOVA), a special case of linear regression that incorporates stratification, to several permutation tests based on linear models that control for pretreatment covariates.
-In simulations using a variety of data-generating processes, some of which violate the parametric assumptions,
+In several simulations of randomized experiments using a model in which certain regression assumptions are violated,
 the permutation tests maintain power comparable to ANCOVA.
-We illustrate the use of these permutation tests alongside ANCOVA with data from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease.
-Given the considerable costs and scientific importance of clinical trials, one may want to include an additional nonparametric method, such as a linear model permutation test, as a robustness check on the statistical inference for the main study endpoints.
+We illustrate the use of these permutation tests alongside ANCOVA using data from a clinical trial comparing the effectiveness of two treatments for gastroesophageal reflux disease.
+Given the considerable costs and scientific importance of clinical trials, an additional nonparametric method, such as a linear model permutation test, may serve as a robustness check on the statistical inference for the main study endpoints.
 \end{abstract}
@@
 Medical experiments use hypothesis testing to assess the evidence that a treatment affects one or more clinically relevant outcomes.
-The simplest version of this experiment involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome, and has been studied for nearly a century (see \cite{fisher_design_1935} and \cite[1990 translation]{neyman_application_1923} for early references).
+The simplest version of this experiment has been studied for nearly a century (see \cite{fisher_design_1935} and \cite[1990 translation]{neyman_application_1923} for early references).
+This experiment involves randomly assigning two treatments to a fixed number of individuals in a group and measuring a single outcome.
 One can conduct hypothesis tests and construct confidence intervals for the estimated treatment effect
@@
 
-Randomized experiments in the real world are rarely this simple:
-if individuals are heterogeneous before the study, then their outcomes may differ for reasons besides the treatment;
-there may be more than two treatments under study;
-patients may not take the treatment they are assigned or may drop out of the study;
-and peer effects, the influence of other members of a treatment group, may affect outcomes beyond the treatment itself.
-Techniques have been developed to deal with each of these issues, invoking additional assumptions to make valid inference possible.
-In this paper, we focus on the issue of heterogeneous pretreatment characteristics and assume that the other issues are not present.
+%Randomized experiments in the real world are rarely this simple:
+%if individuals are heterogeneous before the study, then their outcomes may differ for reasons besides the treatment;
+%there may be more than two treatments under study;
+%patients may not take the treatment they are assigned or may drop out of the study;
+%and peer effects, the influence of other members of a treatment group, may affect outcomes beyond the treatment itself.
+%Techniques have been developed to deal with each of these issues, invoking additional assumptions to make valid inference possible.
+%In this paper, we focus on the issue of heterogeneous pretreatment characteristics and assume that the other issues are not present.
 
-Random assignment of treatment ensures that pretreatment covariates are balanced between treatment groups on average, across all possible randomizations.
+The goal of a randomized experiment is to draw causal inferences about the efficacy of a treatment.
+Thus, it makes sense to analyze experiments in the context of potential outcomes, where every individual has a counterfactual outcome for each of the treatments (\cite{holland_1986_statistics})
+Experiments are an attempt to estimate the counterfactual outcomes by randomly assigning treatments.
+Random assignment of treatment ensures that pretreatment covariates are balanced between treatment groups on average, across all possible randomizations, making treatment groups comparable to each other.
+
 However, in any particular randomization, there may be imbalances.
@@
 A common stratification variable in clinical experiments is location:
-individuals often come from many sites because it is difficult to recruit a sufficient number of participants at one doctor's office or hospital,
+individuals often come from many locations because it is difficult to recruit a sufficient number of participants at one site,
 especially when the object of study is a rare disease or a rare outcome.
-Stratification can have the same beneficial effects as controlling for other pretreatment covariates, 
-as outcomes may vary across strata due to factors unrelated to individuals' treatment and covariates.
 
-Linear regression is another way to control for imbalanced baseline covariates.
+Linear regression is another method to control for imbalanced baseline covariates.
 It is done during the data analysis stage.
@@
 if the true relationship is not linear, then linear regression gives the best linear approximation to the conditional expectation of the outcome.
-It is standard to use analysis of covariance (ANCOVA) to account for stratification in a linear model.
+It is standard to use analysis of covariance (ANCOVA) to incorporate stratification in a linear model.
 ANCOVA is a particular case of linear regression that allows the mean outcome to vary from stratum to stratum.
@@
 
-Hypothesis testing of estimated coefficients requires even stronger assumptions than linearity, namely Gaussian, homoskedastic errors.
-When this holds, a regression coefficient scaled by its estimated variance follows Student's $t$ distribution.
-Thus, if we believe the linear model is correctly specified, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the ANCOVA model, and can be evaluated analytically and efficiently.
-However, the modeling assumptions are clearly violated in many cases:
-when the data are discrete or ordinal, 
-when the treatment has a differential effect across subgroups of individuals and no interaction terms are included, 
-and when the variance of outcomes differs across strata.
-Furthermore, the hypothesis test implicitly assumes that the data were sampled at random from some underlying population, when in fact,
-medical experimenters rarely recruit patients this way (\cite{ludbrook_why_1998}).
+Hypothesis testing of estimated coefficients requires even stronger assumptions which are not guaranteed by the experimental design.
+When the assumptions hold, a hypothesis test for a treatment effect amounts to a hypothesis test of the coefficient for treatment in the ANCOVA model, and can be evaluated analytically and efficiently.
+A fully saturated linear model can yield asymptotically consistent estimates and confidence intervals (\cite{lin_agnostic_2013}).
+This is especially problematic in medical trials, where conditions are not always ideal for the linear model to work well:
+linear models can have substantial bias in small samples (\cite{freedman_regression_2008}),
+outcomes are often discrete or ordinal, 
+the treatment may have a differential effect across subgroups of individuals, 
+and when the distributions of outcomes may differ across strata.
+%Furthermore, the hypothesis test implicitly assumes that the data were sampled at random from some underlying population, when in fact,
+%medical experimenters rarely recruit patients this way (\cite{ludbrook_why_1998}).
 
@@
 Individuals are grouped into strata indexed by $j = 1, \dots, J$, with $n_j$ individuals in stratum $j$ and $\sum_{j=1}^J n_j = N$.
-Of the $n_j$ subjects in stratum $j$, $n_{tj}$ are assigned treatment 1, while the remaining $n_j - n_{tj}$ are assigned treatment 0.
+Of the $n_j$ subjects in stratum $j$, $n_j^T$ are assigned treatment 1, while the remaining $n_j - n_j^T$ are assigned treatment 0.
 Let $Z_{ij}$ indicate the treatment assigned to individual $i$ in stratum $j$.
@@
 For expository clarity, we suppose that $X$ is univariate, but all results are easily extended to the case when $X$ is multivariate.
-Suppose further that there is a categorical pretreatment variable with $J$ levels used to stratify individuals.
+$X$ may be associated with stratum membership.
 
 We are interested in the effect of treatment, measured as differences in potential outcomes $Y_{ij}(1) - Y_{ij}(0)$.
-We may never learn this difference for any particular individual.
-However, a tractable problem is to estimate the mean difference in some group, such as the study sample or in a target population.
-The chosen method of analysis determines which function of differences is considered.
-Various functions of potential outcomes may be of clinical interest; which one prefers depends on the goal of the study.
+We can never learn this difference for any particular individual.
+However, a tractable problem is to estimate the mean difference in the study sample or in a target population.
+Various functions of potential outcomes may be of clinical interest; the goal of the study and the method of analysis determine which function is considered.
 
 We study hypothesis testing for whether these differences are nonzero using parametric ANCOVA and its permutation counterparts, assuming this potential outcomes framework throughout.
-Other valid methods for comparing two groups include using a two-sample $t$-test to test the difference in two means from normal distributions,
+Other valid methods for comparing two groups include using a two-sample $t$ test to test the difference in two means from normal distributions,
 the Wilcoxon rank sum test to test for differences in the medians of two independent groups, 
 and the Kolmogorov-Smirnov test and receiver operating characteristic curve analyses to test whether two groups have different distribution functions (\cite{lehmann_nonparametrics_1975,  vexler_statistical_2016}).
-We focus on testing using the linear model as this is standard in clinical trials, requires fewer distributional assumptions on the data when using the potential outcomes framework, and incorporates control variables to increase power.
+We focus on testing using the linear model as this is standard in clinical trials, 
+requires fewer distributional assumptions on the data when using the potential outcomes framework, 
+deals with averages,
+and incorporates control variables to increase power.
 
@@
 To carry out the standard parametric hypothesis test for a linear model, the following assumptions are needed (\cite{freedman_statistical_2005}):
-\begin{enumerate}
+\begin{itemize}
 \item \textbf{Linearity:} The data $Y$ are related to $X$ and $Z$ linearly.
@@
 \item \textbf{Normality:} The errors are normally distributed.
-\end{enumerate}
+\end{itemize}
 
 The coefficients are estimated using least squares (or equivalently, by maximizing the likelihood).
-The estimated coefficient $\hat{\gamma}$ is the estimated average treatment effect. 
+The coefficient $\hat{\gamma}$ is the estimated average treatment effect. 
 This procedure also yields an estimate $\hat{\sigma}_{\hat{\gamma}}^2$ of the variance of $\hat{\gamma}$.
@@
 The $p$-value for this hypothesis test is the probability, assuming the null hypothesis of zero coefficient is true, that a value drawn from the $t$ distribution is larger in magnitude than the observed $T$.
-This test is equivalent to the $F$-test.
+This test is equivalent to the $F$ test.
 When the model assumptions are true, it is uniformly most powerful.
 
-However, in the context of a randomized experiment, these assumptions will not be satisfied.
-The estimated treatment effect $\hat{\gamma}$ and nominal standard errors $\hat{\sigma}_{\hat{\gamma}}$ can be severely biased (\cite{freedman_regression_2008, lin_agnostic_2013}).
+The linear model is robust to violations of its assumptions, but theoretical guarantees tend to be asymptotic.
+When randomized experiments are analyzed in the parametric framework, which assumes that treatment assignment is fixed and the errors are random,
+the estimated treatment effect $\hat{\gamma}$ and nominal standard errors $\hat{\sigma}_{\hat{\gamma}}$ can be severely biased (\cite{freedman_regression_2008, lin_agnostic_2013}).
 \cite{lin_agnostic_2013} shows that running a regression with a full set of treatment and covariate interaction terms cannot hurt asymptotic precision, and using the Huber-White sandwich standard errors can yield asymptotically valid confidence intervals.
@@
 if the difference in potential outcomes is heterogeneous across strata, then the coefficient $\gamma$ may be attenuated towards 0.
+Thus, it is unclear for any particular dataset whether or not the ANCOVA model will give valid results.
 
 \subsection{Stratified permutation test}\label{subsec:strat_perm_test}
-In randomization inference, the only random quantities are those involving $Z$, the vector of treatment assignments.
-The exact inference is derived from the randomization distribution of statistics involving $Z$.
-In particular, the potential outcomes and covariates $(Y_{ij}(1), Y_{ij}(0), X_{ij})$ are fixed in the finite population under study, while the observed response $Y_{ij}$ changes with $Z_{ij}$.
+All permutation tests essentially have two requirements: a conditioning space, the orbit of a finite group of transformations of the data in which all configurations of the data are equiprobable under the null hypothesis, and a set of sufficient statistics for the data which describe the data under these transformations.(\cite{pesarin_permutation_2010}).
+%We describe these two elements in the context of a stratified randomized experiment.
+In experiments, the only random quantities are those involving $Z$, the vector of treatment assignments.
+In particular, the potential outcomes and covariates are fixed in the finite population under study, while the observed responses change with $Z$.
+The exact permutation inference is derived from the conditioning space which includes all possible values of $Z$.
 
@@
 
-In a randomized experiment with complete randomization, any randomization that assigns $n_j$ individuals to treatment 1 in stratum $j$ for each stratum has equal probability.
+In an experiment with complete randomization, any of the possible allocations of treatment that assign $n_j^T$ out of $n_j$ individuals to treatment 1 in stratum $j$, for each stratum $j=1,\dots,J$, has equal probability.
 We can construct the permutation distribution of any statistic under the null hypothesis by imputing the unobserved potential outcomes using the sharp null hypothesis and by re-randomizing treatment assignments using this principle of equal probabilities.
-The experimental design specifies the permutation scheme; no additional assumptions are necessary.
+The experimental design specifies the permutation scheme; no additional assumptions are needed to find the distribution of any statistic.
+In practice, the $\prod_{j=1}^J {n_j \choose n_j^T}$ possible allocations of treatment may be too numerous to compute the statistic for each one.
+Instead, we typically use Monte Carlo simulations to randomly assign treatment and compute the statistic of interest a large number of times to approximate the randomization distribution to a desired degree of accuracy.
+
+To summarize, the assumptions for the stratified permutation test are the following.
+\begin{itemize}
+\item \textbf{Complete randomization:} the $\prod_{j=1}^J {n_j \choose n_j^T}$ assignments of treatment which assign $n_j^T$ out of $n_j$ individuals to treatment 1 in stratum $j$, for each stratum $j=1,\dots,J$, has equal probability
+\item \textbf{No treatment effect under the null:} both potential outcomes are identical and equal to the observed outcome for each individual $(i, j)$
+\end{itemize}
 
@@
 For NPC, one first conducts each partial test separately, then combines their $p$-values in a way that preserves dependencies.
-This method is equivalent to combining stratum statistics directly when data are from a randomized experiment, because treatment is independent across strata.
+For a randomized experiment, this method is equivalent to combining stratum statistics directly, because treatment is independently assigned across strata.
 
 \subsection{Permutation tests with the linear model}\label{subsec:lm_perm_tests}
+
 The tests in Section~\ref{subsec:strat_perm_test} only use information on the treatment and the outcome.
-However, experimenters record additional covariates that may be predictive of the outcome.
-A test that incorporates these covariates may reduce the variance of the statistic and have more power.
-The permutation tests in this section use the linear model in Equation~\ref{eqn:ancova}, but make different assumptions about the data.
+However, experimenters typically record additional covariates that may be predictive of the outcome.
+Incorporating these covariates can reduce variation in the statistic due to factors unrelated to treatment, thereby increasing power to detect a treatment effect.
+The permutation tests in this section use the linear model in Equation~\ref{eqn:ancova} to control for covariates, but make different assumptions about the data.
 In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes,
@@
 We show two permutation tests developed in this framework.
+They test the same ``sharp'' null hypothesis of no treatment effect.
 
 The first linear model permutation test is a variation on the stratified permutation test described in Section~\ref{subsec:strat_perm_test}.
+The assumptions are the same.
 The only difference is the test statistic:
 this test uses the $t$ statistic for the regression coefficient of $Z$ in the linear model in Equation~\ref{eqn:ancova}.
-Rather than using the $t$ distribution to calculate the significance of the observed $t$ statistic, we construct the permutation distribution of the $t$ statistic.
+Rather than using the $t$ distribution to calculate the significance of the observed $t$ statistic, we construct the permutation distribution of the $t$ statistic by simulation.
 Again, the experimental design specifies the randomization distribution and no additional assumptions are needed:
@@
 Furthermore, under the null hypothesis, the $\varepsilon$ are independent of $Z$ within strata.
-Thus, we can carry out an approximate permutation test by permuting the estimated $\hat{\varepsilon}$ within strata, independently across strata.
 
-The test has two important assumptions:
-the relationship between $Y$ and $X$ is linear and errors are IID within strata.
-Together, these conditions ensure that $\varepsilon$ are exchangeable against $X$ within strata.
-In practice, we must estimate $\alpha$; using $\hat{\alpha}$ to estimate $\hat{Y}$ yields an approximate permutation test.
+We can carry out an approximate permutation test by permuting the estimated $\hat{\varepsilon}$ within strata, independently across strata.
+The test is only approximate because we must estimate $\alpha$ and $\beta$ to construct $\hat{\varepsilon}$.
+The test has two important assumptions pertaining specifically to the linear regression, on top of the two necessary pieces of a permutation test:
+\begin{itemize}
+\item \textbf{Linear model:} the linear model that describes $Y$ using $X$ and the approximation error $\varepsilon$ holds
+\item \textbf{Linear regression fit:} the linear regression residuals $\hat{\varepsilon}$ are a good approximation to the true errors $\varepsilon$ 
+\item \textbf{Conditioning space:} the $\varepsilon$ are exchangeable against $X$ within strata
+\item \textbf{No treatment effect under the null:} the $\varepsilon$ do not depend on $Z$
+\end{itemize}
+
 The $\hat{\varepsilon}$ may not even be approximately exchangeable within strata if the regression does not describe the relation between $Y$ and $X$ well.
 For this reason, \cite{freedman_nonstochastic_1983} advise against using the method if there are large outlier values in the covariate $X$ or when $X$ and $Z$ are highly colinear.
-Randomization does not guarantee this additional exchangeability assumption, so it should be checked.
+Randomization does not guarantee the exchangeability against $X$, so this assumption should be checked.
 One way, suggested by \cite{freedman_nonstochastic_1983}, is to visually inspect a scatterplot of the residuals against $X$.
-One may also run a hypothesis test for association, such as a stratified permutation test using the correlation as a statistic, independently within each stratum and use the nonparametric combination of tests to obtain a single $p$-value (\cite{pesarin_permutation_2010}).
+One may also test for association between $X$ and $\hat{\varepsilon}$, such as a stratified permutation test using the correlation as a statistic, independently within each stratum and use the nonparametric combination of tests to obtain a single $p$-value (\cite{pesarin_permutation_2010}).
 
@@
 We simulated data from a randomized experiment using several different data-generating processes.
-We applied the tests to the data and compared their empirical power over repeated random treatment assignments and random errors.
+The variables included stratification, two randomly assigned treatments, a baseline measure before treatment, and the same variable measured after treatment.
+We tested for a treatment effect and compared the empirical power of the tests over repeated random treatment assignments.
 We compared the following five tests:
@@
  (called ``stratified permutation'' in what follows),
-a stratified permutation test using the mean change scores, defined as the difference between the baseline measure and the outcome (called ``differenced permutation'' in what follows),
+a stratified permutation test using the mean change scores, defined as the difference between the outcome and baseline measure (called ``differenced permutation'' in what follows),
 a stratified permutation test based on the $t$ statistic from the linear regression of outcome on control variables (called ``LM permutation'' in what follows),
 and the Freedman-Lane permutation test.
-The simulations are included in three supplementary files.
+The simulation code is included in four supplementary files.
 
@@
 
-In the first set of simulations, we generated continuous potential outcomes.
+In the first set of simulations, we generated continuous potential outcomes based on a latent variable and disturbance terms, both of which we manipulated.
 We used two scenarios.
-In the first, the treatment effect was constant across strata.
+In the first scenario, the treatment effect was constant across strata.
 We drew a latent random variate $v_{ij}$ from the uniform distribution on $[-4, 4]$.
@@
 
-The baseline value for individual $i,j$ was 
-$$ X_{ij} = \frac{\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}$$.
+We included $n_j=16$ individuals per stratum and treatment assignment was balanced, i.e. 8 people received each treatment in each stratum.
+After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we constructed the baseline value for individual $i,j$ as 
+$$ X_{ij} = \frac{-\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}.$$
 
+\noindent $\gamma$ was a fixed constant.
 Then we generated potential outcomes as
@@
 
-We included $n_j=16$ individuals per stratum and treatment assignment was balanced, i.e. 8 people received each treatment in each stratum.
-After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerated $Z$ and the potential outcomes $Y(Z)$ 10,000 times.
+\noindent This is equivalent to the following linear data-generating process:
+
+$$Y_{ij}(Z_{ij}) = Z_{ij} \gamma e^{v_{ij}} + X_{ij} + (\delta_{ij} - \varepsilon_{ij}).$$
+
+\noindent This model does not exactly match the ANCOVA Equation~\ref{eqn:ancova} and only some of the assumptions are satisfied:
+$Y$ is linear in $(X, Z)$ conditionally on the latent variable $v$, 
+the errors are independent and identically distributed, and are independent of $X$.
+When the latent variables vary in distribution across strata, the assumption of constant slopes is violated.
+
+We regenerated $Z$ and the potential outcomes $Y(Z)$ 10,000 times.
 We repeated this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.
-
 The treatment effect for individual $(i, j)$ was $\gamma e^{v_{ij}}$.
@@
 \includegraphics[width = \textwidth]{fig/continuous_outcomes_simulation_power.pdf}
-\caption{Empirical power curves for the continuous simulated data.
+\caption{Empirical power curves for the continuous simulated data with normally distributed $\varepsilon$ and $\delta$.
 The left panel shows power when the treatment effect was constant across strata and the right panel shows power when the treatment effect was heterogeneous.}
@@
 
-First, we let the $v_{ij}$ have the same distribution across strata, generated normally distributed errors, and varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
+First, we studied how power changed with the average treatment effect.
+We used normally distributed errors and varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
 In the random populations that were generated, this corresponded to population average treatment effects of $0, 0.34, 0.68, 1.02$, and $1.36$ respectively.
 The left panel of Figure~\ref{fig:continuous_outcomes_sim_power} shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
-When the effect size was zero, all five tests had the correct level of 5\%.
-Power for all tests increased with the effect size, though the differenced permutation test lagged behind the rest.
+When the treatment effect was homogeneous across strata, all five tests had the correct level of 5\% (i.e. the power when $\gamma = 0$).
+Power of the differenced permutation test increased at a slower rate than the other tests.
 This is unsurprising, as the correlation between baseline $X$ and outcome $Y$ was low:
-it was highest at $0.33$ when $\gamma=0$ and decreased to approximately $0$ when $\gamma=0.2$.
-The parametric and permutation linear model tests perform comparably and are plotted on top of each other, while the stratified permutation test has slightly more power.
+its highest value was $0.33$ when $\gamma=0$ and it decreased to nearly $0$ when $\gamma=0.2$.
+The parametric and permutation linear model tests performed comparably and are plotted on top of each other, while the stratified permutation test had slightly more power.
 
-Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
-Once again, we varied $\gamma$ and the corresponding population average treatment effects.
-The right panel of Figure~\ref{fig:continuous_outcomes_sim_power} shows the empirical power at level 5\% for these increasing effect sizes.
+The right panel of Figure~\ref{fig:continuous_outcomes_sim_power} shows the empirical power at level 5\% for the same experiment, with heterogeneous treatment effects across strata.
 Four of the five tests had the correct level of 5\%, while the stratified permutation test rejected only 1\% of the time.
@@
 it is beneficial for power to control for the baseline in the linear model, but the correlation is too low to make the differencing method useful.
+
 Finally, we varied the distributions of the errors $\varepsilon$ and $\delta$ and examined the power of the five tests.
-Recall that these are not errors in the linear model framework, but more like disturbances on the potential outcomes and baseline measures for each individual.
-These disturbances affect the distributions of $X$ and $Y$, which may impact the power of the parametric ANCOVA, and affect the variances of $X$ and $Y$, which may make one method of controlling for the baseline more effective than another.
-We fixed $\gamma=0.2$ and let the errors have a standard normal distribution, a $t$ distribution with 2 degrees of freedom, a standard log normal distribution, or an exponential distribution with parameter $1$, shifted to have mean zero.
+Recall that these are not errors in the linear model framework, but rather disturbances on the potential outcomes and baseline measures for each individual.
+These disturbances affect the distributions of $X$ and $Y$, which may impact the power of the parametric ANCOVA and may make one method of controlling for the baseline more effective than another.
+We fixed $\gamma=0.2$ and varied the distributions of $\varepsilon$ and $\delta$.
 
@@
 While the correlation between baseline and outcome was still nearly zero, controlling for it in the linear model tended to increase precision.
-The stratified permutation test lost power for this reason.
+Conversely, the stratified permutation test lost power for this reason.
 The differenced permutation test continued to have low power.
@@
 
-Many outcomes are measured on a discrete, ordinal scale but the same regression-based methods are used to test for a nonzero treatment effect.
-We repeated the procedure above, instead discretizing the covariate $X$ and potential outcomes $Y(Z)$ by taking their floor.
+Many medical outcomes are measured on a discrete, ordinal scale, but the same regression-based methods designed for continuous data are used to test for a nonzero treatment effect.
+These data do not satisfy the ANCOVA assumptions of linearity and of normally distributed errors.
+We repeated the procedure above, instead discretizing the covariate $X$ and potential outcomes $Y(Z)$ by removing their fractional parts.
 We used the same distributions for the latent variables $v$.
@@
 \subsection{Imbalanced treatment groups}
-%Outcomes were generated according to the linear equation 
-%\begin{equation}\label{eqn:dgp}
-%Y_{ij1} =\alpha_j + \beta_0Y_{ij0} + \gamma_j Z_{ij} + \varepsilon_{ij}
-%\end{equation}
-%
-%\noindent for individuals $i = 1, \dots, n_j$, $j = 1, 2, 3$.
-%$\alpha_j$ is the mean effect of being in stratum $j$, 
-%$\beta_0$ is the coefficient for the baseline measurement $Y_{ij0}$, 
-%$Z_{ij}$ is the treatment level, 
-%$\gamma_j$ is the effect of treatment in stratum $j$, 
-%and $\varepsilon_{ij}$ is an error term.
-%We used three strata with $\alpha_1 = 1, \alpha_2 = 1.5,$ and $\alpha_3 = 2$.
-%
-%We used two designs:
-%\begin{itemize}
-%\item Constant additive treatment effect: $\gamma_1 = \gamma_2 = \gamma_3 = \gamma$. This is the implicit assumption when using a linear model.
-%\item Treatment effect in a single stratum: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$.
-This is a constant, additive treatment effect in stratum 1, but no treatment effect in strata 2 and 3.
-This is a simplistic example of a heterogeneous treatment effect.
-The standard ANCOVA model does not account for this scenario.
-%\end{itemize}
-%
-%\noindent We drew the baseline measurements once, treating them as fixed, and conditioned on observing $\{ Y_{ij0} : {i = 1,\dots,n_j, j = 1,\dots, 3}\}$.
-%We randomly assigned treatment to half of the individuals within each stratum, generated random errors, and constructed new outcomes $Y_{ij1}$ using Equation~\ref{eqn:dgp}.
-%We conducted all five tests on this new data, obtaining a two-sided $p$-value for each.
-%We repeated these steps using the same $\{ Y_{ij0} \}$ 1000 times.
-%
-%In our first set of simulations, we assumed that the baseline measurements $Y_{ij0}$ were standard normally distributed and the pairs $(Y_{ij0}, \varepsilon_{ij})$ were independent across $i$ and $j$.
-%We let $\beta_0 = 1$ and the treatment effect be $\gamma = 1$.
-%We used a balanced design with $n_j = 16$ individuals per stratum and balanced treatment assignment, i.e. 8 subjects received each treatment within each stratum.
-%There were four total simulation designs:
-%for each of the two possible treatment effects, we used two distributions for $\varepsilon$.
-%In the first case, we used $\varepsilon \sim N(0, 1)$ to match the parametric assumption of normal errors.
-%In the second case, $\varepsilon$ were drawn from a $t$ distribution with 2 degrees of freedom so the errors were heavy-tailed.
-%
-%Figure~\ref{fig:normal_sim_power} shows the estimated power curves in these four designs.
-%The best case was when the errors were Gaussian and the treatment effect was constant across strata, while the worst case was when the errors are heavy-tailed and the treatment effect only appeared in one stratum.
-%Intuitively, it makes sense that power decreased relative to the Gaussian, constant treatment effect case, as each violated assumption further obscured the treatment effect.
-%A consistent pattern appeared in each design: the stratified permutation test had the lowest power, while the other four tests tended to have similar power.
-%
-%Table~\ref{tab:normal_power} displays the size of the test and the power at level $0.05$.
-%The three tests based on linear models had slightly higher than nominal level, though the margin was within two standard errors of $0.05$.
-%(The number of tests rejected under the null in 1000 trials has a binomial distribution.
-%If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.)
-%These numbers show that actually, for $\alpha=0.05$, the stratified permutation test did not lose a large amount of power when the effect was isolated in a single stratum.
-%In this case, the stratified permutation test may have slightly higher power than ANCOVA at the small significance levels typically used in practice.
-%\begin{figure}[h]
-%\centering
-%\includegraphics[width = \textwidth]{fig/normal_simulation_power.pdf}
-%\caption{Empirical power curves for the Gaussian simulated data}
-%\label{fig:normal_sim_power}
-%\end{figure}
-%%\begin{center}
-%%\input{fig/normal_simulation_power_summary.tex}
-%%\end{center}
-%
-%
-%
-%\begin{table}[]
-%\centering
-%\begin{tabular}{ll|ccccc}
-%\multicolumn{2}{c|}{Design}                                                                  & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
-%Errors                        & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
-%\multirow{3}{*}{Gaussian}     & None                                                        & 0.058                      & 0.042                                                                                & 0.038                                                                                 & 0.062                                                                        & 0.057                                                                        \\
-%                              & Constant                                                    & 0.886                      & 0.545                                                                                 & 0.858                                                                                 & 0.891                                                                        & 0.892                                                                        \\
-%                              & Single stratum                                              & 0.159                      & 0.089                                                                                & 0.160                                                                                 & 0.162                                                                        & 0.170                                                                        \\ \hline
-%\multirow{3}{*}{Heavy-tailed} & None                                                        & 0.032                      & 0.040                                                                                & 0.045                                                                                 & 0.042                                                                        & 0.040                                                                        \\
-%                              & Constant                                                    & 0.355                      & 0.240                                                                                & 0.378                                                                                 & 0.400                                                                        & 0.393                                                                        \\
-%                              & Single stratum                                              & 0.066                      & 0.080                                                                                & 0.082                                                                                 & 0.076                                                                        & 0.080                                                                        \\ \hline
-%\end{tabular}
-%\caption{Empirical power at level $0.05$ for Gaussian simulated data with balanced treatment groups.}
-%\label{tab:normal_power}
-%\end{table}
-%
-%Our second set of simulations used the same design as the previous ones, but modified the sizes of the treatment groups.
-%Within each stratum, 4 individuals received one treatment and 12 received the other.
-%This mimics some real-world experiments, where the more expensive treatment is administered less frequently than the placebo or standard of care.
-%It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exacerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
-%Here, the data had equal variances so we would expect power, not the test level, to be an issue.
-%(See the supplementary Gaussian simulations for simulations using data with heterogeneous variances.)
-%Table~\ref{tab:imbalanced_power} displays the value of these power curves and the size of the test at level $0.05$.
-%In this case, the permutation tests (except for the simple stratified permutation) had slightly higher power than ANCOVA, but the difference was not substantial.
-%
-%%\begin{center}
-%%\input{fig/imbalanced_simulation_power_summary.tex}
-%%\end{center}
-%
-%
-%\begin{table}[ht]
-%\centering
-%\begin{tabular}{ll|ccccc}
-%\multicolumn{2}{c|}{Design}                                                                 & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
-%Errors                        & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
-%\multirow{3}{*}{Gaussian}     & None                                                        & 0.046                      & 0.031                                                                                & 0.028                                                                                 & 0.042                                                                        & 0.045                                                                        \\
-%                              & Constant                                                    & 0.840                      & 0.490                                                                                & 0.810                                                                                 & 0.870                                                                        & 0.860                                                                        \\
-%                              & Single stratum                                              & 0.120                      & 0.050                                                                                & 0.090                                                                                 & 0.100                                                                        & 0.120                                                                        \\ \hline
-%\multirow{3}{*}{Heavy-tailed} & None                                                        & 0.034                      & 0.054                                                                                & 0.055                                                                                 & 0.047                                                                        & 0.042                                                                        \\
-%                              & Constant                                                    & 0.230                      & 0.200                                                                                & 0.270                                                                                 & 0.260                                                                        & 0.270                                                                        \\
-%                              & Single stratum                                              & 0.070                      & 0.080                                                                                & 0.090                                                                                 & 0.090                                                                        & 0.080                                                                        \\ \hline
-%\end{tabular}
-%\caption{Empirical power at level $0.05$ for Gaussian simulated data with imbalanced treatment groups.} 
-%\label{tab:imbalanced_power}
-%\end{table}
-%
-%In our third set of simulations, we made the baseline measurements $Y_{ij0}$ discrete and skewed.
-%This type of data may occur in surveys where individuals are asked to rate their symptom severity on a discrete ordinal scale.
-%We generated the baseline measures using independent draws from a Poisson distribution with mean 4 and censored values at 10.
-%We conditioned on these observed baseline measures after generating them once.
-%Then, for each of 1000 trials, we randomly assigned treatment to half of the individuals in each stratum and generated errors taking on values $0.5$ and $-0.5$ with equal probability.
-%We constructed $Y_{ij1}$ using Equation~\ref{eqn:dgp} with a treatment effect of $\gamma = 0.5$.
-%To ensure that both baseline and outcome were integer scores, we let the $Y_{ij1}$ be unobservable and defined the observed outcome as
-%
-%\begin{displaymath}
-%   \tilde{Y}_{ij1} = \left\{
-%     \begin{array}{ll}
-%       1 & \text{if } {Y}_{ij1} < 1\\
-%       10 & \text{if } {Y}_{ij1} > 10 \\
-%       \lfloor {Y}_{ij1} \rfloor & \text{otherwise}
-%     \end{array}
-%   \right.
-%\end{displaymath}
-%
-%
-%The observed outcomes $\tilde{Y}_{ij1}$, the baseline measures $Y_{ij0}$, and the errors were discrete, violating normality assumptions.
-%
-%We examined the effect of stratum size in this set of simulations.
-%For each of the two treatment effect designs, we considered both equal and unequal stratum sizes.
-%The simulations with equal stratum sizes were generated as before, where each stratum contained $n_j = 16$ individuals.
-%The simulations with unequal stratum sizes were set up so the smallest stratum had only $n_1=8$ individuals, the next had $n_2= 16$, and the largest had $n_3=24$.
-%The design with a nonzero treatment effect in a single stratum and unequal stratum sizes was particularly unfavorable:
-%the nonzero effect only occurred in the smallest stratum.
-%
-%Figure~\ref{fig:skewed_sim_power} shows the estimated power curves in these four designs.
-%When the treatment effect was constant, sample sizes within each stratum didn't matter; 
-%the two power curve plots in the first row of Figure~\ref{fig:skewed_sim_power} look the same up to chance variation.
-%When the treatment effect only occurred in one stratum, there was a substantial loss in power.
-%It was nearly impossible to detect a difference in treatments when the affected stratum was too small.
-%In practice, one does not usually know a priori which individuals will be affected by treatment (otherwise, those not affected would be excluded from the study altogether).
-%This result suggests that when stratifying, one must be cautious not to stratify too finely, as it is impossible to measure an effect with an insufficient sample size.
-%
-%All of the tests appeared to be conservative and to have lower than nominal level.
-%As in the first set of simulations, the stratified permutation test using the raw outcomes had the lowest power of the five tests.
-%The stratified permutation test using the change scores also had lower power than the three linear model tests.
-%Presumably, this was because the differences $\tilde{Y}_{ij1} - Y_{ij0}$ were discrete and limited to a small number of values, so the test statistic could not vary greatly across permutations.
-%Table~\ref{tab:skewed_power} confirms this: power for the three linear model methods was high when the treatment was constant across strata, and power for all of the methods suffered when the effect was isolated in a single stratum.
-%
-%
-%\begin{figure}
-%\centering
-%\includegraphics[width = \textwidth]{fig/skewed_simulation_power.pdf}
-%\caption{Empirical power curves for the skewed, discrete simulated data}
-%\label{fig:skewed_sim_power}
-%\end{figure}
-%
-%%\begin{center}
-%%\input{fig/skewed_simulation_power_summary.tex}
-%%\end{center}
-%\begin{table}[]
-%\centering
-%\begin{tabular}{ll|ccccc}
-%\multicolumn{2}{c|}{Design}                                                               & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
-%Group sizes                 & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
-%\multirow{3}{*}{Balanced}   & None                                                        & 0.036                      & 0.025                                                                                & 0.021                                                                                 & 0.035                                                                        & 0.037                                                                        \\
-%                            & Constant                                                    & 0.974                      & 0.143                                                                                & 0.853                                                                                 & 0.972                                                                        & 0.977                                                                        \\
-%                            & Single stratum                                              & 0.422                      & 0.050                                                                                & 0.275                                                                                 & 0.429                                                                        & 0.424                                                                        \\ \hline
-%\multirow{3}{*}{Imbalanced} & None                                                        & 0.038                      & 0.024                                                                                & 0.019                                                                                 & 0.037                                                                        & 0.039                                                                        \\
-%                            & Constant                                                    & 0.971                      & 0.147                                                                                & 0.860                                                                                 & 0.970                                                                        & 0.972                                                                        \\
-%                            & Single stratum                                              & 0.190                      & 0.070                                                                                & 0.100                                                                                 & 0.190                                                                        & 0.190                                                                        \\ \hline
-%\end{tabular}
-%\caption{Empirical power at level $0.05$ for discrete, skewed simulated data} 
-%\label{tab:skewed_power}
-%\end{table}
-%It makes sense that in all of these simulations, the permutation test using the change scores as the response measure is more powerful than the permutation test using the raw outcomes.
-%The baseline and outcome were highly correlated, so the change score had lower variance than the outcome alone. 
-%In additional simulations, 
-%we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had a correlation of $0.25$.
-%In this case, the result was reversed: 
-%the differenced test had low power while the test using the raw outcomes had a power curve closer to the linear model-based tests.
-%(See the supplementary Gaussian simulations for a more detailed treatment.)
-%This behavior has been studied before by \cite{frison_repeated_1992}, who recommended using the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.
-%
+Many experiments use assign a different number of individuals to each treatment group.
+Our previous simulations assumed that the number in each treatment group was the same.
+We examined the effect of varying the proportion who received treatment in each stratum.
+The tests remain consistent no matter how many individuals are assigned to each treatment group, 
+but power may be affected.
+
+We repeated the simulations with continuous outcomes, Gaussian disturbances, and fixed $\gamma=0.2$ but varied the number of treated in each stratum to be $4$, $8$, or $12$.
+Each of the tests tended to have higher power when the total number of treated units was around half of the population (in this case, 24 out of 48).
+It is interesting to note that all tests had more power when 25\% of units were treated than when 75\% were treated; it seems that power is not a symmetric function.
+In the scenarios with a varying treatment effect, the  effect was largest in the third stratum and smallest in the first stratum.
+As expected, among experiments that fixed the total number of treated units, the experiment with more treatment units in the third stratum tended to have higher power.
+
+The differenced permutation test had substantially lower power than the rest,
+again because the correlation between baseline $X$ and outcome $Y$ was low.
+In this case, the unadjusted stratified test had comparable power to or higher than the linear model based tests.
+ANCOVA had slightly higher power than the linear model permutation tests for all experiments except for when the number of treated units was equal to number of controls and when the number of treated units was 30 or above.
+We refer the reader to the supplementary file for the details of these experiments.
+
+\subsection{Nonlinear model}
+
+The potential outcomes model we used before turned out to be a linear function of $X$ and $Z$, conditional on the latent variable $v$.
+For these simulations, we changed the outcome model to be nonlinear, violating an important ANCOVA assumption.
+Instead of an additive treatment effect, now the effect is multiplicative.
+We generated the latent variable and errors in the same way, but let
+
+\begin{align*}
+X_{ij} &=  \frac{ e^{v_{ij}} + e^{v_{ij}/2}}{2} +\varepsilon_{ij} \\
+Y_{ij}(Z_{ij}) &= (1+\gamma)^{Z_{ij}} X_{ij}  + \delta_{ij} \\
+&= \gamma X_{ij}Z_{ij} + X_{ij} + \delta_{ij}.
+\end{align*}
+
+The treatment effect for individual $(i, j)$ is thus $\gamma X_{ij}$.
+The stratified, unadjusted permutation test failed here:
+the potential outcomes depend on the baseline covariate, so units are not exchangeable without controlling for the baseline.
+The ANCOVA and permutation linear model tests had identical power, while the differenced permutation test had slightly more power.
+The patterns were the same when the latent variables $v$ had the same and different distributions across strata;
+it made little difference because the treatment effect varied within strata, anyway.
 
@@
 These endpoints were measured on a discrete scale.
-There were several additional endpoints, such as daily regurgitation, daily ``hrdq'', and daily dyspepsia, calculated from the survey measures.
+There were several additional endpoints calculated from the survey measures: daily heartburn, daily regurgitation, daily ``hrdq'' (a composite score), and daily dyspepsia.
 Daily ``hrdq'' was the primary endpoint.
 To reduce day-to-day variation, we averaged the measures from each week to obtain two observations per patient, one pre-treatment and one post-treatment.
+These averages are not quite discrete ordinal values, but they are not continuous either: they can only take a finite set of values.
 
 We used site as the stratification variable, as this is the level at which treatment was randomized.
-We did not include country in the model, as a country-level effect should be captured in the site-level effects.
 The model used for the linear regression-based tests was defined as in Equation~\ref{eqn:ancova}.
@@
 \cite{frison_repeated_1992} recommend using change scores only if the correlation is at least $0.5$.
-However, we choose to include the baseline as a covariate in the linear regression rather than using change scores as the dependent variables for illustration and to be consistent with the original analysis of the data.
+However, we choose to include the baseline as a covariate in the linear regression, rather than using change scores as the dependent variables, to illustrate the inclusion of covariates and to be consistent with the original analysis of the data.
 
-Figure~\ref{fig:clinical_distr} shows the distribution of each clinical endpoint for the two treatment groups.
-There is a clear difference in distributions for daily heartburn (``daily\_heart''), daily ``hrdq'' (``daily\_hrdq''), and heartburn frequency (``heart\_freq'').
-The difference is less clear for daily regurgitation (``daily\_regurg''), daily dyspepsia (``daily\_dysp''), regurgitation frequency (``regurg\_freq''), and dyspepsia frequency (``dysp\_freq'').
-The distribution of outcomes for each endpoint is extremely right skewed, calling the assumption of normally distributed error terms into question.
-\todo{discuss other assumptions}
+Table~\ref{tab:clinical_distr} shows the mean and standard deviation of each clinical endpoint for the two treatment groups.
+There is a large difference in means for daily heartburn (``daily\_heart''), daily ``hrdq'' (``daily\_hrdq''), heartburn frequency (``heart\_freq''), and regurgitation frequency (``regurg\_freq'').
+The difference is less clear for daily regurgitation (``daily\_regurg''), daily dyspepsia (``daily\_dysp''), and dyspepsia frequency (``dysp\_freq'').
+The distribution of outcomes for each endpoint is extremely right skewed, which may make it difficult to fit a linear model to the data.
+Residual plots indicate heteroskedastic errors, invalidating the assumption of constant error variance needed for parametric inference.
+The permutation test assumptions are guaranteed by the randomization that was conducted.
+We checked the additional exchangeability conditions for the Freedman-Lane test and found no evidence that they were violated. 
+These checks are included in the supplementary file.
 
@@
 The ANCOVA $p$-values tend to be smaller than the permutation linear model and residual permutation test $p$-values.
-This may be due to the skewed outcome distributions: the parametric test is more sensitive to large values than the permutation tests.
+This may be due to the skewed outcome distributions and heteroskedastic errors: the parametric test is more sensitive to large values than the permutation tests.
 Our conclusions for the heartburn frequency, daily heartburn, and daily ``hrdq'' endpoints differ between ANCOVA and the permutation tests at significance level $0.05$, but not at level $0.1$.
-None of the three tests based on linear models is significant at level $0.1$ for the other four endpoints;
-there is no endpoint which would be deemed significant using a permutation test but insignificant with ANCOVA.
+There is no endpoint which would be deemed significant using a permutation test but insignificant with ANCOVA.
 This suggests that the parametric test correctly discriminates between endpoints that are significantly different and endpoints that are not different between treatment A and treatment B.
+The correspondence between parametric and permutation test results gives some confidence that the results are stable and reliable.
 
-\begin{figure}
+\begin{table}\label{tab:clinical_distr}
 \centering
-\includegraphics[width = \textwidth]{fig/clinical_distr.pdf}
-\caption{Distribution of each endpoint for the two treatments A and B}
-\label{fig:clinical_distr}
-\end{figure}
-
-%
-%\begin{center}
-%\input{fig/results_pvalue_summary.tex}
-%\end{center}
+\begin{tabular}{r|cc|cc}
+  \hline
+ & Mean A & SD A & Mean B & SD B \\ 
+  \hline
+daily\_heart & 0.59 & 0.46 & 0.75 & 0.52 \\ 
+  daily\_regurg & 0.52 & 0.44 & 0.58 & 0.45 \\ 
+  daily\_dysp & 0.56 & 0.51 & 0.55 & 0.49 \\ 
+  daily\_hrdq & 1.11 & 0.84 & 1.33 & 0.86 \\ 
+  heart\_freq & 2.48 & 2.10 & 3.51 & 2.76 \\ 
+  regurg\_freq & 2.30 & 1.97 & 2.58 & 2.16 \\ 
+  dysp\_freq & 2.49 & 2.48 & 2.36 & 2.25 \\ 
+   \hline
+\end{tabular}
+\caption{Mean of each continuous outcome in groups A and B.} 
+\end{table}
 
@@
 Endpoint      & ANCOVA & Stratified Permutation & LM Permutation & Freedman-Lane \\ \hline
-heart\_freq   & 0.035                      & 0.006                                                                                & 0.080                                                                        & 0.082                                                                        \\
-regurg\_freq  & 0.136                      & 0.118                                                                                & 0.280                                                                        & 0.220                                                                        \\
-dysp\_freq    & 0.565                      & 0.948                                                                                & 0.616                                                                        & 0.592                                                                        \\
-daily\_heart & 0.032                      & 0.004                                                                                & 0.056                                                                        & 0.068                                                                        \\
-daily\_regur  & 0.142                      & 0.174                                                                                & 0.286                                                                        & 0.246                                                                        \\
-daily\_hrdq   & 0.043                      & 0.012                                                                                & 0.088                                                                        & 0.098                                                                        \\
-daily\_dysp   & 0.582                      & 0.810                                                                                & 0.756                                                                        & 0.722                                                                        \\
+heart\_freq   & 0.035                      & 0.003                                                                                & 0.070                                                                        & 0.074                                                                       \\
+regurg\_freq  & 0.136                      & 0.157                                                                                & 0.231                                                                        & 0.221                                                                        \\
+dysp\_freq    & 0.565                      & 0.925                                                                               & 0.592                                                                        & 0.579                                                                        \\
+daily\_heart & 0.032                      & 0.010                                                                                & 0.062                                                                        & 0.069                                                                        \\
+daily\_regur  & 0.142                      & 0.195                                                                                & 0.250                                                                        & 0.243                                                                        \\
+daily\_hrdq   & 0.043                      & 0.033                                                                                & 0.086                                                                        & 0.088                                                                        \\
+daily\_dysp   & 0.582                      & 0.803                                                                                & 0.686                                                                        & 0.699                                                                        \\
 \hline
@@
 This paper adds to the literature comparing parametric and nonparametric tests.
+We simulated a variety of data-generating processes under the potential outcomes model, ranging from the ideal case when treatment effects are homogeneous, to cases when data are discrete, effects vary across strata, and treatment assignment is imbalanced within groups.
+The linear regression based tests had comparable power in all circumstances.
+They all suffered a loss of power when the linear model was a poor fit to the data.
+The stratified permutation test appears to perform best when the correlation between the outcome and covariates is low, while using change scores as the dependent variable in the stratified permutation test works well when the correlation is high.
+The linear model permutation tests are a middle ground between these two extremes: they allow one to incorporate covariates with any amount of correlation, assuming that the linear model has a reasonable in-sample fit.
+It is a matter of taste which test one chooses for their experiment: while the parametric test may be robust to violations of its assumptions, it seems reassuring that the permutation test can exactly match the randomization that was done without additional assumptions.
+Applying a linear model based permutation test as a secondary analysis can give insight into how strongly violations of the ANCOVA assumptions affect results
+and give confidence that inferences based on ANCOVA are reliable.
+
 On the one hand, some argue that violations of parametric test assumptions necessitate the use of permutation methods.
-\cite{ludbrook_why_1998} point out that medical trials rarely follow the population sampling model implicit in parametric methods.
-Many people recommend using permutation tests in place of the usual parametric tests for typical analyses, such as ANOVA and generalized linear models (\cite{still_approximate_1981, winkler_permutation_2014}).
+\cite{ludbrook_why_1998} point out that medical trials rarely follow the population sampling model that is implicit in parametric methods, while permutation tests condition on the sample at hand.
+Many people recommend using permutation tests in place of common parametric tests, such as ANOVA and generalized linear models (\cite{still_approximate_1981, winkler_permutation_2014}).
 They argue that there are a myriad of ways that the data may violate the necessary assumptions for the test, and so permutation tests are more robust.
@@
 In these situations, the permutation test strengthens conclusions by giving evidence that the parametric test is robust to departures from its assumptions.
+Our results match those of \citet{vickers_parametric_2005} and \citet{anderson_empirical_1999}, where parametric ANCOVA and regression-based $t$ tests performed the same or better than the comparable nonparametric tests.
 
-Our results match those of \citet{vickers_parametric_2005} and \citet{anderson_empirical_1999}, where parametric ANCOVA and regression-based $t$ tests performed the same or better than the comparable nonparametric tests.
-We simulated a variety of data-generating processes, ranging from the ideal case when all assumptions are met, 
-to the case when data are non-normal, discrete, or skewed.
-In the supplementary simulation files, we also considered data with other possible issues such as heterogeneous treatment effects and unequal sample sizes.
-The linear regression-based tests, both the parametric ANCOVA and the permutation tests, suffered a loss of power when the ANCOVA assumptions were violated.
-However, the tests remained comparable to each other.
-It is a matter of taste which test one chooses for their experiment: while the parametric test may be robust to violations of its assumptions, it is somewhat reassuring that the permutation test can exactly match the randomization that was done while making no distributional assumptions.
-Applying a linear model based permutation test as a secondary analysis can give insight into how ANCOVA results depend on the method's assumptions
-(making it conservative, anti-conservative, or not systematically affecting the $p$-value in a particular direction),
-and give evidence that inferences based on ANCOVA are reliable.
-
-Permutation tests do not come entirely free of assumptions, though: 
-subjects must be exchangeable if we generate the permutation distribution assuming that all possible allocations of treatment which preserve the number of treated units are equally likely.
+In randomized experiments, permutation tests are always valid as long as they match the randomization scheme that was actually conducted.
+The exchangeability of treatments is guaranteed by construction.
+In observational studies, one must prove that subjects are exchangeable.
 \cite{romano_behavior_1990} warns against using permutation tests naively if items are not truly exchangeable. 
@@
 and demonstrates by simulation that the latter has larger than nominal level when the error variances are unequal.
-Randomized experiments mitigate this problem: by definition, treatment assignment is statistically independent of all other variables (possibly conditioning on strata).
+Randomized experiments mitigate this problem: by definition, treatment assignment is statistically independent of all other variables (possibly conditioning on strata) and in expectation over repeated randomizations, the two groups have equal variances.
 
@@
 The onus is on the researcher to make the case that the variable being permuted is uncorrelated with the other variables being held fixed.
-This can be done visually using scatterplots and residual plots (see e.g. \cite{freedman_nonstochastic_1983}) and from knowledge of how the data arose and were collected.
+This can be done visually using scatterplots, residual plots (see e.g. \cite{freedman_nonstochastic_1983}), hypothesis tests, and from knowledge of how the data arose and were collected.
 
-One must be mindful of which covariates to include as controls.
-Ideally, all covariates related to the outcome are known and observed.
-In that case, the coefficient for treatment in a fully saturated linear model (including all covariates and their interactions) is an asymptotically consistent estimate of the average treatment effect (\cite{lin_agnostic_2013}).
-It is rare that one knows \textit{all} relevant covariates in practice; failure to control for all relevant variables can introduce bias in some permutation tests.
- \cite{gail_tests_1988} propose a randomization test based on residuals from an exponential family model and find that omitting relevant covariates leads to tests with higher than nominal level.
-
-The method of controlling for baseline covariates matters, too.
-The naive way is to consider change scores, subtracting baseline measures from the outcome and doing hypothesis testing using the change as the dependent variable.
+Our simulations demonstrate that the method of controlling for baseline covariates matters.
+The naive way to control for repeated measures is to use change scores, the difference between the outcome and baseline measures, as the dependent variable.
 Our simulations confirm the suggestion of \cite{frison_repeated_1992} to use change scores only when there is a strong correlation between baseline and outcome.
@@
 
-\item[Gaussian data simulations:] Simulations using Gaussian data in Section~\ref{sec:simulations}, including code and results. (PDF)
-\item[Imbalanced design simulations:] Simulations using data with different stratum sizes in Section~\ref{sec:simulations}, including code and results. (PDF)
-\item[Skewed data simulations:] Simulations using discrete, skewed data in Section~\ref{sec:simulations}, including code and results. (PDF)
+\item[Continuous outcomes simulations:] Simulations using continuous outcomes in Section~\ref{sec:simulations}, including code and results. (PDF)
+\item[Discrete outcomes simulations:] Simulations using discrete outcomes in Section~\ref{sec:simulations}, including code and results. (PDF)
+\item[Imbalanced treatment simulations:] Simulations using continuous outcomes with varying proportions receiving treatment in Section~\ref{sec:simulations}, including code and results. (PDF)
 \item[Clinical trial data:] Dataset used in Section~\ref{sec:results} to compare methods. (csv)

### 2017-10-05 ms/ms.tex
Complete draft of ms edits

@@
 \item \textbf{Linear regression fit:} the linear regression residuals $\hat{\varepsilon}$ are a good approximation to the true errors $\varepsilon$ 
-\item \textbf{Conditioning space:} the $\varepsilon$ are exchangeable against $X$ within strata
+\item \textbf{Exchangeable errors:} the $\varepsilon$ are exchangeable against $X$ within strata; any configuration of the $\varepsilon$ amongst the units has the same probability
 \item \textbf{No treatment effect under the null:} the $\varepsilon$ do not depend on $Z$
@@
 it is beneficial for power to control for the baseline in the linear model, but the correlation is too low to make the differencing method useful.
+Perhaps surprisingly, the power of the linear model tests was higher when effects were heterogeneous than when they had the same distribution across strata.
 
@@
 These disturbances affect the distributions of $X$ and $Y$, which may impact the power of the parametric ANCOVA and may make one method of controlling for the baseline more effective than another.
-We fixed $\gamma=0.2$ and varied the distributions of $\varepsilon$ and $\delta$.
+We fixed $\gamma=0.2$ and let the errors have a standard normal distribution, a normal distribution with variance depending on the magnitude of $X$ (either standard deviation of 1 if $\lvert X \rvert \leq 1$ or standard deviation of 2 otherwise), a $t$ distribution with 2 degrees of freedom, a standard log normal distribution, or an exponential distribution with parameter $1$, shifted to have mean zero.
 
@@
 The differenced permutation test had substantially less power than the other methods, due to the low correlation between baseline and outcome.
-Power was highest for normally distributed errors and was lowest for $t$ distributed errors.
+Power was highest for the homoskedastic normally distributed errors and was lowest for the heteroskedastic normally distributed errors.
+The scenario with correlated errors violates both the ANCOVA assumption of IID errors and the Freedman-Lane test assumption that errors are exchangeable against $X$.
 
@@
 \multirow{4}{*}{Constant} & Normal & 0.835 & 0.874 & 0.596 & 0.842 & 0.841 \\ 
- & t(2) & 0.286 & 0.316 & 0.146 & 0.292 & 0.291 \\ 
- & Log Normal & 0.450 & 0.472 & 0.178 & 0.463 & 0.465 \\ 
- & Exponential & 0.540 & 0.588 & 0.468 & 0.552 & 0.552 \\ 
+ & Normal, heter. & 0.503 & 0.541 & 0.421 & 0.509 & 0.511 \\  
+ & t(2) & 0.440 & 0.481 & 0.226 & 0.448 & 0.450 \\ 
+ & Log Normal & 0.687 & 0.709 & 0.277 & 0.696 & 0.698 \\ 
+ & Exponential & 0.780 & 0.820 & 0.712 & 0.793 & 0.791 \\ 
    \hline
    \hline
-\multirow{4}{*}{Heterogeneous} & Normal & 0.658 & 0.548 & 0.317 & 0.669 & 0.669 \\ 
-  & t(2) & 0.359 & 0.315 & 0.139 & 0.365 & 0.362 \\ 
-  & Log Normal & 0.487 & 0.361 & 0.155 & 0.507 & 0.509 \\ 
-  & Exponential & 0.631 & 0.543 & 0.407 & 0.636 & 0.633 \\ 
+\multirow{4}{*}{Heterogeneous} & Normal & 0.877 & 0.844 & 0.554 & 0.887 & 0.891 \\ 
+  & Normal, heter. & 0.633 & 0.631 & 0.520 & 0.644 & 0.643 \\ 
+  & t(2) & 0.543 & 0.512 & 0.215 & 0.550 & 0.551 \\ 
+  & Log Normal & 0.711 & 0.608 & 0.252 & 0.730 & 0.731 \\ 
+  & Exponential & 0.814 & 0.786 & 0.651 & 0.820 & 0.816 \\ 
    \hline
@@
 This paper adds to the literature comparing parametric and nonparametric tests.
-We simulated a variety of data-generating processes under the potential outcomes model, ranging from the ideal case when treatment effects are homogeneous, to cases when data are discrete, effects vary across strata, and treatment assignment is imbalanced within groups.
+We simulated a variety of data-generating processes under the potential outcomes model, ranging from the ideal case when treatment effects are homogeneous for all individuals, to cases when errors are heavy-tailed or heteroskedastic, data are discrete, effects vary across strata, and treatment assignment is imbalanced within groups.
 The linear regression based tests had comparable power in all circumstances.
@@
 \item[Imbalanced treatment simulations:] Simulations using continuous outcomes with varying proportions receiving treatment in Section~\ref{sec:simulations}, including code and results. (PDF)
+\item[Nonlinear model simulations:] Simulations using a nonlinear model of potential outcomes in Section~\ref{sec:simulations}, including code and results. (PDF)
 \item[Clinical trial data:] Dataset used in Section~\ref{sec:results} to compare methods. (csv)

### 2017-10-05 simulations/continuous_outcomes.Rmd
Final changes to continuous outcome sims

@@
 The baseline value for individual $i,j$ was 
-$$ X_{ij} = \frac{\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}$$.
+$$ X_{ij} = \frac{-\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}$$.
 
@@
 ```{r generate_sample}
-gen_y1 <- function(gamma, v, error, Z){
+gen_y <- function(gamma, v, error, Z){
   y1 <- 0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error
@@
 
-gen_y0 <- function(gamma, v, error){
+gen_x <- function(gamma, v, error){
   y0 <- 0.5*(-1*gamma*exp(v) + exp(v/2)) + error
@@
   # effect = "same effect" or "heterogeneous"
-  # errors = "normal" or "heavy"
+  # errors = "normal", "t", "lognormal", "exponential", or "x-correlated"
   # n = number of individuals at each stratum
@@
     delta <- rexp(N) - 1
+  } else if (errors == "x-correlated"){
+    epsilon <- rnorm(N)
+    Y0 <- gen_x(gamma, v, epsilon)
+    delta <- rnorm(N)
+    delta[abs(Y0) > 1] <- rnorm(N, sd = 2)
   } else {
@@
 These disturbances affect the distributions of $X$ and $Y$, which may impact the power of the parametric ANCOVA, and affect the variances of $X$ and $Y$, which may make one method of controlling for the baseline more effective than another.
-We fixed $\gamma=0.2$ and let the errors have a standard normal distribution, a $t$ distribution with 2 degrees of freedom, a standard log normal distribution, or an exponential distribution with parameter $1$, shifted to have mean zero.
+We fixed $\gamma=0.2$ and let the errors have a standard normal distribution, a normal distribution with variance depending on the magnitude of $X$ (either standard deviation of 1 if $\lvert X \rvert \leq 1$ or standard deviation of 2 otherwise), a $t$ distribution with 2 degrees of freedom, a standard log normal distribution, or an exponential distribution with parameter $1$, shifted to have mean zero.
 
@@
 The differenced permutation test had substantially less power than the other methods, due to the low correlation between baseline and outcome.
-Power was highest for normally distributed errors and was lowest for $t$ distributed errors.
+Power was highest for homoskedastic normally distributed errors and was lowest when errors were heteroskedastic.
 
@@
 The differenced permutation test continued to have low power.
-Interestingly, when the errors were $t$ or log normally distributed, the power of each test did not change much whether the treatment effects were constant or heterogeneous across strata.
-However, power was lower for normally distributed errors when treatment effects varied across strata.
-In this scenario, the linear model does not fully describe the relationship between baseline and outcome: 
+Interestingly, the power of each of the tests was somewhat higher than when the treatment effects were homogeneous across strata.
+This is somewhat surprising as the linear model does not fully describe the relationship between baseline and outcome: 
 interaction terms between baseline and stratum ID are needed in the model to capture this variation across strata.
@@
 load("continuous_outcomes/exponential_heterogeneous.Rda")
+```
+```{r corr_homogeneous}
+load("continuous_outcomes/corr_errors_homogeneous.Rda")
+```
+```{r corr_heterogeneous}
+load("continuous_outcomes/corr_errors_heterogeneous.Rda")
 
 ```
-
 ```{r power_summary, results = "asis"}
 powers <- list(gaussian_homogeneous,
+               corr_errors_homogeneous,
                t_homogeneous,
@@
                gaussian_heterogeneous,
+               corr_errors_heterogeneous,
                t_heterogeneous,
@@
 summary05 <- t(sapply(powers, function(x) x[5,]))
-summary05 <- data.frame(rep(c("Normal", "t(2)", "Log Normal", "Exponential"), 2), summary05)
+summary05 <- data.frame(rep(c("Normal", "Normal, heter.", "t(2)", "Log Normal", "Exponential"), 2), summary05)
 colnames(summary05) <-  c("Errors", "ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")   
@@
 ```{r power_summary_const_tab, results = "asis"}
-summarytab_constant <- xtable(summary05[1:4,], digits = 3,
+summarytab_constant <- xtable(summary05[1:5,], digits = 3,
        caption = "Empirical power at level $0.05$ for simulated data with constant additive treatment effects",
@@
 ```{r power_summary_het_tab, results = "asis"}
-summarytab_het <- xtable(summary05[5:8, ], digits = 3,
+summarytab_het <- xtable(summary05[6:10, ], digits = 3,
        caption = "Empirical power at level $0.05$ for simulated data with heterogeneous treatment effects",

### 2017-10-05 simulations/continuous_outcomes/corr_errors_heterogeneous.R

@@ -3,12 +3,12 @@ library(doParallel)
 registerDoParallel(cores=5)
 
 set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
-tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "exponential")
+tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "x-correlated")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
   tmp$Y1 <-gen_y(gamma = 0.2, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)
-exponential_heterogeneous <- apply(pvalues, 2, compute_power)
-save(exponential_heterogeneous, file="exponential_heterogeneous.Rda")
+corr_errors_heterogeneous <- apply(pvalues, 2, compute_power)
+save(corr_errors_heterogeneous, file="corr_errors_heterogeneous.Rda")

### 2017-10-05 simulations/continuous_outcomes/corr_errors_homogeneous.R

@@ -1,14 +1,14 @@
 source("simulation_utils.R")
 library(doParallel)
-registerDoParallel(cores=5)
+registerDoParallel(cores=30)
 
 set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
-tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "exponential")
+tmp <- generate_simulated_data(gamma = 0.2, effect = "same effect", errors = "x-correlated")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
   tmp$Y1 <-gen_y(gamma = 0.2, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)
-exponential_heterogeneous <- apply(pvalues, 2, compute_power)
-save(exponential_heterogeneous, file="exponential_heterogeneous.Rda")
+corr_errors_homogeneous <- apply(pvalues, 2, compute_power)
+save(corr_errors_homogeneous, file="corr_errors_homogeneous.Rda")

### 2017-10-05 simulations/continuous_outcomes/exponential_homogeneous.R

@@ -6,7 +6,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "same effect", errors = "exponential")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.2, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-10-05 simulations/continuous_outcomes/gaussian_heterogeneous.R

@@ -6,7 +6,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "normal")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.2, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-10-05 simulations/continuous_outcomes/lognormal_heterogeneous.R

@@ -6,7 +6,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "lognormal")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.2, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-10-05 simulations/continuous_outcomes/lognormal_homogeneous.R

@@ -7,7 +7,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "same effect", errors = "lognormal")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.2, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-10-05 simulations/continuous_outcomes/simulation_utils.R

@@ -127,7 +127,7 @@ generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
   # Input:
   # gamma = multiplier for the magnitude of the treatment effect
   # effect = "same effect" or "heterogeneous"
-  # errors = "normal" or "heavy"
+  # errors = "normal", "t", "lognormal", "exponential", or "x-correlated"
   # n = number of individuals at each stratum
   # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), stratumID (stratum), stratum_effect (beta coefficient per individual), and epsilon (errors)
   
@@ -159,6 +159,11 @@ generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
   } else if (errors == "exponential"){
     epsilon <- rexp(N) - 1
     delta <- rexp(N) - 1
+  } else if (errors == "x-correlated"){
+    epsilon <- rnorm(N)
+    Y0 <- gen_x(gamma, v, epsilon)
+    delta <- rnorm(N)
+    delta[abs(Y0) > 1] <- rnorm(N, sd = 2)
   } else {
     stop("invalid errors parameter")
   }

### 2017-10-05 simulations/continuous_outcomes/t_heterogeneous.R

@@ -6,7 +6,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "heterogeneous", errors = "t")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.2, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-10-05 simulations/continuous_outcomes/t_homogeneous.R

@@ -6,7 +6,7 @@ set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 U
 tmp <- generate_simulated_data(gamma = 0.2, effect = "same effect", errors = "t")
 pvalues <- foreach(i = 1:10000) %dopar% {
   tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
-  tmp$Y1 <-gen_y(gamma = 0.15, v=tmp$v, error=tmp$delta, tmp$Z)
+  tmp$Y1 <-gen_y(gamma = 0.2, v=tmp$v, error=tmp$delta, tmp$Z)
   generate_simulated_pvalues(tmp)
 }
 pvalues <- do.call(rbind, pvalues)

### 2017-10-05 simulations/nonlinear_model.Rmd
Nonlinear model simulations

--- 2017-09-13 simulations/discrete_outcomes.Rmd
@@
 ---
-title: "ANCOVA Comparison Simulations: Discrete Outcomes"
+title: "ANCOVA Comparison Simulations: Nonlinear Model"
 author: "Kellie Ottoboni"
@@
 
-# Discrete data
+The potential outcomes model we used before turned out to be a linear function of $X$ and $Z$, conditional on the latent variable $v$.
+For these simulations, we changed the outcome model to be nonlinear.
+Instead of an additive treatment effect, now the effect is multiplicative.
+We generated the latent variable and errors in the same way, but let
 
-Step 1:
-We generated the covariate sand potential outcomes according to the methods in the continuous outcome section but discretized them by taking the floor.
-Assume that there are $n_j=16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
-After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerate $Z$ 10,000 times.
-We repeat this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.
+\begin{align*}
+X_{ij} &=  \frac{ e^{v_{ij}} + e^{v_{ij}/2}}{2} +\varepsilon_{ij} \\
+Y_{ij}(Z_{ij}) &= (1+\gamma)^{Z_{ij}} X_{ij}  + \delta{ij}.
+\end{align*}
 
-Step 2:
-We regenerate $Z$ and recompute $Y_(Z)$ 10,000 times for each design.
-
-In expectation, the average treatment effect is $\gamma$.
-We compare the empirical power of five tests to detect this treatment effect:
-
-* ANCOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for stratum.
-* Stratified permutation: we permute treatment assignment within stratum, then take the difference in means between treated and control outcomes $Y_1$
-* Differenced permutation: we do the same permutation procedure as the stratified permutation test, except we use the difference between outcome and baseline, $Y_1 - Y_0$
-* Linear model (LM) permutation: we use the same stratified permutation procedure as above, except use the $t$-statistic for the coefficient on treatment in the linear regression of $Y_1$ on $Y_0$, $Z$, and stratum dummies
-* Freedman-Lane test: see the other Rmd document for a full description of this procedure
+The treatment effect for individual $(i, j)$ is thus $\gamma X_{ij}$.
 
@@
 ```{r generate_sample}
-gen_y <- function(gamma, v, error, Z){
-  y <- floor(0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error)
+
+gen_y <- function(gamma, x, error, Z){
+  multiplier <- (1+gamma)^Z
+  y <- multiplier*x + error
   return(y)
@@
 gen_x <- function(gamma, v, error){
-  x <- floor(0.5*(-1*gamma*exp(v) + exp(v/2)) + error)
+  x <- 0.5*(exp(v) + exp(v/2)) + error
   return(x)
@@
   Z <- rep(0:1, length.out = N)
-  Y0 <- gen_x(gamma, v, epsilon)
-  Y1 <- gen_y(gamma, v, delta, Z)
+  Y0 <- gen_y0(gamma, v, epsilon)
+  Y1 <- gen_y1(gamma, v, delta, Z)
   return(data.frame(Y1, Y0, Z, v, stratumID, epsilon, delta))
@@
 
-# Constant additive treatment effect
+# Homogeneous treatment effects
 
 ```{r vary_effect_size_plot, fig.height=4}
-load("discrete_outcomes/vary_gamma_results.Rda")
+load("nonlinear_model/vary_gamma_results.Rda")
 gamma_vec <- seq(0, 0.2, by=0.05)
@@
 Figure 1 shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
-When the effect size was zero, all five tests had the correct level of 5\%, though the stratified and differenced permutation tests rejected 3.5\% of the time.
-The differenced permutation test had substantially lower power than the rest,
-again because the correlation between baseline $X$ and outcome $Y$ was low.
-In this case, the unadjusted stratified test had comparable power to the linear model based tests.
+When the effect size was zero, all five tests had the correct level of 5\%.
+The stratified, unadjusted permutation test failed here:
+the treatment effect is entangled with the baseline covariate, so failing to account for the baseline obscures the treatment effect.
+The parametric and permutation linear model tests perform comparably and are plotted on top of each other, while the differenced permutation test has slightly more power.
 
 # Heterogeneous treatment effect
-
 ```{r vary_effect_size_het_plot, fig.height=4}
-load("discrete_outcomes/vary_gamma_het_results.Rda")
+load("nonlinear_model/vary_gamma_het_results.Rda")
 gamma_vec <- seq(0, 0.2, by=0.05)
@@
 Figure 2 shows the empirical power at level 5\% for these increasing effect sizes.
-The pattern here is nearly identical to Figure 2 in the continuous outcome simulations, so we leave out further comments.
-
-```{r plot_for_ms, include=FALSE}
-p1 <- plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
-  ylim(0, 0.95 ) +
-  theme(legend.position="bottom") + 
-  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
-  report_theme +
-  ggtitle("Constant Treatment \n Effect")
-
-p2 <- plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
-    ylim(0, 0.95) +
-    report_theme +
-    ggtitle("Heterogeneous Treatment \n Effect")
-
-tmp <- ggplot_gtable(ggplot_build(p1))
-leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
-mylegend <- tmp$grobs[[leg]]
-
-p3 <- grid.arrange(arrangeGrob(p1 + theme(legend.position="none"),
-                         p2 + theme(legend.position="none"),
-                         nrow=1),
-             mylegend, nrow=2,heights=c(10, 2))
-
-p3
-ggsave("../ms/fig/discrete_outcomes_simulation_power.pdf", p3, device="pdf", width=8)
-
-```
+Four of the five tests had the correct level of 5\%, while the stratified permutation test rejected only 1\% of the time.
+The same pattern appeared here: the stratified permutation test rarely rejects, while the differenced permutation test dominates all of the linear model based tests.

### 2017-10-05 simulations/nonlinear_model/vary_gamma.R

@@ -1,0 +1,27 @@
+source("simulation_utils.R")
+library(doParallel)
+
+registerDoParallel(cores=30)
+
+gamma_vec <- seq(0, 0.2, by=0.05)
+gamma_res <- foreach(i = seq_along(gamma_vec)) %dopar% {
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "same effect", errors = "normal")
+  pvalues <- replicate(10000, {
+    tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+    tmp$Y1 <-gen_y(gamma = gamma_vec[i], x=tmp$Y0, error=tmp$delta, tmp$Z)
+    generate_simulated_pvalues(tmp, reps=1e3)
+  })
+  return(pvalues)
+}
+
+gamma_res <- lapply(gamma_res, t)
+
+tr_effect <- rep(0, 5)
+for( i in seq_along(gamma_vec)){
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "same effect", errors = "normal")
+  tr_effect[i] <- mean(gamma_vec[i]*tmp$Y0)
+}
+
+save(tr_effect, gamma_res, file="vary_gamma_results.Rda")

### 2017-10-05 simulations/nonlinear_model/vary_gamma_het.R

@@ -1,0 +1,28 @@
+source("simulation_utils.R")
+library(doParallel)
+
+registerDoParallel(cores=30)
+
+gamma_vec <- seq(0, 0.2, by=0.05)
+gamma_res <- foreach(i = seq_along(gamma_vec)) %dopar% {
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "heterogeneous", errors = "normal")
+  pvalues <- replicate(10000, {
+    tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
+    tmp$Y1 <-gen_y(gamma = gamma_vec[i], x=tmp$Y0, error=tmp$delta, tmp$Z)
+    generate_simulated_pvalues(tmp, reps=1e3)
+  })
+  return(pvalues)
+}
+
+gamma_res_het <- lapply(gamma_res, t)
+
+
+tr_effect_het <- rep(0, 5)
+for( i in seq_along(gamma_vec)){
+  set.seed(760682460) # Generated from random.org Timestamp: 2016-11-14 10:21:12 UTC
+  tmp <- generate_simulated_data(gamma = gamma_vec[i], effect = "heterogeneous", errors = "normal")
+  tr_effect_het[i] <- mean(gamma_vec[i] * tmp$Y0)
+}
+
+save(tr_effect_het, gamma_res_het, file="vary_gamma_het_results.Rda")

### 2017-10-05 simulations/imbalanced_ntreat.Rmd
Imbalanced treatment simulations

--- 2017-09-13 simulations/discrete_outcomes.Rmd
@@
 ---
-title: "ANCOVA Comparison Simulations: Discrete Outcomes"
+title: "ANCOVA Comparison Simulations: Imbalanced Treatment Assignment"
 author: "Kellie Ottoboni"
@@
 
-# Discrete data
+# Imbalanced Treatment Groups
 
@@
 We generated the covariate sand potential outcomes according to the methods in the continuous outcome section but discretized them by taking the floor.
-Assume that there are $n_j=16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
+Assume that there are $n_j=16$ individuals per stratum.
+In this set of simulations, we vary the number of treated units in each stratum.
 After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerate $Z$ 10,000 times.
@@
 gen_y <- function(gamma, v, error, Z){
-  y <- floor(0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error)
+  y <- 0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error
   return(y)
@@
 gen_x <- function(gamma, v, error){
-  x <- floor(0.5*(-1*gamma*exp(v) + exp(v/2)) + error)
+  x <- 0.5*(-1*gamma*exp(v) + exp(v/2)) + error
   return(x)
@@
 
-generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
+generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16), ntreat = c(8, 8, 8)){
   # Input:
@@
   # n = number of individuals at each stratum
+  # ntreat = number of treated individuals in each stratum
   # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), stratumID (stratum), stratum_effect (beta coefficient per individual), and epsilon (errors)
@@
   # Generate covariates
-  Z <- rep(0:1, length.out = N)
+  Z <- rep(0, N)
+  for(i in 1:3){
+    Z[stratumID == i] <- rep(1:0, each=c(ntreat[i], n[i] - ntreat[i]))
+  }
   Y0 <- gen_x(gamma, v, epsilon)
@@
 
-plot_power_gamma <- function(powermat_list, gamma_vec, alpha, title){
+plot_power_ntreat <- function(powermat_list, ntreat, alpha, title){
   gamma_power_alpha <- t(sapply(powermat_list, function(x) x[floor(alpha*nrow(x)),]))
@@
   mutate("Method" = Var2) %>%
-  mutate("Effect" = rep(gamma_vec, 5)) %>% 
-  ggplot(aes(x = Effect, y = Power)) +
-  geom_line(aes(color = Method))+
-  xlab("Effect size") +
+  mutate("ntreat" = rep(ntreat, 5)) %>% 
+  ggplot(aes(x = ntreat, y = Power)) +
+  geom_point(aes(color = Method))+
+  xlab("Number treated") +
   theme_bw() +
@@
 
-```{r vary_effect_size_plot, fig.height=4}
-load("discrete_outcomes/vary_gamma_results.Rda")
-gamma_vec <- seq(0, 0.2, by=0.05)
-gamma_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
-plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
+```{r vary_ntreat_plot, fig.height=4}
+load("imbalanced_tr/vary_ntreat_results.Rda")
+ntreat_vec <- list( c(8, 8, 8),
+                 c(4, 4, 4),
+                 c(4, 4, 8),
+                 c(4, 8, 8),
+                 c(4, 8, 12),
+                 c(8, 8, 12),
+                 c(8, 12, 12),
+                 c(12, 12, 12))
+ntreat_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
+ntreat_str <- sapply(ntreat_vec, paste, collapse = ",")
+ntreat_tot <- sapply(ntreat_vec, sum)
+plot_power_ntreat(ntreat_power, ntreat_tot, 0.05, "xx")
 ```
@@
 First, we let the $v_{ij}$ have the same distribution across strata and generated normally distributed errors.
-We varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
-In the random populations that were generated, this corresponded to population average treatment effects of `r round(tr_effect, 2)[1:4]`, and `r round(tr_effect, 2)[5]`, respectively.
-Figure 1 shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
-When the effect size was zero, all five tests had the correct level of 5\%, though the stratified and differenced permutation tests rejected 3.5\% of the time.
+We varied the number of treated units in each group.
+In the random populations that were generated, this corresponded to population average treatment effects of about $1.36$.
+Figure 1 shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% plotted against the total number of treated units.
+Table 1 shows the same information.
+Each of the tests tended to have higher power when the total number of treated units was around half of the population (in this case, 24 out of 48).
+It is interesting to note that all tests had more power when 25\% of units were treated than when 75% were treated; it seems that power is not a symmetric function.
+
 The differenced permutation test had substantially lower power than the rest,
@@
 In this case, the unadjusted stratified test had comparable power to the linear model based tests.
+ANCOVA had slightly higher power than the linear model permutation tests for all experiments except for when the number of treated units was equal to number of controls and when the number of treated units was 30 or above.
+
+```{r vary_ntreat_tab, results="asis"}
+summary05 <- t(sapply(ntreat_power, function(x) x[5,]))
+summary05 <- cbind(ntreat_str, summary05)
+colnames(summary05) <-  c("Number Treated", "ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane") 
+print(xtable(summary05[order(ntreat_tot), ],
+             caption = "Empirical power at level $0.05$ for simulated data with constant additive treatment effects"),
+      include.rownames=FALSE)
+
+```
 
@@
 
-```{r vary_effect_size_het_plot, fig.height=4}
-load("discrete_outcomes/vary_gamma_het_results.Rda")
-gamma_vec <- seq(0, 0.2, by=0.05)
-gamma_power_het <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
-plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
+```{r vary_ntreat_het_plot, fig.height=4}
+load("imbalanced_tr/vary_ntreat_het_results.Rda")
+ntreat_het_vec <- list( c(8, 8, 8),
+                 c(4, 4, 4),
+                 c(4, 4, 8),
+                 c(4, 8, 4),
+                 c(8, 4, 4),
+                 c(4, 8, 8),
+                 c(8, 4, 8),
+                 c(8, 8, 4),
+                 c(4, 8, 12),
+                 c(12, 8, 4),
+                 c(4, 12, 8),
+                 c(8, 12, 8),
+                 c(4, 12, 4),
+                 c(12, 12, 12))
+ntreat_het_power <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
+ntreat_het_str <- sapply(ntreat_het_vec, paste, collapse = ",")
+ntreat_het_tot <- sapply(ntreat_het_vec, sum)
+plot_power_ntreat(ntreat_het_power, ntreat_het_tot, 0.05, "xx")
 ```
 Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
-Once again, we varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
-The corresponding population average treatment effects were of `r round(tr_effect_het, 2)[1:4]`, and `r round(tr_effect_het, 2)[5]`, respectively.
-Figure 2 shows the empirical power at level 5\% for these increasing effect sizes.
-The pattern here is nearly identical to Figure 2 in the continuous outcome simulations, so we leave out further comments.
+Once again, we varied the number of treated units in each stratum and the population average treatment effects were around $1.36$.
+Figure 2 and Table 2 show the empirical power at level 5\% for these experiments.
+The pattern here is similar to Figure 1: power tends to be highest when the number of treated units is around half of the population.
+The treatment effect was largest in the third stratum and smallest in the first stratum;
+as expected, Table 2 shows that among experiments that fixed the total number of treated units, the experiment with more treatment units in the third stratum tended to have higher power.
+These variations are only slight, though.
+As before, the differenced permutation test had low power.
+The linear model permutation tests, ANCOVA, and stratified permutation test, in that order, had the best power.
 
-```{r plot_for_ms, include=FALSE}
-p1 <- plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
-  ylim(0, 0.95 ) +
-  theme(legend.position="bottom") + 
-  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
-  report_theme +
-  ggtitle("Constant Treatment \n Effect")
-
-p2 <- plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
-    ylim(0, 0.95) +
-    report_theme +
-    ggtitle("Heterogeneous Treatment \n Effect")
-
-tmp <- ggplot_gtable(ggplot_build(p1))
-leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
-mylegend <- tmp$grobs[[leg]]
-
-p3 <- grid.arrange(arrangeGrob(p1 + theme(legend.position="none"),
-                         p2 + theme(legend.position="none"),
-                         nrow=1),
-             mylegend, nrow=2,heights=c(10, 2))
-
-p3
-ggsave("../ms/fig/discrete_outcomes_simulation_power.pdf", p3, device="pdf", width=8)
+```{r vary_ntreat_het_tab, results="asis"}
+summary05het <- t(sapply(ntreat_het_power, function(x) x[5,]))
+summary05het <- cbind("Number treated" = ntreat_het_str, summary05het)
+colnames(summary05het) <-  c("Number Treated", "ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane") 
+print(xtable(summary05het[order(ntreat_het_tot),],
+             caption = "Empirical power at level $0.05$ for simulated data with heterogeneous treatment effects"),
+      include.rownames=FALSE)
 

### 2017-10-05 simulations/imbalanced_tr/simulation_utils.R

@@ -1,0 +1,275 @@
+#devtools::install_github("statlab/permuter")
+#require(permuter)
+require(dplyr)
+require(reshape2)
+require(ggplot2)
+require(xtable)
+options(xtable.comment = FALSE, xtable.include.rownames = FALSE, xtable.align = "p{3.5cm}")
+require(grid)
+require(gridExtra)
+
+report_theme <- theme_bw() + theme(
+  axis.text.x = element_text(size = 14, angle = -45),
+  axis.text.y = element_text(size = 14),
+  axis.title = element_text(size = 20),
+  title = element_text(size = 20),
+  legend.title = element_text(size = 14),
+  legend.text = element_text(size = 14),
+  strip.text.x = element_text(size = 14)
+)
+
+permute_within_groups <- function(x, group) {
+  if(is.vector(x)){
+    for(g in unique(group)){
+      gg <- (group == g)
+      x[gg] <- sample(x[gg])
+    }
+  } else if(is.data.frame(x) | is.matrix(x)){
+    for(g in unique(group)){
+      gg <- which(group == g)
+      x[gg, ] <- x[sample(gg), ]
+    }
+  } else {
+    stop("x is an invalid data type")
+  }
+  return(x)
+}
+
+
+
+stratified_two_sample <- function(group, response, stratum, stat = "mean", reps = 1000) {
+  if (!is.vector(group) | !is.vector(response) | !is.vector(stratum)) {
+    stop("inputs must be vectors")
+  }
+  if (!is.numeric(response)) {
+    stop("response must be numeric")
+  }
+  
+  if (length(unique(group)) > 2) {
+    stop("two samples only")
+  }
+  
+  groups <- unique(group)
+  strata <- unique(stratum)
+  
+  ordering <- order(group)
+  response <- response[ordering]
+  stratum <- stratum[ordering]
+  group <- group[ordering]
+  
+  ntreat <- table(group)[1]
+  N <- length(group)
+  
+  # If stat is callable, use it as the test function. Otherwise, look in the
+  # dictionary
+  stats <- list(mean = function(u) {
+    mean(u[1:ntreat], na.rm = TRUE) - mean(u[(ntreat + 1):N], na.rm = TRUE)
+  }, t = function(u) {
+    t.test(u[1:ntreat], u[(ntreat + 1):N], var.equal = TRUE)$statistic
[…]

### 2017-10-09 simulations/continuous_outcomes.Rmd
Create power ratio plots for the simulations

@@
 
+plot_power_ratio_gamma <- function(powermat_list, gamma_vec, alpha, title){
+  gamma_power_alpha <- t(sapply(powermat_list, function(x) x[floor(alpha*nrow(x)),]))
+  colnames(gamma_power_alpha) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+  gamma_power_alpha <- apply(gamma_power_alpha, 2, function(x) x/gamma_power_alpha[,"ANCOVA"])
+  melt(gamma_power_alpha, value.name="Power") %>%
+  mutate("Method" = Var2) %>%
+  mutate("Effect" = rep(gamma_vec, 5)) %>% 
+  filter(Method != "Differenced Permutation") %>%
+  filter(Method != "ANCOVA") %>%
+  ggplot(aes(x = Effect, y = Power)) +
+  geom_line(aes(color = Method))+
+  xlab("Effect size") +
+  ylab("Relative Power") +
+  theme_bw() +
+  theme(
+    axis.text.x = element_text(size = 12),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 16),
+    title = element_text(size = 16),
+    legend.title = element_text(size = 12),
+    legend.text = element_text(size = 14),
+    strip.text.x = element_text(size = 12)
+  ) 
+}
 ```
@@
 plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
+plot_power_ratio_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect")
 ```
@@
 plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect")
+
+plot_power_ratio_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect")
+
 ```
@@
 ```{r plot_for_ms, include=FALSE}
-p1 <- plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
-  ylim(0, 0.9) +
+p1 <- plot_power_ratio_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
+  ylim(0, 1.2) +
   theme(legend.position="bottom") + 
-  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
+  guides(color=guide_legend(nrow=1,byrow=TRUE)) +
   report_theme +
@@
 
-p2 <- plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
-    ylim(0, 0.9) +
+p2 <- plot_power_ratio_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
+  ylim(0, 1.2) +
     report_theme +

### 2017-10-09 simulations/nonlinear_model.Rmd
Create power ratio plots for the simulations

@@
 
+plot_power_ratio_gamma <- function(powermat_list, gamma_vec, alpha, title){
+  gamma_power_alpha <- t(sapply(powermat_list, function(x) x[floor(alpha*nrow(x)),]))
+  colnames(gamma_power_alpha) <- c("ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
+  gamma_power_alpha <- apply(gamma_power_alpha, 2, function(x) x/gamma_power_alpha[,"ANCOVA"])
+  melt(gamma_power_alpha, value.name="Power") %>%
+  mutate("Method" = Var2) %>%
+  mutate("Effect" = rep(gamma_vec, 5)) %>% 
+  filter(Method != "Differenced Permutation") %>%
+  filter(Method != "ANCOVA") %>%
+  ggplot(aes(x = Effect, y = Power)) +
+  geom_line(aes(color = Method))+
+  xlab("Effect size") +
+  ylab("Relative Power") +
+  theme_bw() +
+  theme(
+    axis.text.x = element_text(size = 12),
+    axis.text.y = element_text(size = 12),
+    axis.title = element_text(size = 16),
+    title = element_text(size = 16),
+    legend.title = element_text(size = 12),
+    legend.text = element_text(size = 14),
+    strip.text.x = element_text(size = 12)
+  ) 
+}
+
 ```
@@
 plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
+plot_power_ratio_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect")
+
 ```
@@
 plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect")
+
+plot_power_ratio_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
 ```

### 2017-10-09 ms/ms.tex
Change title, remove differenced perm test from the comparisons in the paper

@@
 {
-  \title{\bf A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Experiments}
+  \title{\bf An Empirical Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Experiments}
 \author{Kellie Ottoboni \\
@@
 We compare analysis of covariance (ANCOVA), a special case of linear regression that incorporates stratification, to several permutation tests based on linear models that control for pretreatment covariates.
-In several simulations of randomized experiments using a model in which certain regression assumptions are violated,
+In simulations of randomized experiments using models which violate some of the parametric regression assumptions,
 the permutation tests maintain power comparable to ANCOVA.
@@
 by exploiting the fact that the difference in average outcomes between the two treatment groups is asymptotically normal with a variance that can be estimated from the data.
-
-%Randomized experiments in the real world are rarely this simple:
-%if individuals are heterogeneous before the study, then their outcomes may differ for reasons besides the treatment;
-%there may be more than two treatments under study;
-%patients may not take the treatment they are assigned or may drop out of the study;
-%and peer effects, the influence of other members of a treatment group, may affect outcomes beyond the treatment itself.
-%Techniques have been developed to deal with each of these issues, invoking additional assumptions to make valid inference possible.
-%In this paper, we focus on the issue of heterogeneous pretreatment characteristics and assume that the other issues are not present.
 
@@
 Section~\ref{sec:methods} introduces the potential outcomes model, shows how this model is inconsistent with the assumptions of the parametric ANCOVA, and describes permutation tests whose assumptions match this model.
-Section~\ref{sec:simulations} presents simulations that suggest that even in this potential outcomes framework, the parametric and permutation tests have comparable power to detect a treatment effect.
+Section~\ref{sec:simulations} presents simulations that suggest that even in this potential outcomes framework when various assumptions for the ANCOVA test are violated, the parametric and permutation tests have comparable power to detect a treatment effect.
 In Section~\ref{sec:results}, we apply each of the tests to data from a clinical trial comparing the performance of two treatments for gastroesophageal reflux disease (GERD).
@@
 the $t$ test from the parametric ANCOVA,
-a stratified permutation test using the difference in mean outcomes
+a stratified permutation test using the difference in mean outcomes as the statistic
  (called ``stratified permutation'' in what follows),
-a stratified permutation test using the mean change scores, defined as the difference between the outcome and baseline measure (called ``differenced permutation'' in what follows),
 a stratified permutation test based on the $t$ statistic from the linear regression of outcome on control variables (called ``LM permutation'' in what follows),
@@
 The simulation code is included in four supplementary files.
+The supplementary files additionally consider a stratified permutation test using the mean change scores, defined as the difference between the outcome and baseline measure;
+we omit the results here.
 
@@
 We used two scenarios.
-In the first scenario, the treatment effect was constant across strata.
+In the first scenario, the treatment effect was homogeneous across strata.
 We drew a latent random variate $v_{ij}$ from the uniform distribution on $[-4, 4]$.
-In the second scenario, the treatment effect varied across strata, and we drew the latent random variable according to
+In the second scenario, the treatment effect was heterogeneous across strata, and we drew the latent random variable according to
 \begin{displaymath}
@@
 \includegraphics[width = \textwidth]{fig/continuous_outcomes_simulation_power.pdf}
-\caption{Empirical power curves for the continuous simulated data with normally distributed $\varepsilon$ and $\delta$.
+\caption{Empirical power ratio curves for the continuous simulated data with normally distributed $\varepsilon$ and $\delta$, showing the ratio of the rejection rate of the test to the rejection rate of the parametric ANCOVA.
 The left panel shows power when the treatment effect was constant across strata and the right panel shows power when the treatment effect was heterogeneous.}
@@
 In the random populations that were generated, this corresponded to population average treatment effects of $0, 0.34, 0.68, 1.02$, and $1.36$ respectively.
-The left panel of Figure~\ref{fig:continuous_outcomes_sim_power} shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
-When the treatment effect was homogeneous across strata, all five tests had the correct level of 5\% (i.e. the power when $\gamma = 0$).
-Power of the differenced permutation test increased at a slower rate than the other tests.
-This is unsurprising, as the correlation between baseline $X$ and outcome $Y$ was low:
-its highest value was $0.33$ when $\gamma=0$ and it decreased to nearly $0$ when $\gamma=0.2$.
-The parametric and permutation linear model tests performed comparably and are plotted on top of each other, while the stratified permutation test had slightly more power.
+When the treatment effect was homogeneous across strata, all the tests had the correct level of 5\% (i.e. the power when $\gamma = 0$).
+The left panel of Figure~\ref{fig:continuous_outcomes_sim_power} shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\%, relative to the empirical power of the ANCOVA, for these increasing effect sizes.
+A ratio of one indicates that the test had the same power as the ANCOVA, while a ratio above or below 1 indicates that power was higher or lower, respectively, than the power of the ANCOVA.
+The permutation linear model tests performed comparably and are plotted on top of each other, while the stratified permutation test had slightly more power.
 
-The right panel of Figure~\ref{fig:continuous_outcomes_sim_power} shows the empirical power at level 5\% for the same experiment, with heterogeneous treatment effects across strata.
-Four of the five tests had the correct level of 5\%, while the stratified permutation test rejected only 1\% of the time.
-Both the stratified and differenced permutation tests had lower power than the linear model tests as the effect size increased.
+The right panel of Figure~\ref{fig:continuous_outcomes_sim_power} shows the empirical power ratios at level 5\% for the same experiment, with heterogeneous treatment effects across strata.
+Three of the four tests had the correct level of 5\%, while the stratified permutation test rejected only 1\% of the time.
+The stratified permutation had lower power than the linear model tests as the effect size increased.
 The correlation between baseline $X$ and outcome $Y$ was low but nontrivial (between $0.04$ and $0.44$), so this result makes some sense:
-it is beneficial for power to control for the baseline in the linear model, but the correlation is too low to make the differencing method useful.
-Perhaps surprisingly, the power of the linear model tests was higher when effects were heterogeneous than when they had the same distribution across strata.
+it is beneficial for power to control for the baseline in the linear model.
 
-Finally, we varied the distributions of the errors $\varepsilon$ and $\delta$ and examined the power of the five tests.
+Finally, we varied the distributions of the errors $\varepsilon$ and $\delta$ and examined the power of the four tests.
 Recall that these are not errors in the linear model framework, but rather disturbances on the potential outcomes and baseline measures for each individual.
-These disturbances affect the distributions of $X$ and $Y$, which may impact the power of the parametric ANCOVA and may make one method of controlling for the baseline more effective than another.
+These disturbances affect the distributions of $X$ and $Y$, which may impact the power of the parametric ANCOVA and may make it beneficial to control for the baseline.
 We fixed $\gamma=0.2$ and let the errors have a standard normal distribution, a normal distribution with variance depending on the magnitude of $X$ (either standard deviation of 1 if $\lvert X \rvert \leq 1$ or standard deviation of 2 otherwise), a $t$ distribution with 2 degrees of freedom, a standard log normal distribution, or an exponential distribution with parameter $1$, shifted to have mean zero.
@@
 All three linear model based tests had roughly the same rejection rate.
-The differenced permutation test had substantially less power than the other methods, due to the low correlation between baseline and outcome.
 Power was highest for the homoskedastic normally distributed errors and was lowest for the heteroskedastic normally distributed errors.
@@
 Conversely, the stratified permutation test lost power for this reason.
-The differenced permutation test continued to have low power.
 Interestingly, when the errors were $t$ or log normally distributed, the power of each test did not change much whether the treatment effects were constant or heterogeneous across strata.
-However, power was lower for normally distributed errors when treatment effects varied across strata.
+Perhaps surprisingly, the power of the linear model tests was higher when effects were heterogeneous than when they had the same distribution across strata.
 In this scenario, the linear model does not fully describe the relationship between baseline and outcome: 
@@
 \begin{adjustbox}{width=\textwidth}
-\begin{tabular}{L | l |LLLLL}
+\begin{tabular}{L | l |LLLL}
   \hline
-Treatment Effect & Errors & ANCOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
+Treatment Effect & Errors & ANCOVA & Stratified Permutation & LM Permutation & Freedman-Lane \\ 
   \hline
-\multirow{4}{*}{Constant} & Normal & 0.835 & 0.874 & 0.596 & 0.842 & 0.841 \\ 
- & Normal, heter. & 0.503 & 0.541 & 0.421 & 0.509 & 0.511 \\  
- & t(2) & 0.440 & 0.481 & 0.226 & 0.448 & 0.450 \\ 
- & Log Normal & 0.687 & 0.709 & 0.277 & 0.696 & 0.698 \\ 
- & Exponential & 0.780 & 0.820 & 0.712 & 0.793 & 0.791 \\ 
+\multirow{4}{*}{Constant} & Normal & 0.835 & 0.874 & 0.842 & 0.841 \\ 
+ & Normal, heter. & 0.503 & 0.541 & 0.509 & 0.511 \\  
+ & t(2) & 0.440 & 0.481 & 0.448 & 0.450 \\ 
+ & Log Normal & 0.687 & 0.709 & 0.696 & 0.698 \\ 
+ & Exponential & 0.780 & 0.820 & 0.793 & 0.791 \\ 
    \hline
    \hline
-\multirow{4}{*}{Heterogeneous} & Normal & 0.877 & 0.844 & 0.554 & 0.887 & 0.891 \\ 
-  & Normal, heter. & 0.633 & 0.631 & 0.520 & 0.644 & 0.643 \\ 
-  & t(2) & 0.543 & 0.512 & 0.215 & 0.550 & 0.551 \\ 
-  & Log Normal & 0.711 & 0.608 & 0.252 & 0.730 & 0.731 \\ 
-  & Exponential & 0.814 & 0.786 & 0.651 & 0.820 & 0.816 \\ 
+\multirow{4}{*}{Heterogeneous} & Normal & 0.877 & 0.844 & 0.887 & 0.891 \\ 
+  & Normal, heter. & 0.633 & 0.631 & 0.644 & 0.643 \\ 
+  & t(2) & 0.543 & 0.512 & 0.550 & 0.551 \\ 
+  & Log Normal & 0.711 & 0.608 & 0.730 & 0.731 \\ 
+  & Exponential & 0.814 & 0.786 & 0.820 & 0.816 \\ 
    \hline
@@
 \end{adjustbox}
-\caption{Empirical power at level $0.05$ for simulated data with constant additive treatment effects (top panel) and heterogeneous treatment effects (bottom panel).} 
+\caption{Empirical power at level $0.05$ for simulated data with homogeneous treatment effects (top panel) and heterogeneous treatment effects (bottom panel).} 
 \label{tab:power_grid1}
@@
 
-The left panel of Figure~\ref{fig:discrete_outcomes_sim_power} shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes when treatment effects were constant.
-When the effect size was zero, all five tests had the correct level of 5\%, though the stratified and differenced permutation tests rejected 3.5\% of the time.
-The differenced permutation test had substantially lower power than the rest,
-again because the correlation between baseline $X$ and outcome $Y$ was low.
+The left panel of Figure~\ref{fig:discrete_outcomes_sim_power} shows the empirical power at level 5\% relative to the empirical power of the ANCOVA for these increasing effect sizes when treatment effects were homogeneous.
+When the effect size was zero, all four tests had the correct level of 5\%, though the stratified permutation test rejected 3.5\% of the time.
 In this case, the unadjusted stratified test had comparable power to the linear model based tests.
-The right panel of Figure~\ref{fig:discrete_outcomes_sim_power} shows the empirical power when treatment effects varied across strata.
+The right panel of Figure~\ref{fig:discrete_outcomes_sim_power} shows the empirical power ratios when treatment effects varied across strata.
 The patterns are similar to when outcomes were continuous.
@@
 \includegraphics[width = \textwidth]{fig/discrete_outcomes_simulation_power.pdf}
-\caption{Empirical power curves for the discrete simulated data.
+\caption{Empirical power ratio curves for the discrete simulated data, showing the ratio of the rejection rate of the test to the rejection rate of the parametric ANCOVA.
 The left panel shows power when the treatment effect was constant across strata and the right panel shows power when the treatment effect was heterogeneous.}
@@
 
-The differenced permutation test had substantially lower power than the rest,
-again because the correlation between baseline $X$ and outcome $Y$ was low.
-In this case, the unadjusted stratified test had comparable power to or higher than the linear model based tests.
+The unadjusted stratified test had comparable power to or higher than the linear model based tests.
 ANCOVA had slightly higher power than the linear model permutation tests for all experiments except for when the number of treated units was equal to number of controls and when the number of treated units was 30 or above.
+Overall, these differences were not practically significant.
 We refer the reader to the supplementary file for the details of these experiments.
@@
 the potential outcomes depend on the baseline covariate, so units are not exchangeable without controlling for the baseline.
-The ANCOVA and permutation linear model tests had identical power, while the differenced permutation test had slightly more power.
+The ANCOVA and permutation linear model tests had identical power.
 The patterns were the same when the latent variables $v$ had the same and different distributions across strata;
-it made little difference because the treatment effect varied within strata, anyway.
+it made little difference because the treatment effect varied within strata due to the dependence on the baseline covariate.
 
@@
 They all suffered a loss of power when the linear model was a poor fit to the data.
-The stratified permutation test appears to perform best when the correlation between the outcome and covariates is low, while using change scores as the dependent variable in the stratified permutation test works well when the correlation is high.
+The stratified permutation test appears to perform best when the correlation between the outcome and covariates is low, while using change scores as the dependent variable in the stratified permutation test works best when the correlation is high.
 The linear model permutation tests are a middle ground between these two extremes: they allow one to incorporate covariates with any amount of correlation, assuming that the linear model has a reasonable in-sample fit.
@@
 The onus is on the researcher to make the case that the variable being permuted is uncorrelated with the other variables being held fixed.
-This can be done visually using scatterplots, residual plots (see e.g. \cite{freedman_nonstochastic_1983}), hypothesis tests, and from knowledge of how the data arose and were collected.
+This can be done visually using scatterplots, residual plots, hypothesis tests, and from knowledge of how the data arose and were collected.
 
@@
 The naive way to control for repeated measures is to use change scores, the difference between the outcome and baseline measures, as the dependent variable.
-Our simulations confirm the suggestion of \cite{frison_repeated_1992} to use change scores only when there is a strong correlation between baseline and outcome.
+Simulations in the supplementary files confirm the suggestion of \cite{frison_repeated_1992} to use change scores only when there is a strong correlation between baseline and outcome.
 Weak correlations between baseline and outcome occur often in practice, as was the case with our GERD dataset.

### 2017-10-13 results/results.Rmd

@@
 To reduce day-to-day variation, we averaged the measures from each week to obtain two observations per patient, one pre-treatment and one post-treatment.
+These averages are no longer discrete ordinal values, but they are not continuous either: they can only take a finite set of values.
 
@@
 We conclude by running the tests for each endpoint.
+
+We show that the data may violate the assumptions for the parametric ANCOVA.
+In particular, the outcome variables and the linear regression residuals are skewed to the right, and there is evidence of heteroskedasticity.
+The small $p$-values that come from the parametric analysis may therefore be unreliable.
+However, the permutation tests yield only slightly larger $p$-values.
+The correspondence between parametric and permutation test results gives some confidence that the results are stable and reliable.
+
+\newpage
 
@@
 Figure \ref{fig:hist_continuous_vars} shows the distribution of the final outcome measures between the two treatment groups.
-Based on these distributions, we would expect to detect a difference in outcomes between the treatments for the `daily_heart`, `daily_hrdq`, and `heart_freq` measures.
+Table 1 shows the means of the distributions.
+Based on these summary statistics, we would expect to detect a difference in outcomes between the treatments for the `daily_heart`, `daily_hrdq`, and `heart_freq` measures.
 
@@
 
-Table 1 shows the correlation between baseline and outcome for each variable.
+```{r summary_tab, results = "asis"}
+outcome_summary <- clinical %>% 
+  group_by(tr) %>% 
+  summarise(mean(daily_heart),
+    mean(daily_regurg), 
+    mean(daily_dysp),
+    mean(daily_hrdq), 
+    mean(heart_freq),
+    mean(regurg_freq), 
+    mean(dysp_freq)) %>%
+  select(-tr) %>%
+  round(2) %>% t()
+outcome_sds <- clinical %>% 
+  group_by(tr) %>% 
+  summarise(sd(daily_heart),
+    sd(daily_regurg), 
+    sd(daily_dysp),
+    sd(daily_hrdq), 
+    sd(heart_freq),
+    sd(regurg_freq), 
+    sd(dysp_freq)) %>%
+  select(-tr) %>%
+  round(2) %>% t()
+outcome_summary <- cbind(outcome_summary[, 1],
+                         outcome_sds[, 1],
+                         outcome_summary[, 2],
+                         outcome_sds[, 2])
+colnames(outcome_summary) <- c("Mean A", "SD A", "Mean B", "SD B")
+rownames(outcome_summary) <- continuous_vars
+print(xtable(outcome_summary, caption = "Mean and standard deviation of each continuous outcome in groups A and B."), include.rownames = TRUE)
+
+```
+
+Table 2 shows the correlation between baseline and outcome for each variable.
 Correlations range between 0.548 and 0.696.
@@
 ```
+\newpage
 
@@
 
+\newpage
+
 ## Parametric ANCOVA
@@
 lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = daily_hrdq)
-print(xtable(summary(aov(lm1))), include.rownames = TRUE)
 daily_hrdq$lm1_resid <- residuals(lm1)
+daily_hrdq$lm1_fitted <- fitted(lm1)
 
 lm2 <- lm(difference ~ tr + factor(SITEID), data = daily_hrdq)
-print(xtable(summary(aov(lm2))), include.rownames = TRUE)
 daily_hrdq$lm2_resid <- residuals(lm2)
+daily_hrdq$lm2_fitted <- fitted(lm2)
 ```
@@
 It is generally safer to change types yourself.)
+
+```{r hrdq_parametric_tab, results = "asis"}
+print(xtable(summary(aov(lm1))), include.rownames = TRUE)
+print(xtable(summary(aov(lm2))), include.rownames = TRUE)
+```
+
+```{r residual_plot_lm_hrdq, fig.align="center", fig.height = 3, fig.cap = "Residual plots for the ANCOVA model of daily hrdq."}
+p01 <- daily_hrdq %>% 
+  ggplot(aes(x = lm1_fitted, y = lm1_resid)) +
+  geom_point() + 
+  geom_hline(yintercept = 0, linetype = "dashed") +
+  xlab("Fitted Values") + ylab("Residual") + 
+  ggtitle("Raw Outcome Model") +
+  theme_bw()
+p02 <- daily_hrdq %>% 
+  ggplot(aes(x = lm2_fitted, y = lm2_resid)) +
+  geom_point() + 
+  geom_hline(yintercept = 0, linetype = "dashed") +
+  xlab("Fitted Values") + ylab("Residual") + 
+  ggtitle("Differenced Model") +
+  theme_bw()
+grid.arrange(p02, p01, nrow=1)
+```
+
+Figure \ref{fig:residual_plot_lm_hrdq} shows the fitted values versus residuals for the two ANCOVA models.
+The residuals of the raw outcome model appear to increase in magnitude with the fitted value, casting doubt on the assumption that the error variance is constant.
+The plot reveals that the differenced model only yields a few predicted values, but does not seem to indicate heteroskedasticity.
 
@@
 
-### Linear model residuals
+### Freedman-Lane test of residuals
 
@@
 To do so, we conduct independent permutation tests within each stratum using the Pearson correlation as the statistic.
-Table 3 shows the results of this test: while the correlation between baseline and residuals appears high, it is not statistically significant at the 5\% level in any stratum.
+Table 5 shows the results of this test: while the correlation between baseline and residuals appears high, it is not statistically significant at the 5\% level in any stratum.
 The Fisher's combined $p$-value for the 8 strata is `r round(fisher_combined_corr, 2)`, so there is no evidence that the independence assumption is violated.
@@
 These are done to check that treatment and covariates are uncorrelated with the regression errors for the Freedman and Lane style permutation test."}
+
 p11 <- daily_hrdq %>% 
@@
 We run this procedure for all seven continuous endpoints.
-Table 5 shows the results.
+Table 6 shows the results.
 
+Figure \ref{fig:residual_plots_lm} displays the residuals versus fitted values for the linear model using each response variable.
+The figures cast doubt on the assumption of constant error variance in the linear model: they indicate that the variance of the residuals increases with the fitted values and the distribution of residuals may have some right skew.
+The assumption of normally distributed errors with constant variance is important for the parametric analysis of the ANCOVA model, but does not apply to the permutation tests.
+Across all tests, the parametric ANCOVA $p$-value is smaller than the linear model permutation test $p$-values.
+Perhaps this is partly due to the heteroskedasticity: the permutation tests make no distributional assumptions, so may be less sensitive to large outcomes or residuals.
 ```{r results_table_make}
@@
 
+```{r residual_plots_lm, fig.align = "center", fig.cap = "Residual plots for the linear regression of outcome on baseline, treatment, and stratum ID for each response variable."}
+
+pplot <- vector("list", 7)
+for(i in seq_along(continuous_vars)){
+  col <- continuous_vars[i]
+  tmpdata <- reshape_data(col)
+  lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = tmpdata)
+  tmpdata <- tmpdata %>% mutate("Residuals" = residuals(lm1), "Fitted" = fitted(lm1))
+  pplot[[i]] <- ggplot(tmpdata, aes(x = Fitted, y = Residuals)) + geom_point() + ggtitle(continuous_vars[i]) +
+  theme_bw()
+}
+do.call(grid.arrange, c(pplot, nrow= 2))
+```
+
 ```{r results_table_print, echo=FALSE, results = "asis", fig.align = "center", fig.cap = "Comparison of $p$-values from four tests, for each continuous endpoint."}

### 2017-10-13 simulations/discrete_outcomes.Rmd

@@
 plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect")
+
+plot_power_ratio_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Constant Treatment Effect")
+
 ```
@@
 ```{r plot_for_ms, include=FALSE}
-p1 <- plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
-  ylim(0, 0.95 ) +
+p1 <- plot_power_ratio_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
+  ylim(0, 1.2 ) +
   theme(legend.position="bottom") + 
-  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
+  guides(color=guide_legend(nrow=1,byrow=TRUE)) +
   report_theme +
@@
 
-p2 <- plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
-    ylim(0, 0.95) +
+p2 <- plot_power_ratio_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
+  ylim(0, 1.2 ) +
     report_theme +

### 2018-03-20 ms/ms.tex
KO final draft edits

@@
 \noindent%
-{\it Keywords:}  Nonparametric methods, linear model, analysis of covariance, analysis of designed experiments
+{\it Keywords:}  Nonparametric methods, linear model, analysis of covariance, analysis of designed experiments, hypothesis testing
 \vfill
@@
 and when the distributions of outcomes may differ across strata.
-%Furthermore, the hypothesis test implicitly assumes that the data were sampled at random from some underlying population, when in fact,
-%medical experimenters rarely recruit patients this way (\cite{ludbrook_why_1998}).
 
@@
 follows the Student $t$ distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
-The $p$-value for this hypothesis test is the probability, assuming the null hypothesis of zero coefficient is true, that a value drawn from the $t$ distribution is larger in magnitude than the observed $T$.
+The $p$-value for this two-sided hypothesis test is the probability, assuming the null hypothesis of zero coefficient is true, that a value drawn from the $t$ distribution is larger in magnitude than the observed $T$.
 This test is equivalent to the $F$ test.
@@
 if the difference in potential outcomes is heterogeneous across strata, then the coefficient $\gamma$ may be attenuated towards 0.
-Thus, it is unclear for any particular dataset whether or not the ANCOVA model will give valid results.
+Thus, it is unclear for any particular dataset whether or not the ANCOVA model will give valid estimates and inference for a treatment effect in finite samples.
 
 \subsection{Stratified permutation test}\label{subsec:strat_perm_test}
-All permutation tests essentially have two requirements: a conditioning space, the orbit of a finite group of transformations of the data in which all configurations of the data are equiprobable under the null hypothesis, and a set of sufficient statistics for the data which describe the data under these transformations.(\cite{pesarin_permutation_2010}).
-%We describe these two elements in the context of a stratified randomized experiment.
+All permutation tests essentially have two requirements: the conditioning space, defined as the orbit of a finite group of transformations of the data in which all configurations of the data are equiprobable under the null hypothesis, and a set of sufficient statistics for the data which describe the data under these transformations (\cite{pesarin_permutation_2010}).
+Exact permutation inference is derived from computing statistics for all (or a random subset) of the conditioning space.
 In experiments, the only random quantities are those involving $Z$, the vector of treatment assignments.
+The conditioning space is made up of all possible values of $Z$ under the randomization done in the experiment.
 In particular, the potential outcomes and covariates are fixed in the finite population under study, while the observed responses change with $Z$.
-The exact permutation inference is derived from the conditioning space which includes all possible values of $Z$.
 
@@
 
-$$H_0: Y_{ij}(1) = Y_{ij}(0), \forall i = 1, \dots, n_j, j = 1,\dots, J.$$
+\begin{equation}\label{eqn:sharp_null}
+H_0: Y_{ij}(1) = Y_{ij}(0), \forall i = 1, \dots, n_j, j = 1,\dots, J.
+\end{equation}
 
@@
 \begin{itemize}
-\item \textbf{Complete randomization:} the $\prod_{j=1}^J {n_j \choose n_j^T}$ assignments of treatment which assign $n_j^T$ out of $n_j$ individuals to treatment 1 in stratum $j$, for each stratum $j=1,\dots,J$, has equal probability
-\item \textbf{No treatment effect under the null:} both potential outcomes are identical and equal to the observed outcome for each individual $(i, j)$
+\item \textbf{Complete randomization:} The $\prod_{j=1}^J {n_j \choose n_j^T}$ assignments of treatment which assign $n_j^T$ out of $n_j$ individuals to treatment 1 in stratum $j$, for each stratum $j=1,\dots,J$, has equal probability.
+\item \textbf{No treatment effect under the null:} Both potential outcomes are identical and equal to the observed outcome for each individual $(i, j)$.
 \end{itemize}
@@
 The permutation tests in this section use the linear model in Equation~\ref{eqn:ancova} to control for covariates, but make different assumptions about the data.
-In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes,
-making several variables exchangeable.
+In a randomized experiment, the treatment assignment is independent of covariates, errors, and potential outcomes.
+This independence makes each of these exchangeable against the treatment.
 We show two permutation tests developed in this framework.
@@
 \begin{itemize}
-\item \textbf{Linear model:} the linear model that describes $Y$ using $X$ and the approximation error $\varepsilon$ holds
-\item \textbf{Linear regression fit:} the linear regression residuals $\hat{\varepsilon}$ are a good approximation to the true errors $\varepsilon$ 
-\item \textbf{Exchangeable errors:} the $\varepsilon$ are exchangeable against $X$ within strata; any configuration of the $\varepsilon$ amongst the units has the same probability
-\item \textbf{No treatment effect under the null:} the $\varepsilon$ do not depend on $Z$
+\item \textbf{Linear model:} The linear model that describes $Y$ using $X$ and the approximation error $\varepsilon$ holds.
+\item \textbf{Linear regression fit:} The linear regression residuals $\hat{\varepsilon}$ are a good approximation to the true errors $\varepsilon$.
+\item \textbf{Exchangeable errors:} The $\varepsilon$ are exchangeable against $X$ within strata; any configuration of the $\varepsilon$ amongst the units has the same probability.
+\item \textbf{No treatment effect under the null:} The $\varepsilon$ do not depend on $Z$.
 \end{itemize}
@@
 We simulated data from a randomized experiment using several different data-generating processes.
-The variables included stratification, two randomly assigned treatments, a baseline measure before treatment, and the same variable measured after treatment.
+The variables included three strata, two levels of a randomly assigned treatment, a baseline measure before treatment, and the same variable measured after treatment.
 We tested for a treatment effect and compared the empirical power of the tests over repeated random treatment assignments.
@@
 and the Freedman-Lane permutation test.
+The goal was to explore the performance of these methods under different violations of the usual linear model assumptions.
+On the one hand, permutation tests assess the sharp null hypothesis, a stronger null hypothesis that implies the null hypothesis of parametric ANCOVA.
+On the other hand, permutation tests make fewer assumptions about the data-generating process and may be more robust to violations of the linear model.
 The simulation code is included in four supplementary files.
@@
 In the first set of simulations, we generated continuous potential outcomes based on a latent variable and disturbance terms, both of which we manipulated.
+Observations were indexed by individual $i = 1, \dots, n_j$ in stratum $j=1, 2, 3$.
+We included $n_j=16$ individuals per stratum and treatment assignment was balanced, i.e. 8 people received each treatment in each stratum.
+These small stratum sizes were chosen to reflect the size of many real-world clinical trials.
+
 We used two scenarios.
@@
 In the second scenario, the treatment effect was heterogeneous across strata, and we drew the latent random variable according to
-\begin{displaymath}
+\begin{equation}
    v_{ij} \sim \left\{
@@
    \right.
-\end{displaymath}
+\end{equation}
 
@@
 
-We included $n_j=16$ individuals per stratum and treatment assignment was balanced, i.e. 8 people received each treatment in each stratum.
 After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we constructed the baseline value for individual $i,j$ as 
-$$ X_{ij} = \frac{-\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}.$$
+\begin{equation}
+X_{ij} = \frac{-\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}.
+\end{equation}
 
@@
 
-$$Y_{ij}(Z_{ij}) = \frac{(2Z_{ij} - 1)\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \delta_{ij}.$$
+\begin{equation}
+Y_{ij}(Z_{ij}) = \frac{(2Z_{ij} - 1)\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \delta_{ij}.
+\end{equation}
 
@@
 
-$$Y_{ij}(Z_{ij}) = Z_{ij} \gamma e^{v_{ij}} + X_{ij} + (\delta_{ij} - \varepsilon_{ij}).$$
+\begin{equation}\label{eqn:sim1_lm}
+Y_{ij}(Z_{ij}) = Z_{ij} \gamma e^{v_{ij}} + X_{ij} + (\delta_{ij} - \varepsilon_{ij}).
+\end{equation}
 
@@
 When the latent variables vary in distribution across strata, the assumption of constant slopes is violated.
+ANCOVA tests the significance of the coefficient on the treatment indicator $Z_{ij}$, but Equation~\ref{eqn:sim1_lm}
+shows that this coefficient varies according to $v_{ij}$.
 
@@
 
-\begin{align*}
+\begin{align}
 X_{ij} &=  \frac{ e^{v_{ij}} + e^{v_{ij}/2}}{2} +\varepsilon_{ij} \\
-Y_{ij}(Z_{ij}) &= (1+\gamma)^{Z_{ij}} X_{ij}  + \delta_{ij} \\
+Y_{ij}(Z_{ij}) &= (1+\gamma)^{Z_{ij}} X_{ij}  + \delta_{ij} \nonumber \\
 &= \gamma X_{ij}Z_{ij} + X_{ij} + \delta_{ij}.
-\end{align*}
+\end{align}
 
@@
 This paper adds to the literature comparing parametric and nonparametric tests.
-We simulated a variety of data-generating processes under the potential outcomes model, ranging from the ideal case when treatment effects are homogeneous for all individuals, to cases when errors are heavy-tailed or heteroskedastic, data are discrete, effects vary across strata, and treatment assignment is imbalanced within groups.
+We simulated a variety of data-generating processes under the potential outcomes model, as opposed to assuming the parametric linear model framework.
+Scenarios ranged from the ideal case when treatment effects are homogeneous for all individuals, to cases when errors are heavy-tailed or heteroskedastic, data are discrete, effects vary across strata, and treatment assignment is imbalanced within groups.
 The linear regression based tests had comparable power in all circumstances.
@@
 and demonstrates by simulation that the latter has larger than nominal level when the error variances are unequal.
-Randomized experiments mitigate this problem: by definition, treatment assignment is statistically independent of all other variables (possibly conditioning on strata) and in expectation over repeated randomizations, the two groups have equal variances.
+Randomized experiments mitigate this problem: 
+by definition, treatment assignment is statistically independent of all other variables (possibly conditioning on strata) and in expectation over repeated randomizations, the two groups have equal variances.
+Randomization guarantees this: it is a feature, not an assumption.
 

### 2018-03-21 ms/ms.tex
missing period

@@
 The goal of a randomized experiment is to draw causal inferences about the efficacy of a treatment.
-Thus, it makes sense to analyze experiments in the context of potential outcomes, where every individual has a counterfactual outcome for each of the treatments (\cite{holland_1986_statistics})
+Thus, it makes sense to analyze experiments in the context of potential outcomes, where every individual has a counterfactual outcome for each of the treatments (\cite{holland_1986_statistics}).
 Experiments are an attempt to estimate the counterfactual outcomes by randomly assigning treatments.

### 2018-03-21 simulations/continuous_outcomes.Rmd

@@
   ggplot(aes(x = Effect, y = Power)) +
-  geom_line(aes(color = Method))+
+  geom_line(aes(linetype = Method))+
   xlab("Effect size") +
@@
   theme(legend.position="bottom") + 
-  guides(color=guide_legend(nrow=1,byrow=TRUE)) +
+  guides(linetype=guide_legend(nrow=1,byrow=TRUE)) +
   report_theme +

### 2018-03-21 ms/ms.tex

@@
 Table~\ref{tab:clinical_pvalues} shows the $p$-values for the four tests and the seven continuous study endpoints.
-Overall, the results confirm our expectations based on visual comparison in Figure~\ref{fig:clinical_distr}:
+Overall, the results confirm our expectations based on Table~\ref{tab:clinical_distr}:
 one or more of the tests reject the null hypothesis that outcomes are the same between treatments for heartburn frequency, daily heartburn, and daily ``hrdq,''
@@
 
-\begin{table}\label{tab:clinical_distr}
+\begin{table}
 \centering
@@
 \caption{Mean of each continuous outcome in groups A and B.} 
+\label{tab:clinical_distr}
 \end{table}
@@
 \centering
-\label{tab:clinical_pvalues}
 \begin{tabular}{l|LLLL}
@@
 \caption{Comparison of p-values from four tests, for each continuous endpoint.} 
+\label{tab:clinical_pvalues}
 \end{table}
